### 2015-07-26 bioc192/tute.tex
adding a couple of tex that i want to work anywhere

\documentclass[a4paper, 11pt]{article}

\begin{document}

\end{document}

### 2015-07-26 science_writing/omsrs_writing.tex
adding a couple of tex that i want to work anywhere

\documentclass[a4paper, 11pt]{article}

\topmargin=-2.5cm
\textheight=10.5in
\oddsidemargin=-10pt
\textwidth=6.5in

\begin{document}

\begin{center}
\LARGE{Otago Medical School Research Society Science Writing Competition}\\
\end{center}

\section*{Why does my research matter?}
\noindent
\begin{itemize}
\item{obesity is a world-wide problem}
\item{cancer is a leading cause of death around the world}
\item{poor cancer prognosis due to subtypes}
\item{effect of chemotherapy varies between individuals}
\item{my research matters because it's looking for individual gene signatures that can be applied to personal medicine}
\end{itemize}

\section*{Explain the research:}
\noindent

\end{document}

### 2015-07-27 bioc192/tute.tex
edited tute.tex
half way through lecture 1

@@
-\documentclass[a4paper, 11pt]{article}
+\documentclass[a4paper, 12pt]{report}
+
+\usepackage{graphicx}
+\usepackage[utf8]{inputenc}
+
+\newcommand{\HRule}{\rule {\linewidth}{0.5mm}}
+
+\newcommand{\mychapter}[2]{
+    \setcounter{chapter}{#1}
+    \setcounter{section}{0}
+    \chapter*{#2}
+    \addcontentsline{toc}{chapter}{#2}
+}
 
@@
 
+\input{./title/title.tex}
+
+\tableofcontents
+\newpage
+
+\mychapter{0}{Introduction}
+
+\mychapter{1}{Lecture 1}
+\section{Protein}
+
+Proteins are the workhorses of the cell -- they are required for every process a cell has to do (e.g. DNA replication, apoptosis, etc.)
+Proteins are very small (50$\sim$100\AA{}, or 5$\sim$10nm big; a ``nanostructure''), yet they all have unique structure and functions that makes them important for cell survival and growth, and therefore often referred to as ``molecular machines''.
+
+Biochemistry is the study of the molecular details of the biology of the cells/living organisms, including proteins, DNA and RNA.
+To understand what components are involved and how they function, you need to know the structure of the thing you're trying to describe -- which, in our case, is the protein.
+
+\section{Amino Acids}
+
+Amino acids are the building blocks of a protein.
+Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (ref to diff. chapter).
+
+There are 20 amino acids in our cells that make up the proteins (insert image of 20 aa).
+All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$) (insert image of amino acid structure).
+Due to the variable side chains, all amino acids except glycine are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
+The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.
+
+\section{Charge of Amino Acids}
+
+Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
+The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
+When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.
+Depending on the pH of the solution, there will be more or less protoneted molecules.
+
+When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to become protonated.
+Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to become deprotonated.
+So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
+
+\begin{center}
+    \textbf
+        {pH \textless{} pK$_a$, then the molecule is protonated \\
+        pH \textgreater{} pK$_a$, then the molecule is deprotonated}
+\end{center}
+
+Some amino acid side chains are also able to be charged, and again, the charge of the amino acid side chain depends on its pK$_a$ value and the pH of the environment.
+
+\section{Amino Acid Side Chains}
+
+As mentioned earlier, all 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
+This means that each of the 20 amino acids have different chemical properties, depending on the side chain group it has.
+Cells take advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster (i.e.
+it gives the protein its functionality, as well as its unique structure).
+
+\section{Amino Acid Subgroups}
+
+Due to the unique chemical properties of the amino acids, they are grouped into subgroups.
+There are four main groups: Non-polar, uncharged polar, negatively charged, and positively charged amino acids.
+
+\subsection{Non-polar}
+
+\subsection{Uncharged Polar}
+
+\subsection{Negatively Charged}
+
+\subsection{Positively Charged}
+
+\mychapter{2}{Lecture 2}
+\section{me}
+
 \end{document}

### 2015-07-27 bioc192/tute.tex
finished lecture 1, starting lecture 2

@@
 
-\section{Charge of Amino Acids}
-
-Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
-The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
-When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.
-Depending on the pH of the solution, there will be more or less protoneted molecules.
-
-When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to become protonated.
-Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to become deprotonated.
-So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
-
-\begin{center}
-    \textbf
-        {pH \textless{} pK$_a$, then the molecule is protonated \\
-        pH \textgreater{} pK$_a$, then the molecule is deprotonated}
-\end{center}
-
-Some amino acid side chains are also able to be charged, and again, the charge of the amino acid side chain depends on its pK$_a$ value and the pH of the environment.
-
 \section{Amino Acid Side Chains}
@@
 
-Due to the unique chemical properties of the amino acids, they are grouped into subgroups.
+Due to the unique chemical properties of the amino acid side chains, they are grouped into subgroups.
 There are four main groups: Non-polar, uncharged polar, negatively charged, and positively charged amino acids.
@@
 
+These amino acids have non-polar (or hydrophobic) side chain, meaning that the side chains cannot be charged under physiological pH.
+The amino acids in this group are: alanine, valine, leucine, isoleucine, glycine, cysteine, phenylalanine, tryptophan, methionine, and proline.
+
 \subsection{Uncharged Polar}
+
+These amino acids have side chains that can be charged, but uncharged under normal physiological pH.
+As you can see, the side chain contains --OH groups and/or --NH$_2$ groups, whichare able to de/protonate under certain pH.
+The amino acids in this group are: serine, threonine, tyrosine, asparagine, and glutamine.
 
@@
 
+These amino acids have side chains that are negatively charged (i.e. deprotonated) side chains at physiological pH.
+The side chain contains a deprotonated --COOH group, giving it a negative charge to the side chain.
+The pK$_a$ of these amino acid side chains ranges from 4$\sim$5.
+The amino acids in this group are: aspartic acid (aspartate) and glutamic acid (glutamate).
+
 \subsection{Positively Charged}
 
+These amino acids have side chains that are positively charged (i.e. protonated) side chains at physiological pH.
+The side chain contains a protonated --NH$_2$ group, giving it a positive charge to the side chain.
+The pK$_a$ of these amino acid side chains ranges from 10$\sim$12, with an exception of histidine, which has a pK$_a$ of 7$\sim$7.5.
+The amino acids in this group are: lysine, arginine, and histidine.
+
+\section{Charge of Amino Acids}
+
+Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
+The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
+When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.
+Depending on the pH of the solution, there will be more or less protonated molecules.
+
+The amount of de/protonated amino acids affect the net charge of the protein, and therefore the isoelectric point (pI) of the protein.
+The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).
+
+When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to become protonated.
+Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to become deprotonated.
+So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
+
+\begin{center}
+    \textbf
+        {pH \textless{} pK$_a$, then the side chain is protonated \\
+        pH \textgreater{} pK$_a$, then the side chain is deprotonated}
+\end{center}
+
+Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
+However, some amino acid side chains may be charged under physiological condition and contribute to the net charge of the amino acid, and therefore the protein.
+This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.
+
+\section{Amino Acid Modifications}
+
+Some amino acids are modified after translation -- this is called the post-translational modification (PTM).
+Types of PTM includes:
+\begin{itemize}
+    \item Disulfide bond formation between cysteines
+    \item Phosphorylation
+    \item Glycosylation
+    \item Methylation
+    \item Adenylation
+    \item Iodination
+    \item Metal binding
+\end{itemize}
+
+\section{Proteins, Peptides, and Peptide Bonds}
+
+Amino acids are covalently linked together by peptide bonds -- the bond formed between the amino group of one amino acid with a carboxyl group of another amino acid.
+A chain of amino acids linked together by peptide bonds are called peptides, and these are usually short.
+Peptides that have a longer chain of amino acids with defined biological function are called proteins.
+(Note: peptides are proteins as well -- it's just a little smaller)
+
+\subsection{Amide (peptide) Bonds}
+
+Amide (peptide) bond has a partial double bond characteristic due to the electron delocalisation/sharing between the oxygen atom and the nitrogen atom.
+This means that the amide bond is rigid, and also have a dipole nature (oxygen is slightly negative, nitrogen is slightly positive).
+
+The amide bonds are predominantly in \textit{trans}-configuration in the protein, although some can be in \textit{cis}-configuration, especially when the bond is preceded by proline.
+This is due to ring structure that the proline side chain forms with its amino group (causes steric clashes).
+
+By convention, the free amino group on the end of the peptide chain is called the N-terminus of the protein, and the free carboxyl group is known as the C-terminus.
+This gives a `directionality' to the peptide sequence, from the N-terminus to the C-terminus, and we draw primary sequence and count the amino acids according to this directionality.
+
 \mychapter{2}{Lecture 2}
-\section{me}
+
+\section{Bond Rotation}
 

### 2015-07-27 bioc192/intro.tex
added intro.tex

This book is intended to provide a basic overview of BIOC192 that will have more than enough details to tutor BIOC192 students.
Please note that to get the maximal use out of this book, and therefore how good you will become at being able to tutor someone in BIOC192, listening to the podcasts and adding your own notes is highly recommended.

Please DO NOT send this book around, especially to the HSFY students, unless you have asked the author (i.e. me) if you could do so.
All rights reserved.
No commercial reproduction, adaptation, distribution or transmission of this book is permitted without the written permission of \ldots me (Riku Takei).
And all the other legal crap people put on their books.

\vspace{5cm}

\noindent
Riku Takei \\
4$^{th}$ year Biochemistry MSc student\\

### 2015-07-27 bioc192/intro.tex
finished lecture 2, onto lecture 3

@@
 This book is intended to provide a basic overview of BIOC192 that will have more than enough details to tutor BIOC192 students.
-Please note that to get the maximal use out of this book, and therefore how good you will become at being able to tutor someone in BIOC192, listening to the podcasts and adding your own notes is highly recommended.
+It is highly recommended to listen to the podcasts and add your own notes to this book to get the maximal use out of this book.
 
@@
 All rights reserved.
-No commercial reproduction, adaptation, distribution or transmission of this book is permitted without the written permission of \ldots me (Riku Takei).
-And all the other legal crap people put on their books.
+No commercial reproduction, adaptation, distribution or transmission (and any other legal crap people put on their books) of this book is permitted without the written permission of the author (Riku Takei).
 

### 2015-07-27 bioc192/lect02.tex
finished lecture 2, onto lecture 3

\section{Levels of Protein Structure}

The specific 3D structure of a protein is very important for the function and biological role of the protein in the cell.

\subsection{Primary Structure}

The Primary structure of a protein is the amino acid sequence of that particular protein.
Note that the sequence of the amino acid (i.e. the primary structure) determines the 3D structure of the protein, since the primary sequence will have all the required amino acids in the right order for protein folding (via hydrophobic interaction).

\subsection{Secondary Structure}

Secondary structure is the local regular folds/structure that are stabilised by H-bonds between the backbone peptide groups of the polypeptide chain.
Examples of secondary structures include the $\alpha$-helix and $\beta$-sheets.
Supersecondary structures are related to the secondary structure, as it consists of combinations of secondary structures to make certain motifs (e.g. greek keys).

\subsection{Tertiary Structure}

Tertiary structure is the structure, or domain, formed by arranging secondary/supersecondary structures.
These usually have a function once it has the correct 3D structure, but it may require other tertiary structures to act as a single protein.

\subsection{Quaternary Structure}

Quaternary structure is where multiple tertiary structure come together to form a single functional protein.
Not all protein form a quaternary structure to function.

\section{Bond Rotation}

Protein structure is created by the main chain and side chain bonds, and requires the main/side chains to be in a certain conformation.
In order to do this, the amino acids must rotate and orientate correctly for the right interaction.

Single bonds between the carbon atoms can be rotated freely, but not the double bonds.
Since the amide bonds have a partial double bond characteristic, it is rigid and forms a plane that does not allow rotation.
However, rotation is allowed at the $\alpha$-carbon (C$_{\alpha}$) atom.
Obviously, even though rotations are allowed at the C$_{\alpha}$ atom, not all angles are allowed due to steric hindrance.

There are four types of bond angle rotations: $\omega, \psi, \phi$, and $\chi$.
These angles have values that range from 0 to $\pm$180$^{\circ}$, and together, it gives the 3D structural information of a protein.

\subsection{Phi ($\phi$) Angle}

Phi angle is defined as the bond angle rotation of the N--C$_{\alpha}$ bond.
This bond rotation leads to O--O atom collision.

\subsection{Psi ($\Psi$) Angle}

Psi angle is defined as the bond angle rotation of the C--C$_{\alpha}$ bond.
This bond rotation leads to the NH--NH collision.

\vspace{1.0cm}

\noindent
\textbf{Remembering phi/psi angles are quite hard, but just remember that, relative to the C$_{\alpha}$ atom, phi angle is the rotation of the amide plane on the amino group side, whereas the psi angle is the rotation of the amide plane on the other side (i.e. the carboxyl side).}

\subsection{Omega ($\omega$) Angle}

mega angle is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).

\subsection{Chi ($\chi$) Angle}

Chi angle is the bond angle rotation of the side chain bond (C$_{\alpha}$--R bond).

\section{Ramachandron Plot}

This plot shows all the possible phi/psi angles (i.e. the two amide planes adjacent to the C$_{\alpha}$) that are allowed in a protein with no steric interference.
Any phi/psi angle pairs that cause steric hindrance are not plotted, so we get some `regions' of allowed angle pairs for any given protein.
The side chains of the amino acids are (usually) staggered to minimise steric hindrance.

\section{Secondary Structure of a Protein}

There are three main secondary structures in a protein: $\alpha$-helix, $\beta$-sheets, and turns.

\subsection{$\alpha$-Helices}

Specifically speaking, $\alpha$-helix is a type of helix that is present in a protein -- there are other types of helices (e.g. polyproline helix), but $\alpha$-helix is the most common type.

In $\alpha$-helix, the carbonyl group of the n$^{th}$ residue forms a H-bond to the amino group of the n+4$^{th}$ residue (e.g. the carbonyl group of the 1$^{st}$ residue H-bonds to the amino group of the 5$^{th}$ residue)
There is about 3.6 residues per turn, which is about 5.4\AA per turn.
The distance (or height) between 2 residues is about 1.5\AA.
The side chains point outwards from the helix (this provides certain chemical properties on the sides of the helix, e.g. hydrophobic side). 
$\alpha$-helices are dipole in nature, due to the amide bond polarity, all of which points at a single direction.

There are some residues that disrupt the helical structure, and these are glycine and proline.
Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
This flexibility can cause disruption in the helical structure.
Proline has a unique side chain that connects back to its amino group, forming an imino group, meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.

\subsection{$\beta$-Sheets}

$\beta$-sheets, or $\beta$-pleated sheets, are formed from H-bond formation between adjacent polypeptide chain (or strands), therefore $\beta$-sheets are formed from 2$\sim$10 strands (but itcan have more strands).
Each strand has about 15 amino acids, and the $\beta$-sheets can be either parallel (both strands go from N to C terminal) or anti-parallel (strands go in an opposite direction).
$\beta$-sheets are extended, but pleated (i.e. it's not completely flat), and the side chains point above and below.

\subsection{Turns}

Turns are sharp hairpin turns in a protein, and it has a high glycine and proline content.
About 30\% of the residues in a protein is involved in turns.
It is common for the residues to form H-bonds across the turn.
There are many types of turns (\textgreater16), but type I and II are common.

\subsection{Relationship between Ramachandron Plots and Secondary Structures}

Secondary structures have certain phi/psi angle pairs, and when you overlay these to the plot, it falls into the allowed regions of the plot at a distinct region.
This is not surprising, because if the secondary structures fall in the forbidden region, how did the structure form in the first place?
It can't -- the structure cannot form due to steric hinderance.

This essentially shows that secondary structures have unique phi/psi angle pairs.

### 2015-07-27 bioc192/lect03.tex
finished lecture 3, onto lecture 4

\section{Supersecondary structure}

Secondary structures such as $\alpha$-helix, $\beta$-sheets, and turns combine together to form a specific supersecondary structures, or `motifs'.

\section{Protein Domains}

Several supersecondary structures then come together to form protein domains (tertiary structure), which often possess a specific biological function.
These domains have hydrophobic core and hydrophilic surface, and is stabilised by non-covalent interactions (but some proteins may have covalent disulfide bonds to stabilise it).
Note: Hydrophobic interactions drive the protein folding process.
Large proteins are often multi-domained, whereas smaller proteins can be made up of a single domain.

\section{Protein Families}

Proteins can be grouped based on the tertiary structure (e.g. $\alpha$/$\beta$-barrel).
There are a lot of protein families.

\section{Quaternary Structure}

Quaternary structure refers to the non-covalent interactions between different domains, which forms a larger protein structure.
The domains are called `subunits', and these assemblies are called dimer (2 subunits), trimers (3 subunits), etc.

\section{Protein Folding Pathway}

Protein folding is directed predominantly by the hydrophobic forces from the non-polar residues.
This is NOT a random process.
Since the proteins are submerged in water, the non-polar residues try to stay away from water, and therefore aggregate together to form secondary structures.
This initial folding process is referred to nucleation.
The hydrophobic regions of the secondary structure comes together to form a larger structure, and stabilised by small non-covalent interactions.

\vspace{0.5cm}

\noindent
The process of protein folding is summarised below:
\begin{enumerate}
\item Short, small secondary structures are formed, driven by hydrophobic forces (nucleation)
\item Nuclei come together to form larger domains
\item Larger domains come together
\item Small conformational changes/adjustments occur to give compact native structure
\end{enumerate}

\section{Stabilisation of Protein Folding}

Proteins are stabilised by non-covalent interactions, which are weak individually, but strong when there are a lot of them.
Covalent bonds such as disulfide bonds can stabilise the protein as well.

\section{Chaperones and Chaperonin}

Some proteins require additional accessory proteins to assist them in folding the proteins properly, or to fold it faster.
If the proteins are not folded properly, it can lead to serious diseases like Alzheimer's and prion.

### 2015-07-27 bioc192/lect04.tex
bed time

\section{Myoglobin}

Myoglobin is the monomeric form of haemoglobin, and it resides in the tissue/muscles for O$_2$ storage.
Myoglobin consists of a globin protein that consists of 8 $\alpha$-helices (numbered for clarity), with a haem group in the middle.
Myoglobin is able to reversibly bind O$_2$, and have high affinity for O$_2$ as well.

\section{Haem (Heme)}

Haem consists of four pyrrole rings that are linked together via methane bridges, with the ferrous (Fe$^{2+}$) attached at the center of the molecule.
This Fe$^{2+}$ is slightly below the plane of the porphyrin ring when O$_2$ is not bound, and is shifted into the plane when it binds (causes conformational change).

The heme is attached to the myoglobin via the His F8 (histidine on the helix F at position 8), and His E7 is on the opposite side to influence the O$_2$ binding affinity.
The His E7 prevents the strong binding of O$_2$ to the heme by kinking the angle at which the O$_2$ binds to the heme.
This allows weaker and therefore reversible binding of the O$_2$ molecule to the myoglobin.

Note that cyanide binds to the Fe$^{3+}$ form, not the Fe$^{2+}$ form.
Cyanide's main inhibitory effect is at the cytochrome Fe$^{3+}$, and affects the electron transport system.

### 2015-07-27 bioc192/lect04.tex
finished lecture 4, onto lecture 5

@@
 The heme is attached to the myoglobin via the His F8 (histidine on the helix F at position 8), and His E7 is on the opposite side to influence the O$_2$ binding affinity.
-The His E7 prevents the strong binding of O$_2$ to the heme by kinking the angle at which the O$_2$ binds to the heme.
+The His E7 prevents the strong binding of O$_2$ to the heme (right angle) by kinking the angle at which the O$_2$ binds to the heme.
 This allows weaker and therefore reversible binding of the O$_2$ molecule to the myoglobin.
@@
 Note that cyanide binds to the Fe$^{3+}$ form, not the Fe$^{2+}$ form.
-Cyanide's main inhibitory effect is at the cytochrome Fe$^{3+}$, and affects the electron transport system.
+Also note that cyanide's main inhibitory effect is at the cytochrome Fe$^{3+}$, and affects the electron transport system.
+
+\section{Haemoglobin}
+
+Haemoglobin is the tetrameric (4 subunit) form of myoglobin, and it is a blood O$_2$ transporter.
+Since haemoglobin is a tetramer, it has 4 haem group in total (one on each subunit), and therefore able to carry 4 O$_2$ molecules.
+Slight conformational change occurs when the O$_2$ binds to the haemoglobin molecule, which will become important in explaining the cooperative effect of the subunits.
+
+\section{O$_2$ Binding Property}
+
+\subsection{Myoglobin}
+
+Myoglobin resides in muscles and tissues, and have high affinity for O$_2$.
+This is shown in the O$_2$ saturation curve.
+At low partial pressure of O$_2$ (pO$_2$), myoglobin is mostly saturated with O$_2$ (i.e. O$_2$ is bound to it).
+This means that myoglobin only releases the O$_2$ molecule when the pO$_2$ is very low.
+
+\subsection{Haemoglobin}
+
+In contrast, haemoglobin's O$_2$ binding affinity is not that high compared with myoglobin.
+As you can see from the O$_2$ saturation curve, it requires quite a high pO$_2$ for haemoglobin to be saturated with O$_2$.
+This makes sense as haemoglobin has to bind to O$_2$ at the lungs where pO$_2$ is relatively high, but also has to release the O$_2$ molecules at the tissue/muscle, where the pO$_2$ is low.
+This allows haemoglobin to be an effective O$_2$ carrier.
+
+One thing to note here from the curve is that the shape of the curve for haemoglobin is sigmoidal (S-shape).
+This indicates that haemoglobin is allosteric (or cooperative), where the subunits assist one another to facilitate the binding/release of O$_2$ molecules.
+Of course, there are other factors that contribute to the allosteric binding/release of O$_2$ molecules.
+
+\section{Sequential and Concerted Model of \-Allosterism}
+
+When O$_2$ binds to a subunit of haemoglobin, it causes a slight conformational change to the subunit.
+This conformational change assists the binding of O$_2$ molecules to the other subunits by cooperativity/allosterism.
+For clarity, the tense (T) state is the state of the subunit when O$_2$ is not bound (low O$_2$ affinity), and the relaxed (R) state is the state of the subunit when O$_2$ is bound (high O$_2$ affinity).
+
+There are two models, sequential and concerted models, of allosterism.
+It is still unknown which model is correct, as there are many observations supporting both models.
+
+\subsection{Sequential Model}
+
+Sequential model is when one O$_2$ molecule binding to a single subunit causes the conformational change in the subunit it binds, and also the subunits that are adjacent to it as well.
+This conformational change makes the other subunits easier to bind to O$_2$ molecules, which will also change conformation of itself and the others.
+
+With respect to the R/T states, binding of O$_2$ molecule causes a single subunit to be in R state, and also changes the conformation of the adjacent subunits so that it is easier to bind to O$_2$ molecules, therefore easier for the subunits to turn into the R state.
+
+\textbf{In sequential model, binding of one substrate changes the conformation of other subunits, which makes it easier for the other subunits to bind the substrate}
+
+\subsection{Concerted Model}
+
+In concerted model, all of the subunits are either in the T state, or the R state, and the states are at equilibrium (i.e. no `in between' state).
+The O$_2$ molecules can bind to either state at any time.
+When there are more substrate (i.e.
+O$_2$) is bound to haemoglobin, the greater the chance for the subunits to be in the R state, which means that the equilibrium is favoured towards the R state when there are more O$_2$ molecules bound to haemoglobin.
+
+\textbf{ In concerted model, the T and R states are at equilibrium and substrates can bind at any time to any conformational state.
+The binding of substrate causes the equilibrium to shift to the R state and the substrates are more likely to bind to the protein.}

### 2015-07-29 bioc192/lect05.tex
finished lecture 5, onto lecture 6

\section{Allosteric Effectors of Haemoglobin}

As discussed earlier in Lecture 4, binding of O$_2$ molecule to the haem group causes a conformational change, and thus promotes the R or T state.
This conformational change is caused by the shift in the Fe$^{2+}$ atom of the haem, which is attached to the His F8 of the haemoglobin subunit, which then cause a slight shift that propagates through the protein.

However, there are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
Since these molecules have effects on haemoglobin, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.

\vspace{0.5cm}

\begin{center}
\textbf{Allosteric effectors are molecules that bind to the allosteric protein at a place other than/different to the normal substrate binds to.}
\end{center}

\section{2,3-Bisphosphoglycerate (BPG)}

2,3-Bisphosphoglycerate (BPG) is a small, negatively charged (-5 charge) molecule that binds to the deoxy-haemoglobin at the center of the tetramer.
Binding of BPG causes the haemoglobin to be in T state, i.e. in the dexygenated state, and decreases the haemoglobin's binding affinity for O$_2$, meaning that the haemoglobin with BPG bound to it will require higher pO$_2$ to oxygenate it.
Binding of BPG at the muscle/tissue site ensures that the O$_2$ molecules are released at the muscles/tissues.

O$_2$ binding to the haemoglobin at the lungs causes the release of BPG molecule, allowing the haemoglobin to carry O$_2$ to the muscles/tissues.

\section{H$^+$ and CO$_2$}

Both H$^+$ and CO$_2$ binds to haemoglobin at a different site as the substrate (O$_2$) and affects the oxygen affinity of haemoglobin.
Both molecules favour the T state of haemoglobin, and therefore favours the deoxygenated state (i.e. decreased O$_2$ affinity).
Note that both of these molecules are related to the pH of the environment, and the blood buffer system is also involved in controlling the concentration of these molecules.

\subsection{The Bohr Effect}

Technically speaking, the Bohr effect is the effect of pH (i.e. [H$^+$]) and [CO$_2$] on the oxygen affinity of haemoglobin.
As pH decreases (therefore [H$^+$] increases), the net charge of haemoglobin changes due to different amino acids getting protonated, which leads to slight conformational changes in the haemoglobin structure, and therefore decrease in haemoglobin's oxygen affinity.
Similarly, as CO$_2$ concentration increases, CO$_2$ is more likely to bind to the haemoglobin and decrease the oxygen affinity of haemoglobin.

\subsection{The Blood Bicarbonate Buffer System}

In the blood, there is an equilibrium between bicarbonate ([HCO$_3^-$]) ion and CO$_2$.

\begin{center}
\ch{H$_2$O + CO$_2$ <=> HCO$_3^-$ + H$^+$}
\end{center}

As you can see, the [H$^+$] concentration and the [CO$_2$] concentration is related to one another due to this buffer system.
This means that at the tissue/muscle where the relative concentration of CO$_2$ is high, it is going to be making the bicarbonate ion, together with H$^+$ ion, and therefore the pH is decreased.
However, at the lungs where the concentration of CO$_2$ is low, there will be less H$^+$ ion present, and therefore have higher pH.

Together with the Bohr effect, this means that the pH is high at the lungs and have low CO$_2$ concentration, and therefore haemoglobin will have high oxygen affinity, and thus get saturated with O$_2$ (oxygenated).
In contrast, the pH will be lower at the tissues/muscles and have high CO$_2$ conference, and therefore haemoglobin will have low oxygen affinity, and thus releases O$_2$ at the tissue/muscle (deoxygenated).

Also note that BPG is released at the lungs when O$_2$ binds to the haemoglobin, and binds at the tissue/muscle when O$_2$ is released and CO$_2$ and H$^+$ binds to it, contributing to the binding and release of O$_2$.(insert the saturation curve of +/- allosteric effectors)

### 2015-07-29 bioc192/lect06.tex
finished lecture 6, onto lecture 7

\section{Haemoglobin Variants}

There are a lot of haemoglobin variants.
Some variants are naturally occurring, whereas other variants are caused by genetic mutations.

\section{Normal Variants}

These variants are normal in the sense that they are naturally occurring, and not caused by abnormal mutations.
These variants are different to the normal haemoglobin in that it uses different type of subunits to make the haemoglobin up.

\subsection{Foetal Haemoglobin (HbF)}

Foetal haemoglobin (HbF) is only present in the foetus blood.
HbF is made up of two $\alpha$ subunits and two $\gamma$ subunits, compared to the adult haemoglobin (HbA) which is made up of two $\alpha$ and two $\beta$ subunits.
Since the foetus has to obtain oxygen from the mother's blood via the placenta, HbF has a higher affinity for oxygen than the HbA.
This is because HbF binds less strongly to BPG due to the Ser143 in the $\gamma$ subunit (His143 in $\beta$ subunit), thereby increasing the oxygen affinity.

The graph below just shows how different subunits are made at different developmental stage.(insert the graph)
As you can see, there are subunits other than $\alpha$, $\beta$ and $\gamma$ subunits.

\section{Abnormal Variants}

These variants are caused by genetic mutations that alter the properties of haemoglobin, and can lead to anaemia or other pathological symptoms.

\subsection{Sickle Cell Haemoglobin (HbS)}

This variant is caused by a single point mutation on the $\beta$ subunit of haemoglobin.
This point mutation causes an amino acid substitution from glutamic acid to valine (E6V mutation), which is on the surface of the protein.

The consequence of this mutation is the formation of deoxy-Hb aggregate/crystal in the red blood cell.
Since valine is a non-polar amino acid, it can stick/bind to other haemoglobin at specific site (in a hydrophobic pocket).

\subsection{Met-Haemoglobin (HbM)}

Methaemoglobin (HbM) is caused by mutations that changes the oxidation state of the iron atom in the haem from Fe$^{2+}$ to Fe$^{3+}$.
This means that the oxygen cannot bind (or have very low affinity for oxygen) to the haem iron, as it is in a different oxidation state.

\subsection{Haemoglobin C (HbC)}

Haemoglobin C (HbC) is common in West Africa, and is caused by a point mutation that causes a substitution mutation.
The substitution is at the same place as HbS: glutamic acid to lysine at position 6 (E6K).
As a consequence of this substitution, the charge is swapped from negatively charged glutamic acid to positively charged lysine.

Malaria is caused by the infection of red blood cells by \textit{Plasmodium \\falciparum}.
Infected red blood cells will contain malaria parasite proteins that produce knobs on the surface of the red blood cells and causes it to adhere to the capillaries, causing tissue damage due to reduced oxygen supply.
However, individuals with homo or heterozygous HbC have less adherence, and thus have some resistance to malaria infection.

### 2015-07-29 bioc192/lect07.tex
finished lecture 7, onto lecture 8

\section{Post-Translational Modification (PTM)}

Most proteins, if not all proteins, in eukaryotic cells are post-translationally modified.
Post-translational modification increases the number of distinct forms of proteins without increasing the number of genes in the genome.
This means that the level of complexity provided by the proteins are also increased as well.

The diversity of the proteins can be expanded by mRNA splicing and PTM.
PTM can be categorised into two categories: the covalent addition of chemical group to amino acid side chain, or covalent hydrolysis of the peptide bond in a protein.

\section{Covalent Modification of Amino Acids}

There are 15 amino acids known to be PTMed by covalent addition of chemical groups.
For each of the amino acids, there are a variety of modifications that can be made to it.
Three PTMs are discussed in this lecture: phosphorylation, gamma-carboxylation of glutamic acid, and hydroxylation.

\section{Phoshphorylation}

Phosphorylation can occur on the hydroxyl (OH) group of the amino acid side chains.
Amino acids that have OH groups are Ser, Thr, and Tyr.
Phosphorylation of the side chains are done by an enzyme called kinase, which adds the large, negatively charged phosphate group to the side chain, and uses ATP during this process.
To remove the phosphate group, an enzyme called phosphatase is used.

The consequence of adding a phosphate group to these amino acid side chains is that it's going to alter the protein structure, due to the electrostatic attraction between the more negatively charged phosphate group with other postitively charged amino acid side chains.
Since the attraction is going to shift the amino acids slightly, it will cause a shift in the whole protein, and therefore alters the conformation of the protein and affects the biological properties of it.

Quite often, phosphorylation state is used to `switch' the conformational state of the protein to enable a biological process in a cell (e.g. facilitating interaction with other proteins).

\subsection{Insulin Receptor}

Binding of insulin to the insulin receptor causes a conformational change in the receptor, and this exposes the tyrosine in the cytosol side of the membrane.
The exposed tyrosine is then targeted by kinases, and get phosphorylated by it.
Phosphorylated tyrosine can then cause responses (e.g. signal transduction, GLUT4 transporter expression).

\subsection{Na$^+$/K$^+$ Pump}

Phosphorylation state affects which molecule (Na$^+$ or K$^+$) to bind to the pump, and therefore be transported.
This is also energy dependent process, as the pump is constantly being phosphorylated and dephosphorylated.

\section{Gamma-carboxy Glutamic Acid (Gla)}

Gamma-carboxy glutamic acid occurs when an additional carboxyl group is added to the glutamic acid.
This PTM is done in some blood coagulation proteins, and it requires vitamin K as a co-factor.
Ten to twelve Gla enables the formation of bidentate Ca$^{2+}$ binding site and allows blood coagulation proteins to interact with platelets, as a part of blood clot formation process.

When you have vitamin K deficiency, you can't gamma-carboxylate the glutamic acids, since the reaction requires vitamin K as a co-factor.
As a consequence, this leads to bleeding disorder.

\section{Hydroxylation}

Hydroxylation mainly occurs on the 4' (or 3') position of the proline, but it can also occur at the 5' position of lysine.
Hydroxylation provides greater possibilities for non-covalent interactions, so it allows for more H-bonding to occur.
As a result, proteins are able to form higher order structure.

Vitamin C is important in hydroxylation, as it is required for the reaction to occur.
Vitamin C keeps the Fe$^{2+}$ in this oxidation form, which is essential for the hydroxylation process.
Since collagen structure requires hydroxylation to keep its higher order structure, low vitamin C causes disruption in this structure.
This disease is known as scurvy.

### 2015-07-29 bioc192/lect08.tex
finished lecture 8, onto lecture 9

\section{Introduction to Enzymes and Enzyme Engineering}

Enzymes (proteins) are able to accelerate and carry out chemical reactions that are most likely impossible to do under physiological/biological conditions.
Without these enzymes, we will not be able to synthesise chemicals required to survive.
It is also interesting that the enzymes have all evolutionarily developed itself to make difficult chemical reactions to faster, but within similar timescale.

Can we utilise the properties/characteristics of enzymes to make it do what we want it to do?
This is the key idea behind the researches being done in the area of enzyme engineering.
We can use enzyme engineering to make enzymes that are able to react with different substrates, or make it more efficient at doing the reaction.

### 2015-07-29 bioc192/lect09.tex
finished lecture 9, onto lecture 10

\section{Introduction to Biotechnology}

For therapeutic and clinical use of proteins, the proteins must be made and extracted from a source.
Biotechnology is the field of research to produce proteins in systems (sort of).

\subsection{The `Old' Way of Obtaining Proteins}

Therapeutic proteins were extracted from relevant sources (e.g. tissues, serum) and was used to treat the patients.
The problems with this approach was the risk of identified and unidentified pathogen infections from the source (e.g. HIV).

\subsection{Recombinant Protein System}

This is the more recent method to produce proteins for therapeutic use.
Plasmids that contain a required protein is expressed in a different organism/system, and the purified and collected.

Recombinant insulin is produced this way, by expressing two separate parts of insulin in \textit{E. coli}.
These parts are extracted and combined to form insulin.

The advantage of this is that it is cheap, has high yield, and it is pathogen free.
However, in a prokaryotic system, the cells are unable to post-trasnlationally modify the protein, which may be essential for the function of some proteins.
Another problem is that the proteins may not fold properly or only partially folded.

\subsection{Eukaryotic System}

This system solves the problem of PTM and protein misfolding, as the proteins will be expressed in a eukaryotic system.

An example of a protein that uses a eukaryotic system is erythropoietin (EPO).
EPO regulated haematopoiesis, and its expression is partially regulated by the blood oxygen level -- EPO acts so that it produce more red blood cells and haemoglobin.
EPO is used therapeutically as well as a performance enhancing drug.
EPO is produced in immortalised Chinese hamster ovary cells.

Different EPOs have different glycosylation profile, which allows us to detect athletes who uses EPO.

\subsection{Recombinant Antibodies}

Recombinant antibodies are designed to control the effects of some proteins in your body (e.g. some cell receptors).
Mouse AB can be made, but it is recognised as foreign material in humans and get rejected.
This lead to the formation of chimeric antibodies, so there will be no immune response.

\subsection{Pharming}

The use of live animals to express/make recombinant proteins, for example, in milk.

### 2015-07-30 bioc192/intro.tex
added graphics

@@
 All rights reserved.
-No commercial reproduction, adaptation, distribution or transmission (and any other legal crap people put on their books) of this book is permitted without the written permission of the author (Riku Takei).
+No commercial reproduction, adaptation, distribution or transmission (and any other legal crap people put on their books) of this book is permitted without the written permission of the author.
 

### 2015-07-30 bioc192/lect01.tex
making separate lecture files
added graphics

--- 2015-07-27 bioc192/tute.tex
@@
-\documentclass[a4paper, 12pt]{report}
-
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc}
-
-\newcommand{\HRule}{\rule {\linewidth}{0.5mm}}
-
-\newcommand{\mychapter}[2]{
-    \setcounter{chapter}{#1}
-    \setcounter{section}{0}
-    \chapter*{#2}
-    \addcontentsline{toc}{chapter}{#2}
-}
-
-\begin{document}
-
-\input{./title/title.tex}
-
-\tableofcontents
-\newpage
-
-\mychapter{0}{Introduction}
-
-\mychapter{1}{Lecture 1}
 \section{Protein}
@@
 Amino acids are the building blocks of a protein.
-Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (ref to diff. chapter).
+Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).
 
-There are 20 amino acids in our cells that make up the proteins (insert image of 20 aa).
-All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$) (insert image of amino acid structure).
+There are 20 amino acids in our cells that make up the proteins.
+All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$).
 Due to the variable side chains, all amino acids except glycine are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
 The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.
+
+\includegraphics[width=\textwidth]{aastructure}
 
@@
 
+\includegraphics[width=\textwidth]{20aa}
+
 \section{Charge of Amino Acids}
@@
 This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.
+
+\begin{center}
+\includegraphics[width=0.8\textwidth]{aapka}
+\end{center}
 
@@
 
-\mychapter{2}{Lecture 2}
-
-\section{Bond Rotation}
-
-\end{document}
+\begin{center}
+\includegraphics[width=\textwidth]{amidebond}
+\end{center}

### 2015-07-30 bioc192/lect10.tex
finished lecture 10, onto lecture 11

\section{Introduction to Enzymes}

In most cases, enzymes are proteins (in some cases, it's RNA enzymes) that acts as biological catalysts.
Enzymes increases the rate of reaction, but does NOT change the equilibrium, nor the thermodynamics of a reaction.
This means that, even though enzymes make the reactions to occur faster, it cannot shift the chemical reaction to favour one product over another.
Note that enzymes are named ``-ases''.

For a chemical reaction to occur, the substrates must be in close proximity of one another, and also have the right orientation.
Enzymes catalyse reactions by doing just that: bind the substrate in the right orientation in close proximity so a chemical reaction can occur at a faster rate.

\section{Active Site of an Enzyme}

Active site is the 3-dimensional site of an enzyme where the reaction occurs.
The active site has amino acid side chains projecting into it to help substrate binding.
The shape of the active site determines the specificity of the enzyme, since enzymes are (usually) very specific.
This means that enzymes only bind to certain substrate, and no other substrates.

\subsection{Geometric Specificity and Stereospecificity}

Geometric specificity is where the substrate must be made of same component/structure for it to be recognised by the enzyme.
Stereospecificity is where the substrate has the required components,but don't have the right orientation, and therefore not recognised by the enzymes.

\section{Enzyme-Substrate Interaction}

\begin{center}
\ch{ E + S <=> ES <=> [{T}] EP <=> E + P }
\end{center}

In an enzyme catalysed reaction, substrate binds to the active site and forms a enzyme-substrate complex (ES), which is stabilised by non-covalent interactions.
Note that this interaction is specific.
As the reaction proceeds, transition state (T), which is neither a substrate nor the product, is formed, then proceeds further into enzyme-product complex (EP).
This complex can then be dissociated to form the original enzyme ready for another reactions, and release the product.

There are two models that describes how the substrate binds to the enzyme: the lock-and-key model, or the induced-fit model.

\subsection{Lock-and-Key Model}

The concept of the lock and key model is fairly simple -- the substrate must have the shape/structure that perfectly matches the enzyme active site.
If the substrate doesn't then, the enzyme cannot react with it and no reaction occurs.
However, this is not really the case in nature as enzymes changes conformation when the substrate binds to it.

\subsection{Induced-fit Model}

This is how an enzyme actually works.
The substrate and the enzyme active site does not match perfectly as it did for the lock and key model.
When the substrate and the enzyme binds, it causes a conformational change in the enzyme, which allows the enzyme to properly bind to the substrate.

Note that one way to increase the rate of reaction is by stabilising the transition state of the substrate.
Enzymes can do this by preferentially binding to the substrate in a transition state conformation.
This is quite often the case, as enzymes usually stabilises the transition state of the substrate during the catalysis.

\section{Enzymes, Activation Energy, and \\Catalytic Mechanisms}

Enzymes can decrease the activation energy of a chemical reaction by various catalytic mechanisms.
There are 6 ways to do this:

\begin{enumerate}
\item Acid-base catalysis
\item Covalent catalysis
\item Metal ion catalysis
\item Electrostatic catalysis
\item Proximity and orientation effects
\item Preferential binding of the transition state complex
\end{enumerate}

In this lecture, only the first three was discussed.

\subsection{Acid-base Catalysis}

This kind of catalysis involves the transfer of H$^+$ ions, for example by glutamic and aspartic acids (COOH group), or by lysine (NH$_3^+$ group).
Note that these side chains require certain pH for it to be in the right ionisation state, and therefore to be able to transfer H$^+$ ions.
In this sense, histidine is very useful as it has a pK$_a$ of about 6.5, which is close to the physiological pH.
This means that with a slight change in pH, it can be either a H$^+$ donor or acceptor, and therefore able to take part in the acid-case catalysis reaction.

\subsection{Covalent Catalysis}

In a covalent catalysis, the enzyme forms a highly reactive, short-lived intermediate which is covalently attached to the enzyme.

\begin{center}
\ch{A-B + E -> A-E + B ->  A + B + E}
\end{center}

The enzyme requires an amino acid with a side chain that can act as a nucleophile (e.g.ser, cys)

\subsection{Metal Ion Catalysis}

About a third of known enzymes require metal ions for their catalytic activity.
Metals can provide substrate orientation, binding energy, and/or sites for redox reactions.
Metals act as a cofactor in these reactions.

\section{Co-factors and Co-enzymes/Co-substrates}

Some enzymes require other factors to catalyse chemical reactions.
These factors can be metal ions or small organic molecules (co-enzymes).
Co-enzymes, or co-substrates, are molecules that help with the reaction, and is changed after the reaction occurs.
Co-substrates are usually derived from vitamin derivatives.
(An example used in the lecture was NAD)

\section{Enzyme Nomenclature}

Since there are MANY enzymes in the world, scientists needed to standardise the naming of the enzymes so other people knew which enzymes they were talking about.
The classes of enzymes are:

\begin{enumerate}
\item Oxidoreductases -- transfer electrons
\item Transferases -- transfer groups
\item Hydrolases -- hydrolysis reactions
\item Lyases -- Addition of groups to double bonds (or reverse)
\item Isomerases -- forms isomeric forms of a molecule
\item Ligases -- formation of C-C, C-S, C-O and C-N bonds (coupled with ATP hydrolysis)
\end{enumerate}

### 2015-07-30 bioc192/lect10.tex
finished lecture 11, onto lecture 12

@@
 
-In an enzyme catalysed reaction, substrate binds to the active site and forms a enzyme-substrate complex (ES), which is stabilised by non-covalent interactions.
+In an enzyme catalysed reaction, substrate binds to the active site and forms an enzyme-substrate complex (ES), which is stabilised by non-covalent interactions.
 Note that this interaction is specific.
 As the reaction proceeds, transition state (T), which is neither a substrate nor the product, is formed, then proceeds further into enzyme-product complex (EP).
-This complex can then be dissociated to form the original enzyme ready for another reactions, and release the product.
+This complex can then be dissociated to form the original enzyme ready for another reaction, and release the product.
 

### 2015-07-30 bioc192/lect11.tex
finished lecture 11, onto lecture 12

\section{Enzyme Kinetics}

It is important for biochemists to be able to compare different enzymes, so they can compare how well an enzyme is to the rest.
One way (or maybe the only way) to do this is to compare the enzyme kinetics -- how well an enzyme can catalyse a reaction.

\section{Measuring Enzyme Catalysis}

Enzyme catalysis of a reaction can be measured from the progress curve of a reaction.
Progress curve is the graph of how much product is made (or how much substrate is used up) over time.
It is important to measure the initial reaction velocity of the reaction from this progress curve in order to calculate some other important measures of enzyme kinetics.

This initial reaction velocity (V$_i$) is the slope of the tangential line of the initial part of the progress curve.
Only the initial part is used, because as you progress through the reaction, some factors start to limit the reaction and becomes unreliable.

\section{Effect of Enzyme Concentration on the Rate of Reaction}

Consider the situation where there is an excess substrate (i.e. unlimited source of substrate).
As the concentration of the enzyme increases, the rate of reaction increases.
\textbf{This means that the velocity of the reaction (V) is proportional to the enzyme concentration, when there is excess substrate.}

This is expected, since there are more enzymes to catalyse the reaction -- as more enzymes are introduced to the system, more enzymes can react with the substrate, and therefore produce more products.

\section{Effect of Substrate Concentration on the Rate of Reaction}

Consider the situation where the concentration of enzyme is fixed.
As we increase the concentration of substrate, the rate of reaction will initially increase linearly, but as more and more substrates are introduced, the rate of reaction reaches a `maximum'.
This is because all of the active site of the enzymes become occupied as the substrate concentration increases -- it reaches the `maximum capacity' of the enzyme, and the enzymes cannot increase the rate of reaction no matter how hard it tries.

In kinetics term, the enzyme shows first-order kinetics initially (where the rate depends on the substrate concentration), but then shows zero-order kinetics (where the rate does not depend on/changes with substrate concentration).
When the velocity of the reaction is plotted with the substrate concentration, this is quite apparent from the hyperbolic curve of the graph.

\subsection{V$_{max}$ and K$_M$}

From this V vs S curve, we can identify two kinetic parameters: V$_{max}$ and K$_M$.
V$_{max}$ is defined as the maximum velocity of the reaction when an enzyme is saturated with substrate.
K$_M$ is defined as the \textbf{substrate concentration} at which the velocity of the reaction is half the maximum velocity ($V = \frac{V_{max}}{2}$).
K$_M$ is also called the Michaelis constant.

Note: do NOT get this mixed up with ``K$_M$ = $\frac{V_{max}}{2}$'' -- remember that K$_M$ is the \textbf{substrate concentration} when V = $\frac{V_{max}}{2}$.

### 2015-08-03 bioc192/lect12.tex
finished lecture 12, onto lecture 13

\section{The Michaelis-Menten Equation}

Using V$_{max}$ and K$_M$, we can define the Michaelis-Menten equation:

\begin{center}
\large{V = $\frac{V_{max} [S]}{K_M + [S]}$}
\end{center}

Few assumptions must be made in order to derive this equation:
\begin{center}
\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
\end{center}
\begin{enumerate}
\item ES formation and breakdown is in rapid equilibrium, i.e. the ES concentration does not change over time ({$\frac{d[ES]}{dt}$} = 0).
\item Enzyme is saturated with substrate, therefore it is at V$_{max}$.
\item Initial rate is used, meaning that the substrate concentration does not change significantly.
\item Initial rate is used, meaning that there is no product present.
\end{enumerate}

Michaelis-Menten equation is a simple model of an enzyme catalysed reaction based on the assumptions stated above, and is applicable to most enzymes.

\section{Lineweaver-Burk Plot}

Since the V vs [S] curve is hyperbolic, you can only estimate the V$_{max}$ from it, and therefore you can only estimate the K$_M$ value.
In order to measure the V$_{max}$ and K$_M$, you can take the double reciprocal of the V vs [S] curve (i.e. take the reciprocal of both V and S).
This gives us the Lineweaver-Burk plot, which you can measure the V$_{max}$ and K$_M$ from the y- and x-intercept, respectively.

\section{Significance of K$_M$}

\begin{center}
\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
\end{center}

K$_M$ characterises a single enzyme-substrate pair.
K$_M$ is independent of the substrate concentration, but is dependent on the substrate, meaning that K$_M$ is different for different substrate.

K$_M$ is defined as the ratio of the breakdown of the ES complex and the formation of the ES complex:

\begin{center}
\large{K$_M$ = $\frac{breakdown[ES]}{formation[ES]}$ = $\frac{k_{-1} + k_2}{k_1}$ = $\frac{k_{-1}}{k_1}$}
\end{center}

Usually, k$_2$ is much slower than k$_{-1}$, so we can ignore k$_2$.
K$_M$ is also called the ES dissociation constant, and its unit is the same as the concentration.

Low K$_M$ indicates that k$_{-1}$ is low and k$_1$ is high, and therefore it has a high affinity between the enzyme and the substrate, and binds to one another tightly, forming a strong ES complex.
High K$_M$ indicates that k$_{-1}$ is high and k$_1$ is low, and therefore it has a low affinity between the enzyme and the substrate, and binds to one another weakly, forming a loose/weak ES complex.

Therefore, you can use K$_M$ as an indicator of substrate preference of an enzyme.

\section{Physiological Significance of K$_M$}

In the cell, the substrate concentration is often below the K$_M$ for a particular enzyme-substrate reaction.
The physiological substrate concentration lies on the either side of the K$_M$, on the linear part of the V vs S curve.
This means that as the substrate concentration increases in the cell, you get a relative increase in the rate of reaction as well, and vice versa.
Since the physiological substrate concentration is around K$_M$, the enzyme is able to tolerate both increase and decrease in the cell's substrate concentration.

When you think about it, if the physiological substrate concentration was at the point where enzyme is working at its V$_{max}$, the enzyme will not be able to cope with the increase in substrate concentration -- it is already working full-on, and cannot process the extra substrate.
This can lead to the accumulation of the substrate in the cell, and could have some effects on the cell.

\section{V$_{max}$ k$_{cat}$}

\begin{center}
\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
\end{center}

\begin{enumerate}
\item k$_2$ is the rate-limiting step of the reaction
\item The rate of the reaction is proportional to the total enzyme concentration ([E$_{total}$]).
\end{enumerate}

This gives us the definition of k$_{cat}$ and V$_{max}$, relative to one another:

\begin{center}
V$_{max}$ = k$_2$ + [E$_{total}$] \ch{->} k$_{cat}$ = $\frac{[E_{total}]}{V_{max}}$
\end{center}

k$_{cat}$ is the turnover number, or the catalytic rate constant of an enzyme.
The unit is in s$^{-1}$.
\textbf{The number of moles of substrate converted into product, per mole of enzyme per second.}

\section{k$_{cat}$, K$_M$ and Catalytic Efiiciency}

\begin{center}
Catalytic Efficiency = $\frac{k_{cat}}{K_M}$
\end{center}

The ratio of k$_{cat}$ and K$_M$ is the measure of the enzyme efficiency -- how quickly a product is formed relative to how well the substrate binds.
The catalytic efficiency will therefore have the highest value when k$_{cat}$ is high (high turnover), and low K$_M$ (high binding affinity).

### 2015-08-04 bioc192/lect13.tex
finished lecture 13, onto lecture 14

\section{Enzyme Inhibitor}

Inhibitor is a small molecule that can bind to the enzyme and reduce its catalytic activity.
These inhibitors are important in various ways: natural inhibitors regulate the metabolism in the cell; many drugs, poisons and toxins target enzymes; and is also used for studying metabolic pathways and mechanisms of an enzyme.
There are two types of enzyme inhibitors: reversible and irreversible inhibitors.

\section{Irreversible Inhibitors}

Irreversible inhibitors are inhibitors that covalently binds to the enzyme, to a specific amino acid side chain.
This means that the inhibitor permanently inactivates an enzyme, as it cannot be unbound.
Examples of irreversible inhibitors include penicillin (targets glycopeptide transpeptidase in bacterial cell wall) and organophosphates (inhibitor of acetylcholinesterase).

\section{Reversible Inhibitors}

There are two types of reversible inhibitors: competitive and non-competitive inhibitors.
Both of these inhibitors bind to the enzyme non-covalently, and can be `knocked off'.
Some inhibitors can be knocked off more easily than the others, depending on how tight the inhibitor binds to the enzyme.

\begin{center}
\ch{ E + I <=> EI }\\

\vspace{0.5cm}

\ch{K$_i$ = $\frac{[E][I]}{[EI]}$}
\end{center}

K$_i$ is the dissociation or inhibition constant -- just like the K$_M$ with normal substrate.
The lower the K$_i$, the stronger the inhibitor binds to the enzyme.
K$_i$ \textless{} $10^{-9}$ mol L$^{-1}$ is strong binding.

\subsection{Competitive Reversible Inhibitor}

These inhibitors bind to the enzyme at the active site, and competes with the substrate molecule.
This means that (usually) the inhibitor will have similar structure as the substrate, as it can bind to the active site.

\begin{center}
\textbf{Competitive inhibitors increases the K$_M$, but does not change the V$_{max}$ of the enzyme.}
\end{center}

This means that the apparent affinity of an enzyme for the substrate is reduced (as the K$_M$ is increased).
The V$_{max}$ is not affected, as the binding of the inhibitor is NOT going to affect the catalytic activity of the enzyme -- the enzyme can still reach the V$_{max}$, it just needs more substrate to out-compete the inhibitor (hence the increase in K$_M$).
(The binding of the inhibitor is not going to affect how fast the reaction proceeds, once the real substrate binds to the enzyme).

\subsection{Non-competitive Reversible Inhibitor}

These inhibitors bind to elsewhere to the active site of the enzymes and causes a conformational change, and therefore affects the efficiency of the catalysis.

\begin{center}
\textbf{Non-competitive inhibitors decreases the V$_{max}$, but does not change the K$_M$ of the enzyme.}
\end{center}

When the inhibitor binds to the enzyme, it does not occupy the active site where the substrate binds to.
This means that the active site is free for the substrate to bind -- it's just that the reaction does not proceed as efficient as without the inhibitor bound to it.
Hence the decreased V$_{max}$, but unchanged K$_M$.
(Another way to think about it -- V$_{max}$ is proportional to the total enzyme concentration and the k$_{cat}$, so if k$_{cat}$ decreases, then V$_{max}$ will also decrease).

\section{Methods of Enzyme Regulation}

\subsection{Allosteric Enzyme Regulation}

Allosteric enzymes generally display co-operativity and usually multi-subunit.
The enzyme's reaction curve shows a sigmoidal curve.
Activity of allosteric enzymes can be regulated by binding molecules (effectors) away from the active site, which can activate or inhibit the enzyme.
Regulatory effect is achieved by conformational changes caused by the effectors binding to the enzyme.
Note that the inhibitors shift the curve to the right, whereas the activator will shift the curve to the left.

\subsection{Covalent Modification of Enzymes}

In this lecture, only phosphorylation was mentioned.
Phosphorylation and de-phosphorylation is the most common, and is catalysed by kinases and phosphatases, respectively.
Phosphorylation state of an enzyme determines whether the enzyme is `on' or `off'.
However, phosphorylation does not necessarily mean that the enzyme is on -- it really does depend on the enzyme you're studying.

\subsection{Activation of Enzymes by Cleavage}

Some enzymes are synthesised as zymogens -- the inactive form of the enzyme.
The enzyme is only active once the enzyme is cleaved at a certain amino acid residue.
Many of the enzymes involved in the digestive processes and blood clotting proteins are made as zymogens.

\section{Isoenzymes/Isozymes}

Multi-subunit enzymes may exist in multiple forms in an organism -- these enzymes are called isoenzymes.
Isoenzymes differ in physical, kinetic and immunological properties, but all of them catalyse the same reaction.
Isoenzymes are formed from two different genes that make slightly different subunit for the enzyme.
Isoenzyme is a form of transcriptional regulation of cellular activity, as some cells in your body has to catalyse the reaction differently to another part of your body.

### 2015-08-05 abstract/abstract.tex
draft introduction

\documentclass[a4paper,12pt]{report}

\usepackage[
    style=authoryear,
    firstinits=true,
    backend=biber,
    uniquename=init,
    isbn=false,
    doi=false,
    url=false,
    mincitenames=1,
    maxcitenames=2,
    minbibnames=7,
    maxbibnames=7,
    natbib=true]{biblatex}
\renewcommand*{\bibfont}{\small}
\renewbibmacro{in:}{}
\renewbibmacro{volume+number+eid}{\printfield{volume}\addcolon}
\renewbibmacro*{name:andothers}{% Based on name:andothers from biblatex.def
    \ifboolexpr{
        test {\ifnumequal{\value{listcount}}{\value{liststop}}}
        and
        test \ifmorenames
    }
        {\ifnumgreater{\value{liststop}}{1}
           {\finalandcomma}
          {}%
            \andothersdelim\bibstring[\emph]{andothers}}
            {}}
\DeclareNameAlias{sortname}{last-first}
\DeclareFieldFormat[article]{title}{#1}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat*{volume}{\mkbibbold{#1}}

\bibliography{../../../References/BibTeX/MSc}

\begin{document}

\begin{center}
\large{\textbf{Body Mass Index and Pathway Dysregulation in Cancer}}\\

\vspace{0.3cm}

Riku Takei (MSc)\\
Supervisor: Assoc. Prof. Mik Black
\end{center}

\section*{Introduction}

It is commonly understood now that cancers are caused by dysregulation of various pathways (known as the ``Hallmarks of Cancer'') that allow the cells to proliferate, survive and migrate \citep{Hanahan2011}.
There have been many reports on the association between obesity and worse cancer prognosis, suggesting that there may be mechanisms that underlie the tumorigenesis and/or cancer progression specifically in obese patients.
\citet{Creighton2012} identified 799 genes that were able to distinguish the obese patients from the non-obese patients in a sample of breast cancer patients.
Furthermore, they showed that small molecules were able to induce or repress these genes, signifying the fact that these obesity specific genes could be used as an indicator for alternative or combinatorial chemotherapy and assist in better clinical decisions.

This study has shown that there were obesity specific genetic signatures in breast cancer, but can these genetic signatures be applied to other cancer types?
A couple of studies have shown that there are genetic signatures common in multiple cancer types, suggesting a possibility of such genetic signatures that can be applied to multiple cancer types, which are also specific to obesity \citep{Alexandrov2013,Lawrence2014}.
However, these studies have not shown any evidence of obesity specific genetic signatures that are common in multiple cancer types.

This research aims to identify genetic signatures that are specific to obesity across multiple cancer types, and to investigate whether there are any pathways being dysregulated in cancers based on these genetic signatures.
Better understanding of the pathways being dysregulated in cancers in obese patients may lead to better clinical decisions and prognosis, and contribute towards personalised medication in the future.
%Furthermore, they were able to use small molecules to induce or repress these obesity specific genes, signifying the fact that these genes may be targeted for treatment in obese patients.

\section*{Methods}

\section*{Results and Discussion}

\printbibliography

\end{document}

### 2015-08-05 bioc192/lect14.tex
finished lecture 14, onto lecture 15

\section{Pharmacology and Toxicology}

Pharmacology is the study of therapeutic drugs and design of these drugs.
It is the study of how drugs work and how it is metabolised in the cell.

Toxicology is the study of poisoning/toxins.
It is the study of the harmful uses drugs in humans, animal and the environment.

\section{Drugs}

A drug is a chemical that changes the behaviour or function of an individual system, organ, tissue or invading organism.
In other words, a chemical that changes the behaviour of the cell.

\subsection{Sources of Drugs}

You can extract and purify drugs from plants and animals, but have some issues around contamination, etc.
Current sources of drugs are made synthetically, so there will be no contamination from purification.
Drug companies can screen for millions of chemicals, and choose the best one and study it (this process is very expensive).
Rational drug design is where you study the actual target and the disease so that you can make a rational decision about how to make or design a drug for that disease.
Another way to make drugs is to use computer to model what kind of drug will be the best to use.
Once identified, it can then be tested in the lab.

\subsection{Targets of Drugs}

Drugs can bind to a molecule and cause an effect on the cell -- the molecule it binds to will be its molecular target.
Understanding the pathways of the disease or the targets of the drug leads to better drug design.

\section{Agonists and Antagonists}

Agonists are the drugs or molecules that binds to a receptor and can produce a full biological response.
On the other hand, antagonists are drugs or molecules that binds to a receptor and prevents the response to occur (i.e. opposite of agonists).

\section{Receptors}

There are two types of receptors described in these lectures -- drug receptors and physiological receptors.

\subsection{Drug Receptor}

Drug receptor is ANYTHING that the drug binds to.
This can be a proteins, lipids, DNA, RNA, or whatever as long as the drug binds to it and have an effect on the cell through the molecule.

\subsection{Physiological Receptor}

Physiological receptors are normal receptors that are present on the cell surface.
Most of these are transmembrane, meaning that it spans the whole membrane.
These receptors allow signalling to occur from the outside (extracellular) to the inside (intracellular) of the cell.
They also regulate cell function by responding to stimuli such as hormones, neurotransmitters or growth hormones and produce a signal for the cell to do something.
Most drugs activate or inhibit these kind of receptors.

There are four types of physiological receptors: ligand-gated/voltage-linked ion channels, G-protein coupled receptors (GPCR), kinase-linked receptors, and Nuclear receptors.

\section{Ligand-Gated Ion Channels}

Ligand-gated ion channels are proteins present in the cell membrane.
This receptor has a receptor and an ion channel domains, and is made up from 5 subunits.
Binding of the ligand induces a conformational change in the protein, which opens up the channel and allow the ions to pass through.

Specific receptors have specific activating molecules, and only certain ions can pass through.
These receptors control the intracellular ion concentration which gives different charges to the cell.
The response times for these receptors are within milliseconds, which is ideal for fast response (e.g. in nerves).

Using nicotinic acetylcholine receptor as an example, normal substrate/agonist of this receptor is acetylcholine.
Another agonist is nicotine, and an antagonist is tubocurarine.
Tubocurarine is an irreversible inhibitor, so it binds to the receptor and won't release, hence you can eat the hunted animal.

\section{Voltage-Gated Ion Channels}

These receptors are made fromm four subunits, and opened in response to changes in voltage across the membrane (e.g. voltage change caused by ligand-gated ion channels).
The mechanism itself is very similar to the ligand-gated channels.
The drugs that can block these receptors literally blocks the ion channel and prevents the ions to move through the pores (e.g. tetrodotoxin).

### 2015-08-06 abstract/abstract.tex
chippin' away at the abstract

@@
 This study has shown that there were obesity specific genetic signatures in breast cancer, but can these genetic signatures be applied to other cancer types?
-A couple of studies have shown that there are genetic signatures common in multiple cancer types, suggesting a possibility of such genetic signatures that can be applied to multiple cancer types, which are also specific to obesity \citep{Alexandrov2013,Lawrence2014}.
+A couple of studies have shown that there are genetic signatures common in multiple cancer types \citep{Alexandrov2013,Lawrence2014}.
 However, these studies have not shown any evidence of obesity specific genetic signatures that are common in multiple cancer types.
 
-This research aims to identify genetic signatures that are specific to obesity across multiple cancer types, and to investigate whether there are any pathways being dysregulated in cancers based on these genetic signatures.
-Better understanding of the pathways being dysregulated in cancers in obese patients may lead to better clinical decisions and prognosis, and contribute towards personalised medication in the future.
+This research aims to identify genetic signatures that are specific to obesity across multiple cancer types, and to investigate whether there are any common pathways being dysregulated in cancers based on these genetic signatures.
+Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to better clinical decisions and prognosis, and contribute towards personalised medication in the future.
 %Furthermore, they were able to use small molecules to induce or repress these obesity specific genes, signifying the fact that these genes may be targeted for treatment in obese patients.
@@
 
+\textit{Downloading data from The Cancer Genome Atlas (TCGA):}
+Clinical data of the patients from 33 different cancer types were downloaded from TCGA database.
+Out of these 33 cancer types, 14 types had the height and weight data of the patients.
+
+\textit{Downloading and processing the ucec RNA-seq data:}
+The RNA-seq data of ucec was downloaded from the International Cancer Genome Consortium (ICGC) database.
+The RNA-seq data was then processed in R for subsequent analysis.\\
+
+\textit{Transforming the ucec data with the obesity metagene from \citet{Creighton2012} paper:}
+To see whether the obesity signatures identified in the Creighton \textit{et al.} paper could be generalised into other cancer types, a transformation matrix was created from the obesity metagene and used to transform the ucec data.
+The transformed ucec data produced a metagene, which was used to assess whether the metagene correlated with the clinical data (i.e. body mass index).
+\\
+
+\textit{Differentially expressed gene analysis in ucec data:}
+Differentially expressed gene analysis was carried out on the ucec data to identify the genes that were differentially expressed in obese patients, compared with the normal/overweight patients.
+
 \section*{Results and Discussion}
+
+The maximum number of patients with both height and weight data was 512 patients in the uterine corpus endometrial carcinoma (ucec) data.\\
 

### 2015-08-12 bioc192/lect15.tex
finished lecture 15, onto lecture 16

\section{Nuclear Receptor}

Nuclear receptors, unlike the transmembrane voltage/ligand-gated channels, reside in the cytosol of the cell (i.e. within the cell).
This is partly because the nuclear receptors act in the nucleus, where it interacts with the DNA for gene transcription.
These nuclear receptors binds and interacts with the DNA at the response element for that receptor, which requires multiple regulation mechanisms.
The binding of the nuclear receptor-ligand complex to the response element causes transcription/gene activation.
Since the effect of the cellular response is observable only after the protein is made, the response time is quite slow -- takes hours to days for the response.

\section{Mechanism of Nuclear Receptor Signalling}

For a ligand to enter the cell, the ligand must be lipophilic/hydrophobic, otherwise it will not be able to enter the cell through the membrane.
Normally, the receptor is bound to a heat shock protein (Hsp; a chaperone), that prevents the receptor from binding to the DNA or other receptors.
When the ligand binds to the receptor, the Hsp is displaced/removed, and the receptor dimerises (i.e. binds to another receptor).
The receptor cannot bind to the DNA unless it is dimerised.
Once bound to the DNA, the dimer recruits/associates with many other proteins to activate the gene (e.g. RNA polymerase) and transcribes the gene for cellular response.

\subsection{Regulation of Nuclear Receptor}

There are multiple regulatory steps in this mechanism to prevent unwanted signal and cell response.
Firstly, the receptor is located inside the membrane, meaning that only the lipophilic/hydrophobic ligands can access/activate the receptor.
The Hsp that is bound to the receptor is also important, because without them, the receptor may be able to bind to the DNA cause unwanted gene activation.
Dimerisation is another level of regulation -- it requires the receptor to be dimerised, or else it's not going to activate the gene.

\section{Structure of Nuclear Receptor}

There are three parts to the receptor.
The first part is the ligand binding site (steroid hormones).
Second part is the DNA binding domain, which is the site where the receptor binds to the DNA.
This part contains the zinc finger motif for DNA binding.
The last part is the transcription regulatory domain, where the transcription factors and other machinery binds to for gene activation.

\section{What Genes are Activated by Nuclear Receptor}

The genes activated by the receptor depends on the tissue site the receptor resides in, what cell type it's in, and what kind of ligand/drug binds to the receptor.
This means that the receptor can activate multiple different genes, and therefore have different cellular response depending on the ligand, tissue site, etc. (i.e. it's not a one-gene-to-one-receptor thing).
For example, estrogen receptor have different cellular response depending on the tissue site.

\section{Nuclear Receptor and Cancer}

Cancer is a very common disease and have high death rates.
Cancers are the cells that divide uncontrollably and proliferate while avoiding cell death.
The growth of the cancer cells are usually driven by the hormones, and therefore driven by the receptors of these hormones.

\subsection{Breast Cancer}

In some breast cancer, the cause is from over activation of estrogen receptor.
In pharmacology terms, estrogen is the agonist of the ER, and over dosage of this can lead to cancer.
Tamoxifen, on the other hand, is an antagonist of ER and prevents the cellular response from occurring.
However, although tamoxifen reduces the growth of breast cancer, it has an agonist side effects in bone (strengthening of bone) and in uterus (increased risk of endometrial cancer).

\subsection{Androgen Receptor and Prostate Cancer}

Testosterone is formed from the testis and adrenal gland (mostly testis).
Testosterone is the ligand for the androgen receptor.

Prostate cancer is (at least initially) by over activation of the androgen receptor by testosterone.
One way to treat it is to remove the testicles.
Another way is to use drugs that reduces the androgen level.
Bicalutamide is one such drug -- it is an androgen receptor antagonist and it also accelerates the AR degradation.
However, it also has a lot of side effects.

\section{Effect of Hormones on Living Organisms}

In a waterway with low levels of estrogen, some male fish developed in to female-like fish.
This was probably due to the chronic exposure to low level of estrogen in the water.
This raises an important question where low levels of hormone may affect the development of humans, if exposed chronically.
Although there has only been speculations, this theory could be true.

### 2015-08-12 bioc192/lect16.tex
finished lecture 16, onto lecture17

\section{G-protein Coupled Receptors (GPCR)}

G-protein Coupled Receptors (GPCRs) are transmembrane receptors that are associated with G-proteins, as the name suggests.
GPCRs acts/responds in the timescale of seconds -- slower than the ligand/voltage gated ion channels, but faster than the nuclear receptors.
GPCRs are important as it is the largest group of the receptors, and therefore an easy target for drugs.
Also, GPCRs are activated by diverse ligands.
About 30 to 50\% of the drugs target the GPCRs, but still only 10\% of the known GPCRs are targeted.

\section{Mechanism of GPCRs}

The ligand binds to the transmembrane receptor on the cell surface.
Before the ligand is attached to the receptor, the receptor is not associated with the G-protein.
Once the ligand binds to the receptor, the conformation of the receptor changes, which allows the G-protein to interact with the receptor.
Interaction with the receptor causes the G-protein to change conformation as well, which increase or decrease the activity of a particular enzyme.
Depending on the activity of the enzyme, the amount of second messenger molecule produced by the enzyme is altered, and therefore causes some cellular response.

\section{Dopamine Receptor}

Dopamine receptor is an example of GPCR.
When the ligand (dopamine) binds to the receptor, it causes a conformational change in the G-protein which decrease the enzyme activity and therefore decreases the amount of second messenger (cAMP) produced.

\subsection{Parkinson's Disease}

Parkinson's disease is caused by decreased level of dopamine, due to the damage in the neural cells that produce dopamine.
Parkinson's can be treated by using levodopa -- a prodrug that is processed into dopamine.
The problem with levodopa is that it is hard to balance/control the effect of levodopa, and leads to psychological effects (such as hallucination/psychosis) and motor complications.

\subsection{Psychosis}

Psychosis occurs when there is too much dopamine around, and over activates the dopamine receptor.
In this case, antipsychotics (antagonists of dopamine receptor) is used to prevent the dopamine to have too much effect on the receptor.
This means that the level of dopamine is stall high, but the effect is inhibited by using antagonists.
Again, however, if you use too much of antipsychotics, you get a Parkinson's like syptoms, signifying the fact that it's very hard to balance the effects of these drugs.

### 2015-08-15 bioc192/lect17.tex
finished lecture 17, onto lecture 18

\section{Membrane Lipids and Proteins}

Membrane lipids allow the cells to make compartments within the cell.
This allows the cell to set up specific functions for different compartments (e.g. setting different pH environment, isolate function/group of proteins).
However, there are some issues with membrane lipids -- they act as a barrier for the molecules to pass through the membrane.
To understand how molecules pass through the membrane, you need to  know what the membrane is made up of.

\section{Membrane Composition}

Membrane lipids are made from amphipathic molecules, meaning that the molecules have both charged and uncharged parts.
The hydrocarbon tails are hydrophobic and are 16$\sim$22 carbons long.
In animals, the length of the hydrocarbon tails are usually even.
These hydrocarbon tails can be either saturated (no double bonds) or unsaturated (with double bonds).
Note that the double bonds introduce a kink in the tail.

In aqueous solution, the polar group of the molecules are exposed to the solution, whereas the hydrophobic part forms the inner part of the membrane.
This forms a well-known lipid bilayer, which is energetically favourable for the molecules to form in aqueous environment.

\section{Membrane Lipid Structure}

Membrane lipids all have a glycerol backbone which two fatty acids and a phosphate group attaches to.
As mentioned, the fatty acids can be of different length, with or without double bonds.
The phosphate group provides the charge for the amphipathic molecule, and often have polar groups added to the phosphate.
This provides the group extra charge which can form patches on the surface of the membrane.

Note that there are other types of membrane lipids present in different tissue of the same organism, or maybe in tissues from another organism, depending on the function of the tissue in its organism.

\section{Membrane Fluidity}

Membrane fluidity is affected by temperature, the type of fatty acids present, and the presence of cholesterol.
All of these factors have an effect on the hydrophobic, non-covalent interaction between the lipids.
Reminder: non-covalent interactions are very weak on its own, but strong when there are a lot of them, and the strength of the interaction is also dependent on the distance/hoe close to each other.

\subsection{Temperature}

Temperature affects the membrane fluidity by disrupting the non-covalent interactions between the hydrocarbon tails.
As the number of the carbon increases in the hydrocarbon tails increases, the melting point also increases, and therefore less fluid.
This is because there are more carbons for hydrophobic interaction that strengthens the interaction.

\subsection{Type of Fatty Acid}

The type of fatty acids, or more like how saturated a fatty acid is, affects the membrane fluidity.
When the hydrocarbon tails are unsaturated, that is, it has double bonds in it, it has a kink.
This kink causes the hydrocarbon tails to be more separated and the packing of the membrane lipids less ordered.
Since the kink is going to spread the hydrocarbon tails apart further apart, this increases the distance between the tails, and therefore weakens the hydrophobic interaction between them (therefore more fluid).

\subsection{Presence of Cholesterol}

Cholesterol has a rigid ring structure that can be used as a ``scaffold''.
This allows the lipids to anchor themselves to it and form a more ordered structure, and therefore makes the membrane less fluid.

\section{Lipid Bilayer is a 2D Fluid}

Lipid bilayer is not a solid, static structure -- it is a fluid.
Lipids are able to migrate rapidly in the plane of the membrane, and ``flip-flopping'' between the planes are very rare.
This is because the polar part will have to move through the non-polar part of the membrane.

\section{Membrane Proteins}

The membrane is jam-packed with proteins.
Proteins are also able to move in the membrane -- it is dynamic.
There are many functions of membrane proteins -- cell to cell contact, surface recognition, cytoskeleton contact, enzymes, transporters, receptors and signalling.

Proteins can be either integral (all or some part of the protein is embedded in the membrane), or peripheral (on the surface of the membrane).
Peripheral proteins can be held in pace on the membrane surface by polar patches from, for example, the polar groups attached to the phosphate group, forming an electrostatic interaction.

Note that the proteins have ``sides'' on the membrane, and are asymmetric.
This means that not all proteins that are exposed to the extracellular surface are also exposed on the intracellular surface.
For example, most (if not all) of the proteins that are glycosylated are exposed on the extracellular surface, and not on the intracellular surface.

\section{How Are Proteins Attached to the Membrane?}

Non-polar amino acids of the protein resides within the lipid bilayer -- it is unfavourable for hydrophobic amino acids to move out of the membrane, so the proteins is anchored to the membrane.
Proteins can also have an anchoring domain that can be used to anchor the domain to the membrane.

Lipidation, a type of post translational modification, adds fatty acid groups to the protein which can insert it into the membrane for anchorage.

### 2015-08-16 bioc192/lect18.tex
finished lecture 18, onto lecture 19

\section{Transport of Non-polar Molecules Through the Membrane}

Transport of non-polar molecules through the membrane is dependent on both the hydrophobicity and the concentration of the molecule.
Unlike polar molecules, non-polar molecules are relatively easier to diffuse through the membrane due to its hydrophobic nature of the molecule.

\subsection{Permeability Coefficient and Oil-Water Partition Coefficient}

The permeability coefficient of a molecule describes how well the molecule is able to pass through the membrane -- i.e. how permeable the molecule is.
The smaller the coefficient is, the harder for the molecule to pass through the membrane, and therefore less permeable.

The oil-water partition coefficient describes how hydrophobic a molecule is.
This means that the higher the oil-water partition coefficient is for that molecule, the more hydrophobic the molecule is.

When these two factors are graphed for different kinds of molecules, you can clearly see that the higher the oil-water partition coefficient (i.e. more hydrophobic the molecule is), the higher the permeability coefficient.
This shows that the more hydrophobic a molecule is, the easier for that molecule to pass through the membrane.

\subsection{Concentration Gradient vs. Flux}

When the  flux  is plotted against the concentration gradient of the molecule across the membrane, it is clear that the flux is proportional to the concentration gradient.
This means that the higher the concentration gradient across the membrane, the higher the flux of that molecule through the membrane (i.e. faster/more will go through).

\section{Transport of Polar Molecules Through the Membrane}

Unlike the non-polar molecules, the lipid bilayer represents an energy barrier for the transport of polar molecules across it.
So, for the polar molecules to pass through the membrane, the energy barrier must be decreased.
Proteins within the membrane make small holes in the membrane to allow polar molecules to pass through.

\subsection{Channels}

Channels, when opened, are open to both the intracellular and the extracellular space (i.e. open to both ends).
These channels can be opened or closed spontaneously, and are ``gated''.
Since the mechanism of the channels are fairly simple, the rate of transport of molecules across the membrane when opened can approach the rate of diffusion -- about 10$^7\sim$10$^8$ molecules per second.
The structure of the holes of the channels provide some specificity of the molecules that can go through the membrane.

\subsection{Transporters}

Unlike the channels, transporters are only open to either the intracellular or the extracellular space.
The binding of the molecule that is to be transported induces a conformational change in the transporter which opens the opposite end, allowing the release of the molecule to the other side.
The release of the molecule causes another conformational change that reverts the transporter back into its original state and the process repeats.
Since the process of transporting the molecule requires conformational change of the transporter itself, the rate of transport is much slower compared to the channels -- about 10$^2\sim$10$^3$ molecules per second.

Another characteristic thing about transporters is that they are very much like an enzyme -- transporters are very selective of what they move through the membrane, and it also shows a Michaelis-Menten-like kinetics.
For example, glucose and mannitol are two very similar structured molecules that have similarly low permeability coefficient without a transporter.
However, when GLUT transporter is present, the permeability coefficient increases a lot for the glucose, but not for mannitol, showing that the transporters are selective in which molecule it transports.

The kinetics of the transporter is represented by the equation below, where J is the flux and [A] is the concentration gradient of the molecule across the membrane.

\begin{center}
\large{$J_A = \frac{J_{max}[A]}{K_M + [A]}$}
\end{center}

\section{Types of Transport Processes}

\begin{description}
\item [Non-mediated] Requires no protein for transport -- free diffusion.
\item [Mediated] Requires a protein. Also called facilitated transport.
\item [Passive] Down a concentration gradient and no energy is required for the transport.
\item [Active] Up a concentration gradient and requires an input of energy. Energy can come from ATP hydrolysis or from the co-transport of another molecule moving down its concentration gradient.
\item [Co-transport] Movement of two molecules by a transporter.
\begin{description}
\item [Symport] Both molecules move through the transporter in the same direction.
\item [Antiport] The two molecules move in the opposite directions to one another.
\end{description}
\end{description}

\vspace{0.5cm}

\textbf{Note: for symport and antiport, both molecules may not be going down the concentration gradient -- one molecule may be going up the concentration gradient, whereas the other maybe going down the concentration gradient.}

### 2015-08-24 bioc192/lect19.tex
finished lecture 19, onto lecture 20

\section{Energy Metabolism and ATP}

We require approximately 1kg of ATP per kg of body fat for the day.
Obviously, we can't carry that much ATP every day, and so we have to make it as we need it.
Anabolism is the use of energy to form/make macromolecules from building blocks (e.g. making proteins).
Metabolism is the opposite of anabolism, where the macromolecules are broken down into smaller constituent/component/building blocks to be oxidised to produce energy, or to be used in anabolic processes.

From the first law of thermodynamics, the total energy within a system is constant -- i.e. energy can neither be created or destroyed.
This means that the energy must be converted from one form to another.

\section{Energy Balance in the Body}

\begin{center}
E$_{intake}$ = E$_{expended}$ + E$_{stored}$\\
\vspace{0.5cm}
E$_{expended}$ = Basal metabolism + Activities
\end{center}

As you can see from the equation, all of the energy taken into the body must be either used up/expended, or stored.
Clearly, to lose weight, you must either decrease your intake, or increase your energy expenditure.
You can increase the energy expenditure by increasing your basal metabolism and/or the amount of exercise/activities.
(Note that small amount of energy is lost in faeces, urine and sweat.)
Weight gain is a problem in developed countries, as they tend to have higher percentage of obese people.

The rest of this lecture will be focusing on each aspect of the equation.

\subsection{Subtopic: Leptin Receptor}

Leptin deficient mice eats too much.
Leptin is a hormone that binds to the leptin receptor, which acts so that the host feels full, and therefore stops eating.
So, if a mouse is leptim deficient, then the mouse will not feel full, and therefore continues eating.

\section{Units of Energy}

Joule (J) is the unit of energy, defined as the \textbf{energy required to push against 1N of force for 1m}.
An older metric is the calories, where 1cal is the energy required to heat 1g of water from 14.5$^\circ$C to 15.5$^\circ$C.
1cal is equivalent to 4.184J.
Daily energy intakes and expenditure usually measured as MJ.

\section{Measuring Energy Intake}

\subsection{Bomb Calorimetry}

You can measure the energy contained in the food by burning the food and measuring the change in temperature of the water.

\begin{center}
\ch{C6H12O6 + 6 O2 -> 6 CO2 + 6 H2O + heat}\\
\ch{C16H32O2 + 23 O2 -> 16 CO2 + 16 H2O + heat}
\end{center}

Carbohydrates releases 2813kJ/mol (glucose), whereas fatty acids release 10024kJ/mol (palmitate).
Since fatty acids are more reduced, it can produce more energy from oxidation.
One thing to note is that not all energy in food is available for the body (e.g. cellulose and nitrogen, which are excreted out).
Therefore, these factors must be taken into account for calculations.

\section{Atwater Factor}

Atwater factors are used to measure how much energy each constituents have.
Fat has 38kJ/g, carbohydrates and proteins have 17kJ/g and ethanol has 29kJ/g.
These atwater factors can be used to calculate how much energy is contained in the food.

\section{Measuring Energy Expenditure}

Bomb calorimetry of a human would be good, but not feasible for obvious reasons.
There are other ways to measure energy expenditure.

\subsection{Direct Calorimetry}

This method relies on measuring heat output from individual in a whole body calorimeter.
This is good for determining the basal metabolic rate (BMR) -- i.e. at rest.
However, it is quite hard to measure the energy used in more physical exercises, due to space constraints and complexity of the calculation and machinery.

\subsection{Indirect Calorimetry}

This method measures the energy expenditure based on the oxygen consumption and carbon dioxide production, which is measured using a respirometer.
Since certain amount of energy is associated with every litre of oxygen consumed (20.9kJ/L), you are able to calculate how much energy was used.
This allows us to calculate the energy expenditure for a range of activities, and also allows the calculation of the respiration exchange ratio (RER).

\subsection{Respiration Exchange Ratio}

\begin{center}
\large{RER = $\frac{CO_2 produced}{O_2 consumed}$}
\end{center}

RER tells us about what kind of fuels are used -- carbohydrates or fat.
Note that the stoichiometry of the oxygen and carbon dioxide is different for carbohydrates and fat.
Since the stoichiometry of oxygen to carbon dioxide is 1:1 in carbohydrates, RER = 1, but this not the case for fat, and therefore have RER = 0.7.

\section{Basal Metabolism}

Basal metabolism is the energy required for the maintenance of life -- general metabolism, muscle concentration, etc.
This is usually defined as the energy expenditure at rest.
BMR is affected by:
\begin{itemize}
\item gender
\item age 
\item body size
\item body composition
\item genetics
\item hormonal status
\item stress levels
\item disease status
\item certain drugs
\end{itemize}

BMR can be increased by: athletic training, fever, late pregnancy, drugs (e.g. caffeine) and hyperthyroidism.
BMR can be decreased by: malnutrition, sleep, drugs (beta blockers) and hypothyroidism.

\subsection{Hyper and Hypothyroidism}

\begin{description}
\item [Hyperthyroidism] Increased metabolic rate due to excessive production of thyroid hormone, which acts to increase the effect of insulin, etc. Symptoms include heat production, anxiety, etc.
\item [Hypothyroidism] Decreased metabolic rate due to thyroid hormone deficiency (from iodine deficiency). Symptoms include coldness, depression, etc.
\end{description}

### 2015-08-25 bioc192/lect20.tex
finished lecture 20, onto lecture 21

\section{Digestion and Absorption of Food}

Food macromolecules have to be broken down into smaller molecules for absorption, so it can be used by the body.

\section{Organs Involved in Digestion}

\subsection{Salivary Glands}

Produces saliva -- neutral pH and contains mucous and amylase that starts the digestion of carbohydrates.

\subsection{Stomach}

For food storage (controlled release into the duodenum) and mixing of the food with gastric juices.
Stomach release highly acidic acid (0.1M HCl) that denature the proteins.
It also secretes pepsinogen (zymogen) for protein digestion, and also mucous layer for protection.

The parietal cells produce the acid (HCl), and the chief cells produce the pepsinogen, which is activated by the acidic environment.

\subsection{Pancreas}

Pancreas is slightly alkaline pH 7, and secretes most digestive enzymes, including amylase, lipase and many proteases.

\subsection{Liver}

Synthesis of bile salts/acids (stored in gall bladder), which is important for fat digestion.

\subsection{Small Intestine}

Where the final phase of digestion and absorption occurs.

\section{Beaumont's Experiment}

Beaumont tested a range of food for digestion using the stomach acid from St Martin.
He showed that the digestion of food was dependent on temperature, mixing, stress, etc.
He also noted that he wasn't able to completely reproduce the acid, most likely due to the enzyme component of the gastric juice.

\section{Digestion}

There 2 main phases involved in digestion:
\begin{enumerate}
\item Hydrolysis of bonds connecting the monomer units in food macromolecules.
\item Absorption of products from the gut into the body.
\end{enumerate}

In the case of carbohydrates, proteins and TAGs, the bond to cleave would be the glycosidic bonds, peptide bonds and ester bonds, respectively.

\section{Carbohydrate Digestion}

Carbohydrates make up 40 to 50\% of the energy intake.
Carbohydrates include starch (amylose and amylopectin), glycogen, simple sugars and fibre (i.e. cellulose).
Cellulose cannot be digested in humans, as we lack the enzyme for breakdown of cellulose.

\subsection{Enzymes Involved in Carbohydrate Digestion}

The digestive process begins in the mouth where the salivary amylase starts breaking down the starch molecules into smaller sugars.
When it enters the small intestine, the pancreatic amylase (secreted from the pancreas) further breaks down the starch molecules into maltose and isomaltose.
The intestinal epithelial cells secretes maltase, isomaltase, lactase and sucrase, which breaks down the disaccharide molecules into monomers (glucose, fructose and galactose).
These monosaccharides are then absorbed into the body via the intestinal villi.

\begin{reactions*}
Maltose/Isomaltose &->[Maltase][Isomaltase] 2 Glucose\\
Sucrose &->[Sucrase] Fructose + Glucose\\
Lactose &->[Lactase] Galactose + Glucose
\end{reactions*}

\subsection{Lactose Intolerance}

This is caused by lactase enzyme deficiency.
Since lactose is not digested, it accumulates inside the intestine where the gut microbes can use it for fermentation.
Lactose intolerance causes bloating, flatulence and diarrhoea due to the fermentation.
Individuals with this should avoid lactose consumption.

\section{Carbohydrate Absorption}

Monosaccharides are absorbed into the body via the villi and microvilli (brush border) of the intestinal wall.
The villi and microvilli provides a large surface area for absorption.

\subsection{Glucose Transport}

Sugars are water soluble, and therefore require a transporter to pass through the cell membrane into the cell.
SGLT1 protein is present in the intestinal side of the cell membrane and transports glucose into the intestinal cell together with the Na ion (symport).
Movement of Na ion from the intestine into the cell is down the concentration gradient, which provides the energy for the glucose transport up the gradient into the cell.

On the other side of the cell membrane, GLUT2 protein is present.
Since the concentration of glucose in the cell is higher than the blood, the glucose can move down the concentration gradient through the GLUT2 transporter.
Note that there is Na/K pump on this side of the membrane to maintain the Na concentration gradient.

Once transported into the blood, other tissues can uptake glucose via GLUT3 (brain) and GLUT4 (muscle and adipose).

\subsection{Coeliac Disease}

Disease of the small intestine, where the body reacts against the wheat protein gluten.
This causes the villi to be flattened and nutrients are not absorbed properly.

### 2015-08-25 bioc192/lect21.tex
finished lecture 21, onto lecture 22

\section{Protein Digestion and Absorption}

Proteins can be used for energy, but it is more for regenerating the depleted amino acids.
It is also important for supplying the body with essential amino acids.
Proteins are the major source of nitrogen for purines, pyrimidines and haem.
The carbon skeleton of the amino acids can be used as fuel/energy production, where the amino group (i.e. nitrogen) is converted into urea and excreted.
Decrease/malnutrition of certain amino acid can lead to Kwashiokhor.

\section{Essential Amino Acids}

We cannot synthesise all 20 amino acids -- require 8 essential amino acids from the diet.
These amino acids are: L, K, T, Y, I, M, F and V.

\section{Regulation of Protein Digestion}

Obviously, we don't want to ``eat ourselves'', so the process is highly regulated by polypeptide hormones.
These hormones are: gastrin (stomach), secretin (duodenum), and cholecystokinin (duodenum).

\section{Enzymes Involved in Protein Digestion}

Before we move onto how proteins are digested, we need to know what kind of enzymes are involved.
The enzymes present are: pepsin (stomach), trypsin, chymotrypsin and carboxypeptidase (3 from pancreas), and aminopeptidase and dipeptidase (2 from small intestine).
All of these enzymes act in the small intestine, except for pepsin.
Note that the protease specificity is determined by the adjacent amino acid side chains of the peptide chain it is breaking.
For example, pepsin, trypsin and chymotrypsin cleaves the polypeptide after aromatic, positively charged and aromatic side chains, respectively.

\section{Protein Digestion}

Digestion of protein involves the hydrolyses of specific peptide bonds, performed by several different proteases.
All of these proteases are secreted as inactive zymogens to prevent them from digesting our own cells, and they are all activated by cleavage of peptides from their structure.

There are two stages of protein digestion:
\begin{enumerate}
\item Endopeptidases attack the peptide bonds within the peptide chain. Pepsin, trypsin and chymotrypsin are all endopeptidases.
\item Exopeptidases attack the peptide bonds at the end of the peptide chain (i.e. from C-term or N-term -- hence amino or carboxypeptidase).
\end{enumerate}

\subsection{Pepsinogen Activation}

Pepsinogen (due to its structure) have its catalytic active site inhibited.
H$^+$ (low pH) changes the conformation of the pepsinogen and allows it to auto-cleave the peptide bond and activate itself.
The activated pepsin can then activate other pepsinogens.
Note that this process is all triggered by the ingestion of food, and the hormones associated with it.

\subsection{Small Intestine Peptidase Activation}

All 3 peptidases (trypsin, chymotrypsin and carboxypeptidase) are secreted in the inactive zymogen forms -- trypsinogen, chymotrypsinogen and procarboxypeptidases.
First, trypsinogen is activated by the membrane bound enterokinase that adds the phosphate group to it.
This causes a conformational change in the trypsinogen structure which allows it to auto-cleave and auto-activate itself.
The activated trypsin activates the chymotrypsinogen and procarboxypeptidase.

As a result of the activation of all of these proteases (including pepsin), dietary proteins are digested into single amino acids, dipeptides and tripeptides.
Di and tripeptidases then digest the di and tripeptides into single amino acid.

\section{Protein Absorption}

\subsection{Amino Acid Transport}

There are 5 main amino acids transporters to transport the amino acids across the luminal membrane.
These are: Neutral, basic, acidic, other (taurine), and Glycine, proline and hydroxyproline.
All of these amino acids are transported in a similar way to the glucose transport -- co-transport with Na.

\subsection{Peptide Absorption}

Very little absorption occurs for peptides that are longer than 4 amino acids.
Di and tripeptides are absorbed in the small intestine via co-transport with H$^+$ ions (PepT1).
These di/tripeptides are then digested in the cell and exported out into the blood.

\subsection{Absorption of Intact Proteins}

This only occurs in few circumstances -- e.g. in newborn for immunoglobulin absorption.

\section{Diseases of Inappropriate Enzyme Activation}

This is associated with inappropriate enzyme action (zymogen activation).
This can lead to the breakdown of mucosal layer and proceed to peptic ulcers.

Pancreatitis is where the pH is altered and degrades/inflame the pancreas.
It can also release the enzymes into the blood.

\section{Diseases of Malabsorption}

Cystic fibrosis patients have abnormally thick mucous secretions due to mutation in chloride transporter.
The thick mucous can block the pancreatic ducts, and therefore no protease activity.

Coeliac disease have been mentioned in the previous lecture.

\section{Digestion of Nucleic Acid}

DNA and RNA are subject to acid hydrolysis in the stomach.
Intestinal nucleases also hydrolyse the phophodiester bonds between the nucleotides.
The individual nucleotides are absorbed via nucleotide transporters (Na co-transport).

It is most unlikely that the DNA/RNA gets incorporated into our body, but it is a possibility as there are evidence of DNA being intact in the gut.

### 2015-08-28 463assignment/463assignment_takei.tex
added assignment for 463

\documentclass[a4paper, 12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage{geometry}

\geometry{
    a4paper,
    left=3cm,
    right=3cm,
    top=3cm,
    bottom=3cm
}

\linespread{1.3}

\begin{document}

\begin{center}
\large{Olfactory Receptors, Copy Number Variations and Evolution}
\end{center}

\section*{\normalsize{Abstract}}

Human olfactory receptors (ORs) make up the largest family of genes in the genome, yet there has been little work done on characterising these receptors.
Recent advances in sequencing technology have allowed many researchers to investigate various aspects of the genome from single nucleotide polymorphisms (SNPs) to large chromosomal rearrangements.
Copy number variations (CNVs) is one such genomic rearrangement whereby some genes get duplicated, inserted, deleted and/or inverted.
Studies have shown that OR genes are highly copy number variable within and between species, but no study has associated this with phenotypic characteristics.

\section*{Introduction}

\section*{Research Proposal}

\section*{Conclusion}

\end{document}

### 2015-08-29 463assignment/463assignment_takei.tex
edited assignment

@@
 Human olfactory receptors (ORs) make up the largest family of genes in the genome, yet there has been little work done on characterising these receptors.
-Recent advances in sequencing technology have allowed many researchers to investigate various aspects of the genome from single nucleotide polymorphisms (SNPs) to large chromosomal rearrangements.
-Copy number variations (CNVs) is one such genomic rearrangement whereby some genes get duplicated, inserted, deleted and/or inverted.
 Studies have shown that OR genes are highly copy number variable within and between species, but no study has associated this with phenotypic characteristics.
+Sensory properties have been linked to satiety and eating behaviours, but has not been described in terms of the individual genotype.
+Next generation sequencing will be used to investigate the effect of OR copy number variation (CNV) on the individual phenotype, and to see whether this affects the sensory-specific satiety and/or food consumption.
 
@@
 
+\subsection*{Intro to olfactory receptors}
+
+Olfaction, or the sense of smell, is mediated by a large group of sensory receptors known as the olfactory receptors (ORs).
+ORs all belong to the G-protein coupled receptor (GPCR) superfamily like other sensory receptors, such as for vision and taste.
+ORs are known to be the largest gene family in the human genome and form clusters on many chromosomes, and consists of about 900 genes of which at least half of these genes are pseudogenes and have no functional activity (ref Glusman et al 2001; Niimura and Nei 2005).
+
+Due to the fact that there are so many OR genes in the human genome, and that these ORs work in a combinatorial manner, it has been difficult for the researchers to decipher the complete code of olfaction (ref Hasin-Brumshtein et al, 2009; Nei et al 2008).
+In addition to this, ORs are highly copy number variable within and between species, adding greater complexity to the olfactory process and remains to be a challenge for the researchers to make a connection between OR CNV with the phenotype (ref Nei et al, 2007).
+
+OR genes also have an important implications in the evolution of different species.
+There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes belong to their respective species (ref original paper or Nei and Niimura 2005).
+The study by Niimura and Nei (2005) showed that the diversity of the OR gene family in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
+Furthermore, they showed that amphibians have retained the diversity of the gene family as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial (Nei and niimura, 2005).
+As both pseudogenization and CNV are evidence of genomic drift (ref Nei et al, 2007).
+
+\subsection*{Intro to copy number variations}
+
+Recent advances in next generation sequencing technology have allowed many researchers to investigate various aspects of the genome, ranging from something as small as single nucleotide polymorphisms (SNPs) to large chromosomal rearrangements (ref NGS papers).
+Copy number variations (CNVs) is one such genomic rearrangement whereby some genes get duplicated, inserted, deleted and/or inverted.
+
+\subsection*{Intro to olfaction and sensory specific satiety}
+
+\subsection*{Significance of the work}
+
 \section*{Research Proposal}
+
+\subsection*{What am I going to do???}
 

### 2015-08-30 463assignment/463assignment_takei.tex
added to do points

@@
 
-\section*{\normalsize{Abstract}}
+\section*{Abstract}
 
@@
 
-\subsection*{Intro to olfactory receptors}
+\subsection*{Olfactory Receptors}
 
@@
 OR genes also have an important implications in the evolution of different species.
-There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes belong to their respective species (ref original paper or Nei and Niimura 2005).
-The study by Niimura and Nei (2005) showed that the diversity of the OR gene family in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
-Furthermore, they showed that amphibians have retained the diversity of the gene family as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial (Nei and niimura, 2005).
-As both pseudogenization and CNV are evidence of genomic drift (ref Nei et al, 2007).
+There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes are found mainly in their respective species (ref original paper or Nei and Niimura 2005).
+The study by Niimura and Nei (2005) showed that the diversity of the OR gene families in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
+Furthermore, they showed that amphibians have retained the diversity of the gene families as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial (Nei and niimura, 2005).
 
-\subsection*{Intro to copy number variations}
+Another study by Liman (2006) showed that pseudogenization and CNVs of the ORs coincided with the trichromatic vision in primates.
+Although this is an evidence of how trichromacy in primates became more important than the ORs for their survival, better understanding of the ligands of the ORs is crucial to support this theory.
+
+%As both pseudogenization and CNV are evidence of genomic drift, it is of great interest for researchers to understand the evolutionary process of ORs by studying the OR CNV (ref Nei et al, 2007).
+
+\subsection*{Next Generation Sequencing and Copy Number Variations}
 
@@
 
+studies have been relatively small -- possibly due to Sequencing limitations.
+talk about 1000 genome project -- provides greater sample size to work with
+limitations of sequencing CNVs -- highly variable, hard to sequence/annotate??
+
 \subsection*{Intro to olfaction and sensory specific satiety}
+
+relate this to obesity and eating disorder
 
@@
 
+evolution and healthier eating
+
 \section*{Research Proposal}
 
-\subsection*{What am I going to do???}
+gather sample to identify the OR genes that could be important -- get samples with known smelling disorder/condition
+sequence the samples and identify the CNV difference and target those genes for further work
+check 1000 genome database for any correlation and/or evidence of evolution
 

### 2015-08-31 463assignment/463assignment_takei.tex
finished assignment

--- 2015-08-28 463assignment/463assignment_takei.tex
@@
 \usepackage[utf8]{inputenc}
+\usepackage{natbib}
 \usepackage{geometry}
+\usepackage{fixltx2e}
 
@@
 \begin{center}
-\large{Olfactory Receptors, Copy Number Variations and Evolution}
+\Large{Olfactory Receptors, Copy Number Variations and Evolution}
 \end{center}
 
-\section*{\normalsize{Abstract}}
+\section*{Abstract}
 
 Human olfactory receptors (ORs) make up the largest family of genes in the genome, yet there has been little work done on characterising these receptors.
-Recent advances in sequencing technology have allowed many researchers to investigate various aspects of the genome from single nucleotide polymorphisms (SNPs) to large chromosomal rearrangements.
-Copy number variations (CNVs) is one such genomic rearrangement whereby some genes get duplicated, inserted, deleted and/or inverted.
 Studies have shown that OR genes are highly copy number variable within and between species, but no study has associated this with phenotypic characteristics.
+Sensory properties have been linked to satiety and eating behaviours, but has not been described in terms of the individual genotype.
+Next generation sequencing will be used to investigate the effect of OR copy number variation (CNV) on the individual phenotype, and to see whether this affects the sensory-specific satiety and/or food consumption.
 
-\section*{Introduction}
+\section*{Background}
+
+\subsection*{Olfactory Receptors}
+
+Olfaction, or the sense of smell, is mediated by a large group of sensory receptors known as the olfactory receptors (ORs).
+ORs all belong to the G-protein coupled receptor (GPCR) superfamily like other sensory receptors, such as for vision and taste.
+ORs are known to be the largest gene family in the human genome and form clusters on many chromosomes, and consists of about 900 genes of which at least half of these genes are pseudogenes and have no functional activity \citep{Glusman2001,Niimura2005}
+
+Due to the fact that there are so many OR genes in the human genome, and that these ORs work in a combinatorial manner, it has been difficult for the researchers to decipher the complete code of olfaction \citep{Hasin-Brumshtein2009, Nei2008}.
+In addition to this, ORs are highly copy number variable within and between species, adding greater complexity to the olfactory process and remains to be a challenge for the researchers to make a connection between OR CNV with the phenotype \citep{Nozawa2007}.
+
+OR genes also have an important implications in the evolution of different species.
+There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes are found mainly in their respective species \citep{Glusman2000}.
+The study by \citet{Niimura2005} showed that the diversity of the OR gene families in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
+Furthermore, they showed that amphibians have retained the diversity of the gene families as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial \citep{Niimura2005}.
+
+Another study by \citet{Liman2003} showed that pseudogenization and CNVs of the ORs coincided with the evolution of trichromatic vision in primates.
+Although this is an evidence of how trichromacy in primates became more important than the ORs for their survival, better understanding of the ligands of the ORs is crucial to support this theory.
+
+\subsection*{Sensory Specific Satiety and Eating Behaviour}
+
+It has been noted for a long time how eating behaviour changed depending on the sensory properties of food, such as the smell and the taste of the food \citep{Rolls1981, Rolls1982}.
+The effect of the sensory properties of food on the ability to change the pleasantness of eating that food was termed ``sensory specific satiety'' \citep{Rolls1981}.
+In humans, the flavour of the food is perceived with both the taste and the smell of the food, and this contributes to the sensory specific satiety \citep{Rolls2005}.
+This means that, although the taste of the food is important, the smell of the food is as important as the taste for the perception of flavour, yet the ligands and mechanism of olfactory process is poorly understood.
+
+Obesity is a significant health problem in many Western countries.
+Prevalence of obesity in United States has not changed significantly since 2003-2004 to 2011-2012, but more than a third of the population are obese and remains a significant health problem \citep{Ogden2014}.
+Better understanding of the olfactory process and their ligands may help understand the eating behaviour of people, and help these people to eat healthier and lose weight.
+
+\subsection*{Next Generation Sequencing and Copy Number Variations}
+
+Recent advances in next generation sequencing (NGS) technology have allowed many researchers to investigate various aspects of the genome, ranging from something as small as single nucleotide polymorphisms (SNPs) to large chromosomal rearrangements \citep{Metzker2010}.
+Copy number variation (CNV) is one such genomic rearrangement whereby some genes get duplicated, inserted, deleted and/or inverted.
+For an overview of copy number variations, please read the review article by \citet{Girirajan2011}.
+
+Traditionally, microarrays have been used to identify variations in the human genome before NGS was made available for lower cost.
+This involved hybridisation of fluorescently labelled DNA sequence from a sample of interest to the complimentary DNA sequence probe on the microarray, and the fluorescence was detected.
+Identification of the variant was inferred from the difference in the intensities of the fluorescence between the control and the test samples.
+However, this method became problematic when studying highly copy number variable regions, as even the highest density microarrays cannot accurately capture the copy number variations \citep{Girirajan2011}.
+
+Due to advances in NGS technology and its lower cost, it is now easier and more feasible to investigate complex genomic variations such as CNVs using these technologies.
+Since the whole genome of the samples would be sequenced and aligned, even the highly variable regions may be annotated and analysed using, for example, paired-end reads \citep{Girirajan2011}.
+Even though NGS technology provides greater accuracy and identification of many variants, alignment of CNVs can still be problematic, especially if the region is highly copy number variable.
+With that said though, single molecule sequencing technology may help define these complex genomic areas, as it would provide a definitive reference sequence for CNV alignment \citep{Rank2009}.
+
+In addition, there have been many studies conducted over the last few years that focussed on genomic variants using NGS.
+The 1000 Genomes Project was one such study, where the whole genome sequences of more than 1000 people around the world have been sequenced to understand the variation across different populations \citep{Durbin2010}.
+Furthermore, these sequence data have been made publicly available for open use, which would be of great use to this project when comparing between different population groups.
+
+\subsection*{Significance of the Work}
+
+Clearly, there is still a significant amount of work to be done on the ORs.
+Due to the high number of OR genes (and therefore correspondingly large number of ORs), it has been difficult to identify the exact ligand-receptor pairs.
+Furthermore, these ORs also work combinatorially and makes it even more difficult to identify these ligand-receptor pairs.
+
+Better understanding of the olfactory process and the ligands of ORs will allow us to further understand the mechanism of sensory specific satiety, and promote healthier diet especially in Western country where obesity is highly prevalent.
+Identification of these ligands and its connection to the CNV of the ORs will also allow us to clarify the significance of these receptors in evolution of mammalian species.
 
@@
 
+The main aim of this project is to identify ORs that are consistently copy number variable between normal samples and the samples with olfactory disorders, and then determine the physiological significance of the identified ORs in terms of sensory specific satiety.
+To do this, samples must be recruited for this study.
+Since the evaluation of olfaction can be difficult to assess, assistance by either a medical practitioner or a psychologist would be essential.
+The samples should be matched by gender, age and ethnicity so that the CNVs are due to the perception of smell of the samples.
+
+Once the samples are recruited, one of the control samples should be sequenced with single molecule real time (SMRT) sequencing, which would be used as the reference genome for the whole study.
+Since ORs are highly copy number variable, it is important to have a good reference genome from a single sample, rather than the reference genome made from multiple different samples.
+This is because we need a definitive CNV reference to identify whether the test samples have more or less copy numbers.
+SMRT sequencing was chosen to make up the reference genome, as other sequencing techniques may still have some problems in aligning the CNVs and/or getting an exact value for the CNVs.
+
+All of the other samples should be sequenced using either the paired-end sequencing method or SMRT sequencing, using the reference genome to align the sequences.
+Although SMRT sequencing would give greater accuracy about the copy number, paired-end sequencing may be sufficient for CNV identification and may be cost efficient with less technical expertise requirement.
+
+Identified ORs can then  be tested for multiple olfactory compounds, as well as their effects on sensory specific satiety.
+As soon as the function of the ORs are uncovered, the data from The 1000 Genomes Project can be used to investigate whether these ORs are significant in other population groups, and use this to provide some evidence for evolutionary process.
+
 \section*{Conclusion}
 
+Although there have been many studies that investigated the effect of sensory properties of food on the pleasantness of food, there has been no link to the genotypic characteristics.
+The use of NGS technology to sequence the whole genome of many samples will allow us to study the genomic variants in details.
+The results from this study can be used to identify the ORs that are essential for the sense of smell, and also provide insight into how the olfactory process works.
+Furthermore, these results can be used to investigate the significance of OR CNVs in terms of  both the sensory specific satiety and evolution of primates.
+
+\newpage
+
+%Referencing:
+\bibliographystyle{BiocRefStyle}
+\bibliography{../../../References/BibTeX/BIOC463}
+
 \end{document}

### 2015-08-31 463assignment/463assignment_takei.tex
edited assignment

@@
 \begin{center}
-\Large{Olfactory Receptors, Copy Number Variations and Evolution}
+    \Large{\textbf{Copy Number Variation of Olfactory Receptors and its Significance on Sensory Specific Satiety and Evolution}}
 \end{center}
@@
 ORs all belong to the G-protein coupled receptor (GPCR) superfamily like other sensory receptors, such as for vision and taste.
-ORs are known to be the largest gene family in the human genome and form clusters on many chromosomes, and consists of about 900 genes of which at least half of these genes are pseudogenes and have no functional activity \citep{Glusman2001,Niimura2005}
+ORs are known to be the largest gene family in the human genome and form clusters on many chromosomes, consisting of about 900 genes of which at least half of these genes are pseudogenes and have no functional activity \citep{Glusman2001,Niimura2005}
 
 Due to the fact that there are so many OR genes in the human genome, and that these ORs work in a combinatorial manner, it has been difficult for the researchers to decipher the complete code of olfaction \citep{Hasin-Brumshtein2009, Nei2008}.
-In addition to this, ORs are highly copy number variable within and between species, adding greater complexity to the olfactory process and remains to be a challenge for the researchers to make a connection between OR CNV with the phenotype \citep{Nozawa2007}.
+In addition to this, ORs are highly copy number variable within and between species, adding greater complexity to the olfactory process which remains to be a challenge for the researchers to make a connection between OR CNV with the phenotype \citep{Nozawa2007}.
 
 OR genes also have an important implications in the evolution of different species.
-There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes are found mainly in their respective species \citep{Glusman2000}.
-The study by \citet{Niimura2005} showed that the diversity of the OR gene families in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
+Two major classes of OR genes have been identified: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes were found mainly in their respective species \citep{Glusman2000}.
+The study by \citet{Niimura2005} showed that the diversity of the OR gene families in fishes was greater than that of the mammalian OR gene families, even though the number of OR genes in a single family were less in fishes.
 Furthermore, they showed that amphibians have retained the diversity of the gene families as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial \citep{Niimura2005}.
@@
 Another study by \citet{Liman2003} showed that pseudogenization and CNVs of the ORs coincided with the evolution of trichromatic vision in primates.
-Although this is an evidence of how trichromacy in primates became more important than the ORs for their survival, better understanding of the ligands of the ORs is crucial to support this theory.
+Although this was an evidence of how trichromacy in primates became more important than the ORs for their survival, better understanding of the ligands of the ORs is crucial to support this theory.
 
@@
 Obesity is a significant health problem in many Western countries.
-Prevalence of obesity in United States has not changed significantly since 2003-2004 to 2011-2012, but more than a third of the population are obese and remains a significant health problem \citep{Ogden2014}.
-Better understanding of the olfactory process and their ligands may help understand the eating behaviour of people, and help these people to eat healthier and lose weight.
+Although the prevalence of obesity in United States has not changed significantly since 2003-2004 to 2011-2012, more than a third of the population are obese and remains a significant health problem \citep{Ogden2014}.
+Better understanding of the olfactory process and their ligands may help understand the eating behaviour of obese people, and promote healthier eating and improve their diet.
 
@@
 Identification of the variant was inferred from the difference in the intensities of the fluorescence between the control and the test samples.
-However, this method became problematic when studying highly copy number variable regions, as even the highest density microarrays cannot accurately capture the copy number variations \citep{Girirajan2011}.
+Although microarray technology was good for identifying SNPs, this method became problematic when studying highly copy number variable regions, as even the highest density microarray was not able to accurately capture the copy number variations in the genome \citep{Girirajan2011}.
 
 Due to advances in NGS technology and its lower cost, it is now easier and more feasible to investigate complex genomic variations such as CNVs using these technologies.
-Since the whole genome of the samples would be sequenced and aligned, even the highly variable regions may be annotated and analysed using, for example, paired-end reads \citep{Girirajan2011}.
+Since the whole genome of the samples would be sequenced and aligned, even the highly variable regions of the genome may be annotated and analysed using NGS technology.
+For example, paired-end sequences can be used to align the genome to a reference and if the sequences aligned abnormally to a region of the genome, this can inform us about the structural rearrangements that may have occurred in that region \citep{Girirajan2011}.
 Even though NGS technology provides greater accuracy and identification of many variants, alignment of CNVs can still be problematic, especially if the region is highly copy number variable.
-With that said though, single molecule sequencing technology may help define these complex genomic areas, as it would provide a definitive reference sequence for CNV alignment \citep{Rank2009}.
+With that said though, single molecule sequencing technology may help define these complex genomic areas, as it would provide a definitive reference sequence for the alignment of CNVs \citep{Rank2009}.
 
 In addition, there have been many studies conducted over the last few years that focussed on genomic variants using NGS.
-The 1000 Genomes Project was one such study, where the whole genome sequences of more than 1000 people around the world have been sequenced to understand the variation across different populations \citep{Durbin2010}.
-Furthermore, these sequence data have been made publicly available for open use, which would be of great use to this project when comparing between different population groups.
+The 1000 Genomes Project was one such study, where the whole genome sequences of more than 1000 people around the world have been sequenced to understand the variation across different population groups \citep{Durbin2010}.
+Furthermore, these sequence data have been made publicly available for open use, which would be of great use to this project when comparing the extent of the OR CNVs between different population groups.
 
@@
 
-The main aim of this project is to identify ORs that are consistently copy number variable between normal samples and the samples with olfactory disorders, and then determine the physiological significance of the identified ORs in terms of sensory specific satiety.
+The main aim of this project is to identify ORs that are consistently copy number variable between normal samples and the samples with olfactory disorders, and to determine the physiological significance of the identified ORs in terms of sensory specific satiety.
 To do this, samples must be recruited for this study.
@@
 
-Once the samples are recruited, one of the control samples should be sequenced with single molecule real time (SMRT) sequencing, which would be used as the reference genome for the whole study.
+Once the samples are recruited, one of the control samples should be sequenced with single molecule real time (SMRT) sequencing and this sequence  would be used as the reference genome for the whole study.
 Since ORs are highly copy number variable, it is important to have a good reference genome from a single sample, rather than the reference genome made from multiple different samples.

### 2015-09-01 bioc192/summary_tute/summarytute.tex
added summary tute

\documentclass[a4paper, 12pt]{report}

\usepackage{chemmacros}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\graphicspath{ {../images/topic1/} {../images/topic2/} {../images/topic3/}}

\newcommand{\HRule}{\rule {\linewidth}{0.5mm}}

\newcommand{\mychapter}[2]{
    \setcounter{chapter}{#1}
    \setcounter{section}{0}
    \chapter*{#2}
    \addcontentsline{toc}{chapter}{#2}
}

\begin{document}

\begin{titlepage}

\begin{center}

\HRule
\vspace{0.4cm}
\Huge{\textbf{BIOC192 Summary Tutorial Notes 2015}}
\HRule

\vfill
\LARGE{\textbf{Compiled by Riku Takei}}

\end{center}

\end{titlepage}

%\tableofcontents
%\newpage

%\mychapter{0}{Introduction}

%\input{../intro.tex}

\mychapter{1}{Lecture 1}

\section{Levels of Protein Structure}

\subsection{Primary Structure}

The Primary structure of a protein is the amino acid sequence of that particular protein.
Note that the sequence of the amino acid (i.e. the primary structure) determines the 3D structure of the protein, since the primary sequence will have all the required amino acids in the right order for protein folding (via hydrophobic interaction).

\subsection{Secondary Structure}

Secondary structure is the local regular folds/structure that are stabilised by H-bonds between the backbone peptide groups of the polypeptide chain.
Examples of secondary structures include the $\alpha$-helix and $\beta$-sheets.
Supersecondary structures are related to the secondary structure, as it consists of combinations of secondary structures to make certain motifs (e.g. greek keys).

\subsection{Tertiary Structure}

Tertiary structure is the structure, or domain, formed by arranging secondary/supersecondary structures.
These usually have a function once it has the correct 3D structure, but it may require other tertiary structures to act as a single protein.

\subsection{Quaternary Structure}

Quaternary structure is where multiple tertiary structure come together to form a single functional protein.
Not all protein form a quaternary structure to function.

\section{Amino Acids}

Amino acids are the building blocks of a protein.
Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).

There are 20 amino acids in our cells that make up the proteins.
All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$).
Due to the variable side chains, all amino acids except glycine are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.

\section{Amino Acid Side Chains}

As mentioned earlier, all 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
This means that each of the 20 amino acids have different chemical properties, depending on the side chain group it has.
Cells take advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster (i.e. it gives the protein its functionality, as well as its unique structure).

\section{Amino Acid Subgroups}

Due to the unique chemical properties of the amino acid side chains, they are grouped into subgroups.
There are four main groups: Non-polar, uncharged polar, negatively charged, and positively charged amino acids.

\subsection{Non-polar}

These amino acids have non-polar (or hydrophobic) side chain, meaning that the side chains cannot be charged under physiological pH.
The amino acids in this group are: alanine, valine, leucine, isoleucine, glycine, cysteine, phenylalanine, tryptophan, methionine, and proline.

\subsection{Uncharged Polar}

These amino acids have side chains that can be charged, but uncharged under normal physiological pH.
As you can see, the side chain contains --OH groups and/or --NH$_2$ groups, whichare able to de/protonate under certain pH.
The amino acids in this group are: serine, threonine, tyrosine, asparagine, and glutamine.

\subsection{Negatively Charged}

These amino acids have side chains that are negatively charged (i.e. deprotonated) side chains at physiological pH.
The side chain contains a deprotonated --COOH group, giving it a negative charge to the side chain.
The pK$_a$ of these amino acid side chains ranges from 4$\sim$5.
The amino acids in this group are: aspartic acid (aspartate) and glutamic acid (glutamate).

\subsection{Positively Charged}

These amino acids have side chains that are positively charged (i.e. protonated) side chains at physiological pH.
The side chain contains a protonated --NH$_2$ group, giving it a positive charge to the side chain.
The pK$_a$ of these amino acid side chains ranges from 10$\sim$12, with an exception of histidine, which has a pK$_a$ of 7$\sim$7.5.
The amino acids in this group are: lysine, arginine, and histidine.

\includegraphics[width=\textwidth]{20aa}

\section{Charge of Amino Acids}

Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.
Depending on the pH of the solution, there will be more or less protonated molecules.

The amount of de/protonated amino acids affect the net charge of the protein, and therefore the isoelectric point (pI) of the protein.
The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).

When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to become protonated.
Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to become deprotonated.
So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).

\begin{center}
    \textbf
        {pH \textless{} pK$_a$, then the side chain is protonated \\
        pH \textgreater{} pK$_a$, then the side chain is deprotonated}
\end{center}

Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
However, some amino acid side chains may be charged under physiological condition and contribute to the net charge of the amino acid, and therefore the protein.
This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.

\begin{center}
\includegraphics[width=0.8\textwidth]{aapka}
\end{center}

\section{Amino Acid Modifications}

Some amino acids are modified after translation -- this is called the post-translational modification (PTM).
Types of PTM includes:
\begin{itemize}
    \item Disulfide bond formation between cysteines
    \item Phosphorylation
    \item Glycosylation
    \item Methylation
    \item Adenylation
    \item Iodination
    \item Metal binding
\end{itemize}

\section{Proteins, Peptides, and Peptide Bonds}

Amino acids are covalently linked together by peptide bonds -- the bond formed between the amino group of one amino acid with a carboxyl group of another amino acid.
A chain of amino acids linked together by peptide bonds are called peptides, and these are usually short.
Peptides that have a longer chain of amino acids with defined biological function are called proteins.
(Note: peptides are proteins as well -- it's just a little smaller)

\subsection{Amide (peptide) Bonds}

Amide (peptide) bond has a partial double bond characteristic due to the electron delocalisation/sharing between the oxygen atom and the nitrogen atom.
This means that the amide bond is rigid, and also have a dipole nature (oxygen is slightly negative, nitrogen is slightly positive).

The amide bonds are predominantly in \textit{trans}-configuration in the protein, although some can be in \textit{cis}-configuration, especially when the bond is preceded by proline.
This is due to ring structure that the proline side chain forms with its amino group (causes steric clashes).

By convention, the free amino group on the end of the peptide chain is called the N-terminus of the protein, and the free carboxyl group is known as the C-terminus.
This gives a `directionality' to the peptide sequence, from the N-terminus to the C-terminus, and we draw primary sequence and count the amino acids according to this directionality.

\begin{center}
\includegraphics[width=\textwidth]{amidebond}
\end{center}

\mychapter{2}{Lecture 2}

\section{Bond Rotation}

Protein structure is created by the main chain and side chain bonds, and requires the main/side chains to be in a certain conformation.
In order to do this, the amino acids must rotate and orientate correctly for the right interaction.

Single bonds between the carbon atoms can be rotated freely, but not the double bonds.
Since the amide bonds have a partial double bond characteristic, it is rigid and forms a plane that does not allow rotation.
However, rotation is allowed at the $\alpha$-carbon (C$_{\alpha}$) atom.
Obviously, even though rotations are allowed at the C$_{\alpha}$ atom, not all angles are allowed due to steric hindrance.

There are four types of bond angle rotations: $\omega, \psi, \phi$, and $\chi$.
These angles have values that range from 0 to $\pm$180$^{\circ}$, and together, it gives the 3D structural information of a protein.

\subsection{Phi ($\phi$) Angle}

Phi angle is defined as the bond angle rotation of the N--C$_{\alpha}$ bond.
This bond rotation leads to O--O atom collision.

\subsection{Psi ($\Psi$) Angle}

Psi angle is defined as the bond angle rotation of the C--C$_{\alpha}$ bond.
This bond rotation leads to the NH--NH collision.

\vspace{1.0cm}

\noindent
\textbf{Remembering phi/psi angles are quite hard, but just remember that, relative to the C$_{\alpha}$ atom, phi angle is the rotation of the amide plane on the amino group side, whereas the psi angle is the rotation of the amide plane on the other side (i.e. the carboxyl side).}

\subsection{Omega ($\omega$) Angle}

mega angle is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).

\subsection{Chi ($\chi$) Angle}

Chi angle is the bond angle rotation of the side chain bond (C$_{\alpha}$--R bond).

\section{Ramachandron Plot}

This plot shows all the possible phi/psi angles (i.e. the two amide planes adjacent to the C$_{\alpha}$) that are allowed in a protein with no steric interference.
Any phi/psi angle pairs that cause steric hindrance are not plotted, so we get some `regions' of allowed angle pairs for any given protein.
The side chains of the amino acids are (usually) staggered to minimise steric hindrance.

\vspace{0.5cm}

\includegraphics[width=\textwidth]{ramachandron}

\section{Secondary Structure of a Protein}

There are three main secondary structures in a protein: $\alpha$-helix, $\beta$-sheets, and turns.

\subsection{$\alpha$-Helices}

Specifically speaking, $\alpha$-helix is a type of helix that is present in a protein -- there are other types of helices (e.g. polyproline helix), but $\alpha$-helix is the most common type.

In $\alpha$-helix, the carbonyl group of the n$^{th}$ residue forms a H-bond to the amino group of the n+4$^{th}$ residue (e.g. the carbonyl group of the 1$^{st}$ residue H-bonds to the amino group of the 5$^{th}$ residue)
There is about 3.6 residues per turn, which is about 5.4\AA per turn.
The distance (or height) between 2 residues is about 1.5\AA.
The side chains point outwards from the helix (this provides certain chemical properties on the sides of the helix, e.g. hydrophobic side). 
$\alpha$-helices are dipole in nature, due to the amide bond polarity, all of which points at a single direction.

There are some residues that disrupt the helical structure, and these are glycine and proline.
Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
This flexibility can cause disruption in the helical structure.
Proline has a unique side chain that connects back to its amino group, forming an imino group, meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.

\begin{center}
\includegraphics[width=0.7\textwidth]{alphahelix}
\end{center}

\subsection{$\beta$-Sheets}

$\beta$-sheets, or $\beta$-pleated sheets, are formed from H-bond formation between adjacent polypeptide chain (or strands), therefore $\beta$-sheets are formed from 2$\sim$10 strands (but itcan have more strands).
Each strand has about 15 amino acids, and the $\beta$-sheets can be either parallel (both strands go from N to C terminal) or anti-parallel (strands go in an opposite direction).
$\beta$-sheets are extended, but pleated (i.e. it's not completely flat), and the side chains point above and below.

\subsection{Turns}

Turns are sharp hairpin turns in a protein, and it has a high glycine and proline content.
About 30\% of the residues in a protein is involved in turns.
It is common for the residues to form H-bonds across the turn.
There are many types of turns (\textgreater16), but type I and II are common.

\subsection{Relationship between Ramachandron Plots and Secondary Structures}

Secondary structures have certain phi/psi angle pairs, and when you overlay these to the plot, it falls into the allowed regions of the plot at a distinct region.
This is not surprising, because if the secondary structures fall in the forbidden region, how did the structure form in the first place?
It can't -- the structure cannot form due to steric hindrance.

This essentially shows that secondary structures have unique phi/psi angle pairs.

\mychapter{3}{Lecture 3}

\section{Protein Folding Pathway}

Protein folding is directed predominantly by the hydrophobic forces from the non-polar residues.
This is NOT a random process.
Since the proteins are submerged in water, the non-polar residues try to stay away from water, and therefore aggregate together to form secondary structures.
This initial folding process is referred to nucleation.
The hydrophobic regions of the secondary structure comes together to form a larger structure, and stabilised by small non-covalent interactions.

\vspace{0.5cm}

\noindent
The process of protein folding is summarised below:
\begin{enumerate}
\item Short, small secondary structures are formed, driven by hydrophobic forces (nucleation)
\item Nuclei come together to form larger domains
\item Larger domains come together
\item Small conformational changes/adjustments occur to give compact native structure
\end{enumerate}

\section{Stabilisation of Protein Folding}

Proteins are stabilised by non-covalent interactions, which are weak individually, but strong when there are a lot of them.
Covalent bonds such as disulfide bonds can stabilise the protein as well.

%\includegraphics[width=\textwidth]{stabilisinginteraction}

\section{Chaperones and Chaperonin}

Some proteins require additional accessory proteins to assist them in folding the proteins properly, or to fold it faster.
If the proteins are not folded properly, it can lead to serious diseases like Alzheimer's and prion.

\mychapter{4}{Lecture 4}

\section{Myoglobin}

Myoglobin is the monomeric form of haemoglobin, and it resides in the tissue/muscles for O$_2$ storage.
Myoglobin consists of a globin protein that consists of 8 $\alpha$-helices (numbered for clarity), with a haem group in the middle.
Myoglobin is able to reversibly bind O$_2$, and have high affinity for O$_2$ as well.

\section{Haem (Heme)}

Haem consists of four pyrrole rings that are linked together via methane bridges, with the ferrous (Fe$^{2+}$) attached at the center of the molecule.
This Fe$^{2+}$ is slightly below the plane of the porphyrin ring when O$_2$ is not bound, and is shifted into the plane when it binds (causes conformational change).

The heme is attached to the myoglobin via the His F8 (histidine on the helix F at position 8), and His E7 is on the opposite side to influence the O$_2$ binding affinity.
The His E7 prevents the strong binding of O$_2$ to the heme (right angle) by kinking the angle at which the O$_2$ binds to the heme.
This allows weaker and therefore reversible binding of the O$_2$ molecule to the myoglobin.

Note that cyanide binds to the Fe$^{3+}$ form, not the Fe$^{2+}$ form.
Also note that cyanide's main inhibitory effect is at the cytochrome Fe$^{3+}$, and affects the electron transport system.

\begin{center}
\includegraphics[width=0.4\textwidth]{haem}
\end{center}

\section{Haemoglobin}

Haemoglobin is the tetrameric (4 subunit) form of myoglobin that consists of two $\alpha$ and two $\beta$ subunits, and it is a blood O$_2$ transporter.
Since haemoglobin is a tetramer, it has 4 haem group in total (one on each subunit), and therefore able to carry 4 O$_2$ molecules.
Slight conformational change occurs when the O$_2$ binds to the haemoglobin molecule, which will become important in explaining the cooperative effect of the subunits.

\section{O$_2$ Binding Property}

\subsection{Myoglobin}

Myoglobin resides in muscles and tissues, and have high affinity for O$_2$.
This is shown in the O$_2$ saturation curve.
At low partial pressure of O$_2$ (pO$_2$), myoglobin is mostly saturated with O$_2$ (i.e. O$_2$ is bound to it).
This means that myoglobin only releases the O$_2$ molecule when the pO$_2$ is very low.

\subsection{Haemoglobin}

In contrast, haemoglobin's O$_2$ binding affinity is not that high compared with myoglobin.
As you can see from the O$_2$ saturation curve, it requires quite a high pO$_2$ for haemoglobin to be saturated with O$_2$.
This makes sense as haemoglobin has to bind to O$_2$ at the lungs where pO$_2$ is relatively high, but also has to release the O$_2$ molecules at the tissue/muscle, where the pO$_2$ is low.
This allows haemoglobin to be an effective O$_2$ carrier.

One thing to note here from the curve is that the shape of the curve for haemoglobin is sigmoidal (S-shape).
This indicates that haemoglobin is allosteric (or cooperative), where the subunits assist one another to facilitate the binding/release of O$_2$ molecules.
Of course, there are other factors that contribute to the allosteric binding/release of O$_2$ molecules.

\begin{center}
\includegraphics[width=0.4\textwidth]{saturationcurve}
\end{center}

\section{Sequential and Concerted Model of \-Allosterism}

When O$_2$ binds to a subunit of haemoglobin, it causes a slight conformational change to the subunit.
This conformational change assists the binding of O$_2$ molecules to the other subunits by cooperativity/allosterism.
For clarity, the tense (T) state is the state of the subunit when O$_2$ is not bound (low O$_2$ affinity), and the relaxed (R) state is the state of the subunit when O$_2$ is bound (high O$_2$ affinity).

There are two models, sequential and concerted models, of allosterism.
It is still unknown which model is correct, as there are many observations supporting both models.

\subsection{Sequential Model}

Sequential model is when one O$_2$ molecule binding to a single subunit causes the conformational change in the subunit it binds, and also the subunits that are adjacent to it as well.
This conformational change makes the other subunits easier to bind to O$_2$ molecules, which will also change conformation of itself and the others.

With respect to the R/T states, binding of O$_2$ molecule causes a single subunit to be in R state, and also changes the conformation of the adjacent subunits so that it is easier to bind to O$_2$ molecules, therefore easier for the subunits to turn into the R state.

\textbf{In sequential model, binding of one substrate changes the conformation of other subunits, which makes it easier for the other subunits to bind the substrate}

\subsection{Concerted Model}

In concerted model, all of the subunits are either in the T state, or the R state, and the states are at equilibrium (i.e. no `in between' state).
The O$_2$ molecules can bind to either state at any time.
When there are more substrate (i.e. O$_2$) is bound to haemoglobin, the greater the chance for the subunits to be in the R state, which means that the equilibrium is favoured towards the R state when there are more O$_2$ molecules bound to haemoglobin.

\textbf{ In concerted model, the T and R states are at equilibrium and substrates can bind at any time to any conformational state. The binding of substrate causes the equilibrium to shift to the R state and the substrates are more likely to bind to the protein.}

\mychapter{5}{Lecture 5}

\section{Allosteric Effectors of Haemoglobin}

As discussed earlier in Lecture 4, binding of O$_2$ molecule to the haem group causes a conformational change, and thus promotes the R or T state.
This conformational change is caused by the shift in the Fe$^{2+}$ atom of the haem, which is attached to the His F8 of the haemoglobin subunit, which then cause a slight shift that propagates through the protein.

However, there are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
Since these molecules have effects on haemoglobin, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.

\vspace{0.3cm}

\begin{center}
\textbf{Allosteric effectors are molecules that bind to the allosteric protein at a place other than/different to the normal substrate binds to.}
\end{center}

\section{2,3-Bisphosphoglycerate (BPG)}

2,3-Bisphosphoglycerate (BPG) is a small, negatively charged (-5 charge) molecule that binds to the deoxy-haemoglobin at the center of the tetramer.
Binding of BPG causes the haemoglobin to be in T state, i.e. in the dexygenated state, and decreases the haemoglobin's binding affinity for O$_2$, meaning that the haemoglobin with BPG bound to it will require higher pO$_2$ to oxygenate it.
Binding of BPG at the muscle/tissue site ensures that the O$_2$ molecules are released at the muscles/tissues.

O$_2$ binding to the haemoglobin at the lungs causes the release of BPG molecule, allowing the haemoglobin to carry O$_2$ to the muscles/tissues.

\section{H$^+$ and CO$_2$}

Both H$^+$ and CO$_2$ binds to haemoglobin at a different site as the substrate (O$_2$) and affects the oxygen affinity of haemoglobin.
Both molecules favour the T state of haemoglobin, and therefore favours the deoxygenated state (i.e. decreased O$_2$ affinity).
Note that both of these molecules are related to the pH of the environment, and the blood buffer system is also involved in controlling the concentration of these molecules.

\subsection{The Bohr Effect}

Technically speaking, the Bohr effect is the effect of pH (i.e. [H$^+$]) and [CO$_2$] on the oxygen affinity of haemoglobin.
As pH decreases (therefore [H$^+$] increases), the net charge of haemoglobin changes due to different amino acids getting protonated, which leads to slight conformational changes in the haemoglobin structure, and therefore decrease in haemoglobin's oxygen affinity.
Similarly, as CO$_2$ concentration increases, CO$_2$ is more likely to bind to the haemoglobin and decrease the oxygen affinity of haemoglobin.

\subsection{The Blood Bicarbonate Buffer System}

In the blood, there is an equilibrium between bicarbonate ([HCO$_3^-$]) ion and CO$_2$.

\begin{center}
\ch{H$_2$O + CO$_2$ <=> HCO$_3^-$ + H$^+$}
\end{center}

As you can see, the [H$^+$] concentration and the [CO$_2$] concentration is related to one another due to this buffer system.
This means that at the tissue/muscle where the relative concentration of CO$_2$ is high, it is going to be making the bicarbonate ion, together with H$^+$ ion, and therefore the pH is decreased.
However, at the lungs where the concentration of CO$_2$ is low, there will be less H$^+$ ion present, and therefore have higher pH.

Together with the Bohr effect, this means that the pH is high at the lungs and have low CO$_2$ concentration, and therefore haemoglobin will have high oxygen affinity, and thus get saturated with O$_2$ (oxygenated).
In contrast, the pH will be lower at the tissues/muscles and have high CO$_2$ conference, and therefore haemoglobin will have low oxygen affinity, and thus releases O$_2$ at the tissue/muscle (deoxygenated).

Also note that BPG is released at the lungs when O$_2$ binds to the haemoglobin, and binds at the tissue/muscle when O$_2$ is released and CO$_2$ and H$^+$ binds to it, contributing to the binding and release of O$_2$.(insert the saturation curve of +/- allosteric effectors)

\includegraphics[width=\textwidth]{bicarbonate}

\mychapter{6}{Lecture 6}

\section{Normal Variants}

These variants are normal in the sense that they are naturally occurring, and not caused by abnormal mutations.
These variants are different to the normal haemoglobin in that it uses different type of subunits to make the haemoglobin up.

\subsection{Foetal Haemoglobin (HbF)}

Foetal haemoglobin (HbF) is only present in the foetus blood.
HbF is made up of two $\alpha$ subunits and two $\gamma$ subunits, compared to the adult haemoglobin (HbA) which is made up of two $\alpha$ and two $\beta$ subunits.
Since the foetus has to obtain oxygen from the mother's blood via the placenta, HbF has a higher affinity for oxygen than the HbA.
This is because HbF binds less strongly to BPG due to the Ser143 in the $\gamma$ subunit (His143 in $\beta$ subunit), thereby increasing the oxygen affinity.

The graph below just shows how different subunits are made at different developmental stage.(insert the graph)
As you can see, there are subunits other than $\alpha$, $\beta$ and $\gamma$ subunits.

\section{Abnormal Variants}

These variants are caused by genetic mutations that alter the properties of haemoglobin, and can lead to anaemia or other pathological symptoms.

\subsection{Sickle Cell Haemoglobin (HbS)}

This variant is caused by a single point mutation on the $\beta$ subunit of haemoglobin.
This point mutation causes an amino acid substitution from glutamic acid to valine (E6V mutation), which is on the surface of the protein.

The consequence of this mutation is the formation of deoxy-Hb aggregate/crystal in the red blood cell.
Since valine is a non-polar amino acid, it can stick/bind to other haemoglobin at specific site (in a hydrophobic pocket).

\subsection{Met-Haemoglobin (HbM)}

Methaemoglobin (HbM) is caused by mutations that changes the oxidation state of the iron atom in the haem from Fe$^{2+}$ to Fe$^{3+}$.
This means that the oxygen cannot bind (or have very low affinity for oxygen) to the haem iron, as it is in a different oxidation state.

\subsection{Haemoglobin C (HbC)}

Haemoglobin C (HbC) is common in West Africa, and is caused by a point mutation that causes a substitution mutation.
The substitution is at the same place as HbS: glutamic acid to lysine at position 6 (E6K).
As a consequence of this substitution, the charge is swapped from negatively charged glutamic acid to positively charged lysine.

Malaria is caused by the infection of red blood cells by \textit{Plasmodium \\falciparum}.
Infected red blood cells will contain malaria parasite proteins that produce knobs on the surface of the red blood cells and causes it to adhere to the capillaries, causing tissue damage due to reduced oxygen supply.
However, individuals with homo or heterozygous HbC have less adherence, and thus have some resistance to malaria infection.

\mychapter{7}{Lecture 7}

\section{Post-Translational Modification (PTM)}

Most proteins, if not all proteins, in eukaryotic cells are post-translationally modified.
Post-translational modification increases the number of distinct forms of proteins without increasing the number of genes in the genome.
This means that the level of complexity provided by the proteins are also increased as well.

The diversity of the proteins can be expanded by mRNA splicing and PTM.
PTM can be categorised into two categories: the covalent addition of chemical group to amino acid side chain, or covalent hydrolysis of the peptide bond in a protein.

\section{Phoshphorylation}

Phosphorylation can occur on the hydroxyl (OH) group of the amino acid side chains.
Amino acids that have OH groups are Ser, Thr, and Tyr.
Phosphorylation of the side chains are done by an enzyme called kinase, which adds the large, negatively charged phosphate group to the side chain, and uses ATP during this process.
To remove the phosphate group, an enzyme called phosphatase is used.

The consequence of adding a phosphate group to these amino acid side chains is that it's going to alter the protein structure, due to the electrostatic attraction between the more negatively charged phosphate group with other postitively charged amino acid side chains.
Since the attraction is going to shift the amino acids slightly, it will cause a shift in the whole protein, and therefore alters the conformation of the protein and affects the biological properties of it.

Quite often, phosphorylation state is used to `switch' the conformational state of the protein to enable a biological process in a cell (e.g. facilitating interaction with other proteins).

\subsection{Insulin Receptor}

Binding of insulin to the insulin receptor causes a conformational change in the receptor, and this exposes the tyrosine in the cytosol side of the membrane.
The exposed tyrosine is then targeted by kinases, and get phosphorylated by it.
Phosphorylated tyrosine can then cause responses (e.g. signal transduction, GLUT4 transporter expression).

\subsection{Na$^+$/K$^+$ Pump}

Phosphorylation state affects which molecule (Na$^+$ or K$^+$) to bind to the pump, and therefore be transported.
This is also energy dependent process, as the pump is constantly being phosphorylated and dephosphorylated.

\section{Gamma-carboxy Glutamic Acid (Gla)}

Gamma-carboxy glutamic acid occurs when an additional carboxyl group is added to the glutamic acid.
This PTM is done in some blood coagulation proteins, and it requires vitamin K as a co-factor.
Ten to twelve Gla enables the formation of bidentate Ca$^{2+}$ binding site and allows blood coagulation proteins to interact with platelets, as a part of blood clot formation process.

When you have vitamin K deficiency, you can't gamma-carboxylate the glutamic acids, since the reaction requires vitamin K as a co-factor.
As a consequence, this leads to bleeding disorder.

\section{Hydroxylation}

Hydroxylation mainly occurs on the 4' (or 3') position of the proline, but it can also occur at the 5' position of lysine.
Hydroxylation provides greater possibilities for non-covalent interactions, so it allows for more H-bonding to occur.
As a result, proteins are able to form higher order structure.

Vitamin C is important in hydroxylation, as it is required for the reaction to occur.
Vitamin C keeps the Fe$^{2+}$ in this oxidation form, which is essential for the hydroxylation process.
Since collagen structure requires hydroxylation to keep its higher order structure, low vitamin C causes disruption in this structure.
This disease is known as scurvy.

\mychapter{8}{Lecture 8}

\section{Introduction to Enzymes and Enzyme Engineering}

Enzymes (proteins) are able to accelerate and carry out chemical reactions that are most likely impossible to do under physiological/biological conditions.
Without these enzymes, we will not be able to synthesise chemicals required to survive.
It is also interesting that the enzymes have all evolutionarily developed itself to make difficult chemical reactions to faster, but within similar timescale.

Can we utilise the properties/characteristics of enzymes to make it do what we want it to do?
This is the key idea behind the researches being done in the area of enzyme engineering.
We can use enzyme engineering to make enzymes that are able to react with different substrates, or make it more efficient at doing the reaction.

\mychapter{9}{Lecture 9}

\section{Introduction to Biotechnology}

For therapeutic and clinical use of proteins, the proteins must be made and extracted from a source.
Biotechnology is the field of research to produce proteins in systems (sort of).

\subsection{The `Old' Way of Obtaining Proteins}

Therapeutic proteins were extracted from relevant sources (e.g. tissues, serum) and was used to treat the patients.
The problems with this approach was the risk of identified and unidentified pathogen infections from the source (e.g. HIV).

\subsection{Recombinant Protein System}

This is the more recent method to produce proteins for therapeutic use.
Plasmids that contain a required protein is expressed in a different organism/system, and the purified and collected.

Recombinant insulin is produced this way, by expressing two separate parts of insulin in \textit{E. coli}.
These parts are extracted and combined to form insulin.

The advantage of this is that it is cheap, has high yield, and it is pathogen free.
However, in a prokaryotic system, the cells are unable to post-trasnlationally modify the protein, which may be essential for the function of some proteins.
Another problem is that the proteins may not fold properly or only partially folded.

\subsection{Eukaryotic System}

This system solves the problem of PTM and protein misfolding, as the proteins will be expressed in a eukaryotic system.

An example of a protein that uses a eukaryotic system is erythropoietin (EPO).
EPO regulated haematopoiesis, and its expression is partially regulated by the blood oxygen level -- EPO acts so that it produce more red blood cells and haemoglobin.
EPO is used therapeutically as well as a performance enhancing drug.
EPO is produced in immortalised Chinese hamster ovary cells.

Different EPOs have different glycosylation profile, which allows us to detect athletes who uses EPO.

\subsection{Recombinant Antibodies}

Recombinant antibodies are designed to control the effects of some proteins in your body (e.g. some cell receptors).
Mouse AB can be made, but it is recognised as foreign material in humans and get rejected.
This lead to the formation of chimeric antibodies, so there will be no immune response.

\subsection{Pharming}

The use of live animals to express/make recombinant proteins, for example, in milk.

\mychapter{10}{Lecture 10}

\section{Introduction to Enzymes}

In most cases, enzymes are proteins (in some cases, it's RNA enzymes) that acts as biological catalysts.
Enzymes increases the rate of reaction, but does NOT change the equilibrium, nor the thermodynamics of a reaction.
This means that, even though enzymes make the reactions to occur faster, it cannot shift the chemical reaction to favour one product over another.
Note that enzymes are named ``-ases''.

For a chemical reaction to occur, the substrates must be in close proximity of one another, and also have the right orientation.
Enzymes catalyse reactions by doing just that: bind the substrate in the right orientation in close proximity so a chemical reaction can occur at a faster rate.

\section{Active Site of an Enzyme}

Active site is the 3-dimensional site of an enzyme where the reaction occurs.
The active site has amino acid side chains projecting into it to help substrate binding.
The shape of the active site determines the specificity of the enzyme, since enzymes are (usually) very specific.
This means that enzymes only bind to certain substrate, and no other substrates.

\subsection{Geometric Specificity and Stereospecificity}

\begin{center}
\includegraphics[width=0.7\textwidth]{geo_stereo}
\end{center}
Geometric specificity is where the substrate must be made of same component/structure for it to be recognised by the enzyme.
Stereospecificity is where the substrate has the required components,but don't have the right orientation, and therefore not recognised by the enzymes.

\section{Enzyme-Substrate Interaction}

\begin{center}
\ch{ E + S <=> ES <=> [{T}] EP <=> E + P }
\end{center}

In an enzyme catalysed reaction, substrate binds to the active site and forms an enzyme-substrate complex (ES), which is stabilised by non-covalent interactions.
Note that this interaction is specific.
As the reaction proceeds, transition state (T), which is neither a substrate nor the product, is formed, then proceeds further into enzyme-product complex (EP).
This complex can then be dissociated to form the original enzyme ready for another reaction, and release the product.

There are two models that describes how the substrate binds to the enzyme: the lock-and-key model, or the induced-fit model.

\begin{center}
\includegraphics[width=0.6\textwidth]{enzymemodels}
\end{center}

\subsection{Lock-and-Key Model}

The concept of the lock and key model is fairly simple -- the substrate must have the shape/structure that perfectly matches the enzyme active site.
If the substrate doesn't then, the enzyme cannot react with it and no reaction occurs.
However, this is not really the case in nature as enzymes changes conformation when the substrate binds to it.

\subsection{Induced-fit Model}

This is how an enzyme actually works.
The substrate and the enzyme active site does not match perfectly as it did for the lock and key model.
When the substrate and the enzyme binds, it causes a conformational change in the enzyme, which allows the enzyme to properly bind to the substrate.

Note that one way to increase the rate of reaction is by stabilising the transition state of the substrate.
Enzymes can do this by preferentially binding to the substrate in a transition state conformation.
This is quite often the case, as enzymes usually stabilises the transition state of the substrate during the catalysis.

\section{Enzymes, Activation Energy, and \\Catalytic Mechanisms}

Enzymes can decrease the activation energy of a chemical reaction by various catalytic mechanisms.
There are 6 ways to do this:

\begin{enumerate}
\item Acid-base catalysis
\item Covalent catalysis
\item Metal ion catalysis
\item Electrostatic catalysis
\item Proximity and orientation effects
\item Preferential binding of the transition state complex
\end{enumerate}

In this lecture, only the first three was discussed.

\subsection{Acid-base Catalysis}

This kind of catalysis involves the transfer of H$^+$ ions, for example by glutamic and aspartic acids (COOH group), or by lysine (NH$_3^+$ group).
Note that these side chains require certain pH for it to be in the right ionisation state, and therefore to be able to transfer H$^+$ ions.
In this sense, histidine is very useful as it has a pK$_a$ of about 6.5, which is close to the physiological pH.
This means that with a slight change in pH, it can be either a H$^+$ donor or acceptor, and therefore able to take part in the acid-case catalysis reaction.

\subsection{Covalent Catalysis}

In a covalent catalysis, the enzyme forms a highly reactive, short-lived intermediate which is covalently attached to the enzyme.

\begin{center}
\ch{A-B + E -> A-E + B ->  A + B + E}
\end{center}

The enzyme requires an amino acid with a side chain that can act as a nucleophile (e.g.ser, cys)

\subsection{Metal Ion Catalysis}

About a third of known enzymes require metal ions for their catalytic activity.
Metals can provide substrate orientation, binding energy, and/or sites for redox reactions.
Metals act as a cofactor in these reactions.

\section{Co-factors and Co-enzymes/Co-substrates}

Some enzymes require other factors to catalyse chemical reactions.
These factors can be metal ions or small organic molecules (co-enzymes).
Co-enzymes, or co-substrates, are molecules that help with the reaction, and is changed after the reaction occurs.
Co-substrates are usually derived from vitamin derivatives.
(An example used in the lecture was NAD)

\mychapter{11}{Lecture 11}

\section{Measuring Enzyme Catalysis}

\begin{center}
\includegraphics[width=0.5\textwidth]{progresscurve}
\end{center}

Enzyme catalysis of a reaction can be measured from the progress curve of a reaction.
Progress curve is the graph of how much product is made (or how much substrate is used up) over time.
It is important to measure the initial reaction velocity of the reaction from this progress curve in order to calculate some other important measures of enzyme kinetics.

This initial reaction velocity (V$_i$) is the slope of the tangential line of the initial part of the progress curve.
Only the initial part is used, because as you progress through the reaction, some factors start to limit the reaction and becomes unreliable.

\section{Effect of Enzyme Concentration on the Rate of Reaction}

\begin{center}
\includegraphics[width=0.5\textwidth]{vvse}
\end{center}

Consider the situation where there is an excess substrate (i.e. unlimited source of substrate).
As the concentration of the enzyme increases, the rate of reaction increases.
\textbf{This means that the velocity of the reaction (V) is proportional to the enzyme concentration, when there is excess substrate.}

This is expected, since there are more enzymes to catalyse the reaction -- as more enzymes are introduced to the system, more enzymes can react with the substrate, and therefore produce more products.

\section{Effect of Substrate Concentration on the Rate of Reaction}

\begin{center}
\includegraphics[width=0.4\textwidth]{vvss}
\end{center}

Consider the situation where the concentration of enzyme is fixed.
As we increase the concentration of substrate, the rate of reaction will initially increase linearly, but as more and more substrates are introduced, the rate of reaction reaches a `maximum'.
This is because all of the active site of the enzymes become occupied as the substrate concentration increases -- it reaches the `maximum capacity' of the enzyme, and the enzymes cannot increase the rate of reaction no matter how hard it tries.

\begin{center}
\includegraphics[width=0.4\textwidth]{zerofirstkinetics}
\end{center}

In kinetics term, the enzyme shows first-order kinetics initially (where the rate depends on the substrate concentration), but then shows zero-order kinetics (where the rate does not depend on/changes with substrate concentration).
When the velocity of the reaction is plotted with the substrate concentration, this is quite apparent from the hyperbolic curve of the graph.

\subsection{V$_{max}$ and K$_M$}

From this V vs S curve, we can identify two kinetic parameters: V$_{max}$ and K$_M$.
V$_{max}$ is defined as the maximum velocity of the reaction when an enzyme is saturated with substrate.
K$_M$ is defined as the \textbf{substrate concentration} at which the velocity of the reaction is half the maximum velocity ($V = \frac{V_{max}}{2}$).
K$_M$ is also called the Michaelis constant.

Note: do NOT get this mixed up with ``K$_M$ = $\frac{V_{max}}{2}$'' -- remember that K$_M$ is the \textbf{substrate concentration} when V = $\frac{V_{max}}{2}$.

\mychapter{12}{Lecture 12}

\section{The Michaelis-Menten Equation}

Using V$_{max}$ and K$_M$, we can define the Michaelis-Menten equation:

\begin{center}
\large{V = $\frac{V_{max} [S]}{K_M + [S]}$}
\end{center}

Few assumptions must be made in order to derive this equation:
\begin{center}
\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
\end{center}
\begin{enumerate}
\item ES formation and breakdown is in rapid equilibrium, i.e. the ES concentration does not change over time ({$\frac{d[ES]}{dt}$} = 0).
\item Enzyme is saturated with substrate, therefore it is at V$_{max}$.
\item Initial rate is used, meaning that the substrate concentration does not change significantly.
\item Initial rate is used, meaning that there is no product present.
\end{enumerate}

Michaelis-Menten equation is a simple model of an enzyme catalysed reaction based on the assumptions stated above, and is applicable to most enzymes.

\section{Lineweaver-Burk Plot}

\begin{center}
\includegraphics[width=0.8\textwidth]{lineweaverburk}
\end{center}

Since the V vs [S] curve is hyperbolic, you can only estimate the V$_{max}$ from it, and therefore you can only estimate the K$_M$ value.
In order to measure the V$_{max}$ and K$_M$, you can take the double reciprocal of the V vs [S] curve (i.e. take the reciprocal of both V and S).
This gives us the Lineweaver-Burk plot, which you can measure the V$_{max}$ and K$_M$ from the y- and x-intercept, respectively.

\section{Significance of K$_M$}

\begin{center}
\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
\end{center}

K$_M$ characterises a single enzyme-substrate pair.
K$_M$ is independent of the substrate concentration, but is dependent on the substrate, meaning that K$_M$ is different for different substrate.

K$_M$ is defined as the ratio of the breakdown of the ES complex and the formation of the ES complex:

\begin{center}
\large{K$_M$ = $\frac{breakdown[ES]}{formation[ES]}$ = $\frac{k_{-1} + k_2}{k_1}$ = $\frac{k_{-1}}{k_1}$}
\end{center}

Usually, k$_2$ is much slower than k$_{-1}$, so we can ignore k$_2$.
K$_M$ is also called the ES dissociation constant, and its unit is the same as the concentration.

Low K$_M$ indicates that k$_{-1}$ is low and k$_1$ is high, and therefore it has a high affinity between the enzyme and the substrate, and binds to one another tightly, forming a strong ES complex.
High K$_M$ indicates that k$_{-1}$ is high and k$_1$ is low, and therefore it has a low affinity between the enzyme and the substrate, and binds to one another weakly, forming a loose/weak ES complex.

Therefore, you can use K$_M$ as an indicator of substrate preference of an enzyme.

\section{V$_{max}$ and k$_{cat}$}

\begin{center}
\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
\end{center}

\begin{enumerate}
\item k$_2$ is the rate-limiting step of the reaction
\item The rate of the reaction is proportional to the total enzyme concentration ([E$_{total}$]).
\end{enumerate}

This gives us the definition of k$_{cat}$ and V$_{max}$, relative to one another:

\begin{center}
V$_{max}$ = k$_2$ + [E$_{total}$] \ch{->} k$_{cat}$ = $\frac{[E_{total}]}{V_{max}}$
\end{center}

k$_{cat}$ is the turnover number, or the catalytic rate constant of an enzyme.
The unit is in s$^{-1}$.
\textbf{The number of moles of substrate converted into product, per mole of enzyme per second.}

\section{k$_{cat}$, K$_M$ and Catalytic Efiiciency}

\begin{center}
Catalytic Efficiency = $\frac{k_{cat}}{K_M}$
\end{center}

The ratio of k$_{cat}$ and K$_M$ is the measure of the enzyme efficiency -- how quickly a product is formed relative to how well the substrate binds.
The catalytic efficiency will therefore have the highest value when k$_{cat}$ is high (high turnover), and low K$_M$ (high binding affinity).

\mychapter{13}{Lecture 13}

\section{Irreversible Inhibitors}

Irreversible inhibitors are inhibitors that covalently binds to the enzyme, to a specific amino acid side chain.
This means that the inhibitor permanently inactivates an enzyme, as it cannot be unbound.
Examples of irreversible inhibitors include penicillin (targets glycopeptide transpeptidase in bacterial cell wall) and organophosphates (inhibitor of acetylcholinesterase).

\section{Reversible Inhibitors}

There are two types of reversible inhibitors: competitive and non-competitive inhibitors.
Both of these inhibitors bind to the enzyme non-covalently, and can be `knocked off'.
Some inhibitors can be knocked off more easily than the others, depending on how tight the inhibitor binds to the enzyme.

\begin{center}
\ch{ E + I <=> EI }\\

\vspace{0.5cm}

\ch{K$_i$ = $\frac{[E][I]}{[EI]}$}
\end{center}

K$_i$ is the dissociation or inhibition constant -- just like the K$_M$ with normal substrate.
The lower the K$_i$, the stronger the inhibitor binds to the enzyme.
K$_i$ \textless{} $10^{-9}$ mol L$^{-1}$ is strong binding.

\subsection{Competitive Reversible Inhibitor}

\begin{center}
\includegraphics[width=0.6\textwidth]{compinh}
\end{center}

These inhibitors bind to the enzyme at the active site, and competes with the substrate molecule.
This means that (usually) the inhibitor will have similar structure as the substrate, as it can bind to the active site.

\begin{center}
\textbf{Competitive inhibitors increases the K$_M$, but does not change the V$_{max}$ of the enzyme.}
\end{center}

This means that the apparent affinity of an enzyme for the substrate is reduced (as the K$_M$ is increased).
The V$_{max}$ is not affected, as the binding of the inhibitor is NOT going to affect the catalytic activity of the enzyme -- the enzyme can still reach the V$_{max}$, it just needs more substrate to out-compete the inhibitor (hence the increase in K$_M$).
(The binding of the inhibitor is not going to affect how fast the reaction proceeds, once the real substrate binds to the enzyme).

\subsection{Non-competitive Reversible Inhibitor}

\begin{center}
\includegraphics[width=0.6\textwidth]{noncompinh}
\end{center}

These inhibitors bind to elsewhere to the active site of the enzymes and causes a conformational change, and therefore affects the efficiency of the catalysis.

\begin{center}
\textbf{Non-competitive inhibitors decreases the V$_{max}$, but does not change the K$_M$ of the enzyme.}
\end{center}

When the inhibitor binds to the enzyme, it does not occupy the active site where the substrate binds to.
This means that the active site is free for the substrate to bind -- it's just that the reaction does not proceed as efficient as without the inhibitor bound to it.
Hence the decreased V$_{max}$, but unchanged K$_M$.
(Another way to think about it -- V$_{max}$ is proportional to the total enzyme concentration and the k$_{cat}$, so if k$_{cat}$ decreases, then V$_{max}$ will also decrease).

\section{Methods of Enzyme Regulation}

\subsection{Allosteric Enzyme Regulation}

\begin{center}
\includegraphics[width=0.6\textwidth]{allosterism}
\end{center}

Allosteric enzymes generally display co-operativity and usually multi-subunit.
The enzyme's reaction curve shows a sigmoidal curve.
Activity of allosteric enzymes can be regulated by binding molecules (effectors) away from the active site, which can activate or inhibit the enzyme.
Regulatory effect is achieved by conformational changes caused by the effectors binding to the enzyme.
Note that the inhibitors shift the curve to the right, whereas the activator will shift the curve to the left.

\subsection{Covalent Modification of Enzymes}

In this lecture, only phosphorylation was mentioned.
Phosphorylation and de-phosphorylation is the most common, and is catalysed by kinases and phosphatases, respectively.
Phosphorylation state of an enzyme determines whether the enzyme is `on' or `off'.
However, phosphorylation does not necessarily mean that the enzyme is on -- it really does depend on the enzyme you're studying.

\subsection{Activation of Enzymes by Cleavage}

Some enzymes are synthesised as zymogens -- the inactive form of the enzyme.
The enzyme is only active once the enzyme is cleaved at a certain amino acid residue.
Many of the enzymes involved in the digestive processes and blood clotting proteins are made as zymogens.

\section{Isoenzymes/Isozymes}

Multi-subunit enzymes may exist in multiple forms in an organism -- these enzymes are called isoenzymes.
Isoenzymes differ in physical, kinetic and immunological properties, but all of them catalyse the same reaction.
Isoenzymes are formed from two different genes that make slightly different subunit for the enzyme.
Isoenzyme is a form of transcriptional regulation of cellular activity, as some cells in your body has to catalyse the reaction differently to another part of your body.

\mychapter{14}{Lecture 14}

\section{Pharmacology and Toxicology}

Pharmacology is the study of therapeutic drugs and design of these drugs.
It is the study of how drugs work and how it is metabolised in the cell.

Toxicology is the study of poisoning/toxins.
It is the study of the harmful uses drugs in humans, animal and the environment.

\section{Drugs}

A drug is a chemical that changes the behaviour or function of an individual system, organ, tissue or invading organism.
In other words, a chemical that changes the behaviour of the cell.

\subsection{Sources of Drugs}

You can extract and purify drugs from plants and animals, but have some issues around contamination, etc.
Current sources of drugs are made synthetically, so there will be no contamination from purification.
Drug companies can screen for millions of chemicals, and choose the best one and study it (this process is very expensive).
Rational drug design is where you study the actual target and the disease so that you can make a rational decision about how to make or design a drug for that disease.
Another way to make drugs is to use computer to model what kind of drug will be the best to use.
Once identified, it can then be tested in the lab.

\subsection{Targets of Drugs}

Drugs can bind to a molecule and cause an effect on the cell -- the molecule it binds to will be its molecular target.
Understanding the pathways of the disease or the targets of the drug leads to better drug design.

\section{Agonists and Antagonists}

Agonists are the drugs or molecules that binds to a receptor and can produce a full biological response.
On the other hand, antagonists are drugs or molecules that binds to a receptor and prevents the response to occur (i.e. opposite of agonists).

\section{Receptors}

There are two types of receptors described in these lectures -- drug receptors and physiological receptors.

\subsection{Drug Receptor}

Drug receptor is ANYTHING that the drug binds to.
This can be a proteins, lipids, DNA, RNA, or whatever as long as the drug binds to it and have an effect on the cell through the molecule.

\subsection{Physiological Receptor}

Physiological receptors are normal receptors that are present on the cell surface.
Most of these are transmembrane, meaning that it spans the whole membrane.
These receptors allow signalling to occur from the outside (extracellular) to the inside (intracellular) of the cell.
They also regulate cell function by responding to stimuli such as hormones, neurotransmitters or growth hormones and produce a signal for the cell to do something.
Most drugs activate or inhibit these kind of receptors.

There are four types of physiological receptors: ligand-gated/voltage-linked ion channels, G-protein coupled receptors (GPCR), kinase-linked receptors, and Nuclear receptors.

\section{Ligand-Gated Ion Channels}

Ligand-gated ion channels are proteins present in the cell membrane.
This receptor has a receptor and an ion channel domains, and is made up from 5 subunits.
Binding of the ligand induces a conformational change in the protein, which opens up the channel and allow the ions to pass through.

Specific receptors have specific activating molecules, and only certain ions can pass through.
These receptors control the intracellular ion concentration which gives different charges to the cell.
The response times for these receptors are within milliseconds, which is ideal for fast response (e.g. in nerves).

Using nicotinic acetylcholine receptor as an example, normal substrate/agonist of this receptor is acetylcholine.
Another agonist is nicotine, and an antagonist is tubocurarine.
Tubocurarine is an irreversible inhibitor, so it binds to the receptor and won't release, hence you can eat the hunted animal.

\section{Voltage-Gated Ion Channels}

These receptors are made fromm four subunits, and opened in response to changes in voltage across the membrane (e.g. voltage change caused by ligand-gated ion channels).
The mechanism itself is very similar to the ligand-gated channels.
The drugs that can block these receptors literally blocks the ion channel and prevents the ions to move through the pores (e.g. tetrodotoxin).

\mychapter{15}{Lecture 15}

\section{Nuclear Receptor}

Nuclear receptors, unlike the transmembrane voltage/ligand-gated channels, reside in the cytosol of the cell (i.e. within the cell).
This is partly because the nuclear receptors act in the nucleus, where it interacts with the DNA for gene transcription.
These nuclear receptors binds and interacts with the DNA at the response element for that receptor, which requires multiple regulation mechanisms.
The binding of the nuclear receptor-ligand complex to the response element causes transcription/gene activation.
Since the effect of the cellular response is observable only after the protein is made, the response time is quite slow -- takes hours to days for the response.

\section{Mechanism of Nuclear Receptor Signalling}

For a ligand to enter the cell, the ligand must be lipophilic/hydrophobic, otherwise it will not be able to enter the cell through the membrane.
Normally, the receptor is bound to a heat shock protein (Hsp; a chaperone), that prevents the receptor from binding to the DNA or other receptors.
When the ligand binds to the receptor, the Hsp is displaced/removed, and the receptor dimerises (i.e. binds to another receptor).
The receptor cannot bind to the DNA unless it is dimerised.
Once bound to the DNA, the dimer recruits/associates with many other proteins to activate the gene (e.g. RNA polymerase) and transcribes the gene for cellular response.

\subsection{Regulation of Nuclear Receptor}

There are multiple regulatory steps in this mechanism to prevent unwanted signal and cell response.
Firstly, the receptor is located inside the membrane, meaning that only the lipophilic/hydrophobic ligands can access/activate the receptor.
The Hsp that is bound to the receptor is also important, because without them, the receptor may be able to bind to the DNA cause unwanted gene activation.
Dimerisation is another level of regulation -- it requires the receptor to be dimerised, or else it's not going to activate the gene.

\section{Structure of Nuclear Receptor}

\begin{center}
\includegraphics[width=0.7\textwidth]{nuclearreceptor}
\end{center}

There are three parts to the receptor.
The first part is the ligand binding site (steroid hormones).
Second part is the DNA binding domain, which is the site where the receptor binds to the DNA.
This part contains the zinc finger motif for DNA binding.
The last part is the transcription regulatory domain, where the transcription factors and other machinery binds to for gene activation.

\section{What Genes are Activated by Nuclear Receptor}

The genes activated by the receptor depends on the tissue site the receptor resides in, what cell type it's in, and what kind of ligand/drug binds to the receptor.
This means that the receptor can activate multiple different genes, and therefore have different cellular response depending on the ligand, tissue site, etc. (i.e. it's not a one-gene-to-one-receptor thing).
For example, estrogen receptor have different cellular response depending on the tissue site.

\mychapter{16}{Lecture 16}

\section{G-protein Coupled Receptors (GPCR)}

G-protein Coupled Receptors (GPCRs) are transmembrane receptors that are associated with G-proteins, as the name suggests.
GPCRs acts/responds in the timescale of seconds -- slower than the ligand/voltage gated ion channels, but faster than the nuclear receptors.
GPCRs are important as it is the largest group of the receptors, and therefore an easy target for drugs.
Also, GPCRs are activated by diverse ligands.
About 30 to 50\% of the drugs target the GPCRs, but still only 10\% of the known GPCRs are targeted.

\section{Mechanism of GPCRs}

The ligand binds to the transmembrane receptor on the cell surface.
Before the ligand is attached to the receptor, the receptor is not associated with the G-protein.
Once the ligand binds to the receptor, the conformation of the receptor changes, which allows the G-protein to interact with the receptor.
Interaction with the receptor causes the G-protein to change conformation as well, which increase or decrease the activity of a particular enzyme.
Depending on the activity of the enzyme, the amount of second messenger molecule produced by the enzyme is altered, and therefore causes some cellular response.

\section{Dopamine Receptor}

Dopamine receptor is an example of GPCR.
When the ligand (dopamine) binds to the receptor, it causes a conformational change in the G-protein which decrease the enzyme activity and therefore decreases the amount of second messenger (cAMP) produced.

\subsection{Parkinson's Disease}

Parkinson's disease is caused by decreased level of dopamine, due to the damage in the neural cells that produce dopamine.
Parkinson's can be treated by using levodopa -- a prodrug that is processed into dopamine.
The problem with levodopa is that it is hard to balance/control the effect of levodopa, and leads to psychological effects (such as hallucination/psychosis) and motor complications.

\subsection{Psychosis}

Psychosis occurs when there is too much dopamine around, and over activates the dopamine receptor.
In this case, antipsychotics (antagonists of dopamine receptor) is used to prevent the dopamine to have too much effect on the receptor.
This means that the level of dopamine is stall high, but the effect is inhibited by using antagonists.
Again, however, if you use too much of antipsychotics, you get a Parkinson's like syptoms, signifying the fact that it's very hard to balance the effects of these drugs.

\mychapter{17}{Lecture 17}

\section{Membrane Composition}

Membrane lipids are made from amphipathic molecules, meaning that the molecules have both charged and uncharged parts.
The hydrocarbon tails are hydrophobic and are 16$\sim$22 carbons long.
In animals, the length of the hydrocarbon tails are usually even.
These hydrocarbon tails can be either saturated (no double bonds) or unsaturated (with double bonds).
Note that the double bonds introduce a kink in the tail.

In aqueous solution, the polar group of the molecules are exposed to the solution, whereas the hydrophobic part forms the inner part of the membrane.
This forms a well-known lipid bilayer, which is energetically favourable for the molecules to form in aqueous environment.

\section{Membrane Lipid Structure}

Membrane lipids all have a glycerol backbone which two fatty acids and a phosphate group attaches to.
As mentioned, the fatty acids can be of different length, with or without double bonds.
The phosphate group provides the charge for the amphipathic molecule, and often have polar groups added to the phosphate.
This provides the group extra charge which can form patches on the surface of the membrane.

Note that there are other types of membrane lipids present in different tissue of the same organism, or maybe in tissues from another organism, depending on the function of the tissue in its organism.

\section{Membrane Fluidity}

Membrane fluidity is affected by temperature, the type of fatty acids present, and the presence of cholesterol.
All of these factors have an effect on the hydrophobic, non-covalent interaction between the lipids.
Reminder: non-covalent interactions are very weak on its own, but strong when there are a lot of them, and the strength of the interaction is also dependent on the distance/hoe close to each other.

\subsection{Temperature}

Temperature affects the membrane fluidity by disrupting the non-covalent interactions between the hydrocarbon tails.
As the number of the carbon increases in the hydrocarbon tails increases, the melting point also increases, and therefore less fluid.
This is because there are more carbons for hydrophobic interaction that strengthens the interaction.

\subsection{Type of Fatty Acid}

The type of fatty acids, or more like how saturated a fatty acid is, affects the membrane fluidity.
When the hydrocarbon tails are unsaturated, that is, it has double bonds in it, it has a kink.
This kink causes the hydrocarbon tails to be more separated and the packing of the membrane lipids less ordered.
Since the kink is going to spread the hydrocarbon tails apart further apart, this increases the distance between the tails, and therefore weakens the hydrophobic interaction between them (therefore more fluid).

\subsection{Presence of Cholesterol}

Cholesterol has a rigid ring structure that can be used as a ``scaffold''.
This allows the lipids to anchor themselves to it and form a more ordered structure, and therefore makes the membrane less fluid.

\section{Membrane Proteins}

The membrane is jam-packed with proteins.
Proteins are also able to move in the membrane -- it is dynamic.
There are many functions of membrane proteins -- cell to cell contact, surface recognition, cytoskeleton contact, enzymes, transporters, receptors and signalling.

Proteins can be either integral (all or some part of the protein is embedded in the membrane), or peripheral (on the surface of the membrane).
Peripheral proteins can be held in pace on the membrane surface by polar patches from, for example, the polar groups attached to the phosphate group, forming an electrostatic interaction.

Note that the proteins have ``sides'' on the membrane, and are asymmetric.
This means that not all proteins that are exposed to the extracellular surface are also exposed on the intracellular surface.
For example, most (if not all) of the proteins that are glycosylated are exposed on the extracellular surface, and not on the intracellular surface.

\section{How Are Proteins Attached to the Membrane?}

Non-polar amino acids of the protein resides within the lipid bilayer -- it is unfavourable for hydrophobic amino acids to move out of the membrane, so the proteins is anchored to the membrane.
Proteins can also have an anchoring domain that can be used to anchor the domain to the membrane.

Lipidation, a type of post translational modification, adds fatty acid groups to the protein which can insert it into the membrane for anchorage.

\mychapter{18}{Lecture 18}

\section{Transport of Non-polar Molecules Through the Membrane}

Transport of non-polar molecules through the membrane is dependent on both the hydrophobicity and the concentration of the molecule.
Unlike polar molecules, non-polar molecules are relatively easier to diffuse through the membrane due to its hydrophobic nature of the molecule.

\subsection{Permeability Coefficient and Oil-Water Partition Coefficient}

The permeability coefficient of a molecule describes how well the molecule is able to pass through the membrane -- i.e. how permeable the molecule is.
The smaller the coefficient is, the harder for the molecule to pass through the membrane, and therefore less permeable.

The oil-water partition coefficient describes how hydrophobic a molecule is.
This means that the higher the oil-water partition coefficient is for that molecule, the more hydrophobic the molecule is.

When these two factors are graphed for different kinds of molecules, you can clearly see that the higher the oil-water partition coefficient (i.e. more hydrophobic the molecule is), the higher the permeability coefficient.
This shows that the more hydrophobic a molecule is, the easier for that molecule to pass through the membrane.

\subsection{Concentration Gradient vs. Flux}

When the  flux  is plotted against the concentration gradient of the molecule across the membrane, it is clear that the flux is proportional to the concentration gradient.
This means that the higher the concentration gradient across the membrane, the higher the flux of that molecule through the membrane (i.e. faster/more will go through).

\section{Transport of Polar Molecules Through the Membrane}

Unlike the non-polar molecules, the lipid bilayer represents an energy barrier for the transport of polar molecules across it.
So, for the polar molecules to pass through the membrane, the energy barrier must be decreased.
Proteins within the membrane make small holes in the membrane to allow polar molecules to pass through.

\subsection{Channels}

Channels, when opened, are open to both the intracellular and the extracellular space (i.e. open to both ends).
These channels can be opened or closed spontaneously, and are ``gated''.
Since the mechanism of the channels are fairly simple, the rate of transport of molecules across the membrane when opened can approach the rate of diffusion -- about 10$^7\sim$10$^8$ molecules per second.
The structure of the holes of the channels provide some specificity of the molecules that can go through the membrane.

\subsection{Transporters}

Unlike the channels, transporters are only open to either the intracellular or the extracellular space.
The binding of the molecule that is to be transported induces a conformational change in the transporter which opens the opposite end, allowing the release of the molecule to the other side.
The release of the molecule causes another conformational change that reverts the transporter back into its original state and the process repeats.
Since the process of transporting the molecule requires conformational change of the transporter itself, the rate of transport is much slower compared to the channels -- about 10$^2\sim$10$^3$ molecules per second.

Another characteristic thing about transporters is that they are very much like an enzyme -- transporters are very selective of what they move through the membrane, and it also shows a Michaelis-Menten-like kinetics.
For example, glucose and mannitol are two very similar structured molecules that have similarly low permeability coefficient without a transporter.
However, when GLUT transporter is present, the permeability coefficient increases a lot for the glucose, but not for mannitol, showing that the transporters are selective in which molecule it transports.

\begin{center}
\includegraphics[width=1.0\textwidth]{kinetics}
\end{center}

The kinetics of the transporter is represented by the equation below, where J is the flux and [A] is the concentration gradient of the molecule across the membrane.

\begin{center}
\large{$J_A = \frac{J_{max}[A]}{K_M + [A]}$}
\end{center}

\section{Types of Transport Processes}

\begin{description}
\item [Non-mediated] Requires no protein for transport -- free diffusion.
\item [Mediated] Requires a protein. Also called facilitated transport.
\item [Passive] Down a concentration gradient and no energy is required for the transport.
\item [Active] Up a concentration gradient and requires an input of energy. Energy can come from ATP hydrolysis or from the co-transport of another molecule moving down its concentration gradient.
\item [Co-transport] Movement of two molecules by a transporter.
\begin{description}
\item [Symport] Both molecules move through the transporter in the same direction.
\item [Antiport] The two molecules move in the opposite directions to one another.
\end{description}
\end{description}

\vspace{0.5cm}

\noindent
\textbf{Note: for symport and antiport, both molecules may not be going down the concentration gradient -- one molecule may be going up the concentration gradient, whereas the other maybe going down the concentration gradient.}

\mychapter{19}{Lecture 19}

\section{Energy Metabolism and ATP}

We require approximately 1kg of ATP per kg of body fat for the day.
Obviously, we can't carry that much ATP every day, and so we have to make it as we need it.
Anabolism is the use of energy to form/make macromolecules from building blocks (e.g. making proteins).
Metabolism is the opposite of anabolism, where the macromolecules are broken down into smaller constituent/component/building blocks to be oxidised to produce energy, or to be used in anabolic processes.

From the first law of thermodynamics, the total energy within a system is constant -- i.e. energy can neither be created or destroyed.
This means that the energy must be converted from one form to another.

\section{Energy Balance in the Body}

\begin{center}
E$_{intake}$ = E$_{expended}$ + E$_{stored}$\\
\vspace{0.5cm}
E$_{expended}$ = Basal metabolism + Activities
\end{center}

As you can see from the equation, all of the energy taken into the body must be either used up/expended, or stored.
Clearly, to lose weight, you must either decrease your intake, or increase your energy expenditure.
You can increase the energy expenditure by increasing your basal metabolism and/or the amount of exercise/activities.
(Note that small amount of energy is lost in faeces, urine and sweat.)
Weight gain is a problem in developed countries, as they tend to have higher percentage of obese people.

The rest of this lecture will be focusing on each aspect of the equation.

\subsection{Subtopic: Leptin Receptor}

Leptin deficient mice eats too much.
Leptin is a hormone that binds to the leptin receptor, which acts so that the host feels full, and therefore stops eating.
So, if a mouse is leptim deficient, then the mouse will not feel full, and therefore continues eating.

\section{Units of Energy}

Joule (J) is the unit of energy, defined as the \textbf{energy required to push against 1N of force for 1m}.
An older metric is the calories, where 1cal is the energy required to heat 1g of water from 14.5$^\circ$C to 15.5$^\circ$C.
1cal is equivalent to 4.184J.
Daily energy intakes and expenditure usually measured as MJ.

\section{Measuring Energy Intake}

\subsection{Bomb Calorimetry}

You can measure the energy contained in the food by burning the food and measuring the change in temperature of the water.

\begin{center}
\ch{C6H12O6 + 6 O2 -> 6 CO2 + 6 H2O + heat}\\
\ch{C16H32O2 + 23 O2 -> 16 CO2 + 16 H2O + heat}
\end{center}

Carbohydrates releases 2813kJ/mol (glucose), whereas fatty acids release 10024kJ/mol (palmitate).
Since fatty acids are more reduced, it can produce more energy from oxidation.
One thing to note is that not all energy in food is available for the body (e.g. cellulose and nitrogen, which are excreted out).
Therefore, these factors must be taken into account for calculations.

\section{Atwater Factor}

Atwater factors are used to measure how much energy each constituents have.
Fat has 38kJ/g, carbohydrates and proteins have 17kJ/g and ethanol has 29kJ/g.
These atwater factors can be used to calculate how much energy is contained in the food.

\section{Measuring Energy Expenditure}

Bomb calorimetry of a human would be good, but not feasible for obvious reasons.
There are other ways to measure energy expenditure.

\subsection{Direct Calorimetry}

This method relies on measuring heat output from individual in a whole body calorimeter.
This is good for determining the basal metabolic rate (BMR) -- i.e. at rest.
However, it is quite hard to measure the energy used in more physical exercises, due to space constraints and complexity of the calculation and machinery.

\subsection{Indirect Calorimetry}

This method measures the energy expenditure based on the oxygen consumption and carbon dioxide production, which is measured using a respirometer.
Since certain amount of energy is associated with every litre of oxygen consumed (20.9kJ/L), you are able to calculate how much energy was used.
This allows us to calculate the energy expenditure for a range of activities, and also allows the calculation of the respiration exchange ratio (RER).

\subsection{Respiration Exchange Ratio}

\begin{center}
\large{RER = $\frac{CO_2 produced}{O_2 consumed}$}
\end{center}

RER tells us about what kind of fuels are used -- carbohydrates or fat.
Note that the stoichiometry of the oxygen and carbon dioxide is different for carbohydrates and fat.
Since the stoichiometry of oxygen to carbon dioxide is 1:1 in carbohydrates, RER = 1, but this not the case for fat, and therefore have RER = 0.7.

\section{Basal Metabolism}

Basal metabolism is the energy required for the maintenance of life -- general metabolism, muscle concentration, etc.
This is usually defined as the energy expenditure at rest.
BMR is affected by:
\begin{itemize}
\item gender
\item age 
\item body size
\item body composition
\item genetics
\item hormonal status
\item stress levels
\item disease status
\item certain drugs
\end{itemize}

BMR can be increased by: athletic training, fever, late pregnancy, drugs (e.g. caffeine) and hyperthyroidism.
BMR can be decreased by: malnutrition, sleep, drugs (beta blockers) and hypothyroidism.

\subsection{Hyper and Hypothyroidism}

\begin{description}
\item [Hyperthyroidism] Increased metabolic rate due to excessive production of thyroid hormone, which acts to increase the effect of insulin, etc. Symptoms include heat production, anxiety, etc.
\item [Hypothyroidism] Decreased metabolic rate due to thyroid hormone deficiency (from iodine deficiency). Symptoms include coldness, depression, etc.
\end{description}

\mychapter{20}{Lecture 20}

\section{Organs Involved in Digestion}

\subsection{Salivary Glands}

Produces saliva -- neutral pH and contains mucous and amylase that starts the digestion of carbohydrates.

\subsection{Stomach}

For food storage (controlled release into the duodenum) and mixing of the food with gastric juices.
Stomach release highly acidic acid (0.1M HCl) that denature the proteins.
It also secretes pepsinogen (zymogen) for protein digestion, and also mucous layer for protection.

The parietal cells produce the acid (HCl), and the chief cells produce the pepsinogen, which is activated by the acidic environment.

\subsection{Pancreas}

Pancreas is slightly alkaline pH 7, and secretes most digestive enzymes, including amylase, lipase and many proteases.

\subsection{Liver}

Synthesis of bile salts/acids (stored in gall bladder), which is important for fat digestion.

\subsection{Small Intestine}

Where the final phase of digestion and absorption occurs.

\section{Digestion}

There 2 main phases involved in digestion:
\begin{enumerate}
\item Hydrolysis of bonds connecting the monomer units in food macromolecules.
\item Absorption of products from the gut into the body.
\end{enumerate}

In the case of carbohydrates, proteins and TAGs, the bond to cleave would be the glycosidic bonds, peptide bonds and ester bonds, respectively.

\section{Carbohydrate Digestion}

Carbohydrates make up 40 to 50\% of the energy intake.
Carbohydrates include starch (amylose and amylopectin), glycogen, simple sugars and fibre (i.e. cellulose).
Cellulose cannot be digested in humans, as we lack the enzyme for breakdown of cellulose.

\subsection{Enzymes Involved in Carbohydrate Digestion}

The digestive process begins in the mouth where the salivary amylase starts breaking down the starch molecules into smaller sugars.
When it enters the small intestine, the pancreatic amylase (secreted from the pancreas) further breaks down the starch molecules into maltose and isomaltose.
The intestinal epithelial cells secretes maltase, isomaltase, lactase and sucrase, which breaks down the disaccharide molecules into monomers (glucose, fructose and galactose).
These monosaccharides are then absorbed into the body via the intestinal villi.

\begin{reactions*}
Maltose/Isomaltose &->[Maltase][Isomaltase] 2 Glucose\\
Sucrose &->[Sucrase] Fructose + Glucose\\
Lactose &->[Lactase] Galactose + Glucose
\end{reactions*}

\subsection{Lactose Intolerance}

This is caused by lactase enzyme deficiency.
Since lactose is not digested, it accumulates inside the intestine where the gut microbes can use it for fermentation.
Lactose intolerance causes bloating, flatulence and diarrhoea due to the fermentation.
Individuals with this should avoid lactose consumption.

\section{Carbohydrate Absorption}

Monosaccharides are absorbed into the body via the villi and microvilli (brush border) of the intestinal wall.
The villi and microvilli provides a large surface area for absorption.

\subsection{Glucose Transport}

Sugars are water soluble, and therefore require a transporter to pass through the cell membrane into the cell.
SGLT1 protein is present in the intestinal side of the cell membrane and transports glucose into the intestinal cell together with the Na ion (symport).
Movement of Na ion from the intestine into the cell is down the concentration gradient, which provides the energy for the glucose transport up the gradient into the cell.

On the other side of the cell membrane, GLUT2 protein is present.
Since the concentration of glucose in the cell is higher than the blood, the glucose can move down the concentration gradient through the GLUT2 transporter.
Note that there is Na/K pump on this side of the membrane to maintain the Na concentration gradient.

Once transported into the blood, other tissues can uptake glucose via GLUT3 (brain) and GLUT4 (muscle and adipose).

\subsection{Coeliac Disease}

Disease of the small intestine, where the body reacts against the wheat protein gluten.
This causes the villi to be flattened and nutrients are not absorbed properly.

\mychapter{21}{Lecture 21}

\section{Protein Digestion and Absorption}

Proteins can be used for energy, but it is more for regenerating the depleted amino acids.
It is also important for supplying the body with essential amino acids.
Proteins are the major source of nitrogen for purines, pyrimidines and haem.
The carbon skeleton of the amino acids can be used as fuel/energy production, where the amino group (i.e. nitrogen) is converted into urea and excreted.
Decrease/malnutrition of certain amino acid can lead to Kwashiokhor.

\section{Essential Amino Acids}

We cannot synthesise all 20 amino acids -- require 8 essential amino acids from the diet.
These amino acids are: L, K, T, Y, I, M, F and V.

\section{Regulation of Protein Digestion}

Obviously, we don't want to ``eat ourselves'', so the process is highly regulated by polypeptide hormones.
These hormones are: gastrin (stomach), secretin (duodenum), and cholecystokinin (duodenum).

\section{Enzymes Involved in Protein Digestion}

Before we move onto how proteins are digested, we need to know what kind of enzymes are involved.
The enzymes present are: pepsin (stomach), trypsin, chymotrypsin and carboxypeptidase (3 from pancreas), and aminopeptidase and dipeptidase (2 from small intestine).
All of these enzymes act in the small intestine, except for pepsin.
Note that the protease specificity is determined by the adjacent amino acid side chains of the peptide chain it is breaking.
For example, pepsin, trypsin and chymotrypsin cleaves the polypeptide after aromatic, positively charged and aromatic side chains, respectively.

\section{Protein Digestion}

Digestion of protein involves the hydrolyses of specific peptide bonds, performed by several different proteases.
All of these proteases are secreted as inactive zymogens to prevent them from digesting our own cells, and they are all activated by cleavage of peptides from their structure.

There are two stages of protein digestion:
\begin{enumerate}
\item Endopeptidases attack the peptide bonds within the peptide chain. Pepsin, trypsin and chymotrypsin are all endopeptidases.
\item Exopeptidases attack the peptide bonds at the end of the peptide chain (i.e. from C-term or N-term -- hence amino or carboxypeptidase).
\end{enumerate}

\subsection{Pepsinogen Activation}

Pepsinogen (due to its structure) have its catalytic active site inhibited.
H$^+$ (low pH) changes the conformation of the pepsinogen and allows it to auto-cleave the peptide bond and activate itself.
The activated pepsin can then activate other pepsinogens.
Note that this process is all triggered by the ingestion of food, and the hormones associated with it.

\subsection{Small Intestine Peptidase Activation}

All 3 peptidases (trypsin, chymotrypsin and carboxypeptidase) are secreted in the inactive zymogen forms -- trypsinogen, chymotrypsinogen and procarboxypeptidases.
First, trypsinogen is activated by the membrane bound enterokinase that adds the phosphate group to it.
This causes a conformational change in the trypsinogen structure which allows it to auto-cleave and auto-activate itself.
The activated trypsin activates the chymotrypsinogen and procarboxypeptidase.

As a result of the activation of all of these proteases (including pepsin), dietary proteins are digested into single amino acids, dipeptides and tripeptides.
Di and tripeptidases then digest the di and tripeptides into single amino acid.

\section{Protein Absorption}

\subsection{Amino Acid Transport}

There are 5 main amino acids transporters to transport the amino acids across the luminal membrane.
These are: Neutral, basic, acidic, other (taurine), and Glycine, proline and hydroxyproline.
All of these amino acids are transported in a similar way to the glucose transport -- co-transport with Na.

\subsection{Peptide Absorption}

Very little absorption occurs for peptides that are longer than 4 amino acids.
Di and tripeptides are absorbed in the small intestine via co-transport with H$^+$ ions (PepT1).
These di/tripeptides are then digested in the cell and exported out into the blood.

\subsection{Absorption of Intact Proteins}

This only occurs in few circumstances -- e.g. in newborn for immunoglobulin absorption.

\section{Diseases of Inappropriate Enzyme Activation}

This is associated with inappropriate enzyme action (zymogen activation).
This can lead to the breakdown of mucosal layer and proceed to peptic ulcers.

Pancreatitis is where the pH is altered and degrades/inflame the pancreas.
It can also release the enzymes into the blood.

\section{Diseases of Malabsorption}

Cystic fibrosis patients have abnormally thick mucous secretions due to mutation in chloride transporter.
The thick mucous can block the pancreatic ducts, and therefore no protease activity.

Coeliac disease have been mentioned in the previous lecture.

\section{Digestion of Nucleic Acid}

DNA and RNA are subject to acid hydrolysis in the stomach.
Intestinal nucleases also hydrolyse the phophodiester bonds between the nucleotides.
The individual nucleotides are absorbed via nucleotide transporters (Na co-transport).

It is most unlikely that the DNA/RNA gets incorporated into our body, but it is a possibility as there are evidence of DNA being intact in the gut.

\mychapter{22}{Lecture 22}

\section{Lipid Digestion and Absorption}

Lipid digestion starts and the mouth, which is continued in the stomach, and lastly in the small intestine (main site of fat digestion).
Lipid digestion is done by the lipase enzymes, present in all three location mentioned above, which breaks down the triacylglycerol (TAG) into glycerol and free fatty acids (FAs).

\section{Bile Acids/Salts}

Dietary lipids must be emulsified/solubilised in the aqueous solution.
This is done by the bile acids/salts, which are made from cholesterol.
Bile acids are synthesised in the liver and stored in the gall bladder as bile, which is secreted in response to cholecystokinin.

Bile acids are powerful detergents with hydrophobic  and hydrophilic surfaces (amphipathic), which allows it to form micelles with TAGs and increase the surface area for lipid digestion by the pancreatic lipase (also associated with the micelles).
Note that bile acids have similar structure to cholesterol, but have OH and COOH groups on them.

\section{Bile Composition}

Bile contains: water, bile acids, electrolytes, cholesterol, phospholipids, and bile pigments (bilirubin).
Note that the ration of electrolytes to cholesterol (electrolytes:cholesterol) is important to avoid gall stones in the gall bladder.

\section{Digestion of Lipids}

Firstly, the lipids are emulsified by the bile salts (in response to cholecystokinin) to form micelles.
Pancreatic lipase binds to the lipid/aqueous interface of the micelles and hydrolyses the TAGs at positions 1 and 3 of the glycerol backbone.
After this, smaller micelles form with bile salts, hydrolysed FFAs, MAGs, and cholesterol.
Micelles are then absorbed across the intestinal cell membrane into the blood.

\section{Fat Malabsorption}

This leads to excess fat and fat soluble vitamins in faeces, and are caused by conditions that interfere with bile or pancreatic lipase secretion (e.g. pancreatitis, gall bladder or liver disease).

\subsection{Orlistat (Xenicol)}

Potent inhibitor of pancreatic lipase, and therefore TAGs get excreted without getting absorbed.

\section{Process of Fat Absorption}

\subsection{Lipoproteins}

Lipoproteins are particles that are made of lipid core (TAGs and cholesterol) surrounded by phospholipids and proteins that sulubilise the lipids in the blood for transport around the body.
In general, the structure of lipoproteins are made of core TAGs and cholesterol esters, surrounded by phospholipids and apoproteins.
The apoprotein component of the lipoproteins provides the specificity of where the lipoproteins are transported.

There are four main classes of lipoproteins: chylomicrons, VLDL, LDL, and HDL.
\begin{description}
\item[Chylomicrons] are the least dense (in terms of TAG content) out of all four classes. These contain mainly TAGs, and the apoproteins associated are: apoB48, C2, C3, and E.
\item[VLDL] are the next least dense class, which contains mainly TAGs (but not as much as chylomicrons), with slightly higher phospholipid content. The main apoproteins associated are: apoB100, C1, C2, C3 and E.
\item[LDL] are second most dense class and consists mainly of cholesterol, with higher protein content. Apoprotein associated with this class is apoB100.
\item[HDL] are the most dense class and consists mainly of proteins. Main apoproteins involved in this class are: apoA1, A2, C1, C2, C3, D and E.
\end{description}

The two main functions of lipoproteins are to make the lipids soluble for transport in the blood, and to provide a delivery system for shifting the lipids in/out of the cells.

\subsubsection{Function of Apoproteins}

Apoproteins have various functions depending on the type of apoprotein.
Some apoproteins are structural for lipoprotein assembly (apoB).
ApoB and E are `ligands' for cell surface receptors (i.e. recognition for lipoprotein absorption).
Some are enzyme cofactors (e.g. apoC2 for lipoprotein lipase).

\subsection{Lipid Transport Pathways}

There  are two major lipid transport pathways.
The exogenous chylomicron pathway transports dietary fat from the GI tract to other parts of the body.
The endogenous VLDL/LDL pathway transports endogenously synthesised fat.

\subsection{Exogenous Pathway}

The FFAs and MAGs that are absorbed into the intestinal cell gets resynthesised back into TAGs.
These TAGs and other lipids combine with apoB in the ER to form chylomicrons and secreted from the intestinal cells into the blood via the lymphatic system.
FFAs are transported to adipose tissues and muscles, and the remnants are transported to the liver where it is recognised by the apoE receptor.

\subsubsection{Lipoprotein Lipase}

Lipoprotein lipase is an enzyme found on the endothelial surface that hydrolyses TAGs in the lipoproteins to glycerol and FFAs.
It has the highest activity in the heart, adipose tissues and skeletal muscles.
Lipoprotein lipase is activated by apoC2 protein, and mutation in apoC2 or lipoprotein lipase leads to increased chylomicrons and plasma TAGs.

\subsection{Endogenous Pathway}

Fat can also be synthesised in the liver as well, and therefore require transport out of the liver cell.
VLDLs are synthesised and exported into the blood from the liver cells, and contains apoC2 to interact with lipoprotein lipase for further TAG breakdown.
VLDL can get ``receycled'' into LDL -- the cholesterol rich lipoprotein that is considered to be ``bad'', and only contains apoB100.
These LDL can get absorbed/taken up by other tissues via  the LDL receptor, but if this receptor is mutated, it can cause increased cholesterol and LDL in the blood.

\subsection{Familial Hypercholesterolemia (FH)}

This is caused by defects in LDL receptor gene and leads to premature atherosclerosis.
It is a dominant disorder and so heterozygotes are also affected by this condition.
This leads to increased LDL level by about 2 to 3 times higher.
FH is treated with statins (drug that inhibits one of the enzymes involved in cholesterol synthesis).

\end{document}

### 2015-09-01 bioc192/summary_tute/summarytute.tex
edited for summary tute

@@
 
+\subsection{Supersecondary structure}
+
+Secondary structures such as $\alpha$-helix, $\beta$-sheets, and turns combine together to form a specific supersecondary structures, or `motifs'.
+
 \subsection{Tertiary Structure}
@@
 
-Amino acids are the building blocks of a protein.
-Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).
+%Amino acids are the building blocks of a protein.
+%Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).
 
@@
 Due to the variable side chains, all amino acids except glycine are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
-The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.
+%The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.
 
@@
 
-As mentioned earlier, all 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
-This means that each of the 20 amino acids have different chemical properties, depending on the side chain group it has.
+%As mentioned earlier, all 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
+Each of the 20 amino acids have different chemical properties, depending on the side chain group it has.
 Cells take advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster (i.e.
@@
 
-Due to the unique chemical properties of the amino acid side chains, they are grouped into subgroups.
-There are four main groups: Non-polar, uncharged polar, negatively charged, and positively charged amino acids.
+%Due to the unique chemical properties of the amino acid side chains, they are grouped into subgroups.
+%There are four main groups: Non-polar, uncharged polar, negatively charged, and positively charged amino acids.
 
@@
 These amino acids have side chains that can be charged, but uncharged under normal physiological pH.
-As you can see, the side chain contains --OH groups and/or --NH$_2$ groups, whichare able to de/protonate under certain pH.
+As you can see, the side chain contains --OH groups and/or --NH$_2$ groups, which are able to de/protonate under certain pH.
 The amino acids in this group are: serine, threonine, tyrosine, asparagine, and glutamine.
@@
 The side chain contains a protonated --NH$_2$ group, giving it a positive charge to the side chain.
-The pK$_a$ of these amino acid side chains ranges from 10$\sim$12, with an exception of histidine, which has a pK$_a$ of 7$\sim$7.5.
+The pK$_a$ of these amino acid side chains ranges from 10$\sim$12, with an exception of histidine, which has a pK$_a$ of about 6 (and therefore only partially ionised).
 The amino acids in this group are: lysine, arginine, and histidine.
@@
 Depending on the pH of the solution, there will be more or less protonated molecules.
-
-The amount of de/protonated amino acids affect the net charge of the protein, and therefore the isoelectric point (pI) of the protein.
 The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).
 
-When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to become protonated.
-Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to become deprotonated.
-So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
+In general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
 
@@
 Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
-However, some amino acid side chains may be charged under physiological condition and contribute to the net charge of the amino acid, and therefore the protein.
-This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.
+Some amino acid side chains may be charged under physiological condition and contribute to the net charge of the amino acid, and therefore the protein.
 
@@
 
-Amino acids are covalently linked together by peptide bonds -- the bond formed between the amino group of one amino acid with a carboxyl group of another amino acid.
-A chain of amino acids linked together by peptide bonds are called peptides, and these are usually short.
-Peptides that have a longer chain of amino acids with defined biological function are called proteins.
-(Note: peptides are proteins as well -- it's just a little smaller)
-
 \subsection{Amide (peptide) Bonds}
@@
 
-\begin{center}
-\includegraphics[width=\textwidth]{amidebond}
-\end{center}
-
 \mychapter{2}{Lecture 2}
@@
 
-Single bonds between the carbon atoms can be rotated freely, but not the double bonds.
 Since the amide bonds have a partial double bond characteristic, it is rigid and forms a plane that does not allow rotation.
@@
 
-\subsection{Phi ($\phi$) Angle}
-
-Phi angle is defined as the bond angle rotation of the N--C$_{\alpha}$ bond.
+\begin{description}
+\item [Phi angle] is defined as the bond angle rotation of the N--C$_{\alpha}$ bond.
 This bond rotation leads to O--O atom collision.
 
-\subsection{Psi ($\Psi$) Angle}
-
-Psi angle is defined as the bond angle rotation of the C--C$_{\alpha}$ bond.
+\item [Psi angle] is defined as the bond angle rotation of the C--C$_{\alpha}$ bond.
 This bond rotation leads to the NH--NH collision.
 
-\vspace{1.0cm}
+\item [Omega angle] is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
+This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).
+
+\item [Chi angle] is the bond angle rotation of the side chain bond (C$_{\alpha}$--R bond).
+
+\end{description}
+
+\vspace{0.3cm}
 
@@
 
-\subsection{Omega ($\omega$) Angle}
-
-mega angle is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
-This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).
-
-\subsection{Chi ($\chi$) Angle}
-
-Chi angle is the bond angle rotation of the side chain bond (C$_{\alpha}$--R bond).
-
-\section{Ramachandron Plot}
-
-This plot shows all the possible phi/psi angles (i.e. the two amide planes adjacent to the C$_{\alpha}$) that are allowed in a protein with no steric interference.
-Any phi/psi angle pairs that cause steric hindrance are not plotted, so we get some `regions' of allowed angle pairs for any given protein.
-The side chains of the amino acids are (usually) staggered to minimise steric hindrance.
-
-\vspace{0.5cm}
-
-\includegraphics[width=\textwidth]{ramachandron}
-
 \section{Secondary Structure of a Protein}
 
-There are three main secondary structures in a protein: $\alpha$-helix, $\beta$-sheets, and turns.
-
 \subsection{$\alpha$-Helices}
-
-Specifically speaking, $\alpha$-helix is a type of helix that is present in a protein -- there are other types of helices (e.g. polyproline helix), but $\alpha$-helix is the most common type.
 
@@
 Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
-This flexibility can cause disruption in the helical structure.
 Proline has a unique side chain that connects back to its amino group, forming an imino group, meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.
@@
 
+This plot shows all the possible phi/psi angles (i.e. the two amide planes adjacent to the C$_{\alpha}$) that are allowed in a protein with no steric interference.
+Any phi/psi angle pairs that cause steric hindrance are not plotted, so we get some `regions' of allowed angle pairs for any given protein.
+The side chains of the amino acids are (usually) staggered to minimise steric hindrance.
+
 Secondary structures have certain phi/psi angle pairs, and when you overlay these to the plot, it falls into the allowed regions of the plot at a distinct region.
@@
 This essentially shows that secondary structures have unique phi/psi angle pairs.
+
+\includegraphics[width=\textwidth]{ramachandron}
 
@@
 
-\vspace{0.5cm}
-
-\noindent
 The process of protein folding is summarised below:
@@
 
-\begin{center}
-\includegraphics[width=0.4\textwidth]{haem}
-\end{center}
-
 \section{Haemoglobin}
@@
 Haemoglobin is the tetrameric (4 subunit) form of myoglobin that consists of two $\alpha$ and two $\beta$ subunits, and it is a blood O$_2$ transporter.
-Since haemoglobin is a tetramer, it has 4 haem group in total (one on each subunit), and therefore able to carry 4 O$_2$ molecules.
 Slight conformational change occurs when the O$_2$ binds to the haemoglobin molecule, which will become important in explaining the cooperative effect of the subunits.
@@
 One thing to note here from the curve is that the shape of the curve for haemoglobin is sigmoidal (S-shape).
-This indicates that haemoglobin is allosteric (or cooperative), where the subunits assist one another to facilitate the binding/release of O$_2$ molecules.
+This indicates that haemoglobin is allosteric and cooperative, where the subunits assist one another to facilitate the binding/release of O$_2$ molecules.
 Of course, there are other factors that contribute to the allosteric binding/release of O$_2$ molecules.
@@
 
-When O$_2$ binds to a subunit of haemoglobin, it causes a slight conformational change to the subunit.
-This conformational change assists the binding of O$_2$ molecules to the other subunits by cooperativity/allosterism.
-For clarity, the tense (T) state is the state of the subunit when O$_2$ is not bound (low O$_2$ affinity), and the relaxed (R) state is the state of the subunit when O$_2$ is bound (high O$_2$ affinity).
+When O$_2$ binds to a subunit of haemoglobin, it causes a slight conformational change to the subunit, and this conformational change assists the binding of O$_2$ molecules to the other subunits by cooperativity/allosterism.
 
 There are two models, sequential and concerted models, of allosterism.
-It is still unknown which model is correct, as there are many observations supporting both models.
 
@@
 This conformational change makes the other subunits easier to bind to O$_2$ molecules, which will also change conformation of itself and the others.
-
-With respect to the R/T states, binding of O$_2$ molecule causes a single subunit to be in R state, and also changes the conformation of the adjacent subunits so that it is easier to bind to O$_2$ molecules, therefore easier for the subunits to turn into the R state.
 
@@
 
-As discussed earlier in Lecture 4, binding of O$_2$ molecule to the haem group causes a conformational change, and thus promotes the R or T state.
-This conformational change is caused by the shift in the Fe$^{2+}$ atom of the haem, which is attached to the His F8 of the haemoglobin subunit, which then cause a slight shift that propagates through the protein.
-
-However, there are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
+There are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
 Since these molecules have effects on haemoglobin, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.
-
-\vspace{0.3cm}
 
@@
 Binding of BPG at the muscle/tissue site ensures that the O$_2$ molecules are released at the muscles/tissues.
-
 O$_2$ binding to the haemoglobin at the lungs causes the release of BPG molecule, allowing the haemoglobin to carry O$_2$ to the muscles/tissues.
@@
 
-\includegraphics[width=\textwidth]{bicarbonate}
-
 \mychapter{6}{Lecture 6}
@@
 \section{Normal Variants}
-
-These variants are normal in the sense that they are naturally occurring, and not caused by abnormal mutations.
-These variants are different to the normal haemoglobin in that it uses different type of subunits to make the haemoglobin up.
 
@@
 
-The graph below just shows how different subunits are made at different developmental stage.(insert the graph)
-As you can see, there are subunits other than $\alpha$, $\beta$ and $\gamma$ subunits.
-
 \section{Abnormal Variants}
@@
 
-This variant is caused by a single point mutation on the $\beta$ subunit of haemoglobin.
-This point mutation causes an amino acid substitution from glutamic acid to valine (E6V mutation), which is on the surface of the protein.
-
+This is caused by a point mutation of an amino acid substitution from glutamic acid to valine (E6V mutation), which is on the surface of the protein.
 The consequence of this mutation is the formation of deoxy-Hb aggregate/crystal in the red blood cell.
@@
 
-The diversity of the proteins can be expanded by mRNA splicing and PTM.
 PTM can be categorised into two categories: the covalent addition of chemical group to amino acid side chain, or covalent hydrolysis of the peptide bond in a protein.
@@
 
-The consequence of adding a phosphate group to these amino acid side chains is that it's going to alter the protein structure, due to the electrostatic attraction between the more negatively charged phosphate group with other postitively charged amino acid side chains.
-Since the attraction is going to shift the amino acids slightly, it will cause a shift in the whole protein, and therefore alters the conformation of the protein and affects the biological properties of it.
-
+The consequence of adding a phosphate group to these amino acid side chains is that it's going to alter the protein structure, due to the electrostatic attraction between the more negatively charged phosphate group with other postitively charged amino acid side chains, and changes the conformation of the protein and its activity/function.
 Quite often, phosphorylation state is used to `switch' the conformational state of the protein to enable a biological process in a cell (e.g. facilitating interaction with other proteins).
@@
 
-When you have vitamin K deficiency, you can't gamma-carboxylate the glutamic acids, since the reaction requires vitamin K as a co-factor.
-As a consequence, this leads to bleeding disorder.
+Vitamin K deficiency leads to bleeding disorder.
 
@@
 The advantage of this is that it is cheap, has high yield, and it is pathogen free.
-However, in a prokaryotic system, the cells are unable to post-trasnlationally modify the protein, which may be essential for the function of some proteins.
+However, in a prokaryotic system, the cells are unable to post-translationally modify the protein, which may be essential for the function of some proteins.
 Another problem is that the proteins may not fold properly or only partially folded.
@@
 
-In most cases, enzymes are proteins (in some cases, it's RNA enzymes) that acts as biological catalysts.
+Enzymes are biological catalysts.
 Enzymes increases the rate of reaction, but does NOT change the equilibrium, nor the thermodynamics of a reaction.
 This means that, even though enzymes make the reactions to occur faster, it cannot shift the chemical reaction to favour one product over another.
-Note that enzymes are named ``-ases''.
 
@@
 
-\begin{center}
-\includegraphics[width=0.7\textwidth]{geo_stereo}
-\end{center}
 Geometric specificity is where the substrate must be made of same component/structure for it to be recognised by the enzyme.
@@
 In an enzyme catalysed reaction, substrate binds to the active site and forms an enzyme-substrate complex (ES), which is stabilised by non-covalent interactions.
-Note that this interaction is specific.
 As the reaction proceeds, transition state (T), which is neither a substrate nor the product, is formed, then proceeds further into enzyme-product complex (EP).
@@
 
-\begin{center}
-\includegraphics[width=0.6\textwidth]{enzymemodels}
-\end{center}
-
 \subsection{Lock-and-Key Model}
@@
 The concept of the lock and key model is fairly simple -- the substrate must have the shape/structure that perfectly matches the enzyme active site.
-If the substrate doesn't then, the enzyme cannot react with it and no reaction occurs.
+If the substrate doesn't then the enzyme cannot react with it and no reaction occurs.
 However, this is not really the case in nature as enzymes changes conformation when the substrate binds to it.
@@
 
-This is how an enzyme actually works.
 The substrate and the enzyme active site does not match perfectly as it did for the lock and key model.
@@
 
-The enzyme requires an amino acid with a side chain that can act as a nucleophile (e.g.ser, cys)
+The enzyme requires an amino acid with a side chain that can act as a nucleophile (e.g. ser, cys)
 
@@
 Co-substrates are usually derived from vitamin derivatives.
-(An example used in the lecture was NAD)
 
@@
 In kinetics term, the enzyme shows first-order kinetics initially (where the rate depends on the substrate concentration), but then shows zero-order kinetics (where the rate does not depend on/changes with substrate concentration).
-When the velocity of the reaction is plotted with the substrate concentration, this is quite apparent from the hyperbolic curve of the graph.
 
@@
 In order to measure the V$_{max}$ and K$_M$, you can take the double reciprocal of the V vs [S] curve (i.e. take the reciprocal of both V and S).
-This gives us the Lineweaver-Burk plot, which you can measure the V$_{max}$ and K$_M$ from the y- and x-intercept, respectively.
 
@@
 Hence the decreased V$_{max}$, but unchanged K$_M$.
-(Another way to think about it -- V$_{max}$ is proportional to the total enzyme concentration and the k$_{cat}$, so if k$_{cat}$ decreases, then V$_{max}$ will also decrease).
+%(Another way to think about it -- V$_{max}$ is proportional to the total enzyme concentration and the k$_{cat}$, so if k$_{cat}$ decreases, then V$_{max}$ will also decrease).
 
@@
 \subsection{Allosteric Enzyme Regulation}
-
-\begin{center}
-\includegraphics[width=0.6\textwidth]{allosterism}
-\end{center}
 
@@
 
-Pharmacology is the study of therapeutic drugs and design of these drugs.
-It is the study of how drugs work and how it is metabolised in the cell.
-
-Toxicology is the study of poisoning/toxins.
-It is the study of the harmful uses drugs in humans, animal and the environment.
-
 \section{Drugs}
@@
 In other words, a chemical that changes the behaviour of the cell.
-
-\subsection{Sources of Drugs}
-
-You can extract and purify drugs from plants and animals, but have some issues around contamination, etc.
-Current sources of drugs are made synthetically, so there will be no contamination from purification.
-Drug companies can screen for millions of chemicals, and choose the best one and study it (this process is very expensive).
-Rational drug design is where you study the actual target and the disease so that you can make a rational decision about how to make or design a drug for that disease.
-Another way to make drugs is to use computer to model what kind of drug will be the best to use.
-Once identified, it can then be tested in the lab.
 
@@
 
-These receptors are made fromm four subunits, and opened in response to changes in voltage across the membrane (e.g. voltage change caused by ligand-gated ion channels).
+These receptors are made from four subunits, and opened in response to changes in voltage across the membrane.
 The mechanism itself is very similar to the ligand-gated channels.
@@
 Also, GPCRs are activated by diverse ligands.
-About 30 to 50\% of the drugs target the GPCRs, but still only 10\% of the known GPCRs are targeted.
 
@@
 These hydrocarbon tails can be either saturated (no double bonds) or unsaturated (with double bonds).
-Note that the double bonds introduce a kink in the tail.
 
@@
 Proteins can also have an anchoring domain that can be used to anchor the domain to the membrane.
-
 Lipidation, a type of post translational modification, adds fatty acid groups to the protein which can insert it into the membrane for anchorage.
@@
 
-The kinetics of the transporter is represented by the equation below, where J is the flux and [A] is the concentration gradient of the molecule across the membrane.
+The kinetics of the transporter is represented by the equation below, where J is the flux and [A] is the concentration of the molecule.
 
@@
 You can increase the energy expenditure by increasing your basal metabolism and/or the amount of exercise/activities.
-(Note that small amount of energy is lost in faeces, urine and sweat.)
-Weight gain is a problem in developed countries, as they tend to have higher percentage of obese people.
-
-The rest of this lecture will be focusing on each aspect of the equation.
 
@@
 \section{Measuring Energy Expenditure}
-
-Bomb calorimetry of a human would be good, but not feasible for obvious reasons.
-There are other ways to measure energy expenditure.
 
@@
 
+DON'T GO THROUGH ALL OF IT -- THEY CAN GO BACK AND LEARN IT.
+
 \subsection{Salivary Glands}
@@
 Carbohydrates make up 40 to 50\% of the energy intake.
-Carbohydrates include starch (amylose and amylopectin), glycogen, simple sugars and fibre (i.e. cellulose).
+Carbohydrates include starch (amylose (no branch) and amylopectin (with branch)), glycogen, simple sugars and fibre (i.e. cellulose).
 Cellulose cannot be digested in humans, as we lack the enzyme for breakdown of cellulose.
@@
 Lactose intolerance causes bloating, flatulence and diarrhoea due to the fermentation.
-Individuals with this should avoid lactose consumption.
 
@@
 
-Before we move onto how proteins are digested, we need to know what kind of enzymes are involved.
 The enzymes present are: pepsin (stomach), trypsin, chymotrypsin and carboxypeptidase (3 from pancreas), and aminopeptidase and dipeptidase (2 from small intestine).
@@
 This is done by the bile acids/salts, which are made from cholesterol.
-Bile acids are synthesised in the liver and stored in the gall bladder as bile, which is secreted in response to cholecystokinin.
+Bile acids are synthesised in the liver and stored in the gall bladder as bile, which is secreted in response to cholecystokinin (duodenum).
 

### 2015-09-05 463assignment/463assignment_takei.tex
commit for merging

--- 2015-08-30 463assignment/463assignment_takei.tex
@@
 
-\section*{Abstract}
+\section*{\normalsize{Abstract}}
 
@@
 
-\subsection*{Olfactory Receptors}
+\subsection*{Intro to olfactory receptors}
 
@@
 OR genes also have an important implications in the evolution of different species.
-There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes are found mainly in their respective species (ref original paper or Nei and Niimura 2005).
-The study by Niimura and Nei (2005) showed that the diversity of the OR gene families in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
-Furthermore, they showed that amphibians have retained the diversity of the gene families as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial (Nei and niimura, 2005).
+There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes belong to their respective species (ref original paper or Nei and Niimura 2005).
+The study by Niimura and Nei (2005) showed that the diversity of the OR gene family in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
+Furthermore, they showed that amphibians have retained the diversity of the gene family as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial (Nei and niimura, 2005).
 
-Another study by Liman (2006) showed that pseudogenization and CNVs of the ORs coincided with the trichromatic vision in primates.
-Although this is an evidence of how trichromacy in primates became more important than the ORs for their survival, better understanding of the ligands of the ORs is crucial to support this theory.
+As both pseudogenization and CNV are evidence of genomic drift (ref Nei et al, 2007).
 
-%As both pseudogenization and CNV are evidence of genomic drift, it is of great interest for researchers to understand the evolutionary process of ORs by studying the OR CNV (ref Nei et al, 2007).
-
-\subsection*{Next Generation Sequencing and Copy Number Variations}
+\subsection*{Intro to copy number variations}
 
@@
 
-studies have been relatively small -- possibly due to Sequencing limitations.
-talk about 1000 genome project -- provides greater sample size to work with
-limitations of sequencing CNVs -- highly variable, hard to sequence/annotate??
-
 \subsection*{Intro to olfaction and sensory specific satiety}
-
-relate this to obesity and eating disorder
 
@@
 
-evolution and healthier eating
-
 \section*{Research Proposal}
 
-gather sample to identify the OR genes that could be important -- get samples with known smelling disorder/condition
-sequence the samples and identify the CNV difference and target those genes for further work
-check 1000 genome database for any correlation and/or evidence of evolution
+\subsection*{What am I going to do???}
 

### 2015-09-05 463assignment/463assignment_takei.tex

--- 2015-08-31 463assignment/463assignment_takei.tex
@@
 
-Due to the fact that there are so many OR genes in the human genome, and that these ORs work in a combinatorial manner, it has been difficult for the researchers to decipher the complete code of olfaction \citep{Hasin-Brumshtein2009, Nei2008}.
-In addition to this, ORs are highly copy number variable within and between species, adding greater complexity to the olfactory process which remains to be a challenge for the researchers to make a connection between OR CNV with the phenotype \citep{Nozawa2007}.
+Due to the fact that there are so many OR genes in the human genome, and that these ORs work in a combinatorial manner, it has been difficult for researchers to decipher the complete code of olfaction \citep{Hasin-Brumshtein2009, Nei2008}.
+In addition to this, ORs are highly copy number variable within and between species, adding greater complexity to the olfactory process which remains to be a challenge for the researchers to make a connection between OR CNV and phenotype \citep{Nozawa2007}.
 
-OR genes also have an important implications in the evolution of different species.
+OR genes also have important implications in the evolution of different species.
 Two major classes of OR genes have been identified: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes were found mainly in their respective species \citep{Glusman2000}.
-The study by \citet{Niimura2005} showed that the diversity of the OR gene families in fishes was greater than that of the mammalian OR gene families, even though the number of OR genes in a single family were less in fishes.
-Furthermore, they showed that amphibians have retained the diversity of the gene families as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial \citep{Niimura2005}.
+The study by \citet{Niimura2005} showed that the diversity of the OR gene families in fish was greater than that of the mammalian OR gene families, even though the number of OR genes in a single family were less in fish.
+Furthermore, they showed that amphibians have retained the same diversity of the gene families as the fish, while also having as large a repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process in the transition from being aquatic to terrestial \citep{Niimura2005}.
 

### 2015-09-05 bioc192/lect22.tex

--- 2015-09-01 bioc192/summary_tute/summarytute.tex
@@
-\documentclass[a4paper, 12pt]{report}
-
-\usepackage{chemmacros}
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc}
-
-\graphicspath{ {../images/topic1/} {../images/topic2/} {../images/topic3/}}
-
-\newcommand{\HRule}{\rule {\linewidth}{0.5mm}}
-
-\newcommand{\mychapter}[2]{
-    \setcounter{chapter}{#1}
-    \setcounter{section}{0}
-    \chapter*{#2}
-    \addcontentsline{toc}{chapter}{#2}
-}
-
-\begin{document}
-
-\begin{titlepage}
-
-\begin{center}
-
-\HRule
-\vspace{0.4cm}
-\Huge{\textbf{BIOC192 Summary Tutorial Notes 2015}}
-\HRule
-
-\vfill
-\LARGE{\textbf{Compiled by Riku Takei}}
-
-\end{center}
-
-\end{titlepage}
-
-%\tableofcontents
-%\newpage
-
-%\mychapter{0}{Introduction}
-
-%\input{../intro.tex}
-
-\mychapter{1}{Lecture 1}
-
-\section{Levels of Protein Structure}
-
-\subsection{Primary Structure}
-
-The Primary structure of a protein is the amino acid sequence of that particular protein.
-Note that the sequence of the amino acid (i.e.
-the primary structure) determines the 3D structure of the protein, since the primary sequence will have all the required amino acids in the right order for protein folding (via hydrophobic interaction).
-
-\subsection{Secondary Structure}
-
-Secondary structure is the local regular folds/structure that are stabilised by H-bonds between the backbone peptide groups of the polypeptide chain.
-Examples of secondary structures include the $\alpha$-helix and $\beta$-sheets.
-Supersecondary structures are related to the secondary structure, as it consists of combinations of secondary structures to make certain motifs (e.g. greek keys).
-
-\subsection{Supersecondary structure}
-
-Secondary structures such as $\alpha$-helix, $\beta$-sheets, and turns combine together to form a specific supersecondary structures, or `motifs'.
-
-\subsection{Tertiary Structure}
-
-Tertiary structure is the structure, or domain, formed by arranging secondary/supersecondary structures.
-These usually have a function once it has the correct 3D structure, but it may require other tertiary structures to act as a single protein.
-
-\subsection{Quaternary Structure}
-
-Quaternary structure is where multiple tertiary structure come together to form a single functional protein.
-Not all protein form a quaternary structure to function.
-
-\section{Amino Acids}
-
-%Amino acids are the building blocks of a protein.
-%Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).
-
-There are 20 amino acids in our cells that make up the proteins.
-All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$).
-Due to the variable side chains, all amino acids except glycine are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
-%The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.
-
-\section{Amino Acid Side Chains}
-
-%As mentioned earlier, all 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
-Each of the 20 amino acids have different chemical properties, depending on the side chain group it has.
-Cells take advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster (i.e.
-it gives the protein its functionality, as well as its unique structure).
-
-\section{Amino Acid Subgroups}
-
-%Due to the unique chemical properties of the amino acid side chains, they are grouped into subgroups.
-%There are four main groups: Non-polar, uncharged polar, negatively charged, and positively charged amino acids.
-
-\subsection{Non-polar}
-
-These amino acids have non-polar (or hydrophobic) side chain, meaning that the side chains cannot be charged under physiological pH.
-The amino acids in this group are: alanine, valine, leucine, isoleucine, glycine, cysteine, phenylalanine, tryptophan, methionine, and proline.
-
-\subsection{Uncharged Polar}
-
-These amino acids have side chains that can be charged, but uncharged under normal physiological pH.
-As you can see, the side chain contains --OH groups and/or --NH$_2$ groups, which are able to de/protonate under certain pH.
-The amino acids in this group are: serine, threonine, tyrosine, asparagine, and glutamine.
-
-\subsection{Negatively Charged}
-
-These amino acids have side chains that are negatively charged (i.e. deprotonated) side chains at physiological pH.
-The side chain contains a deprotonated --COOH group, giving it a negative charge to the side chain.
-The pK$_a$ of these amino acid side chains ranges from 4$\sim$5.
-The amino acids in this group are: aspartic acid (aspartate) and glutamic acid (glutamate).
-
-\subsection{Positively Charged}
-
-These amino acids have side chains that are positively charged (i.e. protonated) side chains at physiological pH.
-The side chain contains a protonated --NH$_2$ group, giving it a positive charge to the side chain.
-The pK$_a$ of these amino acid side chains ranges from 10$\sim$12, with an exception of histidine, which has a pK$_a$ of about 6 (and therefore only partially ionised).
-The amino acids in this group are: lysine, arginine, and histidine.
-
-\includegraphics[width=\textwidth]{20aa}
-
-\section{Charge of Amino Acids}
-
-Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
-The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
-When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.
-Depending on the pH of the solution, there will be more or less protonated molecules.
-The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).
-
-In general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
-
-\begin{center}
-    \textbf
-        {pH \textless{} pK$_a$, then the side chain is protonated \\
-        pH \textgreater{} pK$_a$, then the side chain is deprotonated}
-\end{center}
-
-Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
-Some amino acid side chains may be charged under physiological condition and contribute to the net charge of the amino acid, and therefore the protein.
-
-\begin{center}
-\includegraphics[width=0.8\textwidth]{aapka}
-\end{center}
-
-\section{Amino Acid Modifications}
-
-Some amino acids are modified after translation -- this is called the post-translational modification (PTM).
-Types of PTM includes:
-\begin{itemize}
-    \item Disulfide bond formation between cysteines
-    \item Phosphorylation
-    \item Glycosylation
-    \item Methylation
-    \item Adenylation
-    \item Iodination
-    \item Metal binding
-\end{itemize}
-
-\section{Proteins, Peptides, and Peptide Bonds}
-
-\subsection{Amide (peptide) Bonds}
-
-Amide (peptide) bond has a partial double bond characteristic due to the electron delocalisation/sharing between the oxygen atom and the nitrogen atom.
-This means that the amide bond is rigid, and also have a dipole nature (oxygen is slightly negative, nitrogen is slightly positive).
-
-The amide bonds are predominantly in \textit{trans}-configuration in the protein, although some can be in \textit{cis}-configuration, especially when the bond is preceded by proline.
-This is due to ring structure that the proline side chain forms with its amino group (causes steric clashes).
-
-By convention, the free amino group on the end of the peptide chain is called the N-terminus of the protein, and the free carboxyl group is known as the C-terminus.
-This gives a `directionality' to the peptide sequence, from the N-terminus to the C-terminus, and we draw primary sequence and count the amino acids according to this directionality.
-
-\mychapter{2}{Lecture 2}
-
-\section{Bond Rotation}
-
-Protein structure is created by the main chain and side chain bonds, and requires the main/side chains to be in a certain conformation.
-In order to do this, the amino acids must rotate and orientate correctly for the right interaction.
-
-Since the amide bonds have a partial double bond characteristic, it is rigid and forms a plane that does not allow rotation.
-However, rotation is allowed at the $\alpha$-carbon (C$_{\alpha}$) atom.
-Obviously, even though rotations are allowed at the C$_{\alpha}$ atom, not all angles are allowed due to steric hindrance.
-
-There are four types of bond angle rotations: $\omega, \psi, \phi$, and $\chi$.
-These angles have values that range from 0 to $\pm$180$^{\circ}$, and together, it gives the 3D structural information of a protein.
-
-\begin{description}
-\item [Phi angle] is defined as the bond angle rotation of the N--C$_{\alpha}$ bond.
-This bond rotation leads to O--O atom collision.
-
-\item [Psi angle] is defined as the bond angle rotation of the C--C$_{\alpha}$ bond.
-This bond rotation leads to the NH--NH collision.
-
-\item [Omega angle] is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
-This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).
-
-\item [Chi angle] is the bond angle rotation of the side chain bond (C$_{\alpha}$--R bond).
-
-\end{description}
-
-\vspace{0.3cm}
-
-\noindent
-\textbf{Remembering phi/psi angles are quite hard, but just remember that, relative to the C$_{\alpha}$ atom, phi angle is the rotation of the amide plane on the amino group side, whereas the psi angle is the rotation of the amide plane on the other side (i.e.
-the carboxyl side).}
-
-\section{Secondary Structure of a Protein}
-
-\subsection{$\alpha$-Helices}
-
-In $\alpha$-helix, the carbonyl group of the n$^{th}$ residue forms a H-bond to the amino group of the n+4$^{th}$ residue (e.g.
-the carbonyl group of the 1$^{st}$ residue H-bonds to the amino group of the 5$^{th}$ residue)
-There is about 3.6 residues per turn, which is about 5.4\AA per turn.
-The distance (or height) between 2 residues is about 1.5\AA.
-The side chains point outwards from the helix (this provides certain chemical properties on the sides of the helix, e.g. hydrophobic side). 
-$\alpha$-helices are dipole in nature, due to the amide bond polarity, all of which points at a single direction.
-
-There are some residues that disrupt the helical structure, and these are glycine and proline.
-Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
-Proline has a unique side chain that connects back to its amino group, forming an imino group, meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.
-
-\begin{center}
-\includegraphics[width=0.7\textwidth]{alphahelix}
-\end{center}
-
-\subsection{$\beta$-Sheets}
-
-$\beta$-sheets, or $\beta$-pleated sheets, are formed from H-bond formation between adjacent polypeptide chain (or strands), therefore $\beta$-sheets are formed from 2$\sim$10 strands (but itcan have more strands).
-Each strand has about 15 amino acids, and the $\beta$-sheets can be either parallel (both strands go from N to C terminal) or anti-parallel (strands go in an opposite direction).
-$\beta$-sheets are extended, but pleated (i.e. it's not completely flat), and the side chains point above and below.
-
-\subsection{Turns}
-
-Turns are sharp hairpin turns in a protein, and it has a high glycine and proline content.
-About 30\% of the residues in a protein is involved in turns.
-It is common for the residues to form H-bonds across the turn.
-There are many types of turns (\textgreater16), but type I and II are common.
-
-\subsection{Relationship between Ramachandron Plots and Secondary Structures}
-
-This plot shows all the possible phi/psi angles (i.e. the two amide planes adjacent to the C$_{\alpha}$) that are allowed in a protein with no steric interference.
-Any phi/psi angle pairs that cause steric hindrance are not plotted, so we get some `regions' of allowed angle pairs for any given protein.
-The side chains of the amino acids are (usually) staggered to minimise steric hindrance.
-
-Secondary structures have certain phi/psi angle pairs, and when you overlay these to the plot, it falls into the allowed regions of the plot at a distinct region.
-This is not surprising, because if the secondary structures fall in the forbidden region, how did the structure form in the first place?
-It can't -- the structure cannot form due to steric hindrance.
-
-This essentially shows that secondary structures have unique phi/psi angle pairs.
-
-\includegraphics[width=\textwidth]{ramachandron}
-
-\mychapter{3}{Lecture 3}
-
-\section{Protein Folding Pathway}
-
-Protein folding is directed predominantly by the hydrophobic forces from the non-polar residues.
-This is NOT a random process.
-Since the proteins are submerged in water, the non-polar residues try to stay away from water, and therefore aggregate together to form secondary structures.
-This initial folding process is referred to nucleation.
-The hydrophobic regions of the secondary structure comes together to form a larger structure, and stabilised by small non-covalent interactions.
-
-The process of protein folding is summarised below:
-\begin{enumerate}
-\item Short, small secondary structures are formed, driven by hydrophobic forces (nucleation)
-\item Nuclei come together to form larger domains
-\item Larger domains come together
-\item Small conformational changes/adjustments occur to give compact native structure
-\end{enumerate}
-
-\section{Stabilisation of Protein Folding}
-
-Proteins are stabilised by non-covalent interactions, which are weak individually, but strong when there are a lot of them.
-Covalent bonds such as disulfide bonds can stabilise the protein as well.
-
-%\includegraphics[width=\textwidth]{stabilisinginteraction}
-
-\section{Chaperones and Chaperonin}
-
-Some proteins require additional accessory proteins to assist them in folding the proteins properly, or to fold it faster.
-If the proteins are not folded properly, it can lead to serious diseases like Alzheimer's and prion.
-
-\mychapter{4}{Lecture 4}
-
-\section{Myoglobin}
-
-Myoglobin is the monomeric form of haemoglobin, and it resides in the tissue/muscles for O$_2$ storage.
-Myoglobin consists of a globin protein that consists of 8 $\alpha$-helices (numbered for clarity), with a haem group in the middle.
-Myoglobin is able to reversibly bind O$_2$, and have high affinity for O$_2$ as well.
-
-\section{Haem (Heme)}
-
-Haem consists of four pyrrole rings that are linked together via methane bridges, with the ferrous (Fe$^{2+}$) attached at the center of the molecule.
-This Fe$^{2+}$ is slightly below the plane of the porphyrin ring when O$_2$ is not bound, and is shifted into the plane when it binds (causes conformational change).
-
-The heme is attached to the myoglobin via the His F8 (histidine on the helix F at position 8), and His E7 is on the opposite side to influence the O$_2$ binding affinity.
-The His E7 prevents the strong binding of O$_2$ to the heme (right angle) by kinking the angle at which the O$_2$ binds to the heme.
-This allows weaker and therefore reversible binding of the O$_2$ molecule to the myoglobin.
-
-Note that cyanide binds to the Fe$^{3+}$ form, not the Fe$^{2+}$ form.
-Also note that cyanide's main inhibitory effect is at the cytochrome Fe$^{3+}$, and affects the electron transport system.
-
-\section{Haemoglobin}
-
-Haemoglobin is the tetrameric (4 subunit) form of myoglobin that consists of two $\alpha$ and two $\beta$ subunits, and it is a blood O$_2$ transporter.
-Slight conformational change occurs when the O$_2$ binds to the haemoglobin molecule, which will become important in explaining the cooperative effect of the subunits.
-
-\section{O$_2$ Binding Property}
-
-\subsection{Myoglobin}
-
-Myoglobin resides in muscles and tissues, and have high affinity for O$_2$.
-This is shown in the O$_2$ saturation curve.
-At low partial pressure of O$_2$ (pO$_2$), myoglobin is mostly saturated with O$_2$ (i.e. O$_2$ is bound to it).
-This means that myoglobin only releases the O$_2$ molecule when the pO$_2$ is very low.
-
-\subsection{Haemoglobin}
-
-In contrast, haemoglobin's O$_2$ binding affinity is not that high compared with myoglobin.
-As you can see from the O$_2$ saturation curve, it requires quite a high pO$_2$ for haemoglobin to be saturated with O$_2$.
-This makes sense as haemoglobin has to bind to O$_2$ at the lungs where pO$_2$ is relatively high, but also has to release the O$_2$ molecules at the tissue/muscle, where the pO$_2$ is low.
-This allows haemoglobin to be an effective O$_2$ carrier.
-
-One thing to note here from the curve is that the shape of the curve for haemoglobin is sigmoidal (S-shape).
-This indicates that haemoglobin is allosteric and cooperative, where the subunits assist one another to facilitate the binding/release of O$_2$ molecules.
-Of course, there are other factors that contribute to the allosteric binding/release of O$_2$ molecules.
-
-\begin{center}
-\includegraphics[width=0.4\textwidth]{saturationcurve}
-\end{center}
-
-\section{Sequential and Concerted Model of \-Allosterism}
-
-When O$_2$ binds to a subunit of haemoglobin, it causes a slight conformational change to the subunit, and this conformational change assists the binding of O$_2$ molecules to the other subunits by cooperativity/allosterism.
-
-There are two models, sequential and concerted models, of allosterism.
-
-\subsection{Sequential Model}
-
-Sequential model is when one O$_2$ molecule binding to a single subunit causes the conformational change in the subunit it binds, and also the subunits that are adjacent to it as well.
-This conformational change makes the other subunits easier to bind to O$_2$ molecules, which will also change conformation of itself and the others.
-
-\textbf{In sequential model, binding of one substrate changes the conformation of other subunits, which makes it easier for the other subunits to bind the substrate}
-
-\subsection{Concerted Model}
-
-In concerted model, all of the subunits are either in the T state, or the R state, and the states are at equilibrium (i.e. no `in between' state).
-The O$_2$ molecules can bind to either state at any time.
-When there are more substrate (i.e.
-O$_2$) is bound to haemoglobin, the greater the chance for the subunits to be in the R state, which means that the equilibrium is favoured towards the R state when there are more O$_2$ molecules bound to haemoglobin.
-
-\textbf{ In concerted model, the T and R states are at equilibrium and substrates can bind at any time to any conformational state.
-The binding of substrate causes the equilibrium to shift to the R state and the substrates are more likely to bind to the protein.}
-
-\mychapter{5}{Lecture 5}
-
-\section{Allosteric Effectors of Haemoglobin}
-
-There are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
-Since these molecules have effects on haemoglobin, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.
-
-\begin{center}
-\textbf{Allosteric effectors are molecules that bind to the allosteric protein at a place other than/different to the normal substrate binds to.}
-\end{center}
-
-\section{2,3-Bisphosphoglycerate (BPG)}
-
-2,3-Bisphosphoglycerate (BPG) is a small, negatively charged (-5 charge) molecule that binds to the deoxy-haemoglobin at the center of the tetramer.
-Binding of BPG causes the haemoglobin to be in T state, i.e.
-in the dexygenated state, and decreases the haemoglobin's binding affinity for O$_2$, meaning that the haemoglobin with BPG bound to it will require higher pO$_2$ to oxygenate it.
-Binding of BPG at the muscle/tissue site ensures that the O$_2$ molecules are released at the muscles/tissues.
-O$_2$ binding to the haemoglobin at the lungs causes the release of BPG molecule, allowing the haemoglobin to carry O$_2$ to the muscles/tissues.
-
-\section{H$^+$ and CO$_2$}
-
-Both H$^+$ and CO$_2$ binds to haemoglobin at a different site as the substrate (O$_2$) and affects the oxygen affinity of haemoglobin.
-Both molecules favour the T state of haemoglobin, and therefore favours the deoxygenated state (i.e. decreased O$_2$ affinity).
-Note that both of these molecules are related to the pH of the environment, and the blood buffer system is also involved in controlling the concentration of these molecules.
-
-\subsection{The Bohr Effect}
-
-Technically speaking, the Bohr effect is the effect of pH (i.e. [H$^+$]) and [CO$_2$] on the oxygen affinity of haemoglobin.
-As pH decreases (therefore [H$^+$] increases), the net charge of haemoglobin changes due to different amino acids getting protonated, which leads to slight conformational changes in the haemoglobin structure, and therefore decrease in haemoglobin's oxygen affinity.
-Similarly, as CO$_2$ concentration increases, CO$_2$ is more likely to bind to the haemoglobin and decrease the oxygen affinity of haemoglobin.
-
-\subsection{The Blood Bicarbonate Buffer System}
-
-In the blood, there is an equilibrium between bicarbonate ([HCO$_3^-$]) ion and CO$_2$.
-
-\begin{center}
-\ch{H$_2$O + CO$_2$ <=> HCO$_3^-$ + H$^+$}
-\end{center}
-
-As you can see, the [H$^+$] concentration and the [CO$_2$] concentration is related to one another due to this buffer system.
-This means that at the tissue/muscle where the relative concentration of CO$_2$ is high, it is going to be making the bicarbonate ion, together with H$^+$ ion, and therefore the pH is decreased.
-However, at the lungs where the concentration of CO$_2$ is low, there will be less H$^+$ ion present, and therefore have higher pH.
-
-Together with the Bohr effect, this means that the pH is high at the lungs and have low CO$_2$ concentration, and therefore haemoglobin will have high oxygen affinity, and thus get saturated with O$_2$ (oxygenated).
-In contrast, the pH will be lower at the tissues/muscles and have high CO$_2$ conference, and therefore haemoglobin will have low oxygen affinity, and thus releases O$_2$ at the tissue/muscle (deoxygenated).
-
-Also note that BPG is released at the lungs when O$_2$ binds to the haemoglobin, and binds at the tissue/muscle when O$_2$ is released and CO$_2$ and H$^+$ binds to it, contributing to the binding and release of O$_2$.(insert the saturation curve of +/- allosteric effectors)
-
-\mychapter{6}{Lecture 6}
-
-\section{Normal Variants}
-
-\subsection{Foetal Haemoglobin (HbF)}
-
-Foetal haemoglobin (HbF) is only present in the foetus blood.
-HbF is made up of two $\alpha$ subunits and two $\gamma$ subunits, compared to the adult haemoglobin (HbA) which is made up of two $\alpha$ and two $\beta$ subunits.
-Since the foetus has to obtain oxygen from the mother's blood via the placenta, HbF has a higher affinity for oxygen than the HbA.
-This is because HbF binds less strongly to BPG due to the Ser143 in the $\gamma$ subunit (His143 in $\beta$ subunit), thereby increasing the oxygen affinity.
-
-\section{Abnormal Variants}
-
-These variants are caused by genetic mutations that alter the properties of haemoglobin, and can lead to anaemia or other pathological symptoms.
-
-\subsection{Sickle Cell Haemoglobin (HbS)}
-
-This is caused by a point mutation of an amino acid substitution from glutamic acid to valine (E6V mutation), which is on the surface of the protein.
-The consequence of this mutation is the formation of deoxy-Hb aggregate/crystal in the red blood cell.
-Since valine is a non-polar amino acid, it can stick/bind to other haemoglobin at specific site (in a hydrophobic pocket).
-
-\subsection{Met-Haemoglobin (HbM)}
-
-Methaemoglobin (HbM) is caused by mutations that changes the oxidation state of the iron atom in the haem from Fe$^{2+}$ to Fe$^{3+}$.
-This means that the oxygen cannot bind (or have very low affinity for oxygen) to the haem iron, as it is in a different oxidation state.
-
-\subsection{Haemoglobin C (HbC)}
-
-Haemoglobin C (HbC) is common in West Africa, and is caused by a point mutation that causes a substitution mutation.
-The substitution is at the same place as HbS: glutamic acid to lysine at position 6 (E6K).
-As a consequence of this substitution, the charge is swapped from negatively charged glutamic acid to positively charged lysine.
-
-Malaria is caused by the infection of red blood cells by \textit{Plasmodium \\falciparum}.
-Infected red blood cells will contain malaria parasite proteins that produce knobs on the surface of the red blood cells and causes it to adhere to the capillaries, causing tissue damage due to reduced oxygen supply.
-However, individuals with homo or heterozygous HbC have less adherence, and thus have some resistance to malaria infection.
-
-\mychapter{7}{Lecture 7}
-
-\section{Post-Translational Modification (PTM)}
-
-Most proteins, if not all proteins, in eukaryotic cells are post-translationally modified.
-Post-translational modification increases the number of distinct forms of proteins without increasing the number of genes in the genome.
-This means that the level of complexity provided by the proteins are also increased as well.
-
-PTM can be categorised into two categories: the covalent addition of chemical group to amino acid side chain, or covalent hydrolysis of the peptide bond in a protein.
-
-\section{Phoshphorylation}
-
-Phosphorylation can occur on the hydroxyl (OH) group of the amino acid side chains.
-Amino acids that have OH groups are Ser, Thr, and Tyr.
-Phosphorylation of the side chains are done by an enzyme called kinase, which adds the large, negatively charged phosphate group to the side chain, and uses ATP during this process.
-To remove the phosphate group, an enzyme called phosphatase is used.
-
-The consequence of adding a phosphate group to these amino acid side chains is that it's going to alter the protein structure, due to the electrostatic attraction between the more negatively charged phosphate group with other postitively charged amino acid side chains, and changes the conformation of the protein and its activity/function.
-Quite often, phosphorylation state is used to `switch' the conformational state of the protein to enable a biological process in a cell (e.g. facilitating interaction with other proteins).
-
-\subsection{Insulin Receptor}
-
-Binding of insulin to the insulin receptor causes a conformational change in the receptor, and this exposes the tyrosine in the cytosol side of the membrane.
-The exposed tyrosine is then targeted by kinases, and get phosphorylated by it.
-Phosphorylated tyrosine can then cause responses (e.g. signal transduction, GLUT4 transporter expression).
-
-\subsection{Na$^+$/K$^+$ Pump}
-
-Phosphorylation state affects which molecule (Na$^+$ or K$^+$) to bind to the pump, and therefore be transported.
-This is also energy dependent process, as the pump is constantly being phosphorylated and dephosphorylated.
-
-\section{Gamma-carboxy Glutamic Acid (Gla)}
-
-Gamma-carboxy glutamic acid occurs when an additional carboxyl group is added to the glutamic acid.
-This PTM is done in some blood coagulation proteins, and it requires vitamin K as a co-factor.
-Ten to twelve Gla enables the formation of bidentate Ca$^{2+}$ binding site and allows blood coagulation proteins to interact with platelets, as a part of blood clot formation process.
-
-Vitamin K deficiency leads to bleeding disorder.
-
-\section{Hydroxylation}
-
-Hydroxylation mainly occurs on the 4' (or 3') position of the proline, but it can also occur at the 5' position of lysine.
-Hydroxylation provides greater possibilities for non-covalent interactions, so it allows for more H-bonding to occur.
-As a result, proteins are able to form higher order structure.
-
-Vitamin C is important in hydroxylation, as it is required for the reaction to occur.
-Vitamin C keeps the Fe$^{2+}$ in this oxidation form, which is essential for the hydroxylation process.
-Since collagen structure requires hydroxylation to keep its higher order structure, low vitamin C causes disruption in this structure.
-This disease is known as scurvy.
-
-\mychapter{8}{Lecture 8}
-
-\section{Introduction to Enzymes and Enzyme Engineering}
-
-Enzymes (proteins) are able to accelerate and carry out chemical reactions that are most likely impossible to do under physiological/biological conditions.
-Without these enzymes, we will not be able to synthesise chemicals required to survive.
-It is also interesting that the enzymes have all evolutionarily developed itself to make difficult chemical reactions to faster, but within similar timescale.
-
-Can we utilise the properties/characteristics of enzymes to make it do what we want it to do?
-This is the key idea behind the researches being done in the area of enzyme engineering.
-We can use enzyme engineering to make enzymes that are able to react with different substrates, or make it more efficient at doing the reaction.
-
-\mychapter{9}{Lecture 9}
-
-\section{Introduction to Biotechnology}
-
-For therapeutic and clinical use of proteins, the proteins must be made and extracted from a source.
-Biotechnology is the field of research to produce proteins in systems (sort of).
-
-\subsection{The `Old' Way of Obtaining Proteins}
-
-Therapeutic proteins were extracted from relevant sources (e.g. tissues, serum) and was used to treat the patients.
-The problems with this approach was the risk of identified and unidentified pathogen infections from the source (e.g. HIV).
-
-\subsection{Recombinant Protein System}
-
-This is the more recent method to produce proteins for therapeutic use.
-Plasmids that contain a required protein is expressed in a different organism/system, and the purified and collected.
-
-Recombinant insulin is produced this way, by expressing two separate parts of insulin in \textit{E. coli}.
-These parts are extracted and combined to form insulin.
-
-The advantage of this is that it is cheap, has high yield, and it is pathogen free.
-However, in a prokaryotic system, the cells are unable to post-translationally modify the protein, which may be essential for the function of some proteins.
-Another problem is that the proteins may not fold properly or only partially folded.
-
-\subsection{Eukaryotic System}
-
-This system solves the problem of PTM and protein misfolding, as the proteins will be expressed in a eukaryotic system.
-
-An example of a protein that uses a eukaryotic system is erythropoietin (EPO).
-EPO regulated haematopoiesis, and its expression is partially regulated by the blood oxygen level -- EPO acts so that it produce more red blood cells and haemoglobin.
-EPO is used therapeutically as well as a performance enhancing drug.
-EPO is produced in immortalised Chinese hamster ovary cells.
-
-Different EPOs have different glycosylation profile, which allows us to detect athletes who uses EPO.
-
-\subsection{Recombinant Antibodies}
-
-Recombinant antibodies are designed to control the effects of some proteins in your body (e.g. some cell receptors).
-Mouse AB can be made, but it is recognised as foreign material in humans and get rejected.
-This lead to the formation of chimeric antibodies, so there will be no immune response.
-
-\subsection{Pharming}
-
-The use of live animals to express/make recombinant proteins, for example, in milk.
-
-\mychapter{10}{Lecture 10}
-
-\section{Introduction to Enzymes}
-
-Enzymes are biological catalysts.
-Enzymes increases the rate of reaction, but does NOT change the equilibrium, nor the thermodynamics of a reaction.
-This means that, even though enzymes make the reactions to occur faster, it cannot shift the chemical reaction to favour one product over another.
-
-For a chemical reaction to occur, the substrates must be in close proximity of one another, and also have the right orientation.
-Enzymes catalyse reactions by doing just that: bind the substrate in the right orientation in close proximity so a chemical reaction can occur at a faster rate.
-
-\section{Active Site of an Enzyme}
-
-Active site is the 3-dimensional site of an enzyme where the reaction occurs.
-The active site has amino acid side chains projecting into it to help substrate binding.
-The shape of the active site determines the specificity of the enzyme, since enzymes are (usually) very specific.
-This means that enzymes only bind to certain substrate, and no other substrates.
-
-\subsection{Geometric Specificity and Stereospecificity}
-
-Geometric specificity is where the substrate must be made of same component/structure for it to be recognised by the enzyme.
-Stereospecificity is where the substrate has the required components,but don't have the right orientation, and therefore not recognised by the enzymes.
-
-\section{Enzyme-Substrate Interaction}
-
-\begin{center}
-\ch{ E + S <=> ES <=> [{T}] EP <=> E + P }
-\end{center}
-
-In an enzyme catalysed reaction, substrate binds to the active site and forms an enzyme-substrate complex (ES), which is stabilised by non-covalent interactions.
-As the reaction proceeds, transition state (T), which is neither a substrate nor the product, is formed, then proceeds further into enzyme-product complex (EP).
-This complex can then be dissociated to form the original enzyme ready for another reaction, and release the product.
-
-There are two models that describes how the substrate binds to the enzyme: the lock-and-key model, or the induced-fit model.
-
-\subsection{Lock-and-Key Model}
-
-The concept of the lock and key model is fairly simple -- the substrate must have the shape/structure that perfectly matches the enzyme active site.
-If the substrate doesn't then the enzyme cannot react with it and no reaction occurs.
-However, this is not really the case in nature as enzymes changes conformation when the substrate binds to it.
-
-\subsection{Induced-fit Model}
-
-The substrate and the enzyme active site does not match perfectly as it did for the lock and key model.
-When the substrate and the enzyme binds, it causes a conformational change in the enzyme, which allows the enzyme to properly bind to the substrate.
-
-Note that one way to increase the rate of reaction is by stabilising the transition state of the substrate.
-Enzymes can do this by preferentially binding to the substrate in a transition state conformation.
-This is quite often the case, as enzymes usually stabilises the transition state of the substrate during the catalysis.
-
-\section{Enzymes, Activation Energy, and \\Catalytic Mechanisms}
-
-Enzymes can decrease the activation energy of a chemical reaction by various catalytic mechanisms.
-There are 6 ways to do this:
-
-\begin{enumerate}
-\item Acid-base catalysis
-\item Covalent catalysis
-\item Metal ion catalysis
-\item Electrostatic catalysis
-\item Proximity and orientation effects
-\item Preferential binding of the transition state complex
-\end{enumerate}
-
-In this lecture, only the first three was discussed.
-
-\subsection{Acid-base Catalysis}
-
-This kind of catalysis involves the transfer of H$^+$ ions, for example by glutamic and aspartic acids (COOH group), or by lysine (NH$_3^+$ group).
-Note that these side chains require certain pH for it to be in the right ionisation state, and therefore to be able to transfer H$^+$ ions.
-In this sense, histidine is very useful as it has a pK$_a$ of about 6.5, which is close to the physiological pH.
-This means that with a slight change in pH, it can be either a H$^+$ donor or acceptor, and therefore able to take part in the acid-case catalysis reaction.
-
-\subsection{Covalent Catalysis}
-
-In a covalent catalysis, the enzyme forms a highly reactive, short-lived intermediate which is covalently attached to the enzyme.
-
-\begin{center}
-\ch{A-B + E -> A-E + B ->  A + B + E}
-\end{center}
-
-The enzyme requires an amino acid with a side chain that can act as a nucleophile (e.g. ser, cys)
-
-\subsection{Metal Ion Catalysis}
-
-About a third of known enzymes require metal ions for their catalytic activity.
-Metals can provide substrate orientation, binding energy, and/or sites for redox reactions.
-Metals act as a cofactor in these reactions.
-
-\section{Co-factors and Co-enzymes/Co-substrates}
-
-Some enzymes require other factors to catalyse chemical reactions.
-These factors can be metal ions or small organic molecules (co-enzymes).
-Co-enzymes, or co-substrates, are molecules that help with the reaction, and is changed after the reaction occurs.
-Co-substrates are usually derived from vitamin derivatives.
-
-\mychapter{11}{Lecture 11}
-
-\section{Measuring Enzyme Catalysis}
-
-\begin{center}
-\includegraphics[width=0.5\textwidth]{progresscurve}
-\end{center}
-
-Enzyme catalysis of a reaction can be measured from the progress curve of a reaction.
-Progress curve is the graph of how much product is made (or how much substrate is used up) over time.
-It is important to measure the initial reaction velocity of the reaction from this progress curve in order to calculate some other important measures of enzyme kinetics.
-
-This initial reaction velocity (V$_i$) is the slope of the tangential line of the initial part of the progress curve.
-Only the initial part is used, because as you progress through the reaction, some factors start to limit the reaction and becomes unreliable.
-
-\section{Effect of Enzyme Concentration on the Rate of Reaction}
-
-\begin{center}
-\includegraphics[width=0.5\textwidth]{vvse}
-\end{center}
-
-Consider the situation where there is an excess substrate (i.e. unlimited source of substrate).
-As the concentration of the enzyme increases, the rate of reaction increases.
-\textbf{This means that the velocity of the reaction (V) is proportional to the enzyme concentration, when there is excess substrate.}
-
-This is expected, since there are more enzymes to catalyse the reaction -- as more enzymes are introduced to the system, more enzymes can react with the substrate, and therefore produce more products.
-
-\section{Effect of Substrate Concentration on the Rate of Reaction}
-
-\begin{center}
-\includegraphics[width=0.4\textwidth]{vvss}
-\end{center}
-
-Consider the situation where the concentration of enzyme is fixed.
-As we increase the concentration of substrate, the rate of reaction will initially increase linearly, but as more and more substrates are introduced, the rate of reaction reaches a `maximum'.
-This is because all of the active site of the enzymes become occupied as the substrate concentration increases -- it reaches the `maximum capacity' of the enzyme, and the enzymes cannot increase the rate of reaction no matter how hard it tries.
-
-\begin{center}
-\includegraphics[width=0.4\textwidth]{zerofirstkinetics}
-\end{center}
-
-In kinetics term, the enzyme shows first-order kinetics initially (where the rate depends on the substrate concentration), but then shows zero-order kinetics (where the rate does not depend on/changes with substrate concentration).
-
-\subsection{V$_{max}$ and K$_M$}
-
-From this V vs S curve, we can identify two kinetic parameters: V$_{max}$ and K$_M$.
-V$_{max}$ is defined as the maximum velocity of the reaction when an enzyme is saturated with substrate.
-K$_M$ is defined as the \textbf{substrate concentration} at which the velocity of the reaction is half the maximum velocity ($V = \frac{V_{max}}{2}$).
-K$_M$ is also called the Michaelis constant.
-
-Note: do NOT get this mixed up with ``K$_M$ = $\frac{V_{max}}{2}$'' -- remember that K$_M$ is the \textbf{substrate concentration} when V = $\frac{V_{max}}{2}$.
-
-\mychapter{12}{Lecture 12}
-
-\section{The Michaelis-Menten Equation}
-
-Using V$_{max}$ and K$_M$, we can define the Michaelis-Menten equation:
-
-\begin{center}
-\large{V = $\frac{V_{max} [S]}{K_M + [S]}$}
-\end{center}
-
-Few assumptions must be made in order to derive this equation:
-\begin{center}
-\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
-\end{center}
-\begin{enumerate}
-\item ES formation and breakdown is in rapid equilibrium, i.e. the ES concentration does not change over time ({$\frac{d[ES]}{dt}$} = 0).
-\item Enzyme is saturated with substrate, therefore it is at V$_{max}$.
-\item Initial rate is used, meaning that the substrate concentration does not change significantly.
-\item Initial rate is used, meaning that there is no product present.
-\end{enumerate}
-
-Michaelis-Menten equation is a simple model of an enzyme catalysed reaction based on the assumptions stated above, and is applicable to most enzymes.
-
-\section{Lineweaver-Burk Plot}
-
-\begin{center}
-\includegraphics[width=0.8\textwidth]{lineweaverburk}
-\end{center}
-
-Since the V vs [S] curve is hyperbolic, you can only estimate the V$_{max}$ from it, and therefore you can only estimate the K$_M$ value.
-In order to measure the V$_{max}$ and K$_M$, you can take the double reciprocal of the V vs [S] curve (i.e. take the reciprocal of both V and S).
-
-\section{Significance of K$_M$}
-
-\begin{center}
-\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
-\end{center}
-
-K$_M$ characterises a single enzyme-substrate pair.
-K$_M$ is independent of the substrate concentration, but is dependent on the substrate, meaning that K$_M$ is different for different substrate.
-
-K$_M$ is defined as the ratio of the breakdown of the ES complex and the formation of the ES complex:
-
-\begin{center}
-\large{K$_M$ = $\frac{breakdown[ES]}{formation[ES]}$ = $\frac{k_{-1} + k_2}{k_1}$ = $\frac{k_{-1}}{k_1}$}
-\end{center}
-
-Usually, k$_2$ is much slower than k$_{-1}$, so we can ignore k$_2$.
-K$_M$ is also called the ES dissociation constant, and its unit is the same as the concentration.
-
-Low K$_M$ indicates that k$_{-1}$ is low and k$_1$ is high, and therefore it has a high affinity between the enzyme and the substrate, and binds to one another tightly, forming a strong ES complex.
-High K$_M$ indicates that k$_{-1}$ is high and k$_1$ is low, and therefore it has a low affinity between the enzyme and the substrate, and binds to one another weakly, forming a loose/weak ES complex.
-
-Therefore, you can use K$_M$ as an indicator of substrate preference of an enzyme.
-
-\section{V$_{max}$ and k$_{cat}$}
-
-\begin{center}
-\ch{E + S <=> [k_1][k_{-1}] ES -> [k_2][slow] E + P}
-\end{center}
-
-\begin{enumerate}
-\item k$_2$ is the rate-limiting step of the reaction
-\item The rate of the reaction is proportional to the total enzyme concentration ([E$_{total}$]).
-\end{enumerate}
-
-This gives us the definition of k$_{cat}$ and V$_{max}$, relative to one another:
-
-\begin{center}
-V$_{max}$ = k$_2$ + [E$_{total}$] \ch{->} k$_{cat}$ = $\frac{[E_{total}]}{V_{max}}$
-\end{center}
-
-k$_{cat}$ is the turnover number, or the catalytic rate constant of an enzyme.
-The unit is in s$^{-1}$.
-\textbf{The number of moles of substrate converted into product, per mole of enzyme per second.}
-
-\section{k$_{cat}$, K$_M$ and Catalytic Efiiciency}
-
-\begin{center}
-Catalytic Efficiency = $\frac{k_{cat}}{K_M}$
-\end{center}
-
-The ratio of k$_{cat}$ and K$_M$ is the measure of the enzyme efficiency -- how quickly a product is formed relative to how well the substrate binds.
-The catalytic efficiency will therefore have the highest value when k$_{cat}$ is high (high turnover), and low K$_M$ (high binding affinity).
-
-\mychapter{13}{Lecture 13}
-
-\section{Irreversible Inhibitors}
-
-Irreversible inhibitors are inhibitors that covalently binds to the enzyme, to a specific amino acid side chain.
-This means that the inhibitor permanently inactivates an enzyme, as it cannot be unbound.
-Examples of irreversible inhibitors include penicillin (targets glycopeptide transpeptidase in bacterial cell wall) and organophosphates (inhibitor of acetylcholinesterase).
-
-\section{Reversible Inhibitors}
-
-There are two types of reversible inhibitors: competitive and non-competitive inhibitors.
-Both of these inhibitors bind to the enzyme non-covalently, and can be `knocked off'.
-Some inhibitors can be knocked off more easily than the others, depending on how tight the inhibitor binds to the enzyme.
-
-\begin{center}
-\ch{ E + I <=> EI }\\
-
-\vspace{0.5cm}
-
-\ch{K$_i$ = $\frac{[E][I]}{[EI]}$}
-\end{center}
-
-K$_i$ is the dissociation or inhibition constant -- just like the K$_M$ with normal substrate.
-The lower the K$_i$, the stronger the inhibitor binds to the enzyme.
-K$_i$ \textless{} $10^{-9}$ mol L$^{-1}$ is strong binding.
-
-\subsection{Competitive Reversible Inhibitor}
-
-\begin{center}
-\includegraphics[width=0.6\textwidth]{compinh}
-\end{center}
-
-These inhibitors bind to the enzyme at the active site, and competes with the substrate molecule.
-This means that (usually) the inhibitor will have similar structure as the substrate, as it can bind to the active site.
-
-\begin{center}
-\textbf{Competitive inhibitors increases the K$_M$, but does not change the V$_{max}$ of the enzyme.}
-\end{center}
-
-This means that the apparent affinity of an enzyme for the substrate is reduced (as the K$_M$ is increased).
-The V$_{max}$ is not affected, as the binding of the inhibitor is NOT going to affect the catalytic activity of the enzyme -- the enzyme can still reach the V$_{max}$, it just needs more substrate to out-compete the inhibitor (hence the increase in K$_M$).
-(The binding of the inhibitor is not going to affect how fast the reaction proceeds, once the real substrate binds to the enzyme).
-
-\subsection{Non-competitive Reversible Inhibitor}
-
-\begin{center}
-\includegraphics[width=0.6\textwidth]{noncompinh}
-\end{center}
-
-These inhibitors bind to elsewhere to the active site of the enzymes and causes a conformational change, and therefore affects the efficiency of the catalysis.
-
-\begin{center}
-\textbf{Non-competitive inhibitors decreases the V$_{max}$, but does not change the K$_M$ of the enzyme.}
-\end{center}
-
-When the inhibitor binds to the enzyme, it does not occupy the active site where the substrate binds to.
-This means that the active site is free for the substrate to bind -- it's just that the reaction does not proceed as efficient as without the inhibitor bound to it.
-Hence the decreased V$_{max}$, but unchanged K$_M$.
-%(Another way to think about it -- V$_{max}$ is proportional to the total enzyme concentration and the k$_{cat}$, so if k$_{cat}$ decreases, then V$_{max}$ will also decrease).
-
-\section{Methods of Enzyme Regulation}
-
-\subsection{Allosteric Enzyme Regulation}
-
-Allosteric enzymes generally display co-operativity and usually multi-subunit.
-The enzyme's reaction curve shows a sigmoidal curve.
-Activity of allosteric enzymes can be regulated by binding molecules (effectors) away from the active site, which can activate or inhibit the enzyme.
-Regulatory effect is achieved by conformational changes caused by the effectors binding to the enzyme.
-Note that the inhibitors shift the curve to the right, whereas the activator will shift the curve to the left.
-
-\subsection{Covalent Modification of Enzymes}
-
-In this lecture, only phosphorylation was mentioned.
-Phosphorylation and de-phosphorylation is the most common, and is catalysed by kinases and phosphatases, respectively.
-Phosphorylation state of an enzyme determines whether the enzyme is `on' or `off'.
-However, phosphorylation does not necessarily mean that the enzyme is on -- it really does depend on the enzyme you're studying.
-
-\subsection{Activation of Enzymes by Cleavage}
-
-Some enzymes are synthesised as zymogens -- the inactive form of the enzyme.
-The enzyme is only active once the enzyme is cleaved at a certain amino acid residue.
-Many of the enzymes involved in the digestive processes and blood clotting proteins are made as zymogens.
-
-\section{Isoenzymes/Isozymes}
-
-Multi-subunit enzymes may exist in multiple forms in an organism -- these enzymes are called isoenzymes.
-Isoenzymes differ in physical, kinetic and immunological properties, but all of them catalyse the same reaction.
-Isoenzymes are formed from two different genes that make slightly different subunit for the enzyme.
-Isoenzyme is a form of transcriptional regulation of cellular activity, as some cells in your body has to catalyse the reaction differently to another part of your body.
-
-\mychapter{14}{Lecture 14}
-
-\section{Pharmacology and Toxicology}
-
-\section{Drugs}
-
-A drug is a chemical that changes the behaviour or function of an individual system, organ, tissue or invading organism.
-In other words, a chemical that changes the behaviour of the cell.
-
-\subsection{Targets of Drugs}
-
-Drugs can bind to a molecule and cause an effect on the cell -- the molecule it binds to will be its molecular target.
-Understanding the pathways of the disease or the targets of the drug leads to better drug design.
-
-\section{Agonists and Antagonists}
-
-Agonists are the drugs or molecules that binds to a receptor and can produce a full biological response.
-On the other hand, antagonists are drugs or molecules that binds to a receptor and prevents the response to occur (i.e. opposite of agonists).
-
-\section{Receptors}
-
-There are two types of receptors described in these lectures -- drug receptors and physiological receptors.
-
-\subsection{Drug Receptor}
-
-Drug receptor is ANYTHING that the drug binds to.
-This can be a proteins, lipids, DNA, RNA, or whatever as long as the drug binds to it and have an effect on the cell through the molecule.
-
-\subsection{Physiological Receptor}
-
-Physiological receptors are normal receptors that are present on the cell surface.
-Most of these are transmembrane, meaning that it spans the whole membrane.
-These receptors allow signalling to occur from the outside (extracellular) to the inside (intracellular) of the cell.
-They also regulate cell function by responding to stimuli such as hormones, neurotransmitters or growth hormones and produce a signal for the cell to do something.
-Most drugs activate or inhibit these kind of receptors.
-
-There are four types of physiological receptors: ligand-gated/voltage-linked ion channels, G-protein coupled receptors (GPCR), kinase-linked receptors, and Nuclear receptors.
-
-\section{Ligand-Gated Ion Channels}
-
-Ligand-gated ion channels are proteins present in the cell membrane.
-This receptor has a receptor and an ion channel domains, and is made up from 5 subunits.
-Binding of the ligand induces a conformational change in the protein, which opens up the channel and allow the ions to pass through.
-
-Specific receptors have specific activating molecules, and only certain ions can pass through.
-These receptors control the intracellular ion concentration which gives different charges to the cell.
-The response times for these receptors are within milliseconds, which is ideal for fast response (e.g. in nerves).
-
-Using nicotinic acetylcholine receptor as an example, normal substrate/agonist of this receptor is acetylcholine.
-Another agonist is nicotine, and an antagonist is tubocurarine.
-Tubocurarine is an irreversible inhibitor, so it binds to the receptor and won't release, hence you can eat the hunted animal.
-
-\section{Voltage-Gated Ion Channels}
-
-These receptors are made from four subunits, and opened in response to changes in voltage across the membrane.
-The mechanism itself is very similar to the ligand-gated channels.
-The drugs that can block these receptors literally blocks the ion channel and prevents the ions to move through the pores (e.g. tetrodotoxin).
-
-\mychapter{15}{Lecture 15}
-
-\section{Nuclear Receptor}
-
-Nuclear receptors, unlike the transmembrane voltage/ligand-gated channels, reside in the cytosol of the cell (i.e. within the cell).
-This is partly because the nuclear receptors act in the nucleus, where it interacts with the DNA for gene transcription.
-These nuclear receptors binds and interacts with the DNA at the response element for that receptor, which requires multiple regulation mechanisms.
-The binding of the nuclear receptor-ligand complex to the response element causes transcription/gene activation.
-Since the effect of the cellular response is observable only after the protein is made, the response time is quite slow -- takes hours to days for the response.
-
-\section{Mechanism of Nuclear Receptor Signalling}
-
-For a ligand to enter the cell, the ligand must be lipophilic/hydrophobic, otherwise it will not be able to enter the cell through the membrane.
-Normally, the receptor is bound to a heat shock protein (Hsp; a chaperone), that prevents the receptor from binding to the DNA or other receptors.
-When the ligand binds to the receptor, the Hsp is displaced/removed, and the receptor dimerises (i.e. binds to another receptor).
-The receptor cannot bind to the DNA unless it is dimerised.
-Once bound to the DNA, the dimer recruits/associates with many other proteins to activate the gene (e.g. RNA polymerase) and transcribes the gene for cellular response.
-
-\subsection{Regulation of Nuclear Receptor}
-
-There are multiple regulatory steps in this mechanism to prevent unwanted signal and cell response.
-Firstly, the receptor is located inside the membrane, meaning that only the lipophilic/hydrophobic ligands can access/activate the receptor.
-The Hsp that is bound to the receptor is also important, because without them, the receptor may be able to bind to the DNA cause unwanted gene activation.
-Dimerisation is another level of regulation -- it requires the receptor to be dimerised, or else it's not going to activate the gene.
-
-\section{Structure of Nuclear Receptor}
-
-\begin{center}
-\includegraphics[width=0.7\textwidth]{nuclearreceptor}
-\end{center}
-
-There are three parts to the receptor.
-The first part is the ligand binding site (steroid hormones).
-Second part is the DNA binding domain, which is the site where the receptor binds to the DNA.
-This part contains the zinc finger motif for DNA binding.
-The last part is the transcription regulatory domain, where the transcription factors and other machinery binds to for gene activation.
-
-\section{What Genes are Activated by Nuclear Receptor}
-
-The genes activated by the receptor depends on the tissue site the receptor resides in, what cell type it's in, and what kind of ligand/drug binds to the receptor.
-This means that the receptor can activate multiple different genes, and therefore have different cellular response depending on the ligand, tissue site, etc.
-(i.e.
-it's not a one-gene-to-one-receptor thing).
-For example, estrogen receptor have different cellular response depending on the tissue site.
-
-\mychapter{16}{Lecture 16}
-
-\section{G-protein Coupled Receptors (GPCR)}
-
-G-protein Coupled Receptors (GPCRs) are transmembrane receptors that are associated with G-proteins, as the name suggests.
-GPCRs acts/responds in the timescale of seconds -- slower than the ligand/voltage gated ion channels, but faster than the nuclear receptors.
-GPCRs are important as it is the largest group of the receptors, and therefore an easy target for drugs.
-Also, GPCRs are activated by diverse ligands.
-
-\section{Mechanism of GPCRs}
-
-The ligand binds to the transmembrane receptor on the cell surface.
-Before the ligand is attached to the receptor, the receptor is not associated with the G-protein.
-Once the ligand binds to the receptor, the conformation of the receptor changes, which allows the G-protein to interact with the receptor.
-Interaction with the receptor causes the G-protein to change conformation as well, which increase or decrease the activity of a particular enzyme.
-Depending on the activity of the enzyme, the amount of second messenger molecule produced by the enzyme is altered, and therefore causes some cellular response.
-
-\section{Dopamine Receptor}
-
-Dopamine receptor is an example of GPCR.
-When the ligand (dopamine) binds to the receptor, it causes a conformational change in the G-protein which decrease the enzyme activity and therefore decreases the amount of second messenger (cAMP) produced.
-
-\subsection{Parkinson's Disease}
-
-Parkinson's disease is caused by decreased level of dopamine, due to the damage in the neural cells that produce dopamine.
-Parkinson's can be treated by using levodopa -- a prodrug that is processed into dopamine.
-The problem with levodopa is that it is hard to balance/control the effect of levodopa, and leads to psychological effects (such as hallucination/psychosis) and motor complications.
-
-\subsection{Psychosis}
-
-Psychosis occurs when there is too much dopamine around, and over activates the dopamine receptor.
-In this case, antipsychotics (antagonists of dopamine receptor) is used to prevent the dopamine to have too much effect on the receptor.
-This means that the level of dopamine is stall high, but the effect is inhibited by using antagonists.
-Again, however, if you use too much of antipsychotics, you get a Parkinson's like syptoms, signifying the fact that it's very hard to balance the effects of these drugs.
-
-\mychapter{17}{Lecture 17}
-
-\section{Membrane Composition}
-
-Membrane lipids are made from amphipathic molecules, meaning that the molecules have both charged and uncharged parts.
-The hydrocarbon tails are hydrophobic and are 16$\sim$22 carbons long.
-In animals, the length of the hydrocarbon tails are usually even.
-These hydrocarbon tails can be either saturated (no double bonds) or unsaturated (with double bonds).
-
-In aqueous solution, the polar group of the molecules are exposed to the solution, whereas the hydrophobic part forms the inner part of the membrane.
-This forms a well-known lipid bilayer, which is energetically favourable for the molecules to form in aqueous environment.
-
-\section{Membrane Lipid Structure}
-
-Membrane lipids all have a glycerol backbone which two fatty acids and a phosphate group attaches to.
-As mentioned, the fatty acids can be of different length, with or without double bonds.
-The phosphate group provides the charge for the amphipathic molecule, and often have polar groups added to the phosphate.
-This provides the group extra charge which can form patches on the surface of the membrane.
-
-Note that there are other types of membrane lipids present in different tissue of the same organism, or maybe in tissues from another organism, depending on the function of the tissue in its organism.
-
-\section{Membrane Fluidity}
-
-Membrane fluidity is affected by temperature, the type of fatty acids present, and the presence of cholesterol.
-All of these factors have an effect on the hydrophobic, non-covalent interaction between the lipids.
-Reminder: non-covalent interactions are very weak on its own, but strong when there are a lot of them, and the strength of the interaction is also dependent on the distance/hoe close to each other.
-
-\subsection{Temperature}
-
-Temperature affects the membrane fluidity by disrupting the non-covalent interactions between the hydrocarbon tails.
-As the number of the carbon increases in the hydrocarbon tails increases, the melting point also increases, and therefore less fluid.
-This is because there are more carbons for hydrophobic interaction that strengthens the interaction.
-
-\subsection{Type of Fatty Acid}
-
-The type of fatty acids, or more like how saturated a fatty acid is, affects the membrane fluidity.
-When the hydrocarbon tails are unsaturated, that is, it has double bonds in it, it has a kink.
-This kink causes the hydrocarbon tails to be more separated and the packing of the membrane lipids less ordered.
-Since the kink is going to spread the hydrocarbon tails apart further apart, this increases the distance between the tails, and therefore weakens the hydrophobic interaction between them (therefore more fluid).
-
-\subsection{Presence of Cholesterol}
-
-Cholesterol has a rigid ring structure that can be used as a ``scaffold''.
-This allows the lipids to anchor themselves to it and form a more ordered structure, and therefore makes the membrane less fluid.
-
-\section{Membrane Proteins}
-
-The membrane is jam-packed with proteins.
-Proteins are also able to move in the membrane -- it is dynamic.
-There are many functions of membrane proteins -- cell to cell contact, surface recognition, cytoskeleton contact, enzymes, transporters, receptors and signalling.
-
-Proteins can be either integral (all or some part of the protein is embedded in the membrane), or peripheral (on the surface of the membrane).
-Peripheral proteins can be held in pace on the membrane surface by polar patches from, for example, the polar groups attached to the phosphate group, forming an electrostatic interaction.
-
-Note that the proteins have ``sides'' on the membrane, and are asymmetric.
-This means that not all proteins that are exposed to the extracellular surface are also exposed on the intracellular surface.
-For example, most (if not all) of the proteins that are glycosylated are exposed on the extracellular surface, and not on the intracellular surface.
-
-\section{How Are Proteins Attached to the Membrane?}
-
-Non-polar amino acids of the protein resides within the lipid bilayer -- it is unfavourable for hydrophobic amino acids to move out of the membrane, so the proteins is anchored to the membrane.
-Proteins can also have an anchoring domain that can be used to anchor the domain to the membrane.
-Lipidation, a type of post translational modification, adds fatty acid groups to the protein which can insert it into the membrane for anchorage.
-
-\mychapter{18}{Lecture 18}
-
-\section{Transport of Non-polar Molecules Through the Membrane}
-
-Transport of non-polar molecules through the membrane is dependent on both the hydrophobicity and the concentration of the molecule.
-Unlike polar molecules, non-polar molecules are relatively easier to diffuse through the membrane due to its hydrophobic nature of the molecule.
-
-\subsection{Permeability Coefficient and Oil-Water Partition Coefficient}
-
-The permeability coefficient of a molecule describes how well the molecule is able to pass through the membrane -- i.e. how permeable the molecule is.
-The smaller the coefficient is, the harder for the molecule to pass through the membrane, and therefore less permeable.
-
-The oil-water partition coefficient describes how hydrophobic a molecule is.
-This means that the higher the oil-water partition coefficient is for that molecule, the more hydrophobic the molecule is.
-
-When these two factors are graphed for different kinds of molecules, you can clearly see that the higher the oil-water partition coefficient (i.e.
-more hydrophobic the molecule is), the higher the permeability coefficient.
-This shows that the more hydrophobic a molecule is, the easier for that molecule to pass through the membrane.
-
-\subsection{Concentration Gradient vs. Flux}
-
-When the  flux  is plotted against the concentration gradient of the molecule across the membrane, it is clear that the flux is proportional to the concentration gradient.
-This means that the higher the concentration gradient across the membrane, the higher the flux of that molecule through the membrane (i.e. faster/more will go through).
-
-\section{Transport of Polar Molecules Through the Membrane}
-
-Unlike the non-polar molecules, the lipid bilayer represents an energy barrier for the transport of polar molecules across it.
-So, for the polar molecules to pass through the membrane, the energy barrier must be decreased.
-Proteins within the membrane make small holes in the membrane to allow polar molecules to pass through.
-
-\subsection{Channels}
-
-Channels, when opened, are open to both the intracellular and the extracellular space (i.e. open to both ends).
-These channels can be opened or closed spontaneously, and are ``gated''.
-Since the mechanism of the channels are fairly simple, the rate of transport of molecules across the membrane when opened can approach the rate of diffusion -- about 10$^7\sim$10$^8$ molecules per second.
-The structure of the holes of the channels provide some specificity of the molecules that can go through the membrane.
-
-\subsection{Transporters}
-
-Unlike the channels, transporters are only open to either the intracellular or the extracellular space.
-The binding of the molecule that is to be transported induces a conformational change in the transporter which opens the opposite end, allowing the release of the molecule to the other side.
-The release of the molecule causes another conformational change that reverts the transporter back into its original state and the process repeats.
-Since the process of transporting the molecule requires conformational change of the transporter itself, the rate of transport is much slower compared to the channels -- about 10$^2\sim$10$^3$ molecules per second.
-
-Another characteristic thing about transporters is that they are very much like an enzyme -- transporters are very selective of what they move through the membrane, and it also shows a Michaelis-Menten-like kinetics.
-For example, glucose and mannitol are two very similar structured molecules that have similarly low permeability coefficient without a transporter.
-However, when GLUT transporter is present, the permeability coefficient increases a lot for the glucose, but not for mannitol, showing that the transporters are selective in which molecule it transports.
-
-\begin{center}
-\includegraphics[width=1.0\textwidth]{kinetics}
-\end{center}
-
-The kinetics of the transporter is represented by the equation below, where J is the flux and [A] is the concentration of the molecule.
-
-\begin{center}
-\large{$J_A = \frac{J_{max}[A]}{K_M + [A]}$}
-\end{center}
-
-\section{Types of Transport Processes}
-
-\begin{description}
-\item [Non-mediated] Requires no protein for transport -- free diffusion.
-\item [Mediated] Requires a protein. Also called facilitated transport.
-\item [Passive] Down a concentration gradient and no energy is required for the transport.
-\item [Active] Up a concentration gradient and requires an input of energy. Energy can come from ATP hydrolysis or from the co-transport of another molecule moving down its concentration gradient.
-\item [Co-transport] Movement of two molecules by a transporter.
-\begin{description}
-\item [Symport] Both molecules move through the transporter in the same direction.
-\item [Antiport] The two molecules move in the opposite directions to one another.
-\end{description}
-\end{description}
-
-\vspace{0.5cm}
-
-\noindent
-\textbf{Note: for symport and antiport, both molecules may not be going down the concentration gradient -- one molecule may be going up the concentration gradient, whereas the other maybe going down the concentration gradient.}
-
-\mychapter{19}{Lecture 19}
-
-\section{Energy Metabolism and ATP}
-
-We require approximately 1kg of ATP per kg of body fat for the day.
-Obviously, we can't carry that much ATP every day, and so we have to make it as we need it.
-Anabolism is the use of energy to form/make macromolecules from building blocks (e.g. making proteins).
-Metabolism is the opposite of anabolism, where the macromolecules are broken down into smaller constituent/component/building blocks to be oxidised to produce energy, or to be used in anabolic processes.
-
-From the first law of thermodynamics, the total energy within a system is constant -- i.e. energy can neither be created or destroyed.
-This means that the energy must be converted from one form to another.
-
-\section{Energy Balance in the Body}
-
-\begin{center}
-E$_{intake}$ = E$_{expended}$ + E$_{stored}$\\
-\vspace{0.5cm}
-E$_{expended}$ = Basal metabolism + Activities
-\end{center}
-
-As you can see from the equation, all of the energy taken into the body must be either used up/expended, or stored.
-Clearly, to lose weight, you must either decrease your intake, or increase your energy expenditure.
-You can increase the energy expenditure by increasing your basal metabolism and/or the amount of exercise/activities.
-
-\subsection{Subtopic: Leptin Receptor}
-
-Leptin deficient mice eats too much.
-Leptin is a hormone that binds to the leptin receptor, which acts so that the host feels full, and therefore stops eating.
-So, if a mouse is leptim deficient, then the mouse will not feel full, and therefore continues eating.
-
-\section{Units of Energy}
-
-Joule (J) is the unit of energy, defined as the \textbf{energy required to push against 1N of force for 1m}.
-An older metric is the calories, where 1cal is the energy required to heat 1g of water from 14.5$^\circ$C to 15.5$^\circ$C.
-1cal is equivalent to 4.184J.
-Daily energy intakes and expenditure usually measured as MJ.
-
-\section{Measuring Energy Intake}
-
-\subsection{Bomb Calorimetry}
-
-You can measure the energy contained in the food by burning the food and measuring the change in temperature of the water.
-
-\begin{center}
-\ch{C6H12O6 + 6 O2 -> 6 CO2 + 6 H2O + heat}\\
-\ch{C16H32O2 + 23 O2 -> 16 CO2 + 16 H2O + heat}
-\end{center}
-
-Carbohydrates releases 2813kJ/mol (glucose), whereas fatty acids release 10024kJ/mol (palmitate).
-Since fatty acids are more reduced, it can produce more energy from oxidation.
-One thing to note is that not all energy in food is available for the body (e.g. cellulose and nitrogen, which are excreted out).
-Therefore, these factors must be taken into account for calculations.
-
-\section{Atwater Factor}
-
-Atwater factors are used to measure how much energy each constituents have.
-Fat has 38kJ/g, carbohydrates and proteins have 17kJ/g and ethanol has 29kJ/g.
-These atwater factors can be used to calculate how much energy is contained in the food.
-
-\section{Measuring Energy Expenditure}
-
-\subsection{Direct Calorimetry}
-
-This method relies on measuring heat output from individual in a whole body calorimeter.
-This is good for determining the basal metabolic rate (BMR) -- i.e. at rest.
-However, it is quite hard to measure the energy used in more physical exercises, due to space constraints and complexity of the calculation and machinery.
-
-\subsection{Indirect Calorimetry}
-
-This method measures the energy expenditure based on the oxygen consumption and carbon dioxide production, which is measured using a respirometer.
-Since certain amount of energy is associated with every litre of oxygen consumed (20.9kJ/L), you are able to calculate how much energy was used.
-This allows us to calculate the energy expenditure for a range of activities, and also allows the calculation of the respiration exchange ratio (RER).
-
-\subsection{Respiration Exchange Ratio}
-
-\begin{center}
-\large{RER = $\frac{CO_2 produced}{O_2 consumed}$}
-\end{center}
-
-RER tells us about what kind of fuels are used -- carbohydrates or fat.
-Note that the stoichiometry of the oxygen and carbon dioxide is different for carbohydrates and fat.
-Since the stoichiometry of oxygen to carbon dioxide is 1:1 in carbohydrates, RER = 1, but this not the case for fat, and therefore have RER = 0.7.
-
-\section{Basal Metabolism}
-
-Basal metabolism is the energy required for the maintenance of life -- general metabolism, muscle concentration, etc.
-This is usually defined as the energy expenditure at rest.
-BMR is affected by:
-\begin{itemize}
-\item gender
-\item age 
-\item body size
-\item body composition
-\item genetics
-\item hormonal status
-\item stress levels
-\item disease status
-\item certain drugs
-\end{itemize}
-
-BMR can be increased by: athletic training, fever, late pregnancy, drugs (e.g. caffeine) and hyperthyroidism.
-BMR can be decreased by: malnutrition, sleep, drugs (beta blockers) and hypothyroidism.
-
-\subsection{Hyper and Hypothyroidism}
-
-\begin{description}
-\item [Hyperthyroidism] Increased metabolic rate due to excessive production of thyroid hormone, which acts to increase the effect of insulin, etc. Symptoms include heat production, anxiety, etc.
-\item [Hypothyroidism] Decreased metabolic rate due to thyroid hormone deficiency (from iodine deficiency). Symptoms include coldness, depression, etc.
-\end{description}
-
-\mychapter{20}{Lecture 20}
-
-\section{Organs Involved in Digestion}
-
-DON'T GO THROUGH ALL OF IT -- THEY CAN GO BACK AND LEARN IT.
-
-\subsection{Salivary Glands}
-
-Produces saliva -- neutral pH and contains mucous and amylase that starts the digestion of carbohydrates.
-
-\subsection{Stomach}
-
-For food storage (controlled release into the duodenum) and mixing of the food with gastric juices.
-Stomach release highly acidic acid (0.1M HCl) that denature the proteins.
-It also secretes pepsinogen (zymogen) for protein digestion, and also mucous layer for protection.
-
-The parietal cells produce the acid (HCl), and the chief cells produce the pepsinogen, which is activated by the acidic environment.
-
-\subsection{Pancreas}
-
-Pancreas is slightly alkaline pH 7, and secretes most digestive enzymes, including amylase, lipase and many proteases.
-
-\subsection{Liver}
-
-Synthesis of bile salts/acids (stored in gall bladder), which is important for fat digestion.
-
-\subsection{Small Intestine}
-
-Where the final phase of digestion and absorption occurs.
-
-\section{Digestion}
-
-There 2 main phases involved in digestion:
-\begin{enumerate}
-\item Hydrolysis of bonds connecting the monomer units in food macromolecules.
-\item Absorption of products from the gut into the body.
-\end{enumerate}
-
-In the case of carbohydrates, proteins and TAGs, the bond to cleave would be the glycosidic bonds, peptide bonds and ester bonds, respectively.
-
-\section{Carbohydrate Digestion}
-
-Carbohydrates make up 40 to 50\% of the energy intake.
-Carbohydrates include starch (amylose (no branch) and amylopectin (with branch)), glycogen, simple sugars and fibre (i.e. cellulose).
-Cellulose cannot be digested in humans, as we lack the enzyme for breakdown of cellulose.
-
-\subsection{Enzymes Involved in Carbohydrate Digestion}
-
-The digestive process begins in the mouth where the salivary amylase starts breaking down the starch molecules into smaller sugars.
-When it enters the small intestine, the pancreatic amylase (secreted from the pancreas) further breaks down the starch molecules into maltose and isomaltose.
-The intestinal epithelial cells secretes maltase, isomaltase, lactase and sucrase, which breaks down the disaccharide molecules into monomers (glucose, fructose and galactose).
-These monosaccharides are then absorbed into the body via the intestinal villi.
-
-\begin{reactions*}
-Maltose/Isomaltose &->[Maltase][Isomaltase] 2 Glucose\\
-Sucrose &->[Sucrase] Fructose + Glucose\\
-Lactose &->[Lactase] Galactose + Glucose
-\end{reactions*}
-
-\subsection{Lactose Intolerance}
-
-This is caused by lactase enzyme deficiency.
-Since lactose is not digested, it accumulates inside the intestine where the gut microbes can use it for fermentation.
-Lactose intolerance causes bloating, flatulence and diarrhoea due to the fermentation.
-
-\section{Carbohydrate Absorption}
-
-Monosaccharides are absorbed into the body via the villi and microvilli (brush border) of the intestinal wall.
-The villi and microvilli provides a large surface area for absorption.
-
-\subsection{Glucose Transport}
-
-Sugars are water soluble, and therefore require a transporter to pass through the cell membrane into the cell.
-SGLT1 protein is present in the intestinal side of the cell membrane and transports glucose into the intestinal cell together with the Na ion (symport).
-Movement of Na ion from the intestine into the cell is down the concentration gradient, which provides the energy for the glucose transport up the gradient into the cell.
-
-On the other side of the cell membrane, GLUT2 protein is present.
-Since the concentration of glucose in the cell is higher than the blood, the glucose can move down the concentration gradient through the GLUT2 transporter.
-Note that there is Na/K pump on this side of the membrane to maintain the Na concentration gradient.
-
-Once transported into the blood, other tissues can uptake glucose via GLUT3 (brain) and GLUT4 (muscle and adipose).
-
-\subsection{Coeliac Disease}
-
-Disease of the small intestine, where the body reacts against the wheat protein gluten.
-This causes the villi to be flattened and nutrients are not absorbed properly.
-
-\mychapter{21}{Lecture 21}
-
-\section{Protein Digestion and Absorption}
-
-Proteins can be used for energy, but it is more for regenerating the depleted amino acids.
-It is also important for supplying the body with essential amino acids.
-Proteins are the major source of nitrogen for purines, pyrimidines and haem.
-The carbon skeleton of the amino acids can be used as fuel/energy production, where the amino group (i.e. nitrogen) is converted into urea and excreted.
-Decrease/malnutrition of certain amino acid can lead to Kwashiokhor.
-
-\section{Essential Amino Acids}
-
-We cannot synthesise all 20 amino acids -- require 8 essential amino acids from the diet.
-These amino acids are: L, K, T, Y, I, M, F and V.
-
-\section{Regulation of Protein Digestion}
-
-Obviously, we don't want to ``eat ourselves'', so the process is highly regulated by polypeptide hormones.
-These hormones are: gastrin (stomach), secretin (duodenum), and cholecystokinin (duodenum).
-
-\section{Enzymes Involved in Protein Digestion}
-
-The enzymes present are: pepsin (stomach), trypsin, chymotrypsin and carboxypeptidase (3 from pancreas), and aminopeptidase and dipeptidase (2 from small intestine).
-All of these enzymes act in the small intestine, except for pepsin.
-Note that the protease specificity is determined by the adjacent amino acid side chains of the peptide chain it is breaking.
-For example, pepsin, trypsin and chymotrypsin cleaves the polypeptide after aromatic, positively charged and aromatic side chains, respectively.
-
-\section{Protein Digestion}
-
-Digestion of protein involves the hydrolyses of specific peptide bonds, performed by several different proteases.
-All of these proteases are secreted as inactive zymogens to prevent them from digesting our own cells, and they are all activated by cleavage of peptides from their structure.
-
-There are two stages of protein digestion:
-\begin{enumerate}
-\item Endopeptidases attack the peptide bonds within the peptide chain. Pepsin, trypsin and chymotrypsin are all endopeptidases.
-\item Exopeptidases attack the peptide bonds at the end of the peptide chain (i.e. from C-term or N-term -- hence amino or carboxypeptidase).
-\end{enumerate}
-
-\subsection{Pepsinogen Activation}
-
-Pepsinogen (due to its structure) have its catalytic active site inhibited.
-H$^+$ (low pH) changes the conformation of the pepsinogen and allows it to auto-cleave the peptide bond and activate itself.
-The activated pepsin can then activate other pepsinogens.
-Note that this process is all triggered by the ingestion of food, and the hormones associated with it.
-
-\subsection{Small Intestine Peptidase Activation}
-
-All 3 peptidases (trypsin, chymotrypsin and carboxypeptidase) are secreted in the inactive zymogen forms -- trypsinogen, chymotrypsinogen and procarboxypeptidases.
-First, trypsinogen is activated by the membrane bound enterokinase that adds the phosphate group to it.
-This causes a conformational change in the trypsinogen structure which allows it to auto-cleave and auto-activate itself.
-The activated trypsin activates the chymotrypsinogen and procarboxypeptidase.
-
-As a result of the activation of all of these proteases (including pepsin), dietary proteins are digested into single amino acids, dipeptides and tripeptides.
-Di and tripeptidases then digest the di and tripeptides into single amino acid.
-
-\section{Protein Absorption}
-
-\subsection{Amino Acid Transport}
-
-There are 5 main amino acids transporters to transport the amino acids across the luminal membrane.
-These are: Neutral, basic, acidic, other (taurine), and Glycine, proline and hydroxyproline.
-All of these amino acids are transported in a similar way to the glucose transport -- co-transport with Na.
-
-\subsection{Peptide Absorption}
-
-Very little absorption occurs for peptides that are longer than 4 amino acids.
-Di and tripeptides are absorbed in the small intestine via co-transport with H$^+$ ions (PepT1).
-These di/tripeptides are then digested in the cell and exported out into the blood.
-
-\subsection{Absorption of Intact Proteins}
-
-This only occurs in few circumstances -- e.g. in newborn for immunoglobulin absorption.
-
-\section{Diseases of Inappropriate Enzyme Activation}
-
-This is associated with inappropriate enzyme action (zymogen activation).
-This can lead to the breakdown of mucosal layer and proceed to peptic ulcers.
-
-Pancreatitis is where the pH is altered and degrades/inflame the pancreas.
-It can also release the enzymes into the blood.
-
-\section{Diseases of Malabsorption}
-
-Cystic fibrosis patients have abnormally thick mucous secretions due to mutation in chloride transporter.
-The thick mucous can block the pancreatic ducts, and therefore no protease activity.
-
-Coeliac disease have been mentioned in the previous lecture.
-
-\section{Digestion of Nucleic Acid}
-
-DNA and RNA are subject to acid hydrolysis in the stomach.
-Intestinal nucleases also hydrolyse the phophodiester bonds between the nucleotides.
-The individual nucleotides are absorbed via nucleotide transporters (Na co-transport).
-
-It is most unlikely that the DNA/RNA gets incorporated into our body, but it is a possibility as there are evidence of DNA being intact in the gut.
-
-\mychapter{22}{Lecture 22}
-
 \section{Lipid Digestion and Absorption}
@@
 This is done by the bile acids/salts, which are made from cholesterol.
-Bile acids are synthesised in the liver and stored in the gall bladder as bile, which is secreted in response to cholecystokinin (duodenum).
+Bile acids are synthesised in the liver and stored in the gall bladder as bile, which is secreted in response to cholecystokinin.
 
@@
 FH is treated with statins (drug that inhibits one of the enzymes involved in cholesterol synthesis).
-
-\end{document}

### 2015-09-06 bioc192/lect23.tex
finished lecture 23, onto lecture 24

\section{Vitamins}

\subsection{Fat Soluble Vitamins}

\subsubsection{Vitamin A}

Vitamin A is required for vision, integrity of epithelial cell, embryonic development, and  maintenance of immune system.
Xerophthalmia (``dry-eyes'') is a vitamin A deficiency condition that affects your vision/eyes -- causes ulceration and eventually scarring.
Vitamin A (or retinol) is important for your vision, as it is the substrate that binds to the GPCR that sends visual signals to the brain.

Preformed vitamin A (retinol) can be obtained from meat, or it can be converted from provitamin A (carotenoid) into retinol.
$\beta$-carotene (carotenoid) can be cleaved by dioxygenase into two vitamin A.
The activity of dioxygenase varies between species, due to the difference in the diet and therefore the amount of vitamin A obtained.
Herbivores only get carotenoids so they have high dioxygenase activity, whereas cats have no dioxygenase activity, since all the vitamin A is from meat.
A twelfth of carotenoids from food are converted into vitamin A.

Overdosage of vitamin A can lead to spontaneous abortion and fetal abnormalities, and therefore women should not take retinoid therapy during pregnancy.
The maximum dosage per day as retinol is 3000\si\micro g, but there are no upper level for carotenoids (as it gets excreted).

\subsubsection{Vitamin D}

Obtained/synthesised mainly from sunlight (about 20000IU).
It can also be consumed from your diet (vitamin D$_{2/3}$), although half of dietary vitamin D is lost when fried.
In NZ, you require about 5 to 15 \si\micro g per day, where 1IU = 0.025\si\micro g.
Vitamin D intake decreases during winter and it is also dependent on the ethnicity -- darker skin colour makes it more difficult to produce vitamin D.

Vitamin D deficiency can lead to rickets in children.
People who are at risk are people who have low or limited exposure of sun, have dark skin, and/or people who have fat malabsorption.
Overdosage of vitamin D can lead to hypercalcaemia, but there is no set concentration that is known to have an adverse effect.
However, the ``safe'' concentration is about 100ng/mL (250nmol/L).

In NZ, it is a recommended to expose infants to sunlight for at least 5 to 20 minutes (light and dark skins, respectively) before 11am or after 4pm during spring/summer time (Oct--Mar), unless you have sunprotection in between 11am to 4pm.
During winter, infants should spend time outside.

\subsubsection{Vitamin E}

Vitamin E, or tocopherol, can be obtained mainly from plant oil and in fatty tissues of animals.
Vitamin E is important for the membrane integrity by preventing oxidation of unsaturated fatty acids.
Vitamin E Deficiency is rare, and the RDI is 7 to 10 mg per day (women and men, respectively), and the upper limit is 300mg per day.

\subsubsection{Vitamin K}

Vitamin K is important for blood coagulation, and is offered at birth.
Vitamin K deficient bleeding (VKDB) disorder leads to spontaneous bleeding, and is uncommon.
Adequate intake is about 60 to 70\si\micro g per day (women and men, respectively).

\subsection{Water Soluble Vitamins}

\subsubsection{Vitamin B1}

Vitamin B1, or thiamin, is a co-enzyme for the TPP, which is involved in energy metabolism.
25 to 30mg are stored/used in your body, and excess thiamin is excreted out (therefore no overdosage).
Thiamin deficiency leads to Beri-Beri.

\subsubsection{Vitamin B2}

Vitamin B2, also known as riboflavin, is not linked to any characteristic disease and is therefore not fatal.
1.1 to 1.3mg (women and men, respectively) are the recommended intake.

\subsubsection{Vitamin B3}

Vitamin B3, or niacin, can be made from tryptophan, and therefore not technically a vitamin.
Niacin deficiency leads to pallagra and the 4Ds (diarrhoea, dermatitis, dementia and death).

\subsubsection{Vitamin B4}

Vitamin B4, or pantothenic acid, deficiency is rare, but it can cause fatigue, GI and neurological disturbances.

\subsubsection{Vitamin B6}

Vitamin B6 (pyridoxine) is part of the PLP and PMP co-enzymes.
Deficiency is rare, as it occurs in association with other vitamin B group.
RDI = 1.1mg per day.

\subsubsection{Vitamin B7}

Vitamin B7, or biotin, deficiency is rare and leads to non-specific symptoms.
Avidin (found in raw egg white) is a molecule that binds to biotin and prevents absorption.

\subsubsection{Vitamin B9}

Vitamin B9 (folate) is involved in the one carbon metabolism.
Deficiency is rare in developed countries.
Deficiency leads to megaloblastic anaemia (immature RBC and not enough of them) and neural tube defect in infants.

To protect infants from NTDs, supplements can be used to increase the folate in your body.
This will allow correct dose at the correct time, have max-protection and will have no risk to non-target population.
However, half of all pregnancies are unplanned, and it may be hard to consume the supplement at the right time.

Fortification of the food is another option where people can voluntarily consume products with folate.
Food manufacturer can fortify their food (mandatory fortification) so that their products will have low-level of folate in it (prevent excess intake in off-target groups).

\subsubsection{Vitamin B12}

Vitamin B12 (cobalamin) functions with folate in the one carbon metabolism, and also amino acid and odd chain fatty acid catabolism.
Dietary source is from animal products and deficiency leads to megaloblastic anaemia and nerve damage.

\subsubsection{Vitamin C}

Vitamin C is an antioxidant that is important for hydroxylation.
Deficiency leads to scurvy.

### 2015-09-07 bioc192/lect24.tex

\section{Minerals}

\subsection{Selenium}

There is about a mg to \si\micro g of selenium per kg of body weight.
Selenium functions as a cofactor for enzymes such as glutathione peroxidase and iodothyronine deiodinase.

Selenium deficiency in animals lead to white muscle disease in sheep, where the sheep is still born or die within a few days, and shows unsteady gat and arched back.
In humans, selenium deficiency can cause the Keshan disease, and is also associated with cancer.

NZ soil has selenium deficiency, and can be problematic (use fertiliser, etc.).
50 to 60\si\micro g of selenium per day is required, and the RDI is about 60 to 70\si\micro g per day.
The upper limit is about 400\si\micro g per day.

\subsection{Iodine}

Iodine deficiency can lead to goitre, and iodine cannot be stored in the body (and therefore excreted).
Over 100\si\micro g per L is adequate.
Less than this can lead to hypothyroidism, goitre, impaired mental function and eventually cretinism and abortions/still births.

\subsection{Iron}

Most common deficiency around the world.
Iron is important for Fe$^{2+}$/Fe$^{3+}$ states, oxygen transport, and assists many enzymes in their reactions.

RDI = 8 to 18mg per day (for men and women, respectively).

Factors that influence the iron intake are supply (amount of iron consumed, haem vs non-haem, and promoters/inhibitors of iron absorption) and requirements for iron (increased absorption is iron status is low, menstruation, growth, pregnancy, infections, etc.).

\begin{center}
\begin{tabular}{c | c | c}
& \% intake & \% absorbed \\
\hline
Haem & 10-15 & 25-35 \\
Non-Haem & 85-90 & 2-10\\
\end{tabular}
\end{center}

From this table, you can see that the haem iron is more available and better absorbed than the non-haem irons.

Iron absorption promoters include vitamin C (non-haem) and eating vegetables with meat (non-haem).
Inhibitors include phytates, compounds in coffee/tea, EDTA (preservative) , calcium, zinc and manganese.

\subsection{Zinc}

Zinc is required as a co-factor for over 300 enzymes.
RDI = 8 to 14mg per day (women and men, respectively).
Zinc is low in vegetables, and therefore deficiency is often seen in vegetarians.

Mild zinc deficiency leads to stunted growth in children, decreased taste sensation, and impaired immune functions.
Severe deficiency leads to dwarfism, delayed sexual maturation, and hypopigmented hair.
Causes of zinc deficiency is from diet that is low in meat, staple based on beans, unleavened bred, other wholegrain foods that are high in fibre and phytate.

### 2015-09-07 bioc192/lect25.tex
edited lect 25

\section{Metabolic Pathways and ATP Synthesis}

This part of the lecture is about how the energy from food is released in metabolic pathways and how they are used to make ATP.
As a reminder:
\begin{description}
\item[Anabolism] is where you use energy from ATP hydrolysis to make stuff and keep you alive.
\item[Catabolism] is where you break down food molecules and use that energy to regenerate the used up ATP.
\end{description}

\subsection{Conservation of Energy}

Remember that you can't make or destroy energy -- energy must be converted from one form to another.
If $\Delta$G is less than 0, the reaction is favourable.
If $\Delta$G is equal to 0, the reaction is at equilibrium.
If $\Delta$G is greater than 0, the reaction is unfavourable (but favourable if it goes in the opposite direction).

\subsection{Energy Coupling}

In cells, we need some unfavourable reactions to occur in the pathway, as it could be essential for that pathway.
To do this, you can couple two reactions together so that the total $\Delta$G is less than 0 (and therefore favourable).

\begin{reactions*}
A &-> B\\
C &-> D\\
A + C &-> B + D\\
\end{reactions*}

Enzymes often couple these reactions to drive necessary, but unfavourable reactions.

\begin{center}
\ch{ATP + H2O <=> ADP + Pi + H+} ,  \bioGibbs[exponent=0']{30.5}
\end{center}

As you can see, ATP hydrolysis (and therefore release of energy from the hydrolysis) is favourable, and allows unfavourable reactions to occur if coupled together.
However, this also means that it requires that much energy to regenerate ATP.

\subsubsection{Coupling ATP Hydrolysis with Hexokinse}

\begin{center}
\ch{Glucose + Pi ->[Hexokinase] Glucose-6-phosphate + H2O}
\end{center}

### 2015-09-09 bioc192/lect25.tex
finished lecture 25, onto lecture 26

@@
 \begin{center}
-\ch{Glucose + Pi ->[Hexokinase] Glucose-6-phosphate + H2O}
+    \ch{Glucose + Pi ->[Hexokinase] Glucose{-} 6 {-}phosphate + H2O}
 \end{center}
+
+The reaction above has a \bioGibbs[exponent=0']{+14}, which means that the reaction is unfavourable.
+However, this reaction can occur by coupling this reaction with ATP hydrolysis to make the overall \bioGibbs[exponent=0']{-16}.
+
+\subsection{Oxidising Fuels for Energy}
+
+Both carbohydrates and fats release energy, but we don't want all of the energy to be released/burnt at once.
+Therefore, our body requires a stepwise oxidation of carbohydrates and fats to capture the energy.
+
+In your body, there are 2 main kinds of reactions involved in energy metabolism: redox reactions and phosphorylation reactions.
+
+\subsubsection{Redox Reactions}
+
+Redox reactions are where the fuel molecules get oxidised, and the energy from this oxidation is captured by the co-enzymes which gets reduced from this reaction.
+Eventually, these reduced co-enzymes are used to make ATP.
+
+\subsubsection{Phosphorylation Reactions}
+
+Phosphorylation reactions are for making/generating ATP by directly or indirectly phosphorylating the ADP.
+There are two kinds of phosphorylation reactions:
+\begin{description}
+    \item[Substrate level phosphorylation] This is where the energy to phosphorylate the ATP comes directly from the reactions that break down the molecules.
+        Substrate level phosphorylation occurs in glycolysis and citric acid cycle.
+    \item [Oxidative phosphorylation] This is where the energy to phosphorylate the ATP comes indirectly from the breakdown of the molecules, via co-enzymes.
+        In the redox reactions, the co-enzymes get reduced, and the energy from these reduced co-enzymes can be fed into oxidative phosphorylation.
+\end{description}
+
+\subsection{Reducing Equivalents}
+
+\begin{center}
+    \ch{H = H+ + e-}
+\end{center}
+Biological reactions often involve the transfer of hydrogens (protons + electrons), instead of just electrons in purely chemical reactions.
+The co-enzymes NAD and FAD interact with various enzymes to accept/donate H$^+$/e$^-$.
+Enzymes that catalyse these reactions are often called dehydrogenase, as it removes hydrogen.
+
+\subsection{Co-enzymes}
+
+\subsubsection{NAD}
+
+NAD (nicotinamide adenine dinucleotide) is made up of niacin (vitamin B3).
+NAD accepts hydrogens and electrons in metabolic pathways such as glycolysis, fatty acid oxidation, and citric acid cycle.
+
+\begin{center}
+\ch{NAD+ + 2 H+ + 2 e- <=> NADH + H+}\\
+\vspace{0.5cm}
+\setatomsep{2em}
+NAD$^+$ + \chemfig{R-C(-[2]H)(-[6]OH)-R'}
+\ch{<=>} NADH + \chemfig{R-C(=[6]O)-R'}
++ H$^+$
+\end{center}
+
+As you can see, NAD$^+$ gains a hydrogen atom and two electrons(hence no charge after the reaction).
+
+\subsubsection{FAD}
+
+FAD (flavin adenine dinucleotide) is made from riboflavin (vitamin B2), and it accepts hydrogens in pathways such as fatty acid oxidation and citric acid cycle.
+Flavin co-enzymes are tightly bound to the proteins with which they interact, whereas NAD is free in solution and dissociates after the reaction.
+
+\begin{center}
+    \ch{ FAD + 2 H+ + 2 e- <=> FADH2 }\\
+    \vspace{0.5cm}
+    \setatomsep{2em}
+    FAD + \chemfig{R-C(-[2]H)(-[6]H)-C(-[2]H)(-[6]H)-R'}
+    \ch{<=> FADH2} +
+    \chemfig{R-C(-[2]H)=C(-[2]H)-R'}
+\end{center}
+
+\subsubsection{Co-enzyme A}
+
+Co-enzyme A (CoA) are made from pantothenic acid (vitamin B5), and acts as a carrier of acyl groups (i.e. a fatty acid carrier).
+There are 2 forms: CoASH form (free form) and AcCoA (acyl group attached) form.
+
+\begin{center}
+    \setatomsep{2em}
+    \ch{H+ + CoASH +} \chemfig{CH_3-C(=[2]O)-\chemabove{O}{\scrm}}
+    \ch{<<=> H2O + } \chemfig{CH_3-C(=[2]O)-SCoA}
+\end{center}

### 2015-09-13 bioc192/lect26.tex
finished lecture 26, onto lecture 27

\section{Glycolysis}

Glycolysis is the process where the 6 carbon glucose is split into two 3 carbon pyruvates.
Glycolysis occurs in the cytoplasm and is an ubiquitous pathway, and is found in most cells.
For some cells, glycolysis is essential for the cells' energy requirement (for example RBC, brains cells that use glucose as the major fuel source).
In glycolysis, the energy from this pathway is conserved as ATP and NADH, and the end product (pyruvate is further metabolised for more energy).
There are two main phases in glycolysis: the activation (or energy investment) phase, and the energy pay-off phase.

\subsection{Activation of Glucose}

There are two reactions in glycolysis that add a phosphates onto glucose.
The first phosphorylation (also the first reaction of glycolysis) is done by the enzyme hexokinase, which phosphorylates glucose and get glucose-6-phosphate.
G6P is then rearranged into fructose-6-phosphate by the glucosephosphate isomerase enzyme.
Once in F6P form, phosphofructokinase enzyme adds another phosphate to this molecule to get fructose-1,6-bisphosphate.
FBP is then split in the middle by an aldolase to give DHAP (dihydroxyactone phosphate) and G3P (glyceraldehyde-3-phosphate).
DHAP is then rearranged into G3P by triose phosphate isomerase, as G3P is the form that is fed into the energy pay-off phase.

\begin{reactions*}
Glucose + ATP     & -> G{-} 6 {-}P + ADP\\
G{-} 6 {-}P       & -> F{-} 6 {-}P\\
F{-} 6 {-}P + ATP & -> FBP + ADP\\
FBP               & -> DHAP + G{-} 3 {-}P\\
DHAP &<=> G{-} 3 {-}P
\end{reactions*}

As you can see, we have added two phosphates onto glucose and have splitted this into two 3 carbon molecule with one phosphate on each molecule.
Also note that some of the reactions are slightly unfavourable (e.g. rearrangement of G6P into F6P is \bioGibbs[exponent=0']{1.6}), but the reaction can still go on as the product is constantly being removed, and also that the overall reaction of glycolysis has negative $\Delta$G.
So, when you're looking at the Gibbs free energy of a reaction, think of what is being removed, and also the overall Gibbs free energy of the reaction.

\subsection{Energy Pay-Off/Release Phase}

\subsubsection{Oxidation of G3P and First Substrate Level Phosphorylation}

Once the G3Ps are made, this can then be used to make ATP, but to get a net gain in energy, there must be another phosphate added onto the G3P.
Firstly, a phosphate is added onto G3P to make 1,3-BPG (1,3-bisphosphoglycerate).
The energy required for this phosphorylation is provided by the oxidation of G3P into 1,3-BPG, which also reduces NAD$^+$ into NADH.
Note that this reaction does not require ATP, as all the energy comes from the oxidation of G3P (and therefore this reaction will give us the net gain of ATP).

The 1C phosphate of 1,3-BPG is very reactive, and is removed from the molecule and added onto ADP to produce ATP and 3PG (3-phosphoglycerate).
This is the first substrate level phosphorylation reaction that produces an ATP.

\subsubsection{Arsenic Poisoning}

Arsenic substitutes for phosphate during the oxidation of G3P and is hydrolysed, but does not carry out any substrate level phosphorylation (i.e. no net ATP produced).
This therefore affects the cells that are solely dependent on glycolysis, such as the red blood cells.

\subsubsection{Second Substrate Level Phosphorylation}

Before the second substrate level phosphorylation reaction, 3PG is rearranged to phosphoenolpyruvate (PEP).
Once in PEP form, the phosphate is transferred to ADP by pyruvate kinase (energy comes from the cleavage of phosphate).

\begin{reactions*}
G{-} 3 {-}P + Pi + NAD+ & -> 1 , 3 {-}BPG + NADH + H+\\
1 , 3 {-}BPG + ADP      & -> 3 PG + ATP\\
3 PG ->                 & -> -> PEP\\
PEP + ADP               & -> Pyruvate + ATP\\
\hrulefill\\
Glucose + 2 NAD+ + 2 ADP + 2 Pi &-> 2 Pyruvate + 2 NADH + 2 ATP + 2 H+\\
\end{reactions*}

The overall Gibbs free energy of glycolysis is \bioGibbs[exponent=0']{-73.3}.

\section{Fate of Pyruvate}

The fate of pyruvate depends on the oxygen availability in the cell.

\subsection{Aerobic Condition}

Under aerobic condition, pyruvate is converted into acetyl-CoA (CoA with 2 carbon fatty acid).
This oxidative decarboxylation reaction occurs in the mitochondrial matrix, and requires several co-factors.
Note that this reaction cannot be reversed (pyruvate dehydrogenase complex that catalyses this reaction has an E1 enzyme that has low affinity for carbon dioxide, and therefore the release of carbon dioxide makes it very unlikely for this reaction to be reversed).

\begin{center}
\setatomsep{2em}
\chemfig{CH_3-C([2]=O)-COO\chemabove{}{\scrm}} +
\chemfig{HS-CoA} + \ch{NAD+} \ch{->}
\chemfig{CH_3-C([2]=O)-S-CoA} + \ch{CO2 + NADH + H+}
\end{center}

\subsection{Anaerobic Condition}

Pyruvate cannot be converted into acetyl-CoA under anaerobic condition, and therefore converted into lactate instead.
This reactions is carried out to regenerate the depleted NAD$^+$, and therefore to keep glycolysis going (as NAD concentration is very low in your body, and your cells).

\begin{center}
\ch{Pyruvate + NADH + H+ <=>[Lactate] Lactate + NAD+}
\end{center}

In yeast, anaerobic alcoholic fermentation occurs to regenerate NAD$^+$.

### 2015-09-14 bioc192/lect27.tex
finished lecture 27, ont lecture 28

\section{Fatty Acid as a Fuel}

Fatty acids are the preferred fuel for most tissues in your body.
Fat is the primary energy reserve in your body, making up about 5 to 25\% of body weight in (mammals).
This is because fats are much lighter to store (as it requires no water for storage, and also are more reduced than carbs -- therefore more efficient energy store).

Note that excess glucose and fat are stored in your body as TAGs, and these TAGs are oxidised to provide energy.
However, this does not mean that you can just cut sugar and burn fat, as organs like brain requires glucose as their energy source, not fat.

\section{Delivery of Fatty Acids}

\subsection{From Adipose Tissue to Cells}

In adipose tissue, TAG is broken down into free fatty acids and glycerol by lipase.
FFAs diffuse into blood passively and binds to albumin for transport to other tissues.
Glycerol is transported to the liver.
FFAs get into the cells by diffusion or helped by transporters, and in the cell, FFA binds to the fatty acid binding protein (FABP).

\subsubsection{Fatty Acid Activation}

For fatty acid (or $\beta$) oxidation, the fatty acids must be transported into the mitochondria, as $\beta$-oxidation occurs in the mitochondria.
In order to transport the fatty acids into the mitochondria, FAs are activated in the cytoplasm by attaching CoA onto the FAs.
This generates fatty acyl-CoA, and the energy to do this comes from the hydrolysis of ATP.
Note that the ATP is hydrolysed into AMP, meaning that 2ATP worth of energy was used to accomplish this activation.

\subsection{From Cytoplasm to Mitochondria}

\begin{center}
\includegraphics[width=1.0\textwidth]{fatomt}
\end{center}

Oxidation of fatty acyl-CoA occurs in the mitochondria, but to do this, acyl-CoA must be transported into the mitochondria.
\begin{enumerate}
\item FA-CoA is transported into the intermembrane space via a carrier.
\item FA-CoA is converted into fatty acyl carnitine by carnitine acyl transferase I.
\item Fatty acyl carnitine is carried into the mitochondrial matrix.
\item Fatty acyl carnitine is converted back into FA-CoA by carnitine acyl transferase II.
\end{enumerate}

\subsubsection{Carnitine Supplements}

Note that carnitine is required for the transport of fat into the mitochondrial matrix.
There are supplements of carnitine to try and increase the carnitine concentration (so more fats are transported and burned).
However, you can already make carnitine in your body, and therefore require no extra carnitine (levels of carnitine will be compensated).

\section{$\beta$-Oxidation}

\begin{center}
\includegraphics[width=1.0\textwidth]{betaoxidation}
\end{center}

$\beta$-oxidation (or fatty acid oxidation) oxidises even number carbon fatty acids with no double bonds in it.
No ATP is directly synthesised from this, but energy is captured by co-enzymes (which will eventually be fed into electron transport chain).
The product of $\beta$-oxidation (acetyl-CoA) is fed into citric acid cycle for further oxidation and generation of energy.

$\beta$-oxidation is basically a cycle of oxidation, hydration, oxidation and cleavage, but you end with fatty acid with 2C shorter than what you began with.
Every cycle of $\beta$-oxidation generates FADH$_2$, NADH and an acetyl-CoA (except for the last round, where 2 acetyl-CoA are released).

\begin{center}
$No. of rounds = \frac{no.of C}{2}-1$
\end{center}

### 2015-09-16 bioc192/lect28.tex
finished lecture 28, onto lecture 29

\section{Citric Acid Cycle}

Acetyl-CoA from glycolysis and $\beta$-oxidation is fed into the citric acid cycle, which occurs in the mitochondrial matrix.
All but one reaction (involving FADH$_2$) occurs in the mitochondrial matrix.
Citric acid cycle is a true cycle, as it starts and ends with the same molecule (which is oxaloacetate).

In CAC, there are 4 redox reactions (3NADH and 1 FADH$_2$) and one substrate level phophorylation (GTP/ATP).
Basically, there are two halves in the CAC -- carbon dioxide releasing half, and the rearranging the 4C molecule back to oxaloacetate half.

\subsection{First Half -- Release of CO$_2$}

\subsubsection{Condensation of Acetyl-CoA and Oxaloacetate}

This is the first step of the CAC, where the acetyl-CoA is attached to the oxaloacetate to make the 6C citrate.
Energy to do this comes from the hydrolysis CoA from the acetyl-CoA.
Since CAC is a cycle, the added 2C must be removed at some point of the cycle.

\subsubsection{Isomerisation of Citrate}

Citrate is converted into cis-Aconitate, and then into isocitrate, and these reactions are catalysed by an enzyme called aconitase.
This rearrangement makes the molecule more susceptible for decarboxylation steps.
Poisons such as fluoroacetate and sodium fluoroacetate (1080 poison) targets this step of the CAC.
Fluoroacetate is metabolised to fluorocitrate by binding to the oxaloacetate (similar to the first step of the CAC).
Fluorocitrate then binds to aconitase and inhibits the enzyme, which causes the inactivation of aconitase.
As a result, fluoroacetate decreases the amount of oxaloacetate available AND the activity of aconitase.

\subsubsection{First Oxidative Decarboxylation}

The first decarboxylation step is done in two steps -- oxidation of isocitrate, then the decarboxylation of oxalosuccinate.
The energy from the oxidation is conserved as NADH, and carbon dioxide is released from oxalosuccinate to form $\alpha$-ketoglutarate.

\subsubsection{Second Oxidative Decarboxylation}

This reaction is catalysed by $\alpha$-ketoglutarate dehydrogenase, and is similar to the pyruvate dehydrogenase reaction to add CoA onto pyruvate.
Again, the energy is conserved as NADH, and also releases carbon dioxide to yield a 4C molecule (succinyl-CoA).
This 4C molecule must be rearranged back into oxaloacetate to restart the cycle.

\subsection{Second Half -- Rearranging it back to Oxaloacetate}

\subsubsection{Succinyl-CoA to Succinate}

The conversion of succinyl-CoA to succinate is like the reverse of fatty acid activation.
Removal of CoA releases enough energy to synthesise GTP, which is then used to generate ATP (this reaction has a $\Delta$G close to 0, and therefore close to equilibrium).
This means that this reaction is the third substrate level phosphorylation since glycolysis.

\subsubsection{Succinate to Oxaloacetate}

This conversion is very similar to the $\beta$-oxidation pathway, where succinate is oxidised into fumarate, which is then hydrated into malate, which in turn gets oxidised into oxaloacetate.
The energy from the first oxidation reaction is conserved as FADH$_2$ and the energy from the second oxidation reaction is conserved as NADH.
Note that the first oxidation reaction (which uses succinate dehydrogenase) is shared between CAC and the electron transport chain.
This is in the inner mitochondrial membrane, and is due to the fact that FAD is tightly bound to enzyme, and therefore for ETC to use FADH$_2$, it makes sense to have the enzyme in the inner mitochondrial membrane where ETC occurs.

\begin{center}
\includegraphics[width=1.1\textwidth]{cac}
\end{center}

### 2015-09-17 committee_report/committee_report.tex
added committee report

--- 2015-09-05 463assignment/463assignment_takei.tex
@@
-\documentclass[a4paper, 12pt]{article}
+\documentclass[12pt, a4paper]{article}
 
 \usepackage[utf8]{inputenc}
+\usepackage{natbib}
 \usepackage{geometry}
+\usepackage{fixltx2e}
 
@@
     a4paper,
-    left=3cm,
-    right=3cm,
-    top=3cm,
-    bottom=3cm
+    left=2.5cm,
+    right=2.5cm,
+    top=2.5cm,
+    bottom=2.5cm
 }
@@
 
-\begin{center}
-\large{Olfactory Receptors, Copy Number Variations and Evolution}
-\end{center}
+hello
 
-\section*{\normalsize{Abstract}}
+\citep{Hanahan2011}
 
-Human olfactory receptors (ORs) make up the largest family of genes in the genome, yet there has been little work done on characterising these receptors.
-Studies have shown that OR genes are highly copy number variable within and between species, but no study has associated this with phenotypic characteristics.
-Sensory properties have been linked to satiety and eating behaviours, but has not been described in terms of the individual genotype.
-Next generation sequencing will be used to investigate the effect of OR copy number variation (CNV) on the individual phenotype, and to see whether this affects the sensory-specific satiety and/or food consumption.
-
-\section*{Introduction}
-
-\subsection*{Intro to olfactory receptors}
-
-Olfaction, or the sense of smell, is mediated by a large group of sensory receptors known as the olfactory receptors (ORs).
-ORs all belong to the G-protein coupled receptor (GPCR) superfamily like other sensory receptors, such as for vision and taste.
-ORs are known to be the largest gene family in the human genome and form clusters on many chromosomes, and consists of about 900 genes of which at least half of these genes are pseudogenes and have no functional activity (ref Glusman et al 2001; Niimura and Nei 2005).
-
-Due to the fact that there are so many OR genes in the human genome, and that these ORs work in a combinatorial manner, it has been difficult for the researchers to decipher the complete code of olfaction (ref Hasin-Brumshtein et al, 2009; Nei et al 2008).
-In addition to this, ORs are highly copy number variable within and between species, adding greater complexity to the olfactory process and remains to be a challenge for the researchers to make a connection between OR CNV with the phenotype (ref Nei et al, 2007).
-
-OR genes also have an important implications in the evolution of different species.
-There are two major classes of OR genes: class I and class II genes, or ``fish-like'' and ``mammalian-like'' genes respectively, as the different classes of genes belong to their respective species (ref original paper or Nei and Niimura 2005).
-The study by Niimura and Nei (2005) showed that the diversity of the OR gene family in fishes was greater than that of the mammalian OR gene family, even though the number of OR genes in a single family were less in fishes.
-Furthermore, they showed that amphibians have retained the diversity of the gene family as the fishes, while also having the large repertoire of OR genes in one family as in the mammals, signifying the fact that the CNV was likely due to the adaptation process from being aquatic to terrestial (Nei and niimura, 2005).
-
-As both pseudogenization and CNV are evidence of genomic drift (ref Nei et al, 2007).
-
-\subsection*{Intro to copy number variations}
-
-Recent advances in next generation sequencing technology have allowed many researchers to investigate various aspects of the genome, ranging from something as small as single nucleotide polymorphisms (SNPs) to large chromosomal rearrangements (ref NGS papers).
-Copy number variations (CNVs) is one such genomic rearrangement whereby some genes get duplicated, inserted, deleted and/or inverted.
-
-\subsection*{Intro to olfaction and sensory specific satiety}
-
-\subsection*{Significance of the work}
-
-\section*{Research Proposal}
-
-\subsection*{What am I going to do???}
-
-\section*{Conclusion}
+%Referencing:
+\bibliographystyle{BiocRefStyle}
+\bibliography{../../../References/BibTeX/MSc}
 

### 2015-09-17 bioc192/lect29.tex
finished lecture 29, onto lecture 30

\section{Amino Acids as Fuel}

Amino acids can be used as a fuel source, but its amino group must be removed in order to use it as a fuel.

\section{Deamination of Amino Acids}

Deamination of an amino acid generates a carbon skeleton and a free amino group.
The carbon skeleton can then be used as a fuel.
There are 2 ways of deaminating amino acids: direct deamination into solution by an enzyme, or by transamination of the amino group to a keto acid to form another amino acid.

Direct deamination of the amino acids are simple -- the amino group is removed and released into the solution by an enzyme catalysed reaction.
However, transamination process is a little bit complicated.
In transamination, the amino group of one amino acid is transferred to a keto acid to form a different amino acid, and the leftover carbon skeleton (a keto acid) can be used as a fuel.
For example, glutamate can transfer its amino acid onto a keto acid to form $\alpha$-ketoglutarate and a different amino acid.
These transamination reactions are catalysed by aminotransferases, and requires pyridoxal phosphate (derived from vit B6) to transfer the amino group from one amino acid to another.

\begin{center}
\includegraphics[width=0.8\textwidth]{transamination}
\includegraphics[width=\textwidth]{aminocac}
\end{center}

\subsection{Common Amino/Keto Acid Pairs}

There are common amino acid/keto acid pairs that you should know about.
These are:
\begin{reactions*}
Glutamate &<=> $\alpha$-ketoglutarate\\
Aspartate &<=> Oxaloacetate\\
Alanine &<=> Pyruvate
\end{reactions*}

Different amino acid carbon skeleton can feed into the citric acid cycle at different places, which may or may not produce energy.
For example, aspartate can form oxaloacetate, but going into the CAC here will not directly give you energy (it will provide more oxaloacetate for the cycle).

\section{Removal of Ammonium}

Free ammonium released from the direct deamination of an amino acid is toxic, and must be removed from the cell.
This can be done by adding the free ammonium to a keto acid ($\alpha$-ketoglutarate) to form an amino acid (glutamate).
This amino acid can then transaminate the amino group onto pyruvate to form alanine, which can then be tranported to the liver.
In the liver, alanine is transminated onto $\alpha$-ketoglutarate, and then directly deaminated from glutamate, and the liver cell can process this ammonium into urea.

\begin{center}
\includegraphics[width=0.7\textwidth]{ammonium}
\end{center}

### 2015-09-18 bioc192/lect31.tex
temp update

\section{Oxidative Phosphorylation -- Part 2}

\subsection{Complex I}

Complex I is where the electrons from NADH are accepted/transferred (i.e. where NADH is oxidised).
Two electrons are fed into the ETC from NADH, via this complex.
During this process of transferring the electrons, four protons are pumped into the intermembrane space of the mitochondria.

\subsection{Complex II}

Complex II is where the electrons from FADH$_2$ are accepted/transferred (i.e. where FADH$_2$ is oxidised).
Two electrons are fed into the ETC from FADH$_2$, via this complex.
However, no proton is pumped through this complex -- just electrons through the ETC.
Note that succinate dehydrogenase from CAC is associated with this complex, allowing the enzyme-bound FADH$_2$ to be oxidised and feed into the ETC.
Another thing to mention is that the FADH$_2$ from $\beta$-oxidation (done in the matrix) transfers its electrons via the electron transferring protein, which then feeds into the ubiquinone and not through complex II.

\subsection{Co-Enzyme Q (Ubiquinone)}

Co-enzyme Q (or ubiquinone) is a mobile electron carrier that transfers electrons from complex I/II to complex III.
Co-Q has a characteristic of being able to transfer one electron at a time, which NADH or FADH$_2$ cannot do.
Since complex III requires the electrons to be transferred one at a time, CoQ is an ideal electron carrier from complexes I/II.
The state of ubiquinone when only one electron is accepted is called semi-quinone.

\subsection{Complex III}

Complex III accepts one electron at a time from CoQ and release/transfer this electron to cytochrome c (again, one at a time).
During this transfer process, complex III pumps four protons into the intermembrane space.

\subsection{Cytochrome c}

Cytochrome c contains a haem, and carries one electron at a time to complex IV via a reversible Fe$^{2+}$/Fe$^{3+}$ redox reactions.

\subsection{Complex IV}

Biologically, the production of water occurs as follows:

\begin{center}
\ch{4H+ + 4e- + O2 -> 2H2O}
\end{center}

This shows that the reaction requires four electrons to carry this reaction out.
Complex IV is able to ``hold'' four electrons and wait for the electrons to accumulate, and then do the reaction to produce water.
This complex also pumps two protons into the intermembrane space.

\subsection{Energy Accounting}

To account for the energy, the number of protons must be counted for each reduced co-enzymes.
For NADH, two electrons are transferred to complexes I, III, and IV, meaning that 10 protons in total are pumped into the intermembrane space.
For FADH$_2$, two electrons are transferred to complex II, III, and IV, meaning that 6 protons in total are pumped into the intermembrane space.

\section{Poisoning the ETC}

### 2015-09-21 bioc192/lect30.tex

\section{Oxidative Phosphorylation -- Part 1}

Oxidative phosphorylation is the coupled reaction/process of the electron transport chain and the ATP generation by the ATP synthase.
These processes are coupled by the proton motive force (proton gradient).
The electron transport chain forms this gradient by pumping proton into the intermembrane space of the mitochondria, and the ATP synthase uses this gradient to generate ATP.

\section{Electron Transport Chain}

Electron transport chain (ETC) is located in the inner mitochondrial membrane, as the ETC must have access to all of the reduced co-enzymes from other pathways.
In addition to this, since the ETC must form a proton gradient, this inner mitochondrial membrane acts as the barrier for the gradient.
(Side note: you can centrifuge the cells to isolate certain components of the cell, and use this to do experiments on one particular organelle).

\begin{center}
\includegraphics[width=0.7\textwidth]{etc}
\end{center}

In ETC, electrons are transported through the ETC via the electron carriers (complexes I, II, III, IV and mobile carriers), through series of redox reactions.
The trasnported electrons are ultimately used to reduce oxygen into water, meaning that oxygen is the terminal electron acceptor.
Note that protons are pumped into the intermembrane space as electrons are transported through the complexes.

\subsection{Electron Carriers}

As mentioned before, electrons are passed through various electron carriers.
These carriers are arranged in the ETC as complexes, numbered from I to IV.
There are also 2 mobile electron carriers (mobile in the sense that they are able to communicate between the complexes), called ubiquinone (Co-enzyme Q) and cytochrome c.
Note that cytochrome c is associated with complex III on the intermembrane space side of the inner mitochondrial membranei, and is actally not free in solution.
This means that the electrons are transferred from cytochrome c to complex IV by direct contact (remember that membrane is a fluid).

\subsection{Movement of Electrons Through ETC}

The movement of the electrons from the reduced co-enzymes to oxygen through the ETC requires series of electron carriers (mentioned above) to undergo series of redox reactions.
Electrons move from low redox potential to higher redox potential -- i.e. from reduced co-enzyme to oxygen (highest redox potential).
During the electron movement process, energy is released in a stepwise manner through series of carriers, which allows us to capture energy for pumping protons into the intermembrane space.

\begin{center}
\includegraphics[width=\textwidth]{redoxpotential}
\end{center}

### 2015-09-21 bioc192/lect31.tex
finished lecture 31, onto lecture 32

@@
 \section{Poisoning the ETC}
+
+\subsection{Rotenone}
+
+Rotenone inhibits electron transfer from complex I to coenzyme Q
+
+\subsection{Cyanide}
+
+Cyanide binds to cytochrome a3, which is in complex IV.
+
+\subsection{Carbon Monoxide}
+
+Carbon monoxide binds to where oxygen binds on complex IV, and therefore prevents the electron transfer.
+
+\subsection{Consequences}
+
+All of these will stop the electrons from flowing through the ETC.
+This will in turn prevent the oxidation of NADH and FADH$_2$, which means that these co-enzymes cannot get recycled back into its oxidised form (NAD$^+$ and FAD).
+If there are no NAD$^+$ or FAD around, $\beta$-oxidation and citric acid cycle shuts down as well.
+As a result, protons don't get pumped and therefore no (or decreased) ATP produced.
+Reactive oxygen species can also be produced when the ETC is inhibited, as the electrons have to be passed onto somewhere.
+Since this transfer is from another complex, the electron transfer is incomplete and results in the production of highly damaging reactive oxygen species.

### 2015-09-21 bioc192/lect32.tex
finished lecture 32, onto lecture 33

\section{History of ATP Synthase}

For years, scientists were not able to find any ``energised'' intermediates for substrate level phosphorylation.
Peter Mitchell proposed the chemiosmotic coupling theory, which is a theory that the ETC forms a proton gradient between the inner mitochondrial membrane, and this proton gradient drives the ATP synthesis.
This theory was proposed on the basis that the inner mitochondrial membrane was impermeable to proton, ETC was pumping protons into the intermembrane space from the matrix, and this proton gradient drove the ATP synthesis.

\section{Proton Motive Force}

\begin{center}
\includegraphics[width=0.6\textwidth]{pmf}
\end{center}

Proton Motive Force is the electrochemical gradient formed by the proton gradient, and this drives the synthesis of ATP.
This proton gradient produces 2 kinds of energetic gradients: pH (or chemical) gradient sue to the difference in proton gradient, and electrical gradient as a result of the charge difference across the membrane (positive (\ch{H3O+}) in intermembrane space, and negative (\ch{OH-}) in the matrix).
The energy of this gradient (about 180mV) drives the ATP synthesis.

\section{Experimental Proof of pmf}

\subsection{Experiment with Mitochondria with no Outer Membrane}

When there was no outer membrane of the mitochondria, there was no ATP produced, even though the ETC was still active.
If an energised intermediate was present, then ATP would have still been produced.

\subsection{Experiment using an Uncoupler}

Uncouplers dissipate the pmf by shuttling the protons across the inner mitochondrial membrane, and thus uncouples the proton gradient from ATP synthesis.
When uncouplers were introduced, ATP synthesis was stopped, showing that the pmf was necessary for ATP synthesis.

\subsubsection{Dinitrophenol (DNP)}

\begin{center}
\includegraphics[width=0.6\textwidth]{dnp}
\end{center}

The use of dinitrophenol (an uncoupler) causes your body to heat up, as the proton gradient is dissipated as heat.
You would feel lethargic as well, as you're not making as much ATP as you're supposed to (due to no proton gradient for ATP synthesisdue to no proton gradient for ATP synthesis).
There is no known lethal dose, but it's probably a good idea not to use it.

\subsection{Liposome with Light-Inducible Proton Pump}

\begin{center}
\includegraphics[width=0.3\textwidth]{lipomt}
\end{center}

In this experiment, a liposome was created with a light-inducible proton pump and F$_o$F$_1$ ATP synthase.
When the light was shone on the liposome, the proton was pumped into the liposome, forming a proton gradient.
When this gradient was formed, they observed the production of ATP by the ATP synthase.
This experiment showed that only the pmf was sufficient to drive ATP synthesis, and does not require ETC (as long as there was a proton gradient).

\section{F$_o$F$_1$ ATP Synthase}

\begin{center}
\includegraphics[width=0.4\textwidth]{atpsynthase}
\end{center}

F$_O$ refers to the membrane part of the ATP synthase, and the F$_1$ part refers to the matrix part of the ATP synthase.
ATP synthase is literally a molecular motor, driven by the proton motive force.

\subsection{Mechanism of ATP Synthesis}

There are 2 parts to the ATP synthase: the rotor subunits which rotate/move ($\gamma$ subunit), and the stator subunits which do not move/rotate ($\alpha$ and $\beta$ subunits).
The proton flow through the rotor subunits drives the rotor movement, and this in turn causes a conformational changes in the stator subunits and drives the ATP synthesis.

The $\alpha/\beta$ pairs of F$_1$ binds to ATP or ADP, depending on its conformation.
When it's in an open (O) state, the pair is able to release ATP or bind ADP + Pi.
When it's in a tight (T) state, the catalysis of ATP is able to occur.
When it's in the loose (L) state, the pair is able to hold the ADP and Pi in preparation for catalysis in the T state.
The conformations rotate from O, then to L, then to T, and repeats.

\section{Energy Accounting}

ATP synthase makes an ATP per 4 protons being shuttled across the membrane.
So, for every NADH, 2.5 ATP are made, and for every FADH$_2$, 1.5 ATP are made.

### 2015-09-28 bioc192/lect33.tex
finished lecture 33, onto lecture 34

\section{Photosynthesis}

Photosynthesis is the process in which the energy of sunlight is used by organisms (especially green plants) to synthesise carbohydrates from carbon dioxide and water.
The overall simplified equation of this reaction is:
\ch{6 CO2 + 6 H2O ->[Light{} energy] C6H12O6 + 6 O2}

The energy stored in the carbohydrate is then available to the biosphere through respiration (i.e. carbs used to respire).

\section{History of Photosynthesis}
\begin{description}
\item[Joseph Priestly] Found that plants made ``air''.
When the plant was placed with a candle and/or a mouse, both remained alive.
However, when there was no plant in the jar, the mouse died.
\item[Jan Ingenhousz] Showed that light was required for photosynthesis and suggested the equation: \ch{CO2 + H2O ->[light] (CH2O)n + O2}.
\item[Cornels van Neil] Hypothesised that the oxygen came from the water (or its equivalent part), not the carbon dioxide.
\item[Robin Hill] Experimentally proved that van Neil's hypothesis was right by doing an experiment on isolated thylakoids.
He removed carbon dioxide from the system and added Fe$^{3+}$.
This caused a reaction water and produced oxygen (Hill reaction).
\item[Sam Ruben and Martin Kamen] Used radio-labelled oxygen to track where the oxygen was coming from, and therefore provided direct evidence that the oxygen was coming from the water, not carbon dioxide.
\item[Calvin and Benson] Used 14C atom to track the path of the carbon and proposed the famous Calvin-Benson cycle.
In this cycle, 3 Ribulose-1,5-bisphosphate (RuBP) is carboxylated to make 6 3-phosphoglycerate (PGA or 3PG), which is catalysed by rubisco (the most abundant protein in the world).
3PG are then rearranged into 1,3-BPG, then into glyceraldehyde-3-phosphate (PGAL or G3P).
One of these G3P is a net gain, where as other 5 are use to regenerate the starting RuBP.
\item[Sam Wildman] Discovered the enzyme Rubisco, which catalyzes the first reaction of the Calvin-Benson cycle.
\end{description}

\section{Mechanism of Photosynthesis}

For the first part of the Calvin-Benson cycle to occur, it needs ATP and NADPH -- where do they come from?
The Calvin-Benson cycle is ``coupled'' to the photosystems that utilises light to split water (the light reaction).
There are different types of antenna pigments in different photosynthetic organisms.
This allows them to exploit different niches in the electromagentic spectrum.

\subsection{Antenna and Reaction Center Concept}

Emerson and Arnold came up with the antenna and reaction center concept.
They found that they required 10 photons to split a single oxygen molecule, and there were about 2500 cholorphylls present to split the oxygen.
The idea was that the antenna complex (made of multiple pigment molecules) transferred the energy from the light to the reaction center, where the electron is transferred to molecules.

\subsection{Z Scheme}

The Z scheme was proposed by Robin Hill (the same guy who did the Hill reaction), where the light energy was transferred to specialised pigments in photosystems I or II (note that different wavelengths for different photosystems).
Once the energy is transferred to the photosystem, it can use this to transfer electrons into the electron transport chain (and split water in the process, if it's photosystem II).
Eventually, this electron is accepted by NADP$^+$ and forms NADPH, which is used in the first set of reactions in Calvin-Benson cycle.
The electrochemical gradient formed from the splitting of water and the electron transport chain is then used to make ATP to drive the first set of reactions.

\subsection{Chemiosmosis Theory}

The first evidence of this theory proposed by Peter Mitchel was from an experiment with chloroplasts.
Isolated thylakoids were added to a pH 4 solution and equilibrised.
The thylakoids were then transferred into a pH 8 solution and observed in dark (to prevent any ATP production via light reaction).
This showed that the pH gradient was enough to make ATP.

### 2015-09-28 bioc192/lect34.tex
finished lecture 34, onto lecture 35

\section{Alcohol Metabolism}

\subsection{Alcohol}

\begin{center}
\setatomsep{2em}
\chemfig{H-C(-[2]H)(-[6]H)-C(-[2]H)(-[6]H)-OH}
\end{center}

Alcohol has 29 kJ/g of energy, and takes about 1 hour to get into the blood.

\subsection{Alcohol Dehydrogenase}

Alcohol is processed into acetaldehyde by alcohol dehydrogenase enzyme, which can process about 10g of alcohol per hour (i.e. one drink per hour).

\begin{center}
\setatomsep{2em}
\chemfig{H-C(-[2]H)(-[6]H)-C(-[2]H)(-[6]H)-OH} \ch{+ NAD+ ->[Alcohol{}Dehydrogenase]}
\chemfig{H-C(-[2]H)(-[6]H)-C(=[1]O)(-[7]H)} \ch{+ NADH + H+}
\end{center}

\subsubsection{Kinetics of Alcohol Dehydrogenase}

Alcohol dehydrogenase has a K$_M$ of about 1mM, where 2 to 3mM of alcohol is equivalent to about 2 to 3 drinks.
This means that alcohol dehydrogenase is already at V$_{max}$ after a couple of drinks, and therefore your body starts to accumulate alcohol.

\subsection{Acetaldehyde Dehydrogenase}

\begin{center}
\setatomsep{2em}
\chemfig{H-C(-[2]H)(-[6]H)-C(=[1]O)(-[7]H)} \ch{+ NAD+ ->[Acetaldehyde{}Dehydrogenase]}
\chemfig{H-C(-[2]H)(-[6]H)-C(=[1]O)(-[7]\chemabove{O}{\scrm})} \ch{+ NADH + H+}
\end{center}

Acetaldehyde is toxic to your body and is quickly converted into acetate mainly in the mitochondria (some in cytoplasm).
Glu487Lys mutation in mitochondrial acetaldehyde dehydrogenase causes asian flush, and affects 40\% of the eastern population.

\subsection{Acetyl-CoA Synthetase}

\begin{center}
\ch{Acetate + CoA + ATP -> Acetyl-CoA + AMP + PPi}
\end{center}

Acetate is then converted into acetyl CoA, which can then be used to generate energy, or it can be converted into fatty acids (and cause fatty liver).

\subsection{Fate of Acetyl-CoA from Alcohol}

The obvious path for the acetyl CoA is the CAC/ETC/oxidative phosphorylation for ATP synthesis.
However, if there is enough ATP/energy, the accumulated acetyl CoA is used to make fatty acids.
Since alcohol has no regulatory system, all alcohol is converted into acetyl CoA and it has to be used as energy or stored as fatty acids.

\subsection{Consequences of Alcohol Metabolism}

Alcohol metabolism causes an increase in NADH and ATP in the cell.
This causes the CAC, $\beta$-oxidation, ETC, pyruvate dehydrogenase, and glycolysis (PFK) to slow down.
It also causes the fatty acids to be esterified back to TAGs, and cause fatty liver.
The production of lactate is also favoured, as the NAD$^+$ source is depleted.
This then causes the pH to lower and inhibits gluconeogenesis in the liver (cause low blood glucose and coma).

\section{Metabolism of Alcohol as a Toxin}

Note that this is a different enzymatic system as the one described above.
This pathway uses the microsomal ethanol oxidising system, where the oxidase enzyme (e.g. cytochrome P450) converts alcohol into acetaldehyde.
This means that the speed of alcohol to acetaldehyde conversion is increased.
However, this also means that it metabolises some drugs faster (since alcohol can directly increase the transcription of cyt P450 enzyme, and therefore faster drug clearance).

\begin{center}
\includegraphics[width=0.6\textwidth]{chronicalcmet}
\end{center}

### 2015-09-28 bioc192/lect35.tex
finished lecture 35, onto lecture 36

\section{Fuel Stores}

Energy must be made on demand at the cells.
Fuel stores are required since we cannot store ATP in our body -- ATP must be made in the cell at the time it's needed, at the rate it's needed by oxidising fuels.
We need fuel stores to:
\begin{itemize}
\item To maintain a supply of glucose between meals (especially for brain)
\item To provide immediate fuel for increased activity
\item For long periods when food intake may be inadequate (i.e. starvation)
\end{itemize}
In a fit, 70kg person, there are about 15kg of fat, 200 to 400g of glycogen, and almost no circulating glucose.

\section{Fat Storage}

TAGs are stored as fat droplets in adipose tissues.
Excess fat and carbs from diet can be converted to stored fat, and note that there is no limit to the amount of fat being stored in your body.
TAGs can be various form as well -- different length fatty acids, double bonds, etc.

\section{TAG Synthesis}

Fatty acids from chylomicrons and VLDL are transported to the adipose tissue, with the help of lipoprotein lipase.
The fatty acids are activated by forming acyl-CoA with free CoASH in the adipose tissue (note that this process uses 2 ATP worth of energy, as it makes AMP).
These activated fatty acids are then added to the glycerol backbone (glycerol 3 phosphate), which is made from DHAP in glycolysis (note that it's different to glyceraldehyde 3 phosphate, which goes through the rest of the glycolysis).
All of these processes are under the control of the hormone insulin.

\section{Mobilisation of TAGs}

To mobilise the TAGs, the fatty acids need to be cleaved off the TAGs by hormone sensitive lipase, which is stimulated by adrenaline and glucagon.
This allows the release of free fatty acids and glycerol.

\section{Glycogen}

Glycogen is a branched polysaccharide that is connected to one another via the $\alpha_{1-4}$ and $\alpha_{1-6}$ linkages.
Glycogen is mainly stored in liver and muscles as granules.

\section{Glycogen Synthesis}

Occurs mainly in liver and muscle, immediately after a meal.
Glycogen synthesis requires energy inputs as ATP/UTP.
Glucose is activated to form the high energy precursor UDP-glucose, and this process is done by the glycogen synthase enzyme.
Branching enzyme cleaves long glycogen branches and adds it to the growing branch.
Again, this process is stimulated by insulin.

\begin{center}
\ch{Glucose + ATP ->[Hexokinase] G6P + ADP <=>[Mutase] G1P + UTP -> UDP-Glucose + PPi}\\
\ch{Glycogen + UDP-Glucose ->[Glycogen\ synthase] Glycogen(n+1) + UDP}
\end{center}

Since UDP-glucose is a high energy intermediate, the addition of this to the glycogen is irreversible and requires a different enzyme to break it down.
As mentioned before, the branching enzyme cleaves a chain with more than 6 glucose molecules, and attaches it to the main branch.
This is stimulated by insulin.

\section{Mobilisation of Glycogen}

Glycogen is degraded by glycogenolysis.
Liver glycogen is released as glucose into the blood, to be used by other parts of the body.
Muscle glycogen releases the fuel for glycolysis within the muscle cells.

\section{Formation of Fatty Acids from Glucose}

Excess glucose carbon is converted to fatty acids, which occurs mainly in liver.
This is a complex and energy-requiring process, which involves the addition of acetyl-CoA onto the fatty acids (until 16C long)
These fatty acids are exported as TAGs (16C long) in VLDL, and stored las TAGs in the adipose tissue.
This is stimulated by insulin.

### 2015-09-29 committee_report/committee_report.tex
added some specifications for committee report
added methods and 50-word summary

@@
 \usepackage{fixltx2e}
+\usepackage{enumitem}
+\usepackage{amsmath}
 
@@
 
-hello
+\begin{center}
+\Large{\textbf{Body Mass Index and Pathway Dysregulation in Cancer}}\\
+\end{center}
 
-\citep{Hanahan2011}
+\begin{center}
+Riku Takei (MSc)\\
+Supervisor: Assoc. Prof. Mik Black
+\end{center}
 
-%Referencing:
-\bibliographystyle{BiocRefStyle}
-\bibliography{../../../References/BibTeX/MSc}
+\subsection*{Abstract (50-words)}
+
+Recent studies in breast cancer have shown that there are genetic signatures that are specific to obesity, and these signatures were further investigated in other cancer types (Creighton and fuentes mattei paper).
+The obesity genes identified by Creighton \textit{et al.} did not correlate with any of the cancer types investigated.
+
+\subsection*{Introduction}
+
+\newpage
+
+\subsection*{Methods}
+
+\begin{description}[leftmargin=0pt]
+
+\item[Validation of the methodology used.]
+
+The original CEL files of all the samples used by Creighton \textit{et al.} were downloaded from Gene Expression Omnibus (GEO) database (website ref).
+The microarray expression data for all 103 samples were loaded into R/RStudio and were RMA normalised using the package '????'.
+The 799 obesity gene probes were extracted from the data and singular value decomposition was applied to this matrix to generate a metagene of the obesity genes.
+The metagene was compared with the BMI data of the samples which was plotted as a boxplot (BMI status) and a dotplot (BMI values).
+
+\item[Identification of samples with BMI data.]
+
+The clinical data for all cancer types (33 in total) were downloaded from The Cancer Genome Atlas (TCGA) database and were checked whether the samples had both height and weight data.
+
+\item[BMI calculation and categorisation.]
+
+    BMI were calculated using the samples' height and weight data, where:
+    \begin{equation*}
+    BMI = \frac{Weight(kg)}{Height^2(cm^2)}
+\end{equation*}
+Samples were then categorised into normal weight (BMI \textless{ }25), overweight (25 $\le$ BMI $<$ 30) and obese (BMI $\ge$ 30).
+
+\item[Assessing whether the metagene is associated with BMI.]
+
+Once the metagenes were obtained for each cancer type, the metagene was plotted against BMI status and raw BMI value.
+
+\item[Processing the cancer data.]
+
+RNA-seq data for all of the cancer types with both height and weight data (8 in total) were downloaded from the International Cancer Genome Consortium (ICGC).
+All of these data were loaded into R/RStudio and were processed into a matrix form.
+The data were then logged to the base 10, and standardised so that each gene had a mean of 0 and standard deviation of 1.
+
+\item[Transforming the cancer data.]
+
+Transformation matrix was used to obtain the metagene for the other cancer data.
+In order to transform the data, the transformation matrix from the original data must have the same number of genes as the transforming data, so the genes present in both the Creighton \textit{et al.} data and all of the other cancer types were identified.
+Transformation matrix was made from the Creighton \textit{et al.} data using these common genes, and then this transformation matrix was used to transform all of the cancer types.
+
+\end{description}
+
+\newpage
+
+\subsection*{Results and Discussion}
+
+Since the project was based on the results presented by Creighton \textit{et al.} (2012), that there were obesity-specific genetic signatures in breast cancer, the methods used in this project was applied to the raw data from their paper and the results were compared.
+
+must not exceed 6 pages (with figures and references).
+
+%\citep{Hanahan2011}
+
+%%Referencing:
+%\bibliographystyle{BiocRefStyle}
+%\bibliography{../../../References/BibTeX/MSc}
 

### 2015-09-30 committee_report/committee_report.tex
finished results and discussion

@@
 
-\subsection*{Abstract (50-words)}
+\subsection*{50 Words Summary}
 
-Recent studies in breast cancer have shown that there are genetic signatures that are specific to obesity, and these signatures were further investigated in other cancer types (Creighton and fuentes mattei paper).
+Recent studies in breast cancer have shown that there are obesity-specific genetic signatures, and these signatures were further investigated in other cancer types (Creighton and fuentes mattei paper).
 The obesity genes identified by Creighton \textit{et al.} did not correlate with any of the cancer types investigated.
@@
 
-\newpage
+%\newpage
 
@@
 
-Since the project was based on the results presented by Creighton \textit{et al.} (2012), that there were obesity-specific genetic signatures in breast cancer, the methods used in this project was applied to the raw data from their paper and the results were compared.
+Since the project was based on the results presented by Creighton \textit{et al.} (2012), the methods used in this project was applied to the raw data from their paper and the results were compared to validate whether the methodolgy to be used in this project is applicable to other data.
 
-must not exceed 6 pages (with figures and references).
+After the Creighton \textit{et al.} data were normalised using RMA method, singular value decomposition was applied to the data to obtain the obesity metagene and plotted on a heatmap (fig1a).
+The resulting heatmap showed that the overall expression of the obesity-specific genes correlated with the metagene.
+To assess whether the metagene was actually correlating with the sample BMI, the metagene was plotted against the sample BMI status and sample BMI value (fig1b and c).
+These plots clearly showed that the metagene was able to identify and separate the obese samples from the normal and overweight samples.
+Together, these plots confirmed that the methodology applied to the data set was able to recapitulate the results presented by Creighton \textit{et al.}.
+
+The metagenes for other cancer types were generated by using the obesity-specific genes from breast cancer.
+Transformation matrix was applied to each cancer type to generate the metagene and was plotted on the heatmap (fig2).
+All of the cancer types showed significant correlation with the metagene created from the transformation, where the high metagene values corresponded with high overall gene expression, and vice versa.
+
+To confirm whether the association was in fact due to BMI of the samples, plots of metagene against BMI status and BMI value were created.
+In contrast to the Creighton \textit{et al.} plots, no cancer types showed any significant separation between the obese samples and the normal/overweight samples.
+These results suggest that the obesity-specific genes from breast cancer were not specific enough to be used as a genetic marker for other cancer types.
+
+Although these current results do not show any significant correlation of the obesity-specific genes and the sample BMI in different cancers, there are many places for improvements.
+The data processing of the raw RNA-seq data was carried out using a program written by myself, which could have processed the data incorrectly.
+There are other programs to process these raw data, so the data processing procedure should be repeated and compared with the current results.
+
+The obesity-specific genes in each cancer type was not checked with the BMI values of the samples.
+The results presented here only showed the association of the metagene with the BMI value and status, and it would be good to double-check whether the expression of the genes show any correlation with the BMI of the samples.
+
+For future experiments, differential expression analysis for each cancer type should be carried out to identify the genes that are dysregulated in obese samples but not in the normal or overweight data.
+These genes can then be compared with the genes from other cancer types to identify genes that are  commonly up or down regulated in different cancers.
+More recent paper by Fuentes-Mattei \textit{et al.} (2014) have also found obesity-specific genes in breast cancer, and these genes should be investigated as well.
 

### 2015-10-03 bioc192/lect36.tex
finished lecture 36, onto lecture 37

\section{Starvation}

During starvation, your body needs to supply the brain with fuels (i.e. 120g of glucose per day).
Fatty acids can be utilised by other tissues.
Proteins can also be used as a fuel, but we must conserve as much proteins as possible during starvation, as it would affect the structure and function of the protein, leading to death/trouble.

Glucagon is the main hormone responsible during starvation, and is produced in the $\alpha$ cells of the pancreas.

\subsubsection*{Quick Calculation}

Imagine that we have about 10 kg of fat in our body, and the atwater factor of fat is 38kJ/g (40kJ/g to make the calculation easier).
This means that we have about 400000kJ of energy available to us.
Since we require about 100000kJ of energy per day, we can survive for about 40 days with just fat.
However, our brain needs glucose, not fat, for fuels, so we need an alternative mechanism to provide the brain with energy.

\section{Liver Glycogen}

There is about 90 to 120 g of glycogen that we can mobilise it back to glucose (stimulated by glucagon).
This glycogen store can be utilised and provides enough glucose for the brain for one day.
Glycogen can be broken down into glucose by the enzyme glycogen phosphorylase and debranching enzyme.
The resulting G1P is then converted into G6P, which can then be fed into glycolysis (in muscle) or be used to make new glucose (in liver).
Note that there are glycogen in muscle cells, but these cannot be converted back into glucose as muscle cells don't have the enzyme glucose-6-phosphatase enzyme that is present in the liver cells.

\begin{center}
\includegraphics[width=0.8\textwidth]{gluconeogenesis}
\end{center}

\section{Gluconeogenesis}

As mentioned earlier, this occurs mainly in liver (some in kidney cortex), and is stimulated by glucagon.
Glucose is synthesised using three fuel molecules: lactate (from RBC and muscle), alanine (from muscle protein), and glycerol (from TAG in adipose tissue).
Gluconeogenesis is an energy requiring process, and this energy is provided by fatty acid oxidation.
Most of the glucose produced is used by the brain.

\begin{center}
\includegraphics[width=0.8\textwidth]{gluconeogenesis2}
\end{center}

Just a quick note that both lactate and alanine feed in at the pyruvate level, and so requires energy input to make glucose, but glycerol can feed in at the DHAP level and require no energy to make new glucose.

\section{Maintaining Fuel Supply to the Brain}

Our glycogen store is able to support the brain for a day -- what do we use after we've used up the glycogen?
We use up about 200g of TAGs per day when we are starving, and we can use 20g of glycerol from TAGs to make new glucose (which accounts for about 20g of glucose).
This means that we still need about 100g of glucose to fuel our brain -- where does it come from?

\subsection{Protein Stores}

We can break down protein to provide the source of new glucose.
We have got about 10 to 15kg of protein inour body, but we don't have any specific storage proteins for fuel.
This means that some proteins must be degraded to make glucose.
However, this can only last for up to 2 weeks before we die from other complications(i.e. we cannot rely on protein too heavily).
So, we can break down proteins, but we have to conserve as much protein as possible to prevent structural and functional damage to your body.

\subsection{Ketone Bodies}

\begin{center}
\includegraphics[width=0.7\textwidth]{moleculesfasting}
\end{center}

An experiment done Cahill and Owen in 1967 showed that the blood glucose in your body was maintained during fasting and there was an increase in FFA concentration due to TAG breakdown.
They also noted that $\beta$-hydroxybutyrate and acetoacetate concentration have increased by 100 times during fasting, and showed that about 50g of these ketone bodies were used by the brain.
This meant that only 50g of glucose was needed to be made from the breakdown of protein.

\subsubsection{Ketone Body Synthesis}

Ketone bodies are synthesised in the liver from fatty acids.
Fatty acids are activated into acyl CoA, which is then broken down into acetyl CoA via $\beta$-oxidation.
Acetyl CoA is then converted into ketone bodies (acetoacetate and $\beta$-hydroxybutyrate) in the process of ketogenesis.
These ketone bodies can cross the blood brain barrier, converted back into acetyl CoA in the brain, which is then used in the CAC to make energy.
Note that these ketone bodies can be spontaneously decomposed into acetone (hence the sweet smell in you breath).

\begin{center}
\includegraphics[width=0.6\textwidth]{ketogenesis}
\end{center}

\section{Metabolic Adaptations to Starvation}

To summarise, the body uses all of the mechanisms described above to keep the body alive and adapt to starvation.
Fatty acids are used by all aerobic tissues (except brain) for energy.
Once we use up all of the glycogen stores, excess fatty acids are converted into ketone bodies for the brain to use, and therefore prevent/conserve the amount of proteins broken down (and therefore stay alive for longer).

### 2015-10-03 bioc192/lect37.tex
finished lecture 37, onto lecture 38

\section{Exercise Metabolism}

Different pathways and fuels are used for different kinds of exercise.
There is about \SI{5}{\micro\mol} of ATP per g of muscle, which can generate enough energy for an exercise for a second.
This means that we need to make more ATP to keep doing exercise, and depending on the type of exercise, we have to utilise different pathways and fuels at different time.

\section{Anaerobic Exercise}

Anaerobic exercise is defined as the high intensity exercise that requires rapid generation of energy in short period of time where oxygen (although available) is not used as an electron acceptor (i.e. not making energy through CAC).
In anaerobic exercise, it mostly uses phosphocreatine and glycogen to make energy for the exercise.

\subsection{Phosphocreatine}

Phosphocreatine is the on site, fast fuel.
There is about 20mol of phosphocreatine per g of muscle, and it's a high energy compound.
Phosphocreatine is made in the liver from arginine and glycine, which is then transported to msucles.
The phosphate from the phosphocreatine is transferred onto ADP to make ATP, which is catalysed by creatine kinase.

Some studies show that if you are able to maintain the creatine in your muscle, you can exercise for a longer time.

\subsection{Glycogen}

Glycogen phosphorylase breaks down the glycogen into glucose-1-phosphate in muscle, which is converted back into glucose-6-phosphate, and fed into glycolysis.

\subsection{Anaerobic Glycolysis}

Muscle glycogen is used as fuel and no oxygen is required for this.
ATP is generated from the substrate level phosphorylation reaction, and NAD$^+$ is regenerated by the pyruvate to lactate conversion.
This allows rapid generation of energy, but only for a short time.

\subsection{Regulation of Glycolysis in Exercising Muscle}

Glycogen mobilisation is stimulated by Ca$^{2+}$ and adrenaline (when you're exercising you release calcium ions).
Phosphofructokinase activity is upregulated by AMP and inorganic phosphates (produced from ATP usage into ADP; Pi from making ATP from two ADP).
This allows increased flux through the glycolytic pathway.

\section{Aerobic Exercise}

Aerobic exercise depends more on the oxidation of glucose and fatty acids, using oxygen (i.e. CAC etc.).
After 20sec of anaerobic exercise, aerobic exercise kicks in.
At this point, 50\% of energy comes from glucose and 50\% of energy comes from fatty acids.

\begin{center}
\includegraphics[width=0.7\textwidth]{exercisemetabolism}
\end{center}

\section{Aerobic Training}

Aerobically trained people are able to rely less on glycogen and therefore able to exercise for a longer period of time (i.e. don't have to rely on glycogen as much).
When you train, you increase the mitochondrial ability in the cells, and therefore able to oxidise lipids and carbohydrates more efficiently.

\section{Muscle Fibre Types}

\begin{description}
\item[Type I] is the red (slow twitch) fibre, and mainly uses oxidative phosphorylation for energy production.
\item[Type II] is the white (fast twitch) fibre, and mainly uses glycolytic paths for energy production.
\end{description}

\section{Myostatin}

Myostatin acts to inhibit muscle growth and differentiation.
People with defects in this gene therefore have increase muscle production/development (due to increase in certain transcription factors for muscle development).

### 2015-10-03 bioc192/lect38.tex
finished lecture 38, onto lecture 39

\section{Diabetes}

Symptoms of diabetes include:
\begin{itemize}
\item fatigue
\item weight loss
\item intense thirst
\item frequent urination
\item hyperglycaemia
\item glucosuria
\item ketones
\end{itemize}

There are two types of diabetes:
\begin{description}
\item[Type I diabetes] is where the individual has an autoimmune disease of the pancreatic $\beta$ cells, where insulin is made.
Since the patient is responsive to insulin (just that they can't make it themselves), treatments include insulin injection.
This type is mainly genetic.
\item[Type II diabetes] is where the individual is not responding to the insulin (i.e. insulin resistance).
This means that insulin injection would not work on these patients and require alternative medication (e.g. hypoglycaemic drugs), if there is any (or just change lifestyle).
This type is both genetic and environmental.
\end{description}

When the blood glucose level is too low, it can cause serious problems (coma and death).
Although unless you have very high blood glucose for a long period of time, increase in blood glucose is not too serious.
However, it can cause many problems chronically (e.g. glycation of proteins, nreve impairment, retinopathy, etc.).
Glucose tolerance test can be used to test the effect of insulin in the patient.

\section{Insulin}

Insulin is a peptide hormone synthesised from the $\beta$ cells in the pancreas, and is secreted in response to increase blood glucose.
Since insulin is a peptide hormone, you cannot ingest it (as the GI would break it down).
Insulin acts on the whole body, but mainly the liver, muscle and adipose tissue to stop the fuel breakdown/mobilisation and stimulates the uptake/synthesis of fuel stores.

Consequences of no or very little insulin are:
\begin{description}
\item[Decreased glucose uptake and storage] due to no action by insulin.
\item[Increased glycogen mobilisation] as your cells are not taking up any glucose.
\item[Increased glucose synthesis] as the liver thinks that you're starving.
\item[Increased lipolysis] to try and increase the energy production.
\item[Increased ketogenesis] to compensate for the ``lack'' of glucose by your brain.
\item[Decreased removal of TAGs from blood] as no signal from insulin.
\item[Increased protein breakdown] due to the body thinking it's starving.
\end{description}

\begin{center}
\includegraphics[width=0.8\textwidth]{insulin}
\includegraphics[width=0.8\textwidth]{effectofinsulin}
\end{center}

### 2015-10-03 bioc192/lect39.tex
finished lecture 39, onto lecture 40

\section{Obesity}

New Zealand is the third most obese country in the world.
Body Mass Index (BMI) is measured as:
\begin{center}
$BMI = \dfrac{Weight (kg)}{Height^2 (m^2)}$
\end{center}
and define people with BMI $>$ 30 obese, 25 $<$ BMI $\leq$ 30 overweight, 20 $<$ BMI $\leq$ 25 normal, and BMI $\leq$ 20 underweight.

\begin{center}
\includegraphics[width=0.9\textwidth]{energyflow}
\end{center}

\section{Metabolic Rate}

Metabolic rate is fixed, and partially genetic.
However, if we can increase the metabolic rate, then we would be able to lose weight without exercising.
How do we do this?

\section{Uncoupling Protom Motive Force from Oxidative Phosphorylation}

By uncoupling the proton motive force from the oxidative phosphorylaiton will dissipate the energy as heat and burn more fat.

\subsection{Dinitrophenol}

Dinitrophenol can be used to uncouple the proton motive force by shuttling the proton across the inner mitochondrial membrane.

\subsection{Brown Adipose Tissue}

Brown adipose tissue (BAT) is a special thermogenic tissue found in hibernating animals, babies, and some in adults.
BAT keeps the animals warm by generating heat (by uncoupling the pmf).
These cells contain many mitochondria and small fat droplets.

\subsubsection{Uncoupling Protein (UCP)}

UCPs were found in BAT, and are present in the inner mitochondrial membrane.
UCPs are regulated proton channels in the membrane and uncouples the ATP synthesis.
This means that heat is generated due to this uncoupling, and increases the BMR.

\subsubsection{Diet-induced Thermogenesis}

In rodents, activated UCP in BAT burnd off excess dietary energy.
Humans have active BAT, which may be used to burn fat/excess energy.
There are also related UCPs in WAT and muscles, which could potentially be used to burn fat and excess energy.

\subsubsection{Strategies of BAT Usage}

\begin{enumerate}
\item Stimulate existing BAT
\item Switch on BAT differentiation and growth
\item Transplantation of engineered BAT
\end{enumerate}

\section{Genetic Component of Obesity}

There are three genes (ob = obese, db = diabetic obese, and fa = fatty) that make mice fat.

\subsection{Obese Gene (Leptin)}

The ob/ob gene codes for leptin -- hormone secreted from the ``fat'' adipose tissue cells.
Leptin acts on the hypothalamus to let the body know that it is ``full'', and therefore require no more food.
This leads to a decrease in appetite and stimulates energy expenditure (therefore maintains normal body weight).

\begin{center}
\includegraphics[width=0.8\textwidth]{leptin}
\end{center}

In a mouse model, they injected leptin into the ob/ob (obese) mouse.
This stopped the mouse to eat more, and the body weight returned to normal (i.e. lost weight).

Defects in either leptin or leptin receptor (in hypothalamus and other tissue) can lead to increased weight (absent in db/db and fa/fa mouse).

\subsection{Leptin in Humans}

Humans also have leptin and its receptors, and mutations in these can lead to obese phenotype.
This means that the individual will have no leptin or non-responsive to leptin, and therefore always feel hungry.
Most obese humans appear to be resistant to leptin.

\section{Treatments}

\begin{description}
\item[Food breakdown] Target pancreatic lipase (Xenicol) so less fat is absorbed.
\item[Satiety signals] Increase the leptin levels or target leptin receptors.
\item[Mitochondria and brown fat] Uncouple oxidative phosphorylation from electron transport chain by upregulating UCPs and/or increase functional BAT.
\end{description}

### 2015-10-05 bioc192/lect40.tex
finished BIOC192

\section{Integration of Metabolism}

\subsection{Brain Metabolism}

Brain aerobically oxidises glucose and ketone bodies (during starvation).
The key pathways used by the brain are glycolysis, CAC, ETC and oxidative phosphorylation, in the presence of oxygen.

\subsection{Skeletal Muscle Metabolism}

Skeletal muscle can aerobically or anaerobically oxidise fuel molecules, depending on the energy requirement.
It can aerobically oxidise glucose, FAs, ketone bodies, and TAGs (to make FAs).
During anaerobic metabolism, lactate is produced and is released into the blood (goes to liver and used as fuel or gluconeogenesis).
Proteins can be broken down into amino acids under starving condition and used as fuel, or it can be released into the blood.

\subsection{Hear Muscle Metabolism}

Heart is completely aerobic and able to oxidise pretty much all fuels -- glucose, FAs, ketone bodies and lactate (which is converted into pyruvate for CAC).
Key pathways include glycolysis, $\beta$-oxidation, CAC, ETC, and oxidative phosphorylation.

\subsection{Adipose Tissue Metabolism}

This is the major TAG store, and also where TAGs are stored/mobilised.
Mobilisation of FAs are done by hormone sensitive lipase and FFA are released into the blood for other cells to use, whereas the glycerol backbone is sent to the liver for gluconeogenesis.
Adipose tissue also secretes various hormones such as leptin, but it can also synthesise and secrete many other hormones (referred to as adipokines).

\subsection{Liver Metabolism}

Major metabolic processing organ, and is the first place to have access to the fuels after digestion of food.
It also synthesise and stores glycogen, and is also the major site of fat synthesis (which is exported as TAGs in VLDL).
Glycogenolysis and gluconeogenesis also occurs in the liver.
Synthesis of ketone bodies and conversion of toxic excess ammonia to urea is also done here.

\section{Signalling Your Body}

\subsection{Endocrine System}

Chemical signals that are synthesised in the endocrine gland, and is released into the blood to act on distant target tissues.
These regulates metabolic activities and cellular processes.

\subsection{Nervous System}

Neurotransmitter chemicals are synthesised and released from the nerve endings into the synapse.
These act on the post synaptic cells (e.g. nerve/muscle).

\subsection{Types of Hormones}

\begin{description}
\item[Steroid hormones] e.g. estrogen, testosterone, etc.
\item[Polypeptide/peptide hormones] e.g. insulin, glucagon, etc.
\item[Amino acid derivatives] most neurotransmitters.
\end{description}

\subsection{Functions Regulated by Signal Molecules}

Pretty much everything that the cell needs to do to keep us alive -- making energy, growing, wound healing, metabolism, etc.
Hormones and other chemical signals:
\begin{description}
\item[Stimulate/inhibit the activity of enzymes or other proteins in cells] It can regulate metabolic pathways and secretory processes.
\item[Alter transcription of specific genes] and therefore regulate the amount of enzymes and proteins produced.
This means that it regulates growth and differentiation.
\end{description}

\subsection{Signal Transduction and Diseases}

\begin{description}
\item[Type II Diabetes] Insulin signal transduction is impaired due to resistance/loss of insulin.
New drugs target the signal transduction pathway of insulin to stimulate the pathway.
\item[Cancers] Overactive signal transduction stimulates the cells to divide and leads to tumour.
New drugs to block unwanted signal transduction.
\end{description}

### 2015-10-11 notes/rhodopsin.tex
added rhodopsin notes

\documentclass[a4paper,12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{fixltx2e}
\usepackage{enumitem}
\usepackage{amsmath}

\geometry{
    a4paper,
    left=2.5cm,
    right=2.5cm,
    top=2.5cm,
    bottom=2.5cm
}

\begin{document}

\begin{center}
\LARGE{Rhodopsin Notes}
\end{center}

\section*{Rhodopsin}

\subsection*{\normalsize{Kim et al. 2004}}

Retinal is covalently bound to the receptor via the protonated Schiff linkage to the lysine 296 (K296), which is on the transmembrane helix 7.
The buried Schiff base is stabilised by an electrostatic interaction called the salt bridge with the carboxylate group of the glutamic acid 113 (E113).
Isomerization of the rhodopsin molecule from cis to all trans causes this salt bridge to be disrupted and leads to the conformational changes that activate the rhodopsin receptor.

Direct (E113Q) or indirect (G90D or A292E) mutations causes the disruption of this salt bridge and leads to a constitutive activation of the retinal-unbound form of rhodopsin, opsin.
The E113Q mutation will obviously lead to the disruption of the salt bridge, as this amino acid is essential for the stabilisation of the Schiff base that the retinal is bound to.
G90D and A292E mutations cause these amino acid side chains to competitively interact with the Schiff base, and thereby rupturing the salt bridge.

There are other mutations that causes a constitutive activation of opsin that does not seem to affect the salt bridge.
These mutations are M257Y and E134Q.
This study was investigating what the structural changes associated with each of these mutations.

\noindent\rule{\textwidth}{0.4pt}

The results showed that the even though the salt bridge mutations caused a change in conformation of rhodopsin and activated the receptor to a degree, this was not enough to activate the receptor in dark state.
Likewise, the non-salt bridge mutations caused a change in conformation, but not enough to fully activate the receptor.
Both salt bridge mutations and non-salt bridge mutations were not additive, meaning that more than two salt bridge mutations did not significantly increased the activation, likewise with non-salt bridge mutations.
These results showed that one type of mutation was not enough to induce the conformation changes required for full activation of the receptor.

However, when the salt bridge mutation was combined with the M257Y (non-salt bridge mutation), it showed a significant increase in the activity of the rhodopsin (with retinal bound), almost to a level of light induced state, likely caused by the TM6 movement.

\subsection*{\normalsize{Ernst et al 2005}}

This study was to determine what component of retinal was important for the conformatioal changes caused by light induction and how the information was effectively transduced.
It was known that in the meta II state of rhodopsin, the polyene chain of all trans retinal was firmly locked in this state and prevented the re-isomerisation of the all trans retinal into the cis retinal.
It was thought that the cyclic part of the retinal was important for the movement of the TM6 out of the helix bundle and activate the receptor.
They used acyclic (ac) retinal to test the role of this cyclic part of the retinal.

\noindent\rule{\textwidth}{0.4pt}

They found that the ac retinal was able to form an active rhodopsin, but not as efficiently as normal retinal.
This was due to the change in pKa observed between the normal rhodopsin (pKa = 7.5) and the ac retinal bound rhodopsin (pKa = 5.2).
In order to achieve an active state, the rhodopsin must be able to be protonated in order to hydrolyse and release the trans retinal.
This means that it is much harder for the ac rhodopsin to protonate, as it has lower pKa.

They also found that even when ac rhodopsin becomes active, it was only short-lived compared to the normal rhodopsin.
After its activation, the polyene chain is tightly fitted in the receptor and is only able to be translocated/dissociated from the complex.
This translatory movement is made easier for the ac retinal due to the loss of the cyclic ring structure, and also the acyclic structure is not able to interact as much as the TM6, therefore not able to maintain the active conformation as much as the all trans retinal.

Taken together, the acyclic retinal causes less formation of the active meta II as well as the increased degradation of the meta II.

\section*{Transducin/G-protein}

\subsection*{\normalsize{Yeagle and Albert 2003}}

This study was looking at how the interaction of the G-protein (transducin) with rhodopsin activated it and produce signals.
Transducin is a heterotrimeric protein that consist of an $\alpha$, $\beta$ and $\gamma$ subunits.
Beta and gamma subunits are tightly bound to one another and cannot be dissociated (unless under extreme denaturing condition).
The alpha subunit has GTPase domain at the interface of the complex with beta/gamma subunit, and have positive N-term helix and C-terminal end which binds to the negatively charged groove that is opened up upon light activation of rhodopsin (i.e. in meta II state).

The C-terminus of alpha subunit binds into the groove formed upon activation.
This binding places the lysine (K341) and aspartic acid (D341) near the conserved D(E)RY sequence of the rhodopsin, and stabilises the active conformation of rhodopsin in meta II state.

N-terminus of alpha subunit is ordered when it is bound to the beta/gamma subunit of the G-protein and when it is bound to GDP instead of GTP.
This flexible N-term helix undergoes a significant change in conformation upon binding to the receptor, and allows the GDP/GTP exchange to occur, and also favours the release of the beta/gamma subunit dissociation.

\subsection*{Bohm et al 1997}

Review of the structures of transducin.
The alpha subunit has a GTPase domain which is at the interface of the alpha-beta/gamma subunit.
This GTPase domain contains three switches (switches I to III) which are conformationally sensitive to the gamma phosphate of GDP/GTP.
These switches are bound/interacts with the GTP, but these are flexible when GDP is bound.
The switch region, especially switch II, interacts with the beta/gamma subunit when there is no phosphate (i.e. in the GDP bound form).
However, when GTP is bound, the switches are more tightly bound to the GDP than the beta/gamma subunits and decreases the binding site for the beta/gamma subunits, and therefore the complex dissociates.

\begin{description}
\item[Phosducin] is a regulator of beta/gamma subunits and it prevents the re-association with the alpha subunit.
Since the beta/gamma subunits are required for the GDP/GTP exchange, phosducin prevents the activation of the alpha subunit.
\item[Regulator of G-protein Signalling (RGS)] binds to the switch regions and increases the alpha subunit's GTPase acitivity (therefore decreases the duration).
\end{description}

\subsection*{Slep et al 2001}

This paper looked at how the alpha subunit of the transducin was able to activate the peophodiesterase (PDE) enzyme, which converts cGMP into GMP and closes the calcium channel.

PDE has an inhibitory gamma subunit which inhibits the PDE activity.
The C-terminal peptide of this domain inhibits PDE activity.
This C terminal inhibitory peptide can also bind to/interact with the switch II of the alpha subunit, which allows the alpha/beta subunit of PDE to catalyse the cGMP into GMP reaction.

RGS binding to the alpha subunit of transducin is enhanced by the binding of the PDE gamma subunit binding.
This allows the increase in the GTP hydrolysis by the alpha subunit by positioning and stabilising the switch regions and catalytic groups for efficient hydrolysis.

\section*{Rhodopsin kinase and recoverin}

\subsection*{Arshavsky 2002}

This paper was to address the mechanisms proposed by different researchers about the inactivation process of  rhodopsin.
Rieke and Baylor (1998) proposed that the inactivation of rhodopsin is a multi-step process rather than a single stochastic event so the inactivation process is uniform and therefore able to produce reliable electric response.

Different studies show different amount of phosphorylation on the C-terminal end of rhodopsin to inactivate rhodopsin.
One group proposed that the inactivation process occurs in a 1 to 7 phosphorylation step mechanism by rhodopsin kinase.
This lead to the hypothesis where phosphorylation event itself accounted for the reproducibility of the response, which was mediated by the calcium feedback due to the activation of the signal (calcium feedback hypothesis).
The phosphorylation is done by rhodopsin kinase, which is mediated by recoverin (constantly regulated by calcium concentration).

Multiple phosphorylation hypothesis, on the other hand, proposes that rather than the individual phosphorylation event, the number of phosphorylation on rhodopsin provides the reproducibility of the signal.
This is based on the observation where arrestin binds to rhodopsin and inactivate it, but binds to rhodopsin with higher affinity when it is phosphorylated more.
Thus the arrestin binding provides the reproducible signal response.

Kennedy et al (2001) showed that the phosphorylation of  rhodopsin was sequential and in a highly ordered pattern, and likewise with dephosphorylation.
This supports the multiple phosphorylation hypothesis.
However, this does not mean that the calcium feedback hypothesis is incorrect -- these two mechanisms may work together to produce a reproducible signal.

Another thing mentioned was that there were two phases of dark adaptation in mouse.
The first phase occurs within minutes (2min), whereas the other occurs after 40 minutes.
These corresponded to the time frames of the phosphorylation of rhodopsin and the slow decay of the photoactivated rhodopsin and the subsequent regeneration of the cis retinal bound form of rhodopsin, as all of the decay product can activate the signal.

\subsection*{Koch et al 2003}

This study was looking at the relevance of the N-terminal myristoylation of recoverin.
At high calcium concentration, recoverin is able to bind to rhodopsin kinase and inhibit its activity.
When the rods are illuminated, the signal via rhodopsin leads to a decrease in the calcium concentration in the cell, which causes the activation of rhodopsin kinase.

\begin{description}
\item[T state]is where the recoverin is in a calcium free state and the myristoyl is buried within the hydrophobic pocket.
This state activates rhodopsin kinase.
\item[R state]is where the recoverin have calcium bound to it, which causes the myristoyl to protrude out.
\end{description}

There are two calcium binding domain in recoverin -- EF2 which binds to calcium with a low affinity, and EF3 which binds to calcium with high affinity.
This means that calcium binds to the EF3 first, and then onto EF2.
Furthermore, the myristoyl is suggedted to be an allosteric regulator that affects the sequential binding of calcium onto recoverin (EF3 first, then EF2).

Koch et al investigated what the myristoyl modification was important for.

\noindent\rule{\textwidth}{0.4pt}

With a non-myristoylated recoverin, they showed that the myristoylation modification was not necessary for the inhibition of the rhodopsin kinase, but the calcium binding to the EF2 was (E85Q mutant that could not bind to calcium was not able to inhibit rhodopsin kinase).

They showed that the calcium binding to EF3 was sufficient to cause the conformational changes from T to R state by destabilising the T state of the N-terminal domain of recoverin.
This N-terminal destabilisation is counteracted by the myristoyl group, where the myristoyl group stabilises the N term in the T state, and thus prevent inhibition of rhodopsin kinase.
When the calcium binds to the EF2 domain, this myristoyl stabilisation of the N term is relieved, and the full conformational change from T state to the R state.
The E85Q mutant which could not bind calcium in EF2 therefore was not able to adopt the full R state conformation, and therefore was not able to inhibit rhodopsin kinase.

They also showed that upon calcium binding, a hydrophobic patch appears.
When calcium binds to the EF3 domain, most of the patch appears, but there is a hydroxyl group (from tyrosine) in the middle of the patch that prevents the interaction with rhodopsin kinase.
This hydroxyl group is completely buried when calcium binds to the EF2 domain, again showing that calcium binding to the EF2 domain is essential for the inhibition of rhodopsin kinase.

\subsection*{Higgins et al 2006}

This paper was looking at how recoverin inhibited rhodopsin kinase.
Rhodopsin kinase is known to be able to phosphorylate rhodopsin as well as itself (autophosphorylation).
It has been known that the phosphorylation site of rhodopsin by itself has low affinity by rhodopsin kinase, suggesting that there must be additional interaction site to recognise the activated rhodopsin as a substrate.

\noindent\rule{\textwidth}{0.4pt}

Higgins et al showed that the first 15 N-terminal residues of rhodopsin kinase were essential for the phosphorylation of rhodopsin, and also for recoverin interaction.
However, they showed that even though these residues were essential for phosphorylation, it was not essential for rhodopsin kinase's catalytic activity (i.e. they were able to autophosphorylate itself without the residues).
This suggests that these residues are important for the recognition of the rhodopsin kinase by other proteins but not the kinase activity.

They predicted that these 15 residues formed an amphipathic helix that was able to be recognised by various proteins.
Mutations that altered the hydrophobic residues of this amphipathic helix abolished the ability of rhodopsin kinase recognition by other proteins, signifying the importance of these residues.

Higgins et al concluded that the N terminal residues are essential for the recognition of rhodopsin by other proteins, and the binding of antibody to these residues blocked the phosphorylation of rhodopsin by preventing the rhodopsin kinase interacting with rhodopsin.

\section*{Arrestin}

\subsection*{Burns et al 2006}

This study was done to establish the roles of the arrestin splice variant.
\begin{description}
\item[p48 arrestin] is the full length arrestin that is 10 times as more abundant that p44 splice variant, and is excluded from the rod outer segment that is dark adapted.
Under constant light illumination, arrestin moves to the outer segment from the inner segment.
\item[p44 arrestin] is the C terminal truncated arrestin that can turn off rhodopsin more efficiently than p48.
\end{description}
This lead to the hypothesis where the p44 arrestin is more important than p48 in deactivating rhodopsin under dark adapted conditions.

\noindent\rule{\textwidth}{0.4pt}

First they showed that under light condition, the binding of p48 to rhodopsin does not cause significant increase in the p44 concentration -- shows that the p44 variant is not due to the proteolysis of p48 variant.

Constant activation of rhodopsin can cause degeneration of the photoreceptor.
They showed that both p48 and p44 arrestin was able to prevent the photoreceptor degeneration under normal condition (where the exposure to light was alternated with darkness).
However, p48 was able to prevent the degeneration of the photoreceptor under constant light condition, whereas the p44 splice variant could not.
This suggested that p48 was able to prevent light induced retinal damage.

Previous reports on the abrupt turn off of rhodopsin in rhodopsin kinase knock out cells was thought to be due to the p44 arrestin binding to the non-phosphorylated rhodopsin.
They showed that the arrestin and rhodopsin kinase double knockout cells still had the abrupt turn off (but after a longer time).
They also showed that the presence of p48 made the response look similar to the WT cells, but the presence of p44 did not, suggesting that p48 is able to deactivate the non-phosphorylated receptor, but not p44.
They concluded that p48 was important for decreasing the time required for dark adaptation by stabilising the decay product by binding to rhodopsin, but not p44.

Arrestin mechanism:

\begin{enumerate}
\item Arrestin has two domains, activation recognition domain (low affinity to activated rhodopsin), and phosphorylation recognition domain (low affinity to phosphorylated rhodopsin).
\item Encounter with the activated and phosphorylated rhodopsin disrupts the interaction that stabilise the basal state of arrestin (arrestin activation).
\item This leads to a conformation change and exposes the secondary binding site of arrestin that has a high affinity for the activated and phophorylated rhodopsin.
\end{enumerate}

\end{document}

### 2015-10-12 notes/kinase.tex
added kinase notes

\documentclass[a4paper,12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{fixltx2e}
\usepackage{enumitem}
\usepackage{amsmath}

\geometry{
    a4paper,
    left=2.5cm,
    right=2.5cm,
    top=2.5cm,
    bottom=2.5cm
}

\begin{document}

\begin{center}
\LARGE{Kinase Notes}
\end{center}

\section* {RAF Transactivation}

\subsection*{\normalsize{Hatzivassiliou et al 2010}}

One of the first evidence that V600E mutant inhibitors activated the WT BRAF/KRAS paradoxically.

\noindent\rule{\textwidth}{0.4pt}

CRAF knockout was able to prevent the pathway activation, but BRAF knockout could not.
This suggested that CRAF was essential for the pathway activation, but not BRAF.
They mutated the CRAF so that the inhibitor was unable to bind to it (T421N mutant), and showed that this was also able to prevent the activation of the pathway, signifying the fact that the inhibitor binding to the active site was essential for effective downstream signalling.

They further investigated the inhibitor AZ-628 and compared it with the other inhibitors such as PLX 4720 and GDC 6879.
\begin{description}
\item[AZ-628] binds to the kinase in an inactive conformation of the active site, and has a very slow/irreversible off rate.
\item[GDC6879 and PLX4720] both have a very rapid off rate, and therefore unable to occupy the ATP binding site for long.
This means that these inhibitors cannot inhibit CRAF kinase activity effectively.
\end{description}

They showed that AZ628 was able to inhibit the pathway acitvation effectively, but the others could not.
They hypothesised that this was due to the ability of AZ628 to occupy the active site of CRAF for a long period of time, and thus able to inhibit the kinase activity of CRAF.

They also showed that the inhibitor binding to the active site affected the dimerisation of BRAF/CRAF, and that the dimerisation was essential for the pathway activation via the recruitment to the membrane in a RAS-dependent manner.
They also showed that the mutants that could not be dimerised were not able to activate the pathway, showing further evidence of dimer dependent pathway activation.

\subsection*{\normalsize{Poulikakos et al 2010}}

Another evidence suggesting that the BRAF inhibitor activates WT BRAF/CRAF.
The inhibitor is effective against the BRAF V600E mutants, but not the WT BRAF.

\noindent\rule{\textwidth}{0.4pt}

First, they showed that the BRAF KO, but not CRAF KO activated the pathway in a dose dependent manner when inhibitor was added (i.e. activation occurred even without BRAF in the cell).
This result shows how CRAF is essential for the pathway activation, but not BRAF.

Secondly, they created a mutant (T421M) that was not able to bind to the PLX inhibitors, but was able to bind to sorafenib.
In this mutant, they showed that the PLXs were not able to activate the pathway, but sorafenib was able to acitvate it.
This meant that direct binding of the inhibitor to the ATP binding site of the kinase was essential for the activation of the pathway.

Next, they tested whether the kinase activity was required for the pathway acitvation.
They generated a JAB sensitive mutant which is also a kinase dead mutant, and co-transfected this mutant with various cells (e.g. WT, kinase dead, etc.).
To these cells, they added JAB inhibitor which is going to inhibit the JAB sensitive but not the WT CRAF.
The results showed that when it was co transfected with the WT kinase, the pathway was activated, but was not activated when together with either a dimer mutant (R401H) or the kinase dead (D486N) mutant.
This suggested that the pathway was activated by transactivating the kinase partner, rather than the kinase that gets inhibited by the inhibitor, and that dimerisation is essential for the transactivation.

Lastly, since the dimerisation of the RAF kinases are RAS dependent, they tested whether over activation of RAS would cause the re-activation of the pathway in V600E cells.
As predicted, the pathway was activated even with the inhibitor added to the cells, and showed that the drug resistance can be accomplished by overactivating the RAS.

\subsection*{\normalsize{Hu et al 2013}}

Previous studies have shown that it was the transactivation that caused the pathway activation, and this activation required dimerisation, kinase activity in the transactivating partner, and a competitive inhibitor binding.
RAF is known to have two phosphorylation sites that must be phosphorylated in order to activate itself.
One of these sites is in/at the activation loop, and the other phosphorylation site is at the N terminal acidic (NtA) motif.
This paper showed that the N term acidic residues are essential for this transactivation.

\noindent\rule{\textwidth}{0.4pt}

They first showed that the phosphorylation of the NtA motif was required for the dimer dependent transactivation.
The serine residue in the NtA (SSDD residues) of the BRAF kinase are constitutively phosphorylated, and mutation in these residues lead to decrease pathway activation.
In contrast, the corresponding residues in CRAF (SSYY) are not constitutively phosphorylated, and this phosphorylation is required for the activation.
Mutation that mimicked the phosphorylation state (changing it to acidic residues) stimulated the pathway, demonstrating that the phosphorylation of these residues are essential for pathway activation.
Furthermore, when the RAFs were mutated so that they were unable to form dimers, this transactivation process was abolished, even with the constitutively transactivating mutants.

Secondly, they tested whether this NtA phosphorylation was required by both the activator and the receiver (where activator is kinase dead, but receiver is not).
The results showed that the receiver with mutated NtA was still able to be transactivated by an activator, suggesting that the phosphorylation of NtA is necessary on the activator, but not on the receiver.

Next, they tested the relevance of the conserved tryptophan after the NtA motif.
Substitution mutation of this tryptophan impaired the ability for the BRAF/CRAF to be an activator or a receiver.
They showed that this impairment was not due to the lack of dimerisation, and concluded that this tryptophan was important to place the NtA correctly at the dimer interface.

They also showed that the dimerisation of the activator and the receiver resulted in the phosphorylation of the activation loop.
This result suggested that, since the activator is kinase dead, the activation loop phosphorylation is from the auto phosphorylation by the receiver.
This meant that, in order to autophosphorylate itself, the kinase must be in the active conformation before this phosphorylation, and that the transactivation process changes the conformation into the active conformation.

They hypothesised that the NtA motif placement caused the hydrophobic regulatory spine (R spine) to change into the active DFG ``in'' positionm rather than the DFG ``out'' inactive conformation.
As predicted, the mutation that constitutively changed the R loop into the active DFG in conformation activated itself without dimer formation and/or NtA phosphorylation.

Finally, they showed that the NtA phosphorylation was MEK dependent.
This meant that BRAF activates B/CRAF via transactivation, which phosphorylates MEK.
Phosphorylated MEK then in turn phosphorylates CRAF at the NtA motif, which makes it an activator.

\section*{Tumour Resistance}

\subsection*{\normalsize{Thakur et al 2013}}

They used V600E primary tumour from humans and transplanted/xenografted it into mice, and the applied vemurafenib until resistance developed.
This resistant tumour was transplanted into a different mouse and used to test the effect of drugs and compared it with the WT tumour that is sensitive to the drug.

\noindent\rule{\textwidth}{0.4pt}

First, they showed the tumour responded to the drug (vemurafenib) in a dose dependent manner, but drug resistant tumour did not.
They also showed that the resistant tumour had constantly higher pERK level compared to the primary tumour, and still responded to the drug, but less effective.
This suggested that the V600E tumour still required this mutation (i.e. BRAF V600E) for its growth, and that the resistance was not due to mutations in RAS or MEK (since no secondary mutations were found).
They showed that this resistance was not due to the splice variant of BRAF, but due to the overexpression of BRAF by elevated mRNA and/or copy number variation of the BRAF gene.

Next, to test whether elevated level of BRAF was the cause, they made a cell culture with the primary tumour cells.
In doing so, they showed that either too much or too little vemurafenib caused a decrease in the tumour growth, showing that the tumour was reliant on the drug.
After they were able to grow the tumour cells, they tested whether the BRAF knockdown had an effect on growth.
They found that complete knockdown of BRAF resulted in the suppression of proliferation but partial suppression of BRAF (to the level of wild type) caused re-sensitization to the drug.
This result showed that the tumour resistance was due to increased levels of BRAF and is also dependent on the signal made from V600E mutant, rather than another mechanism (e.g. RAS/MEK mutations).

They showed that the drug withdrawal, although it increased the pERK level, decreased the growth of the tumour.
This decreased growth was reverted/overcome after a while as it got used to the drug concentration.
This showed how the tumour showed fitness deficit in the absence of the drug (i.e. reliant on the drug).
The withdrawal symptoms were applicable to other BRAF mutant cell lines, proving that these tumours were reliant on the drug.

Finally, they showed that intermittent drug application increased the period until drug resistance emerged.
This signified how drug treatment plans could be altered to prevent rapid drug resistance.

\noindent\rule{\textwidth}{0.4pt}

\begin{enumerate}
\item Vemurafenib resistance of V600E tumours rely on the re-activation of the ERK pathway via BRAF, even though that is the component that is being inhibited.
This was achieved by increasing the amount of BRAF in the cell and/or the up/downstream components of the pathway.
\item Fitness deficit from drug withdrawal was due to the elevated ERK activation.
This lead to the decrease of cell division/cycle and decreased apoptosis.
\item Emergence of drug resistance can be controlled by intermittent drug application/administration.
\end{enumerate}

\subsection*{\normalsize{Lito et al 2012}}

This paper focussed on the V600E mutant cells to figure out the mechanism in which the resistance to drugs arises.

\noindent\rule{\textwidth}{0.4pt}

Firstly, they found that in V600E cells the RAS level was very low compared to the normal cells.
This meant that V600E acts independently of RAS, due to the overactive negative feedback mechanism by ERK (via Spry2 and DUSP6 inhibits the signal pathway from RTKs to RAS).

Secondly, they showed that the activation of RAS was followed by ERK activation, even though vemurafenib was added.
This was due to the relief of the ERK mediated feedback inhibition of RAS (by constitutively active BRAF), which leads to the increased activity of RAS.
The activation of RAS causes the BRAF to dimerise with CRAF and transactivate it, which can stimulate the pathway.

From this result, they hypothesised that if the re-activation was caused by the transactivated CRAF, then a MEK inhibitor should be able to inhibit this re-activated pathway.
In fact, MEK inhibitor was able to inhibit the pathway.
They also showed that the combination of the RAF and MEK inhibitor was able to reduce the pathway activation significantly, compared to either drug alone.

ERK over activation by V600E caused the de-sensitization of the receptor's ability to recognise the external signal, and so hypothesised that the relief of ERK mediated feedback inhibition would re-sensitize the receptor to signals.
They proved this by showing that ERK was phosphorylated after 2 hours of vemurafenib treatment with EGF and NRG.
This showed that the rebound of the pathway was partly due to the presence of the ligands that stimulated the pathway and therefore re-activation of the pathway via RAS.

\section*{Unravelling Complex Pathway}

\subsection*{\normalsize{Toettcher et al 2013}}

This study used optogentically activatible SOS that specifically activated RAS to determine what specific downstream component is activated by RAS.
There are 2 models of signalling:
\begin{description}
\item[Combinatorial] is where different stimuli from different signal(s) in combination leads to a downstream effect, which is specific to the combination of the stimuli.
\item[Dynamic] is where the period of time in which the component is stimulated by a particular signal may be detected and cause a downstream effect.
Since different signals will have different different timescale, it will be detected slightly differently.
\end{description}
The output was measured as the amount of BFP-ERK in the nucleus after stimulus, as well as the recruitment of YFP-SOS to the membrane.

\noindent\rule{\textwidth}{0.4pt}

They showed that optogentics approach was able to mimic (and comparable to) the signal stimuli such as PDGF and NGF.
There were high variability within the population, but the signal/response from optogentics was consistent in a single cell (likely due to the variable expression between the cells).
With this approach, they showed that the RAS specifically activated MAPK/ERK pathway, but not the PI3K pathway, suggesting that PI3K pathway activation requires more than RAS by itself.

They also showed that the RAS signalling was a high bandwidth and low pass filter pathway, meaning that RAS/ERK pathway is able to respond to the stimulus over a broad range of time period (from 4min to 2hrs) and filters the transient minute scale inputs (hence low pass filter).
This means that only long lasting stimuli are transmitted efficiently through RAS/ERK pathway and short stochastic stimuli are ignored.

From the high-throughput protein array screen, they managed to identify STAT3 being activated as a result of prolonged (more than 2hrs stimulus) RAS, and therefore ERK activation.
Furthermore, they showed that STAT3 activation resulted from the release of a cytokine from the cell that has active ERK pathway and affects the nearby cells in paracrine fashion.
However, they showed that the cytokine released did not activate STAT3 in the cells that produced it (i.e. did not act in an autocrine fashion).
This was possible in a population of cells, as some cells had lower expression levels than other cells, allowing the STAT3 sensitivity even if it was activating ERK pathway (by a small amount).

\subsection*{\normalsize{Dar et al 2012}}

First of all, they noted that the most successful drugs for cancers are the ones that target more than one part of the pathway.
However, these discoveries were due to luck more than the design of the drug.
So, they proposed a whole-body/phenotypic screening method using Drosophila that allowed them to identify these poly pharmacological drugs that target multiple pathways.

\noindent\rule{\textwidth}{0.4pt}

They made mutant flies such that the flies were expressing the oncogenic receptor tyrosine kinase RET.
They calibrated these flies so that 50\% of the flies were able to develop into pupae, but none were able to grow into adult flies.
Thus, to get adult flies, this RET receptor had to be inhibited by a drug.
They discovered AD57 which could rescue the flies, and also found AD58 to have a toxic effect (i.e. made it worst).

They focussed on these drugs to identify why one was such a good drug, whereas the other drug was not.
They showed that different drugs inhibited various pathway by a different degree.
This provided some targets/anti-targets to test why the inhibition of some pathway led to toxicity (AD58), but not in the other (AD57).

They found that AD58 toxicity was due to the over activation of the ERK pathway by relieving the inhibition of RAF by dTor.
Since AD58 inhibits dTor, RAF is not inhibited by it, and therefore leads to the activation of ERK.
Since AD57 also inhibited dTor, they tested whether making a drug that did not inhibit dTor would make AD57 a better inhibitor.
AD80 was designed as a result, and it worked the best out of all the drugs identified, and it inhibited all of the pathways without significantly activating other pathways (e.g. by dTor).

\section*{Insight into the complex mechanism of the ERK pathway}

\subsection*{\normalsize{Haling et al 2014}}

This paper showed the additional mechanism of how the RAF/MEK/ERK pathway is activated/regulated.

\noindent\rule{\textwidth}{0.4pt}

They showed that in WT  or RAS mutant cells, the RAF/MEK complexes are enriched, suggesting that RAF is mainly in a complex with MEK in the cytosol and is quiescent and awaiting activation.
They also showed that the BRAF dimerisation was required for the interaction with RAS-GTP, but not with MEK, meaning that the complex is quiescent until activation by RAF dimerisation through RAS-GTP.

They noted that in the RAF/MEK complex, MEK adopts an inactive conformation but RAF is in an active conformation where the MEK activation loop is able to access the active site of the RAF, ready for phosphorylation.
This phosphorylation event may lead to the dissociation of the complex due to the steric and electrostatic interactions.

Mutations that prevented the RAF/MEK complex interaction caused a decrease in the ERK phosphorylation.
This was due to the inability of MEK to interact with RAF, and therefore no phosphorylation event, eventually leading to no activation of ERK.

The P-loop mutation in BRAF resulted in the weakening of the RAF/MEK complex formation, and therefore increased the BRAF/CRAF dimerisation (instead with MEK), and as a consequence, lead to the activation of the pathway via CRAF transactivation.

Since RAF is in an active conformation when bound to MEK, they tested whether the conformation affected the complex formation.
Inhibitors such as vemurafenib that keeps RAF in an inactive conformation increased the ERK activation (by promoting the dissociation from MEK and transactivation of CRAF).
However, mutations that kept the active conformation was able to inhibit the activation (as the complex was still held in place).

\noindent\rule{\textwidth}{0.4pt}
\begin{enumerate}
\item Normally, BRAF is in complex with MEK in the cytosol and is primed but sequestered.
\item Extracellular signal causes dimerisation of BRAF with BRAF/CRAF and allows catalytic activity and phosphorylates MEK (which is already attached to RAF).
\item Phosphorylated MEK then dissociates from the RAF and phosphorylates ERK, which activates the pathway.
\end{enumerate}

\end{document}

### 2015-10-16 committee_report/committee_report.tex
changed font to Times New Roman
finished writing draft of committee meeting

--- 2015-09-29 committee_report/committee_report.tex
@@
 
+\usepackage{graphicx}
+\usepackage{subcaption}
+\usepackage[labelfont=bf,font=footnotesize]{caption}
 \usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage{natbib}
@@
 \usepackage{amsmath}
+\usepackage{mathptmx}
 
@@
 
+\defcitealias{TCGA}{NCI and NHGRI, 2015}
+\defcitealias{ICGC}{ICGC, 2015}
+\defcitealias{GEO}{NCBI, 2015}
+
+\setlength{\bibsep}{5pt plus 0.3ex}
+
+\graphicspath{{./CommitteeMeeting2015/}}
+
 \begin{document}
@@
 \Large{\textbf{Body Mass Index and Pathway Dysregulation in Cancer}}\\
+%\end{center}
+
+%\begin{center}
+\normalsize{
+    Riku Takei (MSc)\\
+        Supervisor: Assoc. Prof. Mik Black
+}
 \end{center}
 
-\begin{center}
-Riku Takei (MSc)\\
-Supervisor: Assoc. Prof. Mik Black
-\end{center}
+\vspace{-30pt}
 
-\subsection*{Abstract (50-words)}
+\subsection*{50 Word Summary}
 
-Recent studies in breast cancer have shown that there are genetic signatures that are specific to obesity, and these signatures were further investigated in other cancer types (Creighton and fuentes mattei paper).
-The obesity genes identified by Creighton \textit{et al.} did not correlate with any of the cancer types investigated.
+Recent breast cancer studies have shown that there were obesity-specific genetic signatures.
+These signatures were further investigated in other cancer types.
+Obesity genes identified by previous studies did not correlate with any other cancer types investigated.
+
+\vspace{-11pt}
 
@@
 
-\newpage
+It is now commonly understood that cancers are caused by dysregulation of various mechanisms or pathways (known as the ``Hallmarks of Cancer'') that allow the cells to proliferate, survive and migrate \citep{Hanahan2011}.
+There have been many reports on the association between obesity and worsened cancer prognosis, suggesting that there may be distinct changes that underlie tumorigenesis and/or cancer progression that are specific to obese patients.
+The exact mechanism in which obesity accelerates cancer progression is unknown, but obesity is thought to disrupt the levels of various hormones (such as insulin, leptin, adeponectin, and other adipokines) that eventually activate tumorigenic pathways (for a quick overview, see the recent review by \citet{Renehan2015}).
+
+\citet{Creighton2012} identified 662 genes that were able to distinguish breast cancer from obese patients from those of non-obese patients.
+Furthermore, they showed that small molecules were able to induce or repress these genes, signifying the fact that these obesity specific genes could be used as an indicator for alternative or combinatorial chemotherapy and assist in better clinical decisions.
+
+More recent study by \citet{Fuentes-Mattei2014} showed that tumours in obese mice developed and progressed much faster than those in control mice.
+Differential gene expression analysis showed that the genes overexpressed were the target of protein kinase B (AKT), and the activation of the AKT/mammalian target of rapamycin (mTOR) was observed.
+Importantly, they were also able to control the AKT/mTOR pathway activation by using inhibitors that target the downstream effectors.
+
+Both studies showed that there were obesity specific genetic signatures in breast cancer, but can these genetic signatures be applied to other cancer types?
+Other studies have shown that there are genetic signatures common in multiple cancer types \citep{Alexandrov2013,Lawrence2014}.
+However, these studies have not shown any evidence of obesity-specific genetic signatures that are common across multiple cancer types.
+
+This research aims to determine whether gene expression signatures exist that are specific to obesity across multiple cancer types, and to investigate whether there are any common pathways being dysregulated in cancers based on these genetic signatures.
+Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to improved clinical decisions, and contribute towards personalised treatment in the future.
+
+%\newpage
+
+\vspace{-11pt}
 
@@
 
-The original CEL files of all the samples used by Creighton \textit{et al.} were downloaded from Gene Expression Omnibus (GEO) database (website ref).
-The microarray expression data for all 103 samples were loaded into R/RStudio and were RMA normalised using the package '????'.
+The original CEL files of all the samples used by \citet{Creighton2012}  were downloaded from Gene Expression Omnibus (GEO) database \citepalias{GEO}.
+The microarray expression data for all 103 samples were loaded into R/RStudio and were RMA normalised using the  ``affy'' package.
 The 799 obesity gene probes were extracted from the data and singular value decomposition was applied to this matrix to generate a metagene of the obesity genes.
 The metagene was compared with the BMI data of the samples which was plotted as a boxplot (BMI status) and a dotplot (BMI values).
+The same method was applied to \citet{Fuentes-Mattei2014} data.
 
@@
 
-The clinical data for all cancer types (33 in total) were downloaded from The Cancer Genome Atlas (TCGA) database and were checked whether the samples had both height and weight data.
+The clinical data for all cancer types (33 in total) were downloaded from The Cancer Genome Atlas (TCGA) database and were checked whether the samples had both height and weight data \citepalias{TCGA}.
 
@@
     \begin{equation*}
-    BMI = \frac{Weight(kg)}{Height^2(cm^2)}
+    BMI = \frac{Weight(kg)}{Height^2(m^2)}
 \end{equation*}
@@
 
-\item[Assessing whether the metagene is associated with BMI.]
-
-Once the metagenes were obtained for each cancer type, the metagene was plotted against BMI status and raw BMI value.
-
 \item[Processing the cancer data.]
 
-RNA-seq data for all of the cancer types with both height and weight data (8 in total) were downloaded from the International Cancer Genome Consortium (ICGC).
+RNA-seq data for all of the cancer types with both height and weight data (8 in total) were downloaded from the International Cancer Genome Consortium (ICGC) database \citepalias{ICGC}.
 All of these data were loaded into R/RStudio and were processed into a matrix form.
@@
 
+\item[Assessing whether the metagene is associated with BMI.]
+
+Once the metagenes were obtained for each cancer type, the metagene was plotted against BMI status and raw BMI value.
+
 \end{description}
 
-\newpage
+%\newpage
+
+\vspace{-30pt}
 
@@
 
-Since the project was based on the results presented by Creighton \textit{et al.} (2012), that there were obesity-specific genetic signatures in breast cancer, the methods used in this project was applied to the raw data from their paper and the results were compared.
+Since the project was based on the results presented by \citet{Creighton2012}, the methods used in this project was applied to the raw data from their paper and the results were compared to validate whether the methodolgy to be used in this project is applicable to other data.
 
-must not exceed 6 pages (with figures and references).
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.6\textwidth]{creightonsvdheatmap}
+\caption{\textbf{Heatmap of the gene expression of all the samples in Creighton \textit{et al.} data.}
+Histogram on the top left hand corner shows the scale of the gene expression level, where red  and blue corresponds to high or low expression, respectively.
+The single row on top of the heatmap represents the metagene produced from the singular value decomposition of the Creighton \textit{et al.} data.
+The graph to the left of the heatmap is the correlation tree of the genes, and the labels to the right of the heatmap are the genes used.
+The labels at the bottom of the heatmap are the TCGA sample ID, ordered by the metagene score (lowest score to the left, highest score to the right).}
+\label{crheatmap}
+\end{figure}
 
-%\citep{Hanahan2011}
+After the Creighton \textit{et al.} data were normalised using RMA method, singular value decomposition was applied to the data to obtain the obesity metagene and plotted on a heatmap (Figure \ref{crheatmap}).
+The resulting heatmap showed that the overall expression of the obesity-specific genes correlated with the metagene.
+To assess whether the metagene was actually correlating with the sample BMI, the metagene was plotted against the sample BMI status and sample BMI value (Figure \ref{crmetavsbmi}).
+These plots clearly showed that the metagene was able to identify and separate the obese samples from the normal and overweight samples.
+Together, these plots confirmed that the methodology applied to the data set was able to recapitulate the results presented by \citet{Creighton2012}.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.9\textwidth]{crmetavsbmi}
+\caption{\textbf{Box and dot plots of the Creighton \textit{et al.} metagene against BMI status and BMI value of the samples, respectively.}
+The y-axis represents the metagene score and the x-axis represents the BMI status (for box plot) and BMI values (for dot plot).
+Metagene was able to separate the obese samples from the normal/overweight samples, as shown in the box plot.
+}
+\label{crmetavsbmi}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.47\textwidth]{hm8-0}
+\includegraphics[width=0.47\textwidth]{out-0}
+\caption{\textbf{Heatmap of the gene expression of all the sample with BMI data in bladder cancer (BLCA) data using the transformation matrix from Creighton \textit{et al.} data  (left) and a box plot of the metagene against BMI status of the samples (right).}
+The parameters, scales and axes used in the heatmap are identical to that from Figure \ref{crheatmap}.
+The overall gene expression of the samples were captured by the metagene, as shown by the heatmap, but there was no correlation with the BMI status of the samples.
+Similar observations were made for other cancer types (data not shown).
+}
+\label{blcameta}
+\end{figure}
+
+The metagenes for other cancer types were generated by using the obesity-specific genes from breast cancer.
+Transformation matrix was applied to each cancer type to generate the metagene and was plotted on the heatmap (Figure \ref{blcameta}).
+All of the cancer types showed significant correlation with the metagene created from the transformation, where the high/low metagene values corresponded with high/low overall gene expression, respectively.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.47\textwidth]{out1}
+\includegraphics[width=0.47\textwidth]{out2}
+\caption{\textbf{Heatmap of the gene expression of all the sample with BMI data in bladder cancer (BLCA) data using the transformation matrix from Fuentes-Mattei \textit{et al.} data  (left) and a box plot of the metagene against BMI status of the samples (right).}
+The parameters, scales and axes used in the heatmap are identical to that from Figure \ref{crheatmap}.
+The overall gene expression of the samples were captured by the metagene, as shown by the heatmap, but there was no correlation with the BMI status of the samples.
+Similar observations were made for other cancer types (data not shown).
+}
+\label{blcameta}
+\end{figure}
+
+To confirm whether the association was in fact due to BMI of the samples, plots of metagene against BMI status and BMI value were created.
+In contrast to the  plots made from the Creighton \textit{et al.} data, no cancer types showed any significant separation between the obese samples and the normal/overweight samples (Figure \ref{blcameta}).
+These results suggest that the obesity-specific genes from breast cancer were not specific enough to be used as a genetic marker for other cancer types.
+
+More recent paper by \citet{Fuentes-Mattei2014} have also found  obesity-specific genes (111 unique genes) in breast cancer, and these genes were also investigated.
+Similar method was applied to the Fuentes-Mattei \textit{et al.} data and metagenes were obtained for each cancer type.
+Again, the metagene seemed to correlate with the gene expression level of the cancer types, but there was no association with BMI value or BMI status.
+However, it is interesting how the metagenes created from both Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} data were able to capture the overall gene expression profile, but not BMI.
+It may be that the obesity-specific genes identified by both groups may have identified genes that were specific to a different clinical variable and/or molecular marker that was somehow related to the sample BMI, but not BMI itself.
+Further analyses must be conducted to conclude these results with certainty.
+
+Although current results do not show any significant correlation of the obesity-specific genes and the sample BMI in different cancers, there are many places for improvements.
+Firstly, the data processing of the raw RNA-seq data was carried out using a program written by myself which could have processed the data incorrectly.
+There are other programs to process these raw data, so the data processing procedure should be repeated and compared with the current results.
+
+Secondly, the obesity-specific genes specific to each cancer type were not checked with the BMI values of the samples.
+The results presented here only showed the association of the metagene (created from a different cancer type and data) with the BMI value and status, and it would be good to double-check whether the expression of the genes show any correlation with the BMI of the samples in other cancer types.
+
+For future experiments, differential expression analysis for each cancer type should be carried out to identify the genes that are dysregulated in obese samples but not in the normal or overweight data.
+These genes can then be compared with the genes from other cancer types to identify genes that are  commonly up or down regulated in different cancer types, which are specific to BMI status.
+
+\vspace{-11pt}
 
 %%Referencing:
-%\bibliographystyle{BiocRefStyle}
-%\bibliography{../../../References/BibTeX/MSc}
+\bibliographystyle{BiocRefStyle}
+\bibliography{../../../References/BibTeX/MSc,../../../References/BibTeX/nonpapercitations}
 

### 2015-10-16 committee_report/committee_report.tex
minor edit

@@
 Secondly, the obesity-specific genes specific to each cancer type were not checked with the BMI values of the samples.
-The results presented here only showed the association of the metagene (created from a different cancer type and data) with the BMI value and status, and it would be good to double-check whether the expression of the genes show any correlation with the BMI of the samples in other cancer types.
+The results presented here only showed the association of the metagene, created from a different cancer type and data (Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} data), with the BMI value and status, and it would be good to double-check whether the expression of the genes show any correlation with the BMI of the samples in other cancer types.
 

### 2015-10-19 committee_report/committee_report.tex
final copy

--- 2015-10-16 committee_report/committee_report.tex
@@
 
-Recent breast cancer studies have shown that there were obesity-specific genetic signatures.
-These signatures were further investigated in other cancer types.
-Obesity genes identified by previous studies did not correlate with any other cancer types investigated.
+Recent breast cancer studies have demonstrated the existence of that there were obesity-specific genetic signatures.
+In this work, these signatures were investigated in other cancer types.
+Obesity genes identified by previous studies did not correlate with obesity in any other cancer types investigated.
 
@@
 
-It is now commonly understood that cancers are caused by dysregulation of various mechanisms or pathways (known as the ``Hallmarks of Cancer'') that allow the cells to proliferate, survive and migrate \citep{Hanahan2011}.
+It is now commonly understood that cancers are caused by dysregulation of various mechanisms or pathways (known as the ``Hallmarks of Cancer'') that allow tumour cells to proliferate, survive and migrate \citep{Hanahan2011}.
 There have been many reports on the association between obesity and worsened cancer prognosis, suggesting that there may be distinct changes that underlie tumorigenesis and/or cancer progression that are specific to obese patients.
-The exact mechanism in which obesity accelerates cancer progression is unknown, but obesity is thought to disrupt the levels of various hormones (such as insulin, leptin, adeponectin, and other adipokines) that eventually activate tumorigenic pathways (for a quick overview, see the recent review by \citet{Renehan2015}).
+The exact mechanism in which obesity accelerates cancer progression is unknown, but obesity is thought to disrupt the levels of various hormones (such as insulin, leptin, adeponectin, and other adipokines) that eventually activate tumorigenic pathways (for an overview, see the recent review by \citet{Renehan2015}).
 
-\citet{Creighton2012} identified 662 genes that were able to distinguish breast cancer from obese patients from those of non-obese patients.
+\citet{Creighton2012} identified 662 genes that were able to distinguish breast tumour from obese patients from those of non-obese patients.
 Furthermore, they showed that small molecules were able to induce or repress these genes, signifying the fact that these obesity specific genes could be used as an indicator for alternative or combinatorial chemotherapy and assist in better clinical decisions.
 
-More recent study by \citet{Fuentes-Mattei2014} showed that tumours in obese mice developed and progressed much faster than those in control mice.
+A more recent study by \citet{Fuentes-Mattei2014} showed that tumours in obese mice developed and progressed much faster than those in control mice.
 Differential gene expression analysis showed that the genes overexpressed were the target of protein kinase B (AKT), and the activation of the AKT/mammalian target of rapamycin (mTOR) was observed.
-Importantly, they were also able to control the AKT/mTOR pathway activation by using inhibitors that target the downstream effectors.
+Importantly, they were also able to control AKT/mTOR pathway activation by using inhibitors that targeted  downstream effectors.
 
 Both studies showed that there were obesity specific genetic signatures in breast cancer, but can these genetic signatures be applied to other cancer types?
-Other studies have shown that there are genetic signatures common in multiple cancer types \citep{Alexandrov2013,Lawrence2014}.
-However, these studies have not shown any evidence of obesity-specific genetic signatures that are common across multiple cancer types.
+Other studies have shown that there are genetic signatures common in multiple cancer types \citep{Alexandrov2013,Lawrence2014}, however, these studies have not shown any evidence of obesity-specific genetic signatures that are common across multiple cancer types.
 
 This research aims to determine whether gene expression signatures exist that are specific to obesity across multiple cancer types, and to investigate whether there are any common pathways being dysregulated in cancers based on these genetic signatures.
-Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to improved clinical decisions, and contribute towards personalised treatment in the future.
+Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to improved clinical decisions, and could thus contribute towards personalised treatment in the future.
 
@@
 
-The original CEL files of all the samples used by \citet{Creighton2012}  were downloaded from Gene Expression Omnibus (GEO) database \citepalias{GEO}.
+The raw gene expression data files of all the samples used by \citet{Creighton2012}  were downloaded from the  Gene Expression Omnibus (GEO) database \citepalias{GEO}.
 The microarray expression data for all 103 samples were loaded into R/RStudio and were RMA normalised using the  ``affy'' package.
-The 799 obesity gene probes were extracted from the data and singular value decomposition was applied to this matrix to generate a metagene of the obesity genes.
+The 799 obesity-related  gene probes were extracted from the data and Singular Value Decomposition was applied to this matrix to generate a metagene summary of these probes.
 The metagene was compared with the BMI data of the samples which was plotted as a boxplot (BMI status) and a dotplot (BMI values).
@@
 
-The clinical data for all cancer types (33 in total) were downloaded from The Cancer Genome Atlas (TCGA) database and were checked whether the samples had both height and weight data \citepalias{TCGA}.
+The clinical data for all cancer types (33 in total) were downloaded from The Cancer Genome Atlas (TCGA) database and were checked to determine whether the samples had both height and weight data \citepalias{TCGA}.
 
@@
 
-Transformation matrix was used to obtain the metagene for the other cancer data.
-In order to transform the data, the transformation matrix from the original data must have the same number of genes as the transforming data, so the genes present in both the Creighton \textit{et al.} data and all of the other cancer types were identified.
-Transformation matrix was made from the Creighton \textit{et al.} data using these common genes, and then this transformation matrix was used to transform all of the cancer types.
+A transformation matrix was used to obtain the metagene for the eight downloaded cancer datasets.
+In order to transform the data, the transformation matrix from the original data must have the same number of genes as the new data, so the genes present in both the Creighton \textit{et al.} data and all of the other cancer types were identified.
+The transformation matrix was made from the Creighton \textit{et al.} data using these common genes, and then this transformation matrix was used to transform all of the cancer types.
 
@@
 The graph to the left of the heatmap is the correlation tree of the genes, and the labels to the right of the heatmap are the genes used.
-The labels at the bottom of the heatmap are the TCGA sample ID, ordered by the metagene score (lowest score to the left, highest score to the right).}
+The labels at the bottom of the heatmap are the sample ID, ordered by the metagene score (lowest score to the left, highest score to the right).}
 \label{crheatmap}
@@
 
-After the Creighton \textit{et al.} data were normalised using RMA method, singular value decomposition was applied to the data to obtain the obesity metagene and plotted on a heatmap (Figure \ref{crheatmap}).
+After the Creighton \textit{et al.} data were normalised using the RMA method, singular value decomposition was applied to the data to obtain the obesity metagene and plotted on a heatmap (Figure \ref{crheatmap}).
 The resulting heatmap showed that the overall expression of the obesity-specific genes correlated with the metagene.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{crmetavsbmi}
+\includegraphics[width=0.4\textwidth]{crflipped}
+\includegraphics[width=0.4\textwidth]{crflipped2}
 \caption{\textbf{Box and dot plots of the Creighton \textit{et al.} metagene against BMI status and BMI value of the samples, respectively.}
 The y-axis represents the metagene score and the x-axis represents the BMI status (for box plot) and BMI values (for dot plot).
-Metagene was able to separate the obese samples from the normal/overweight samples, as shown in the box plot.
+The metagene was able to separate the obese samples from the normal/overweight samples, as shown in the box plot.
 }
@@
 \centering
-\includegraphics[width=0.47\textwidth]{hm8-0}
-\includegraphics[width=0.47\textwidth]{out-0}
+\includegraphics[width=0.47\textwidth]{crmetastd1}
+\includegraphics[width=0.47\textwidth]{crmetastdclin1}
 \caption{\textbf{Heatmap of the gene expression of all the sample with BMI data in bladder cancer (BLCA) data using the transformation matrix from Creighton \textit{et al.} data  (left) and a box plot of the metagene against BMI status of the samples (right).}
@@
 \centering
-\includegraphics[width=0.47\textwidth]{out1}
-\includegraphics[width=0.47\textwidth]{out2}
+\includegraphics[width=0.47\textwidth]{fmmetaheat1}
+\includegraphics[width=0.47\textwidth]{fmmetaclin1}
 \caption{\textbf{Heatmap of the gene expression of all the sample with BMI data in bladder cancer (BLCA) data using the transformation matrix from Fuentes-Mattei \textit{et al.} data  (left) and a box plot of the metagene against BMI status of the samples (right).}
@@
 
-More recent paper by \citet{Fuentes-Mattei2014} have also found  obesity-specific genes (111 unique genes) in breast cancer, and these genes were also investigated.
-Similar method was applied to the Fuentes-Mattei \textit{et al.} data and metagenes were obtained for each cancer type.
+The recent paper by \citet{Fuentes-Mattei2014} also found  obesity-specific genes (111 unique genes) in breast cancer, and these genes were also investigated.
+Similar methods were applied to the Fuentes-Mattei \textit{et al.} data, and metagenes were obtained for each cancer type.
 Again, the metagene seemed to correlate with the gene expression level of the cancer types, but there was no association with BMI value or BMI status.
@@
 
-Although current results do not show any significant correlation of the obesity-specific genes and the sample BMI in different cancers, there are many places for improvements.
-Firstly, the data processing of the raw RNA-seq data was carried out using a program written by myself which could have processed the data incorrectly.
+Although current results do not show any significant correlation of the obesity-specific genes and the sample BMI in different cancers, there are many places for improvement.
+%Firstly, the data processing of the raw RNA-seq data was carried out using a program written by myself which could have processed the data incorrectly.
+Firstly, the data processing of the raw RNA-seq data was carried out using only one program, which may have processed the data incorrectly.
 There are other programs to process these raw data, so the data processing procedure should be repeated and compared with the current results.
 
-Secondly, the obesity-specific genes specific to each cancer type were not checked with the BMI values of the samples.
-The results presented here only showed the association of the metagene (created from a different cancer type and data) with the BMI value and status, and it would be good to double-check whether the expression of the genes show any correlation with the BMI of the samples in other cancer types.
+Secondly, the obesity-specific genes specific to each cancer type were not individually checked (i.e. per gene) with the BMI values of the samples.
+The results presented here only showed the association of the metagene, created from a different cancer type and data (Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} data), with the BMI value and status, and it would be good to double-check whether the expression of the individual genes show any correlation with the BMI of the samples in other cancer types.
 

### 2016-03-10 thesis/thesis.tex
starting my thesis - just the backbone structure for now

\documentclass[a4paper, 12pt, oneside]{report}

% Preamble:
\usepackage[T1]{fontenc}       % package for specifying output font encoding
\usepackage[utf8]{inputenc}    % package for specifying input font encoding
\usepackage{graphicx}          % package for graphics handling (i.e. figure images)
\usepackage{mathptmx}          % package for Times New Roman font
\usepackage{setspace}          % package for setting line space

% If you want to compile specific files, uncomment the line below and add the filenames in the {} :
%\includeonly{}

\begin{document}

\include{./title}

\pagenumbering{roman} % set page numbering to roman numeral

\include{./abstract}

\include{./acknowledgement}

\tableofcontents

\addcontentsline{toc}{section}{\contentsname}

\clearpage

\addcontentsline{toc}{section}{\listfigurename}

\listoffigures

\clearpage

\addcontentsline{toc}{section}{\listtablename}

\listoftables

% May need to include abbreviations and glossary here, before any chapter begins

\clearpage

\pagenumbering{arabic} % revert the page numbering to normal numbering

\include{./intro}

\include{./methods}

\include{./results}

\include{./discussion}

\end{document}

### 2016-03-10 thesis/title.tex
starting my thesis - just the backbone structure for now

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title page for my thesis - code is pretty self explanatory
% Required packages: setspace
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{titlepage}

\centering

\rule[2.0mm]{\textwidth}{0.5mm}

\begin{doublespace}
    {\Huge \scshape Body Mass Index and Pathway Dysregulation in Cancer}
\end{doublespace}

\rule[2.0mm]{\textwidth}{0.5mm}

\vspace{2.0mm}
{\Large \scshape Riku Takei}

\vfill

{\normalsize \scshape A Thesis Submitted For The Degree Of\\}
{\Large \scshape Master Of Science\\}
\vspace{2.0mm}
{\normalsize \scshape At The University Of Otago, Dunedin, New Zealand \\}

\vspace{10.0mm}
{\normalsize \today}

\end{titlepage}

### 2016-03-22 tips_and_notes/vim/intro.tex
started writing up my tips book

\chapter{Quick Introduction}

I used to use emacs during my undergrad years, but I never really mastered  it (or even become advanced at it).
The farthest I got with emacs was cutting/pasting lines, undoing mistakes, and moving cursor to the start/end of the line, which was enough to get through all the coding needs I had at the time.

I never hated emacs (I actually quite liked it), but I did notice the awkwardness of all the shortcuts that emacs provided for the user.
Most, if not all, shortcuts were prefixed with a \verb|<C-x>|, and some of the letters that followed were not as informative to remember.
Furthermore, emacs uses not only the control-key, but alt-key, shift-key, super-key, and other modifier keys as well (and each modifier key in combination with letters has different meanings!).
This really confused me (and annoyed me a bit), as you had to remember the right key combinations to carry out whatever you wanted it to do, and if you happened to accidentally press another modifier key, well, you are in for a surprise!

My point here is that emacs provides the user with so many shortcuts that will allow the user to do everything and anything by using all the different key combinations that you'll either forget or mistype.
There's just too many of them!!

When I began my postgrad study, I thought, ``If I was to keep coding throughout my study, I would need to be able to code faster''.
But when I started looking at the shortcuts that emacs provided, I was not motivated at all (for the very reason I mentioned earlier).
That is when I heard about vim and how efficient it was.

So, I started looking at how vim worked and what could be done with it.
What are the advantages?
How are they useful?
What can you do with it?
At first, I was weirded out by the way vim worked, like the modal interface, but I soon became used to it and agreed with the idea of not leaving the home row.
I also discovered multiple plugins that gave the features of the emacs that I loved in the days of undergrad comp sci courses.
The emacs features were based on the configuration file on their computer that I could not reproduce myself, hence I was getting frustrated with the raw emacs that I have installed.

As I read more and more about vim, the urge to learn vim properly became larger.
I had plenty of other options to learn (for example, another programming language), but I chose vim to be my first target - if I could master how to edit text faster, then whatever that follows will also (probably) become faster.
So, I decided to learn vim using the vim tips book ``Practical Vim'' written by Drew Neil.
This part will mostly contain the tips that I thought was very useful from Drew Neil's book, so if you want more details, go and read his book.

### 2016-03-22 tips_and_notes/vim/intro.tex
started vim tips intro

@@
 \chapter{Quick Introduction}
+
+\section{Why I started using Vim}
 
@@
 I also discovered multiple plugins that gave the features of the emacs that I loved in the days of undergrad comp sci courses.
-The emacs features were based on the configuration file on their computer that I could not reproduce myself, hence I was getting frustrated with the raw emacs that I have installed.
+The emacs features were based on the configuration file on the comp sci computer (which I could not reproduce myself), hence I was getting frustrated with the raw emacs that I have installed.
 
 As I read more and more about vim, the urge to learn vim properly became larger.
-I had plenty of other options to learn (for example, another programming language), but I chose vim to be my first target - if I could master how to edit text faster, then whatever that follows will also (probably) become faster.
+I had plenty of other options to learn before vim (for example, another programming language), but I chose vim to be my first target - if I could master how to edit text faster, then whatever that follows will also (probably) become faster.
 So, I decided to learn vim using the vim tips book ``Practical Vim'' written by Drew Neil.
-This part will mostly contain the tips that I thought was very useful from Drew Neil's book, so if you want more details, go and read his book.
+This part of the book will mostly contain the tips that I thought was, or could be, very useful from Drew Neil's book, so if you want more details, go and read his book.
+
+Also, just a note before I start - most sentences will probably be incomplete sentences (bullet-point like), as I am not writing this as a step-by-step guide to vim.
+This book is just a bunch of tips I pulled out from his book.

### 2016-03-23 tips_and_notes/vim/bys.tex
up to act, repeat, reverse

\section{Before you start}

Firstly, make sure you can touch type.
It doesn't have to be super fast or anything like that - just know where to find the keys that you are looking for (including special symbols like \%).
A good website I found was \url{www.keybr.com}.
It's free to sign up, and keeps all the record of what you have typed and what keys you need to work on.

Secondly, do the Vim tutor on command line (type \verb|vimtutor| in terminal).
This tutorial will get you through the basics of vim and how to get around files, save files, close files, etc.
Vim tutor will be a good starting point before using vim.

Lastly, have a google around for the spf13-vim plugin.
This plugin downloads other plugins that make vim awesome (e.g. tab completion).
Starting off with this plugin will make your life a lot easier, but remember to make your own .vimrc file once you get the hang of all the commands and (re-)mappings.

### 2016-03-23 tips_and_notes/vim/philosophy.tex
up to act, repeat, reverse

\chapter{The ``Vim Way''}

\section{Get used to the Modal Interface}

Vim is set up in such a way that every key on the keyboard has a different meaning depending on the current ``mode'' of the text editor.
There are three major modes in vim (all of which we will talk about later):
\begin{description}
    \item [Normal Mode] The default mode of vim for editing text.
    \item [Insert Mode] Primary mode for inserting/writing text into the buffer/file.
    \item [Visual Mode] Allows you to select text in a buffer/file. The selected text can be used for copy/paste/manipulation/substitution/etc.
\end{description}

Each mode has a specific purpose and the keys are laid out to make it easier to carry it out (although most keys have the same, if not similar, meanings -- which makes it easier to remember than emacs!).
For example in insert mode, all the keys are set out so that we can type in the buffer.
On the other hand, the primary purpose of normal mode is to move around the buffer and manipulate the text, so each key has a different meaning (e.g. h/j/k/l to move around).

This allows us to focus on one task at a time, and also allows us to choose the right tool to carry out specific tasks.

\section{Never leave the home row}

One of the most annoying thing while programming is having to go back and forth between the mouse and the keyboard.
For example, you start coding for a bit, and you found a wee bug few line up the screen.
You pause and use your mouse to click where you want to delete/add text, fix your mistakes and click back to your original position to carry on coding.

This doesn't sound too bad.
But what if you compiled your program and found multiple errors within a file and also in other files as well?
Sure, you can open multiple screens or windows of your favourite text editor and use your mouse to scroll up and down the screen, clicking, and selecting for the piece of text you want to edit.
Important thing to note here is that every single time you leave the keyboard to reach for the mouse, you are not typing.
Going back and forth might not take up that much time, but it will add up.
Also, you'll probably get annoyed if you had to change a tiny bit of text at a time, especially if you had hundreds of them to go through.

Luckily, vim has a modal interface -- modes like normal mode and visual mode are designed so that you don't have to reach for your mouse (or the arrow keys) as much, if not at all.
We will meet these modes later on.

\section{Act, Repeat, Reverse}

### 2016-03-23 tips_and_notes/vim/chapter2/philosophy.tex
up to chapter 3, modes; separated all the chapters in different directories

--- 2016-03-23 tips_and_notes/vim/philosophy.tex
@@
 \section{Act, Repeat, Reverse}
+
+When using vim, everything should be set up so that you can ``act, repeat, reverse''.
+What I mean by this is that you should use vim in such a way that you edit a bit of text (act), then repeat the same edit on different parts of the text (repeat), and make all of his reversible (reverse).
+We will meet the dot-command in the next chapter.
+
+When going through this book, you should always think about how you are going to type less, automate tasks, and how to prevent yourself typing the same text over and over again.
+The main goal is to do a single edit, then make this repeatable so that you don't have to re-type it again later, and if you happen to carry out the changes too many time, you should be able to reverse the change.
+
+\section{Edit your text in chunks}
+
+In vim, it is easy to write everything in one single burst in the insert mode (like in other text editors and MS Word), but this should be avoided.
+Since vim's undo command (\verb|u|) reverts the last change made to  the text, if you type everything all at once you run the risk of undoing the whole text.
+Rather than doing this, you should try and  change the text bit by bit to make sure your undo command doesn't undo everything, but only the small incremental changes you have made to the document.
+
+\section{Navigate in chunks}
+
+Hopefully, you know how to navigate simply in vim by now (the \verb|hjkl| keys).
+When you're navigating yourself in vim, you should try and use more keys that skips more than one letter (i.e. use \verb|h| and \verb|l| keys less).
+You should use the keys that allow you to skip multiple words or lines at a time, like the \verb|w/b| keys and \verb|<ctrl-f>/<ctrl-b>| keys.

### 2016-03-24 tips_and_notes/vim/chapter3/normal.tex
pre-leaving for chch

\section{Normal Mode}

Normal mode can be accessed by pressing  the \verb|<esc>| or \verb|<ctrl-[>| command, and is probably the mode that you will be in most of the time.
This is because most of your editing will start in this mode, then switch into another mode to complete your editing task.
In some cases, you probably won't have to leave the normal mode at all (for example deleting a word or a line).

An analogy used in the book was of a painter -- painters, when drawing up a nice piece of artwork, do not leave their paintbrush on the canvas.
Instead, they rest their paint brush, look at it and think about it for a while, and then decide what to do next.
Whatever they do next can be anything from picking a different colour of paint to paint in, or it could be a different kind of brush to give different texture.
Vim's normal mode works just like that -- it gives you ``resting'' mode to think about what to write, what words to edit, how to edit, etc.

\subsection{Chunk your undos}

As mentioned already, you should chunk your undos as finely as you can by going back into normal mode.
Vim defines a ``change'' as any change in the text.
This could range from deleting a single letter to deleting the whole document (and everything you can think of in between those range).
When using insert mode, vim will define the change as whatever you type in insert mode until you go back into normal mode (i.e. until you have pressed the \verb|<esc>| key).

You should find your own definition of a chunk you are comfortable with and stick with it.
For example, I usually write a sentence up (or part of a sentence up), then always go back to normal mode to keep line-wise chunks rather than any smaller/bigger chunks.

By having smaller chunks for undoing your edit, you are able to go back to your previous text in a finer detail, rather than having to undo a huge chunk and rewriting most of the text you have just undone.

\subsection{Compose repeatable changes: Meet the Dot Command}

When you make a change to your text/document, vim allows you to repeat this change as well as undoing the change (as in previous section).
This means that any changes you make to the text can be repeated on a different part of the text, and therefore won't have to re-type the exact change or commands you have typed earlier (never retype!!).

### 2016-03-30 tips_and_notes/vim/chapter3/command.tex
onto Ex mode

\section{Command Line (Ex) Mode}

This mode is essentially an extension of normal mode, but provides vim with more functionality and super effective at editing the whole document (if used correctly).
With normal mode by itself, you can only edit so much with the given operator and motion keys.
You can increase and specify the range with visual modes, but this will only affect locally, within the selection.
If you wanted to edit the whole document, you can use \verb|G| as  the motion, but the operator keys may not be specific enough to do the task you want it to do.

Ex mode provides the minor adjustments and specificity required by you to carry out the exact task you want vim to do for the whole document, or just part of it.

### 2016-03-30 tips_and_notes/vim/chapter3/insert.tex
onto Ex mode

\section{Insert Mode}

This mode will be self explanatory -- it is the mode you type text into the document.
However, there are tricks and work-arounds that you can use in insert mode so that you don't have to always go back into normal mode.
To go back into normal mode, you can either press \verb|<esc>| or \verb|<ctrl-[>|.

\subsection{Deleting from insert mode}

\begin{tabular}{c|l}
    \verb|<ctrl-h>| & delete one character\\
    \verb|<ctrl-w>| & delete one word\\
    \verb|<ctrl-u>| & delete one line\\
\end{tabular}

\subsection{Insert-normal mode}

Insert-normal mode is a mode that allows you to go back into normal mode for just one command, and then switches automatically back into insert mode.
To use it, while in insert mode, press \verb|<ctrl-o>|.
This will put you into insert-normal mode, and after you have entered a command in this mode, it will switch you back into insert mode.
\verb|zz| is a normal mode command that redraws the screen so that the current line is positioned in the middle of the screen.
This can be quite useful in insert-normal mode, especially if you want to reposition the screen and get back into the typing immediately.

\subsection{Paste text while in insert mode}

You can paste text from certain registers by using \verb|<ctrl-r>{register}|.
This will paste the text stored inside the register that you have specified into the current cursor position.
Use \verb|<ctrl-p>{register}| to paste the text literally without indentation, and other nonsense (e.g. carriage return).

\subsection{Use expression register to insert calculations and other complicated stuff}

The ``='' symbol is a special register that allows you to paste the result of the expression typed in.
When you type \verb|<ctrl-r>=|, this will open up a prompt which you can type expression into.
For example, you could type 3576*235 into this, and once you hit the enter button, the result should be inserted into the text (by the way, the answer to that equation is 840360).

### 2016-03-30 tips_and_notes/vim/chapter3/normal.tex
onto Ex mode

@@
 Vim's normal mode works just like that -- it gives you ``resting'' mode to think about what to write, what words to edit, how to edit, etc.
+
+\subsection{Operator + Motion = Action}
+
+This formula is the most powerful tool and rule you will use in normal mode.
+In normal mode, there are keys known as the operator keys, and keys known as the motion keys.
+Together, they can perform an action over certain parts of text, depending on the combination of the operator and the motion keys you have pressed.
+
+\subsubsection{Operator keys}
+
+Operator keys are the core components of all the editing commands you carry out in normal mode -- and there are a lot of them!
+The easiest way to learn them (other than using the keys repeatedly) is to learn them by their mnemonics (or maybe what they look like).
+\newline
+
+\begin{tabular}[c]{c|l}
+    c            & change\\
+    d            & delete\\
+    y            & yank\\
+    g$\sim$      & swap case\\
+    gU           & make uppercase\\
+    gu           & make lowercase\\
+    =            & autoindent\\
+    \textgreater & shift to right\\
+    \textless    & shift to left\\
+\end{tabular}
+\newline
+
+Also note that vim will wait for you to press a motion key after you have pressed the operator key -- this is called the ``operator-pending mode''.
+This means that you can press the operator key and have enough time to think about what motion you will need to do the right edit.
+
+\subsubsection{Motion keys}
+
+Motion keys are the keys that allow you to move around the document while in normal mode.
+It is also the means to specify how far the operator commands should take action on.
+Like the operator keys, you should learn it by their mnemonics.
+\newline
+
+\begin{tabular}[c]{c|l}
+    h/j/k/l & default cursor movement\\
+    w/W     & word or WORD\\
+    b/B     & go back a word or WORD\\
+    e/E     & go to the end of the word or WORD\\
+    p       & paragraph\\
+    0/\$    & beginning/end of the current line\\
+    gg/G    & start/end of the current file\\
+    t/T     & (un)til the character, searching forward/backward\\
+    f/F     & find the character, searching forward/backward\\
+\end{tabular}
+
+\subsubsection{Text objects}
+
+Text objects define a regions of text by structure, and consists of two characters.
+The first character is either an ``i'' or an ``a'', and the second character can be anything that describes either a bounded or delimited text object.
+
+\begin{description}
+    \item [Bounded text objects] are defined by boundaries (e.g. words, sentences, and paragraphs).
+    \item [Delimited text objects] are anything in between a pair of matching symbols (e.g. (), [], \{\}, and xml tags).
+\end{description}
+
+The ``i'' stands for inside, whereas ``a'' stands for around/all.
+With the bounded text objects, ``i'' will capture the current word and ``a'' will capture the word with a single white space (if any).
+With the delimited text objects, ``i'' will capture everything inside the specified matching symbols and ``a'' will capture everything inside the matching symbols and the symbols you have specified.
+Together with other motion keys or specific matching symbols, text objects can be very useful in defining and capturing parts of the texts that you want to edit.
+
+\subsubsection{Action}
+
+When you combine the operator and the motion keys and/or text objects together, it will edit the document as demanded.
+For example, \verb|dw| will delete the word (from the current position of the cursor to the end of the word), and \verb|cw| will change the word (from the current position of the cursor to the end of the word).
+If you use the text object, it's even better -- it will delete/change the word your cursor is currently positioned, instead of going from the current cursor position to the end of the word.
+
+If you press the operator key twice (e.g. \verb|dd|), it will do the action on the current line.
+So, \verb|dd| will delete the current line, and \verb|>>| will indent the current line to your right.
 
@@
 When you make a change to your text/document, vim allows you to repeat this change as well as undoing the change (as in previous section).
-This means that any changes you make to the text can be repeated on a different part of the text, and therefore won't have to re-type the exact change or commands you have typed earlier (never retype!!).
+This means that any changes you make to the text can be repeated on a different part of the text, and therefore won't have to re-type the exact change or commands you have typed earlier.
+Ideally, you should type one editing command (e.g. daw to delete a word) then repeat this command by using the dot (\verb|.|).
+
+When you press \verb|.|, vim repeats the last editing command that altered the text.
+For example, if you opened a new line and typed ``hello'' then escaped, the dot command will repeat your edit -- i.e. it will make a new line saying ``hello''.
+Making a new line saying ``hello'' is not that helpful, but with other useful editing commands, the dot command can be very useful.
+
+For example, if you wanted to make a word uppercase, you can do so by typing \verb|gUaw| or \verb|gUw| while the cursor is on the word you want to uppercase.
+What if you wanted to uppercase a bunch of words, spread out throughout the document?
+You are not going to retype \verb|gUw| every single time, but instead you can move to the word you want to change the case and press the \verb|.| -- it will remember the last editing command (\verb|gUw|) and uses it on the word.
+Pretty neat, huh?
+
+\subsection{Use counts to do simple addition and subtraction}
+
+If you have numbers in your document which you would like to increment or decrement, use \verb|<ctrl-a>| and \verb|<ctrl-x>|.
+When you have the cursor on the number, press \verb|<ctrl-a>/<ctrl-x>| to increment/decrement respectively.
+If you are at the beginning of the line, pressing these keys will jump to the first number (if there is any), and does the incrementing/decrementing for you.
+
+\subsection{Combine numbers with commands}
+
+If you know exactly how many words/lines/paragraphs to edit, you can prefix the command with numbers to repeat the commands n times.
+For example, when you type \verb|c2w|, you would drop into insert mode after deleting two words that are ahead of the cursor.
+It may be quicker in some situations to use the dot command.
+
+\subsection{Build custom operator and motion keys}
+
+You can build your custom motion keys by using the \verb|omap| command in the vimrc (I think).
+See \verb|:h :map-operator| and \verb|:h :omap-info|.

### 2016-03-30 tips_and_notes/vim/chapter3/visual.tex
onto Ex mode

\section{Visual Mode}

Visual mode is very similar to normal mode, but it allows you to define text selection using your cursor position(s).
You can enter visual mode by pressing \verb|v| in normal mode.
To enter a line-wise visual mode, press \verb|V|, and for a block-wise visual mode, press \verb|<ctrl-v>|.

To reselect the previous selection, press \verb|gV|.
If you want to swap the cursor position from one end of a selection to another, press \verb|o| while in insert mode.

If you want to carry out a (ex) command on the selection, press \verb|:| while in insert mode with the text selected (or just simply type the operator command if it's not complicated).
This will drop you into ex-command line prompt, where you can type in whatever you want to do with the selection.

### 2016-03-31 tips_and_notes/vim/chapter3/command.tex
tmp commit

@@
 
-Ex mode provides the minor adjustments and specificity required by you to carry out the exact task you want vim to do for the whole document, or just part of it.
+Ex mode provides the minor adjustments and specificity you cannot get in normal mode alone to carry out the exact task you want vim to do for the whole document, or just part of it.
+To enter Ex mode, press \verb|:| while in normal mode, and you will be given a command prompt starting with ``:''.
+
+\subsection{Basic command editing in Ex mode}
+
+As in insert mode, you can use \verb|<ctrl-w/u>| to delete a word/line in the command line.
+You can also paste characters from registers with \verb|<ctrl-r>{register}|.
+To insert a word or WORD in the Ex command prompt, type \verb|<ctrl-r><ctrl-w/a>|.
+
+\newpage
+
+\subsection{Ex mode syntax}
+
+The basic command syntax in Ex mode is as follows:
+
+\begin{center}
+    \verb|:[start],[end] [normal] command [address]|
+\end{center}
+
+\subsubsection{Range}
+
+The \verb|start| and the \verb|end| defines the range of the lines the command specified is to act on.
+You can define the range either relatively or absolutely.
+Absolute range is self explanatory -- you define your range by using the line number of the file.
+Relative range is used relative to particular parts of the document, with or without offsets.
+Some useful relative addresses are:
+\newline
+
+\begin{tabular}{c|l}
+    .  & current line\\
+    \$ & end of file\\
+    \% & all of the lines in the current file\\
+    0  & virtual line above the first line of the file\\
+\end{tabular}
+\newline
+
+With these relative addresses, you can define where you want the changes to occur, from where ever you are in the file.
+You can also give these addresses offsets (using + or - sign) so you don't have to worry about where in the file you are.
+%For example, if you wanted to substitute a pattern from the start of the file to five lines before the end of the file, you can type \verb|:1,$-5s/{word}/{substitute}/g|.
+
+\subsubsection{Visual mode range}
+
+If you are too lazy to type out the range or if you can't be bothered thinking where the range should start and end, an alternative is to use the visual selection as the range.
+When you select a part of the text you want to use as the range, you can press \verb|:| while in visual mode to open up an Ex command prompt with the range prepopulated with weird marks like this \verb|:'<,'>|.
+
+These symbols represents the start and the end of the visual selection, which you can use it in the Ex command.
+
+\subsubsection{Using patterns as a range}
+
+Instead of specifying ranges with visual selection and addresses, you can also use search patterns as a range.
+To start (and end) a search pattern, you have to type \verb|/| with the search pattern you are looking for in between the slash.
+If there is a special character in the search, then you'll have to escape the characters with a backslash (\verb|\|).
+
+\subsection{Copy/move lines to specific address}

### 2016-04-01 tips_and_notes/vim/chapter3/command.tex
finished chapter 3

@@
 \subsection{Copy/move lines to specific address}
+
+To copy or move some lines to a specific address in your document, type \verb|:[range]copy{address}| or \verb|:[range]move{address}|
+You can type t (for copy) or m (for move) instead of typing it out fully.
+Also, the address where the text is placed is below the specified address (so if you specified it to paste it to the current line, it will paste one line below it).
+
+\subsection{Execute normal mode commands}
+
+To execute any normal mode command on the lines specified, type \\\verb|:[range]normal{command}|.
+This will execute the command you have specified on the range of lines.
+If you have forgotten to place ``;'' after each line, then you could type \verb|[range]normal A;| to append ``;'' to each of the lines you have specified.
+This can also be applied to un/commenting lines of texts as well.
+
+\subsection{Repeating Ex command}
+
+To repeat the last Ex command you have used, press \verb|@:|.
+If you want to repeat it again, press \verb|@@|.
+You can also look at the history of the Ex command you have used by using the arrow keys while in the Ex command line prompt.
+
+\subsection{Command-line window}
+
+Command-line window is a mode where you could look through the history of all the commands you have used in a separate buffer.
+You can navigate in this window using the \verb|j/k| keys as in normal mode.
+To execute the command, simply press \verb|<CR>| while the cursor is on the command you want to repeat.
+
+Good thing about this mode is that you are able to edit these commands using vim's editing power.
+This means that you are able to connect/edit/alter previous commands you have used to suit the current document, or for streamlining your workflow.
+To switch from command-line prompt to command-line window anytime by pressing \verb|<ctrl-f>|.
+To exit the command-line window, press \verb|:q| (i.e. quit out of the buffer).
+
+You can also look up your search history by typing \verb|q/| instead of \verb|q:|.
+
+\subsection{Using shell commands and shell in Vim}
+
+To run shell commands within vim, type \verb|!| once you're in the Ex command prompt.
+For example, if you wanted to compile your current \LaTeX{} file into pdf, you can type \verb|:!pdflatex %| (where \% is a shorthand for the current filename).
+
+If you wanted to do more than one command while in vim and you're tired of keep typing !, then you could open up a shell session within vim (type \verb|:shell|).
+Note that ! is a one-off command, whereas \verb|:shell| is a full shell session.
+An alternative approach to this is to suspend vim by pressing \verb|<ctrl-z>|, and then when you are done with the shell, you could resume vim by typing \verb|fg| in the shell.
+
+\subsubsection{Some useful shell commands in Vim}
+
+\begin{tabular}{c|l}
+    \verb|:read !{command}|         & directs standard output into a buffer\\
+    \verb|:write !{command}|        & directs buffer content as standard input for\\
+                                    & the command\\
+    \verb|:[range]write !{command}| & execute command in shell using the range\\
+                                    & of lines as input\\
+    \verb|:[range]!{filter}|        & use range as standard input for the filter\\
+                                    & and overwrite the range (i.e. filter the range\\
+                                    & using the program filter)\\
+\end{tabular}
+\newline
+
+Note that the positioning of the whitespace can be important for distinguishing when to use the text as standard input for a shell command, or for filtering.

### 2016-04-01 tips_and_notes/vim/chapter4/files.tex
finished chapter 3

\chapter{Files, Windows, and Tabs}

\section{Files}

### 2016-04-01 tips_and_notes/vim/chapter4/window.tex
finished chapter 4

\section{Windows}

Vim allows us to split the window with same or different buffers for editing.
An alternative to this would be to use tmux.
You can use windows to look at different parts of the document, so that you can have a window open to refer to the text elsewhere of the document, and a window open for editing the document.

\subsection{Split and navigate through windows}

\begin{tabular}{c|l}
\verb|<ctrl-w>s|              & split window horizontally\\
\verb|<ctrl-w>h|              & split window vertically\\
\verb|:sp {file}|             & split window horizontally and load file\\
\verb|:vsp {file}|            & split window vertically and load file\\
\verb|<ctrl-w><ctrl-w>|       & cycle through open windows\\
\verb|<ctrl-w><ctrl-h/j/k/l>| & focus the window to left/down/up/right \\
                              & window\\
\verb|:cl(ose)|               & close the active window\\
\verb|:on(ly)|                & close all windows except current window\\
\verb|<ctrl-w>=|              & equalize width and height of all windows\\
\verb|<ctrl-w>_|              & maximize height of the active window\\
\verb=<ctrl-w>|=              & maximize width of the active window\\
\verb|[N]<ctrl-w>_|           & set active window height to N rows\\
\verb=[N]<ctrl-w>|=           & set active window width to N columns\\
\end{tabular}

### 2016-04-04 tips_and_notes/vim/chapter4/files.tex

@@
 \section{Files}
+
+First of all, note that the buffers are not the same as the files -- we edit the buffers in vim, and then save the changed buffer as a file.
+In vim, we can edit multiple buffers simultaneously by giving vim more than one file names to edit (\verb|vim file1 file2 ...|).
+
+\subsection{Navigating between different buffers}
+
+There are few commands you can use to navigate yourself to different buffer you have opened in vim.
+\newline
+
+\begin{tabular}{c|l}
+    \verb|:bnext| (\verb|:bn|)  & next buffer\\
+    \verb|:bprev| (\verb|:bp|)  & previous buffer\\
+    \verb|:bfirst| (\verb|:bf|) & first buffer\\
+    \verb|:blast| (\verb|:bl|)  & last buffer\\
+\end{tabular}
+\newline
+
+It will be a good idea to map these to different key bindings to access different buffers quickly.\\
+E.g. \verb|nnoremap <silent> {key_combination} :bprevious<CR>|
+
+\subsection{Buffer list}
+
+The \verb|:ls| command opens up the buffer list to see what files you have opened, and which file you are currently working with.
+The \verb|\%| represents the current buffer, and \verb|\#| indicates the alternative buffer.
+You can switch between the current buffer and the alternative buffer by pressing \verb|<ctrl-^>| (or \verb|<ctrl-6>|).
+
+\subsection{Argument list}
+
+Argument list represents the list of files that was passed as an argument when you run the \verb|vim| command in the shell.
+To open the argument list, type \verb|:args|.
+Argument list can be changed by using \verb|:args {arglist}| (e.g. \verb|:args index.html app.js|).
+It can also take wildcard characters (globs) like * (wildcard) and ** (recurse downwards into other directories).
+
+Files can be specified using the backtick (\verb|`|) expansion as well.
+Backtick executes the text within it in the shell and lets you use the output.
+So, you can use \verb|:args `cat files.txt`| to use the content of files.txt as the argument list.

### 2016-04-04 tips_and_notes/vim/chapter4/tab.tex

\section{Tabs}

Tabs in vim can hold a collection of windows (like a workspace in  linux).
The \verb|:lcd {path}| command sets the working directory for the current window (not the tab).
This means that you can assign different working directory to different windows/tabs (i.e. separate workspace).
\verb|:tabedit {filename}| will open a new tab with filename.
If filename is empty, tab is created with an empty buffer.
\verb|<ctrl-w>T| will move the current window into a new tab page.
\verb|:tabclose| will close the tab page, even if it has got multiple windows open.
\verb|:tabonly| will close all other tabs except for the current tab.
Finally, \verb|gt/gT| will ``go to'' the next/previous tab.
You can specify the number of tabs you want to jump past (e.g. 3gt).

### 2016-04-05 tips_and_notes/vim/chapter5/efficient.tex
wrote navigation summary

\section{Efficient navigation}

\section{Navigation Summary}

\begin{center}
\begin{tikzpicture}[node distance = 1.5cm]
    \node (o) at (0,0) {};
    \node [left=of o] (h) {h/b/B/ge/gE/0/T/F/?};
    \node [right=of o] (l) {l/w/W/e/E/\$/t/f/\verb|/|};
    \node [above=of o] (k) {k/\verb|<ctrl-b>|};
    \node [below=of o] (j) {j/\verb|<ctl-f>|};
    \draw [<->, very thick] (h) edge (l) (k) edge (j);
\end{tikzpicture}
\end{center}

### 2016-04-05 tips_and_notes/vim/chapter5/efficient.tex
finished navigation

@@
 \section{Efficient navigation}
+
+The basic navigation will get the work done quite efficiently (faster than MS Office at least), but you can go faster by using other commands to navigate within and between files.
+Logically, to go any faster than the basic movement introduced in the previous section, you will have to be able to navigate through the document faster than word-wise navigation -- i.e.
+navigations that are larger than a word.
+This section will show you how to do this (if used correctly).
+
+\subsection{Navigate by specific letters}
+
+To move your cursor to (or close to) a specific letter/character, press \verb|f/t{char}|.
+\verb|f| will search for the character you have pressed after pressing \verb|f| from the current cursor position to the end of the line.
+If the character is found, the cursor will jump to the character, and if it is not found, the cursor remains at its position.
+
+\verb|t| key does a similar thing to the \verb|f| key, but moves the cursor to one position before the character you searched for.
+Again, if the character is not present, the cursor doesn't move from the current position.
+To repeat the search/jumping forward to the character, press \verb|;|.
+To reverse the search/jump backward to the character, press \verb|,|.
+
+Note that \verb|f/t| keys  both search forward in the line, but if you wanted to search backward to start off with, then you could use the uppercase version -- \verb|F/T|.
+Again, you can repeat the jump using the \verb|;/,| keys.
+
+\subsubsection{Think like a scrabble player}
+
+Remember that there are some letters that appear more often than other letters, so choosing a good letter for jumping will make your navigation faster.
+When using the character search commands, it's better to choose target characters with a low frequency of occurrences.
+
+\subsection{Search to navigate}
+
+Note that \verb|f/F/t/T| keys only work on a single character at a time, and only on the current line.
+Unless your line is massive, then this method of navigation doesn't work as efficiently as you would expect.
+
+Another way for navigating around the document is to use the search function to search for the specific words in the document.
+To start a search, simply press the \verb|/| key -- this will give you a prompt where you can type the word you are looking for.
+Depending on how you have set up vim, the search results will come up as you type, and automatically moves the cursor to the beginning of the first match of your search.
+
+Once you press the \verb|<CR>|, the cursor will move to the first match in the document.
+To abort/cancel the search, press \verb|<esc>|.
+To go to the next/previous matching word, press \verb|n/N|, respectively.
+If you wanted to search backwards instead of searching forward, you can press \verb|?| instead of \verb|/|.
+
+An advantage of the search command is that you can use the search while you are in visual mode, as well as operator-pending mode.
+So, you could use \verb|c/d/gU| and other keys with the search command.
+Note that the search command is an exclusive command, so it doesn't include the letter under the cursor (i.e. the first letter of your first match of your search).
+
+\subsection{Jump between matching parenthesis}
+
+You can jump between matching parentheses using the \verb|\%| symbol.
+This is useful if you want to visually select the inside the parentheses.
+
+\subsection{Mark your place and jump to your marks}
+
+You might want to set a mark in the file so that you can come back to it instantly.
+To set a mark somewhere in the document, press \verb|m{a|$\sim$\verb|z}|, where the specified letter will store the mark that you have set in the document.
+
+There are two ways you could jump to the marks that you have set:
+\begin{enumerate}
+    \item \verb|'{mark}| moves to the line where the mark was set in the document (the cursor is placed at the beginning of the line).
+    \item \verb|`{mark}| moves to the exact position of the mark where you have set the mark in the document.
+\end{enumerate}
+
+\subsubsection{Default marks in Vim}
+
+Other than the marks you have specified and explicitly set in the document, there are marks that vim automatically keeps track of.
+\newline
+
+\begin{tabular}{c|l}
+    \verb|``| & position before the last jump within the current file\\
+    \verb|`.| & location of last change\\
+    \verb|`^| & location of last insertion\\
+    \verb|`[| & start of last change or yank\\
+    \verb|`]| & end of last change or yank\\
+    \verb|`<| & start of last visual selection\\
+    \verb|`>| & end of last visual selection\\
+\end{tabular}
+\newline
+
+Some of these marks will work with a single quote as well.
+
+\subsubsection{Uppercase vs lowercase marks}
+
+Lowercase marks are local to each individual buffer, but uppercase marks are globally accessible.
+This means that you are able to jump to the document with uppercase mark set anytime you like, even if you haven't opened that file when you started vim.
+By default, global marks are persisted between editing sessions, meaning that you can always come back to a specific file if a mark is set somewhere in the document.
+
+\subsection{Navigate between files with jumps}
+
+Vim has a thing called the jump list (\verb|:jumps|).
+Jump list records any command that changes the active file for the current window (known as ``jumps'').
+\verb|<ctrl-o>| goes back one jump, and \verb|<ctrl-i>| goes forward one jump.
+
+Note that motions move around within a file, whereas jumps can move between files.
+Generally, long-range motions are classified as jumps, but short-range motions are classified as motions.
+
+Some examples of jumps include:
+\newline
+
+\begin{tabular}{c|l}
+    \verb|[count]G| & jump to line number\\
+    searching (\verb|/|/\verb|?|/n/N) & jump to next/previous occurrence of a pattern\\
+    \verb|\%| & jump to matching parentheses\\
+    \verb|(|/\verb|)| & jump to start/next sentence\\
+    \verb|{|/\verb|}| & jump to start/next paragraph\\
+    \verb|H/M/L| & jump to top/middle/bottom of screen\\
+    \verb|gf| & jump to file name under cursor\\
+    \verb|<ctrl-]>| & jump to definition of keyword under the cursor \\
+\end{tabular}
+
+\subsection{Change list}
+
+Just like jump list, change list is a list of the modifications made to each buffer during an editing session.
+\verb|g;/g,| allows you to traverse forward/backward through the change list, respectively.
 

### 2016-04-06 tips_and_notes/vim/chapter6/copyandpaste.tex
started chapter 6 for vimtips

\chapter{Copy and Paste}

Copying and pasting text are essential features of any text editors.
I have already mentioned how to yank/change/delete/paste (\verb|y/c/d/p|) in vim, but there is a little bit more to it than just simply copy/deleting/pasting.

\section{Registers}

There are some useful features in vim that will allow you to store and paste multiple different pieces of text in your document.
The key to this feature are the registers.
Any key on the keyboard can be used as a register to store pieces of text, and you can pull out the text that is stored in the specified register.

To specify which register to store/use, you have to specify the register by pressing \verb|"{register}|, kind of like when you are setting marks in documents (see \textit{\ref{subsec:marks}:\nameref{subsec:marks}}).
For example, to yank a word into the register \verb|a|, you would press \verb|"ayiw| while the cursor is on the word you want to store.

The basic syntax of copying/pasting (and some other commands) using a register is \verb|"{register}y/p[motion]|.
When no register is specified for paste/copy, the \verb|"| register is used by default (known as the unnamed register).

\section{Named registers}

Vim has a register for each letter of the alphabet, meaning that we have 26 unique registers that we can store different pieces of text.
This is especially useful when you want to copy/paste/delete multiple text, but want to use it again another time.

\section{Special registers}

There are registers that have special function in vim and reserved for specific purposes.

\subsection{Unnamed register (\texttt{""})}

Unnamed register is used by \verb|x/s/d/c/y| commands, and stores the text that has been deleted or yanked as a result of the command.
So, whenever you remove a text from normal mode, the text will be saved to the unnamed register.

\subsection{Yank register (\texttt{"0})}

When \verb|y{motion}| is used, it saves not only to the unnamed register, but also saves it to the yank register (\verb|"0|).
This register is only used by the \verb|y{motion}|, and therefore unaffected by other deleting/substituting commands like \verb|s/d/x/c|.
Meaning that you are able to copy a piece of text, then change a different bit of text and paste the text you have yanked using the yank register.

\subsection{Black hole register (\texttt{"\_})}

The underscore register is the place where nothing is returned (i.e. nothing is saved).
This is useful when you don't want to overwrite a different register (e.g. unnamed register).

\subsection{System clipboard register (\texttt{"+})}

If you want to copy/paste to/from an external program, you need to access the system clipboard.
You can paste text from another place by using \verb|"+p| (or \verb|<ctrl-r>+| in insert mode).
If you want to use a text from vim, you can store the text in \verb|"+| and use paste as usual (\verb|<ctrl-v>|) in another place.

If it doesn't work on ubuntu properly, try installing the vim-gnome package.

\subsection{Expression register (\texttt{"=})}

When you access this register, it drops down into Ex command prompt.
Vim will evaluate the expression you have typed and stores the result in the register.
You can paste the result in insert mode using the \verb|<ctrl-r>=| keys.

\subsection{Read-only register}

### 2016-04-07 tips_and_notes/vim/chapter6/copyandpaste.tex
started on chapter 7

@@
 I have already mentioned how to yank/change/delete/paste (\verb|y/c/d/p|) in vim, but there is a little bit more to it than just simply copy/deleting/pasting.
+Just a quick note before we get started: \verb|gp/gP| does the same thing as the \verb|p/P| command, except that it leaves the cursor at the end of the pasted text.
 
@@
 \subsection{Read-only register}
+
+There are some registers in vim that are reserved for specific purposes:
+\newline
+
+\begin{tabular}{c|l}
+    \verb|"%| & name of the current file you are editing\\
+    \verb|"#| & name of the alternate file\\
+    \verb|".| & last inserted text\\
+    \verb|":| & last ex command\\
+    \verb|"/| & last search pattern\\
+\end{tabular}
+\newline
+
+\section{Visual selection and registers}
+
+When you paste while in visual mode, the text which has been selected will be replaced with the text in the unnamed register.
+This means that the selection that was overwritten by the paste is now in the unnamed register (i.e. the content of the unnamed register is swapped with the visual selection).
+This can be useful if you want to rearrange words/lines/selection in your document.
+
+\section{paste from register}
+
+Sometimes it can be a little confusing whether to use \verb|p| or \verb|P|, as one will paste before and the other will paste after the cursor, respectively.
+It may be helpful to go into Insert mode and use \verb|<ctrl-r>{register}|, as it will always paste the text after the cursor (as if we have typed it).
+Also, together with \verb|c{motion}| (and others), you can make a useful dot command with it.

### 2016-04-07 tips_and_notes/vim/chapter7/macros.tex
started on chapter 7

\chapter{Macros}

From the Practical Vim: \textit{``Like the game of Othello, Vim's macros takes a minute to learn and a lifetime to master.''}

\section{Record and execute a macro}

The \verb|q| key acts as both the start and stop command for recording a macro\verb|q| key acts as both the start an stop command for recording a macro,
\verb|q{register}| starts recording our keystrokes to be used as a macro.
Press \verb|q| again to stop recording the keystrokes.
Note that anything stored in the register will be overwritten after you have recorded the macro to the register.

To execute

### 2016-04-08 tips_and_notes/vim/chapter7/macros.tex
finished chapter 7

@@
 
-To execute
+To execute the macro in a specific register, type \verb|@{register}|.
+To replay the most recent macro, type \verb|@@|.
+
+\section{How to construct a macro - Normalise, Strike, Abort}
+
+The golden rule of setting up macro is to ensure that every command is repeatable.
+Firstly, you have to normalise the cursor position, then strike your target with a repeatable motion, and lastly, the macro should abort when that motion fails.
+
+\subsection{Normalise the cursor position}
+
+Before you do anything, make sure your cursor is positioned so that the next command does what you expect and where you expect it.
+For example, you might want to move the cursor to the next search match (\verb|n|), or start of the current line (\verb|0|), or first line of the file (\verb|gg|)
+This makes it easier for you to strike the right target every  time you run the macro.
+
+\subsection{Strike your target with a repeatable motion}
+
+Navigate by search, use text objects, and exploit the full arsenal of vim's motions to make your macro as flexible and repeatable as you can.
+Never use the \verb|h/j/k/l| keys when you can use other useful motion (e.g. \verb|0|/\verb|gg|).
+
+\subsection{Abort when motion fails}
+
+If a motion fails while a macro is executing, then vim aborts the rest of the macro by default.
+For example, if a macro changed the word you are searching, then it will abort (i.e. do nothing) as soon as there are no more words that matches the search pattern.
+This means that you can repeat it as many times without worrying about over doing the macro in a document.
+
+\section{Play back a macro with a count}
+
+Unlike the dot command, you can repeat a macro using a count (e.g. \verb|100@q| repeats the macro stored in the register \verb|q| 100 times).
+
+\subsection{Running a macro in series or parallel}
+
+Think of it as an electric circuit -- series will stop working if one part of the circuit blows, but a parallel circuit can still work even if one bit blows up.
+When you run \verb|100@{register}|, this will do it in series and it will repeat the macro 100 times until it aborts.
+If you select the text first, then do \verb|:'<,'>normal@q|, you will repeat the macro in parallel for those lines, which will ignore the lines that it cannot change, rather than aborting the whole thing as soon as one aborts.
+
+\section{Appending commands to macro}
+
+If you forgot to include a command at the end, you can append it by using the uppercase letter of the register (e.g. \verb|qA|).
+
+\section{Running macros on multiple files}
+
+You can set up the macro to go to the next file \verb|:bn|) for series, or set up an argument list with all the files (\verb|:args *{filetype}|) for parallel.
+In parallel, you can do \verb|:argdo normal@{register}| to use the macro on all the files in \verb|:args|.
+After you have run this, you will have to save all the buffers (\verb|:wa(ll)| to write all, or you could incorporate \verb|:wnext| (write, then next buffer) into your macro).
+
+\section{Macro for incrementing numbers}
+
+In vim, you can set variables by using \verb|:let| command (e.g. \verb|:let i=1|).
+\verb|:echo {variable}| will echo the value stored in the variable (\verb|:echo i| will give 1).
+You can increment \verb|i| by saying \verb|:let i+=1|.
+To insert the incremented number into the text, you will have to record/save \verb|i| into an expression register: \verb|<ctrl-r>=i<CR>|.
+In order for this to work, you will have to set the variable outside the macro first, then use the variable and increment the variable within the macro, so it will save/adjust the value of the variable as the macro is replayed.
+
+\section{Edit the contents of the macro}
+
+Just like yank and put, macro is also saved in the register.
+To edit the command, you have to paste the macro into the document then edit it.
+Once you have finished editing the command, paste it back onto the register using the \verb|"{register}y$| command, which will yank the command without the \verb|<CR>|.
+Finally, delete the line when you are done.

### 2016-04-08 tips_and_notes/vim/chapter8/patterns.tex
started chapter 8

\chapter{Search and pattern matching}

\section{Basic searching}

As discussed before, \verb|/| searches forward and \verb|?| searches backward for a pattern, and \verb|n/N| will jump to the next/previous search pattern.
Repeat the search by leaving the search field blank and press \verb|<CR>|
Note that \verb|n| goes forward for \verb|/| search but ``backward'' (forward in the backward direction) for the \verb|?| search.
To switch the direction, just use \verb|/| or \verb|?| with no search pattern -- this will re-use the previous pattern to search forward/backward.
You can go through your search history by using the \verb|<up>| key in the search prompt (or use \verb|q/|).

\subsection{Mute search highlighting}

\verb|:noh(lsearch)| will mute the highlighting for current search, but will enable it for the next search.
Either use this as is (\verb|:noh|) or map it to a useful shortcut.

\section{Pattern matching in search}

There are various tricks and tips you can add to your search pattern to hone in  on the target text that you are actually interested in.

\subsection{Case sensitivity in search}

You can set the case sensitivity globally by tweaking the \verb|:ignorecase| setting, but you may want to set the case sensitivity differently, depending on your needs.
The \verb|\c| and \verb|\C| items in your search pattern will allow you to do this.

### 2016-04-11 tips_and_notes/vim/chapter10/advancedsub.tex
finished chapter 10

\chapter{Advanced substitution}

This chapter includes some cases of substitution that require little bit more thinking than normal substitute commands.

\section{Rearrange CSV fields using submatches}

You can use submatches in search to rearrange one or more words.\\
\verb|:%s/{search pattern with submatches}/\{submatch 1-9}|
See pg 228 in Practical Vim for a detailed example.

\section{Perform arithmetic on the replacement}

You can perform arithmetics on the matches and replace the old matches with the result.
For example, \verb|:%s//\=submatch(0)-1/g| will minus 1 to whatever matches in the search.
\verb|submatch({number})| returns the \verb|{number}|'th submatch of the matched text.
When 0 is used, the whole matches text is returned.
\verb|submatch()| can only be used in the \verb|:s| command or the \verb|substitute()| function.
Another thing to note is that you have to use this function to access the submatches in vim script.

\section{Swap two or more words}

You can use a dictionary data type for this \verb|:let x={"dog":"man", "man":"dog"}|\\
You can access it by \verb|x["key"]| (e.g. \verb|:echo x["dog"]| will give man).
You also need to be able to match the words that you want to swap.\\
\verb-/\v(<"word1">|<"word2">...)- will capture the match into the submatches.
To swap  the words, type:\\
\verb|:%s//\={dictionary of words to be swapped}[submatch(1)]/g|\\
This uses the submatch to look up the matching keyword in the dictionary you have defined, and use the other key value as the replacement string.
As a result, you would have swapped the words.
This approach will allow you to swap 2 or more words in a document with a single hit.
(checkout the plugin discussed on pg 232 in Practical Vim).

\section{Find and replace across multiple files}

You need to first set up the substitute command, then use the argument list and \verb|:argdo| command to apply it to all the files.
You can use the \verb|e| flag to suppress the error messages when it can't find any matches in a file.
This method will attempt to substitute in all  the files, whether or not it has the match (see pg 235-236 for an alternative solution).

### 2016-04-11 tips_and_notes/vim/chapter8/patterns.tex
finished chapter 10

@@
 
+\subsection{Preview the first match before execution}
+
+Enable \verb|:incsearch| to see what the search pattern your search pattern is picking up.
+
+\subsection{Auto-complete the search field}
+
+You can auto-complete the search field by using \verb|<ctrl-r><ctrl-w>|.
+
 \section{Pattern matching in search}
@@
 The \verb|\c| and \verb|\C| items in your search pattern will allow you to do this.
+The \verb|\c| makes the search to ignore case, and \verb|\C| makes the search case sensitive.
+These items can appear anywhere in the search pattern, so if you forgot to put on case sensitivity at the start, you can append \verb|\c| of \verb|\C| at the end of your search (e.g.
+\verb|/foo\c| or \verb|/foo\C|).
+
+\subsection{Smart case sensitivity}
+
+The \verb|:smartcase| option allows you to search case insensitive if everything is lower case, but if it contains an uppercase, it switches to case sensitive search.
+For example, if you search \verb|/hello| with smart case on, it will match \verb|hello, Hello, hEllo,| etc., but when you search for \verb|/Hello|, it will only match \verb|/Hello|.
+
+\subsection{Very magic search}
+
+The \verb|\v| in a search tells the search to switch into a ``very magic'' search, where all characters are assumed to have a special meaning, except letters, digits, and underscore (\verb|_|).
+This makes the rules about escaping characters in the search a bit easier.
+
+\subsection{Verbatim search}
+
+Sometimes you want to search for words with \verb|.|, \verb|*|, \verb|?| and other characters that have special meanings.
+The \verb|\V| item switches on the ``very nomagic'' search, where nothing has a special meaning (unless backslashed within the pattern).
+For example, a normal search for \verb|/a.k.a| could match \verb|backward|, but with verbatim search (\verb|/\Va.k.a|) will only match \verb|a.k.a|.
+
+In general, use \verb|\v| for regular expression and \verb|\V| for literal search.
+
+\subsection{Hex character}
+
+The \verb|\x| character stands for \verb|[0-9a-fA-F]| (i.e. hex characters).
+
+\section{Specifying the word boundary}
+
+Use \verb|<| and \verb|>| in the search to specify the boundary of the word.
+For example, \verb|the| can come up in \verb|there, these,| etc., but \verb|<the>| will only match the word \verb|the|.
+
+\section{Use parentheses to capture submatches}
+
+Any pattern matches within the parentheses is ``captured'' or stored in a temporary place.
+You can access the captured text using \verb|\{digit}|.
+If multiple parentheses are used, it is stored in \verb|\1|, then in \verb|\2|, and so on.
+Note that \verb|\0| always refer to the entire match, even if you are not using a parentheses.
+These captured text can be used later in other commands such as substitution.
+If you don't want to store the text, type \verb|%| before the parentheses -- this will prevent vim to store the pattern.
+
+\section{Other characters}
+
+\verb|\w| character matches word characters, including letters, numbers, and the underscore.
+\verb|\W| character matches everything except for word characters.
+The \verb|\zs/\ze| characters marks the start/end of a match in a pattern.
+
+\section{Patterns vs. Matches}
+
+Patterns can be anything you are searching for, whereas matches are the particular word/characters you want to highlight/change.
+For example, you might be interested in the word ``Vim'' that comes after ``Practical'' (i.e. want to look for the occurrence of \verb|Practical Vim| and only highlight/change the \verb|Vim| part).
+
+If you search just for Vim, it will match all the Vim in the document, whereas only few will match with Practical Vim (but you only want to change the Vim part, and leave the Practical part alone).
+To do this, use \verb|\zs| and \verb|\ze| within a pattern where you want the match -- \verb|/Practical \zsVim\ze|.
+This should highlight Vim that comes after the Practical.
+
+\section{Characters to escape during searches}
+
+When searching forward, you will have to escape the \verb|/| character.
+When searching backward, you will have to escape the \verb|?| character.
+You will have to escape \verb|\| all the time.
+You can use \verb|escape({str}, {char1}, {char2}, ...)| function to automatically escape specified characters.
+
+\section{Count the matches for current pattern}
+
+\verb|:%s///gn| will count the number of matches in the current document.
+The \verb|n| flag after \verb|g| suppresses the usual substitute behaviour.
+By leaving the search field blank, we tell Vim to use the current pattern.
+It may be a good idea to map this to a shortcut.
+
+\section{Using an offset in searches}
+
+The \verb|/e| in the search pattern tells vim to place the cursor at the end of the matching search pattern.
+For example, \verb|/lang/e| will place the cursor on the \verb|g| of the word lang.
+
+\section{Operate on search matches that vary in length}
+
+First, you need to find a match that will match the text.
+For example, if you are looking for either \verb|Xml| or \verb|Xhtml|, you could match them using \verb|/\vX(ht)?ml\C|.
+Then, you can use the previous match pattern as a motion to work on the search match.
+Again, for example, if you want to uppercase those words, you could type \verb|gU//e| (``make everything until the end of the match uppercase'') to make the words uppercase.
+
+The \verb|n| key wil jump to the next match, but places the cursor at the end of the match, as \verb|//e| was the last search done.
+Therefore you will have to use \verb|//| (without the \verb|e|) to go to the next match.
+
+The ``textobj-lastpat'' by Kana Natsuno adds \verb|i/| text object to operate on the search matches.
+This allows you to do the same thing with \verb|gUi/|, and also allows you to use \verb|n| to go to the next match (and therefore dot formula will work).
+
+\section{Workflow for constructing complex patterns/regular expressions}
+
+Example sentences to make a complex search:
+\newline
+
+\begin{center}
+    \verb|This string contains a 'quoted' word.|
+    \verb|This string contains 'two' quoted 'words.'|
+    \verb|This 'string doesn't make things easy.'|
+\end{center}
+
+\subsection{A broad match}
+
+You will have to start off with a broad match that matches a lot of text in the document.
+The search \verb|/\v'.+'| will match \verb|'| followed by any character until the last \verb|'| mark.
+(\verb|.| = anything, \verb|+| = as many times).
+
+\subsection{Refinement}
+
+After the first crude search, you will have to edit and refine the search so that it matches the pattern more specifically.
+You can change the \verb|.| to \verb|[^']| so it matches any characters but \verb|'|.
+
+\subsection{Iterate until perfect match}
+
+You will need to keep refine the match until you get a perfect match.
+You can change the search to \verb-/\v'([^']|'\w)+'- to ignore the \verb|'| used as apostrophe.
+You can change this search into \verb-/\v'(([^']|'\w)+)'- to store it in \verb|\1| to use it later for substitution.
+\verb|:%s//"\1"/g| would change all single quote marks into double quotes.
+
+\section{Search for the visually selected text}
+
+coming soon

### 2016-04-12 tips_and_notes/vim/chapter11/global.tex
finished chapter 11

\chapter{Global commands}

The global command combines the Ex commands with Vim's pattern matching abilities.
It allows you to run Ex command on each line that matches a specific pattern.

\section{Syntax of global command}

{\large\verb|:[range] global [!] /{pattern}/ [range for cmd] [cmd]|}
\newline

The default range of the \verb|:global| command is the whole file (\verb|%|).
The \verb|{pattern}| can be left blank to use the previous search pattern.
\verb|[cmd]| can be any Ex command (except for another \verb|:global| command), and the default command is the \verb|:print| command.
The \verb|!| tells it to invert the command (i.e. do the \verb|cmd| on everything that doesn't match the pattern).

\section{Delete or keep matching lines}

\begin{tabular}{c|l}
\verb|:g/{regular exp}/d| & this would delete all the lines that contain the regular expression\\
\verb|:v/{regular exp}/d| & \verb|:v| stands for \verb|:vglobal| (inverted global, i.e. \verb|:global| with \verb|!|).\\
& This will delete all the lines without the regular expression.\\
\end{tabular}

\section{Collect certain lines from a file}

For example, you need to extract all ``TODO'' comments in a file.
You can do this by \verb|:g/TODO|, which prints it out.
If you want to save it in a register, do \verb|:g/TODO/yank A|, which will save it to the register \verb|a| (capital A tells it to append to the specified register).
Now you can paste or do whatever using this register.

\section{Alphabetise a selection}

You can do this by using the Vim's sort command.
Say if you needed to do this to a block of lines, and there were multiple blocks in a file, which will be difficult to do it one by one, you can hit all of them at once using the global command.
Note that the \verb|cmd| can take a range as well, independent of the \verb|range| for the \verb|:global| command.
The \verb|.| address stands for the line that the cursor is positioned, but in \verb|:g|, it stands for the line that matches the \verb|pattern|.

If the block was in between \verb|{| and \verb|}|, you can do \verb|:g/{/ .+1,/}/-1 sort|.
The \verb|+1| and \verb|-1| is the offset to specify only the content of the block.
Note that \verb|/{/| searches for the closing bracket.

In general: \verb|:g/{start}/ .,{finish} [cmd]| will do the work to each of the specified range.

\section{Muting messages from \texttt{cmd}}

Sometimes \verb|cmd| sends out a message, which can be annoying.
You can silent this by using \verb|:silent| (or \verb|:sil|) to suppress the messages.
For example, \verb|:g/{/sil .+1,/}/-1 >| will indent the lines specified with no messages.

### 2016-04-13 tips_and_notes/vim/chapter12/list.tex
kind of finished chapter12 and vimtips

\chapter{Quick fix and location list}

\section{Compile code and use quick fix list}

You are able to run \verb|:make| in Vim, instead of having to run make in terminal.
When errors are given, vim automatically parses each line and extracts the filename, line number, and error messages.
For each warning, vim creates a record in the quick fix list.
You can jump through the quick fix list, which takes you to the location of the error.
Some of the commands are listed below (see \verb|:h quickfix| for full list):\\

\begin{tabular}{c|l}
    \verb|:cnext|     & jump to the next item\\
    \verb|:cprevious| & jump to the previous item\\
    \verb|:cfirst|    & jump to the first item\\
    \verb|:clast|     & jump to the last item\\
    \verb|:cfile|     & jump to the first item in the next file\\
    \verb|:cnfile|    & jump to the first item in the next file\\
    \verb|:cpfile|    & jump to the last item in the next file\\
    \verb|:cc N|      & jump to the N$^{th}$ item\\
    \verb|:copen|     & open the quick fix list\\
    \verb|:cclose|    & close the quick fix list\\
\end{tabular}
\newline

You can prepend the \verb|:cnext| and other commands with numbers to go through them $n$ times at a time (e.g. \verb|:5cnext| will go through 5 errors at a time).
If you want to skip all errors in current file and start with the next file, use \verb|:cnfile|.
You can go through and choose the errors visually by using the quick fix window, and use \verb|k/j| keys to scroll.
You can also use search within this window.
To close the window, type \verb|:q| or \verb|:cclose|.

Vim holds onto the last ten quick  fix list,  and you can load them up using \verb|:colder| or \verb|:cnewer| commands.
Again, this can be used with numbers to go back to the n$^{th}$ quick fix list.

### 2016-04-15 thesis/intro/intro.tex
added a bit of plan in my thesis; added few more file types in .gitignore; added reference style file

\chapter{Introduction}

\section{Obesity}

General intro to obesity\\
Implications of obesity -- disease associated with it and/or cancer\\

\section{Cancer}

What is cancer\\

\subsection{Hallmarks of Cancer}

What is it caused by (11 hallmarks of cancer)\\

\subsection{Relationship between obesity}

Why is it important in obesity context\\

\section{Obesity Associated Genetic Signatures}

Literature review, focussing on \citet{Creighton2012} and \citet{Fuentes-Mattei2014}\\

\section{Aim Of The Project}

### 2016-04-15 thesis/methods/methods.tex
added a bit of plan in my thesis; added few more file types in .gitignore; added reference style file

\chapter{Methods}

\section{Download Data from publicly available source}

\subsection{The Cancer Genome Atlas}

\subsection{International Cancer Genome Consortium}

\section{Validation Of Methodology}

\subsection{Normalisation of \citet{Creighton2012} data}

where was the original data from?\\
(rma normalisation before any analysis)

\subsection{Differential Gene Expression Analysis}

(mention WHO classification of obese/overweight/lean)\\
(also mention limma package)

\subsection{Singular Value Decomposition}

\section{Applying metagene to other cancer types}

\subsection{Transformation matrix}

\section{Pathway Enrichment Analysis}

### 2016-04-15 thesis/results/results.tex
added a bit of plan in my thesis; added few more file types in .gitignore; added reference style file

\chapter{Results from \citet{Creighton2012} metagene}

\section{SVD}

### 2016-04-15 thesis/thesis.tex
organised the directory structure
added a bit of plan in my thesis; added few more file types in .gitignore; added reference style file

@@
 \usepackage[utf8]{inputenc}                  % package for specifying input font encoding
+\usepackage{natbib}                          % package for reference management
 \usepackage{graphicx}                        % package for graphics handling (i.e. figure images)
 \usepackage{mathptmx}                        % package for Times New Roman font
-\usepackage{setspace}          % package for setting line space
+\usepackage{chngcntr}                        % package for making continuous figure numbering, instead of section-based numbering
+\usepackage{setspace}                        % package for setting line space (used in title page)
+\usepackage[hidelinks, colorlinks]{hyperref} % package for hyperlinking references - hidelinks option removes the ugly boxes around the huperlinks
+\usepackage[md]{titlesec}                    % package for tweaking the chapter/section/subsection titles format
+
+% Set up the hyperlinks - hange the colour of the links to blue:
+\hypersetup{allcolors=blue}
+
+% change figure numbering from section-based to continuous:
+\counterwithout{figure}{chapter}
+\counterwithout{table}{chapter}
 
@@
 
+% Change the title of bibliography to references:
+\renewcommand{\bibname}{References}
+
 \begin{document}
 
-\include{./title}
+\include{./misc/title}
 
@@
 
-\include{./abstract}
+\include{./misc/abstract}
 
-\include{./acknowledgement}
+\include{./misc/acknowledgement}
 
@@
 
-\include{./intro}
+\include{./intro/intro}
 
-\include{./methods}
+\include{./methods/methods}
 
-\include{./results}
+\include{./results/results}
 
-\include{./discussion}
+\include{./results/results2}
+
+\include{./discussion/discussion}
+
+% Use the reference style specified in BiocRefStyle.bst:
+\bibliographystyle{BiocRefStyle}
+
+% Print out the bibliography:
+\bibliography{MSc}
 

### 2016-04-16 thesis/methods/methods.tex
started working on methods section

@@
 \chapter{Methods}
+
+\section{Body Mass Index (BMI)}
+
+\subsection{BMI Calculation}
+
+BMI was calculated from the height and weight data of the samples using the following equation:\\
+\begin{equation}
+    BMI = \frac{weight (kg)}{Height^2(m^2)}
+\end{equation}
+
+\subsection{BMI Classification}
+
+Samples were classified based on the WHO definition:
+\begin{table}[h]
+    \caption{WHO defined BMI classification}\\
+    \label{tab:who-bmi-class}
+    \begin{center}
+        \begin{tabular}{|c|c|}
+            \hline
+            \bfseries {Classification} & \bfseries {BMI}\\
+            \hline
+            Underweight & \textless 20.0\\
+            Normal weight/lean & 20.0$\sim$24.9\\
+            Overweight & 25.0$\sim$29.9\\
+            Obese & \textgreater 30.0\\
+            \hline
+        \end{tabular}
+    \end{center}
+\end{table}
 
@@
 
-\subsection{The Cancer Genome Atlas}
+Data related to multiple different cancer types were downloaded from free and publicly available source such as The Cancer Genome Atlas and the International Cancer Genome Consortium.
 
-\subsection{International Cancer Genome Consortium}
+\subsection{The Cancer Genome Atlas (TCGA)}
 
-\section{Validation Of Methodology}
+The clinical data for all cancer types (33 types in total) were downloaded from TCGA database and were checked for the height and weight data for each sample.
+Any cancer type with no weight and/or height data of the samples were excluded from the project, as no BMI information can be obtained without these data.
+This gave a total of eight cancer types (**insert all the names of cancer types here) to be used for further analyses.
+
+\subsection{International Cancer Genome Consortium (ICGC)}
+
+RNA-seq data for the eight cancer types with sample BMI information were downloaded from the ICGC database.
+
+\section{Validation of methodology used by \citet{Creighton2012}}
+
+\subsection{Raw data from \citet{Creighton2012} paper}
+
+The raw microarray gene expression data files of all the samples used by \citet{Creighton2012} were downloaded from the Gene Expression Omnibus (GEO) database.
+Raw data were imported into R/RStudio for further analyses.
 
@@
 
-where was the original data from?\\
-(rma normalisation before any analysis)
+The raw microarray data for all 103 samples in the study were normalised using Robust Multi Array (RMA) method from the \textit{affy} package in R.
 
-\subsection{Differential Gene Expression Analysis}
+\subsection{Differential gene expression analysis}
 

### 2016-04-18 thesis/intro/intro.tex
tmp commit

@@
 \chapter{Introduction}
+\label{ch:intro}
 
 \section{Obesity}
+\label{sec:obesity}
 
-General intro to obesity\\
-Implications of obesity -- disease associated with it and/or cancer\\
+General intro to obesity.
+Implications of obesity -- disease associated with it and/or cancer.
 
 \section{Cancer}
+\label{sec:cancer}
 
-What is cancer\\
+What is cancer.
 
 \subsection{Hallmarks of Cancer}
+\label{subsec:cancerhallmarks}
 
-What is it caused by (11 hallmarks of cancer)\\
+What is it caused by (11 hallmarks of cancer).
 
 \subsection{Relationship between obesity}
+\label{subsec:obsbackground}
 
-Why is it important in obesity context\\
+Why is it important in obesity context.
 
-\section{Obesity Associated Genetic Signatures}
+\section{Obesity associated genetic signatures}
+\label{sec:obsgene}
 
-Literature review, focussing on \citet{Creighton2012} and \citet{Fuentes-Mattei2014}\\
+Literature review, focussing on \citet{Creighton2012} and \citet{Fuentes-Mattei2014}.
 
-\section{Aim Of The Project}
+\section{Aim of the project}
+\label{sec:aim}
+
+This research project aims to determine whether gene expression signatures exist that are specific to obesity across multiple different cancer types, and to investigate whether there are any common pathways being dysregulated in cancers based on these genetic signatures.
+Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to improved clinical decisions, and could thus contribute towards personalised treatment in clinical settings in the future.

### 2016-04-18 thesis/methods/methods.tex
tmp commit

@@
 \chapter{Methods}
+\label{ch:methods}
 
 \section{Body Mass Index (BMI)}
+\label{sec:bmi}
 
 \subsection{BMI Calculation}
+\label{subsec:bmicalc}
 
-BMI was calculated from the height and weight data of the samples using the following equation:\\
+BMI was calculated from the height and weight data of the samples using the following equation:
+
 \begin{equation}
-    BMI = \frac{weight (kg)}{Height^2(m^2)}
+    \label{eq:bmicalc}
+    BMI = \frac{weight (kg)}{Height^2(m^2)}\\
 \end{equation}
@@
 \subsection{BMI Classification}
+\label{subsec:bmiclassification}
 
-Samples were classified based on the WHO definition:
-\begin{table}[h]
-    \caption{WHO defined BMI classification}\\
-    \label{tab:who-bmi-class}
+Samples were classified based on the WHO definition, as shown in Table~\ref{tab:whobmiclass}.
+\begin{table}[hb]
+    \caption{WHO defined BMI classification}
+    \label{tab:whobmiclass}
     \begin{center}
@@
 \section{Download Data from publicly available source}
+\label{sec:data}
 
@@
 \subsection{The Cancer Genome Atlas (TCGA)}
+\label{subsec:tcga}
 
@@
 \subsection{International Cancer Genome Consortium (ICGC)}
+\label{subsec:icgc}
 
-RNA-seq data for the eight cancer types with sample BMI information were downloaded from the ICGC database.
+RNA-seq data for the eight cancer types with sample BMI information available were downloaded from the ICGC database.
+The RNA-seq data for each of the eight cancer types were imported into R.
 
 \section{Validation of methodology used by \citet{Creighton2012}}
+\label{sec:methodvalidation}
 
-\subsection{Raw data from \citet{Creighton2012} paper}
+\subsection{Raw data from the \citet{Creighton2012} study}
+\label{subsec:rawdatacr}
 
@@
 
-\subsection{Normalisation of \citet{Creighton2012} data}
+(add data/sample composition -- race, grade, etc -- maybe in table form)
+
+\subsection{Normalisation of raw \citet{Creighton2012} data}
+\label{subsec:normcrdata}
 
@@
 
+\subsection{Obesity associated gene probes from the \citet{Creighton2012} study}
+\label{subsec:crobsgene}
+
+799 obesity associated gene probes identified by \citet{Creighton2012} were downloaded from (where? - i think the same place as microarray data).
+The list of gene probes were imported into R for further analyses and validation of methodology.
+
 \subsection{Differential gene expression analysis}
+\label{subsec:deg}
 
-(mention WHO classification of obese/overweight/lean)\\
-(also mention limma package)
+All 103 samples from the \citet{Creighton2012} study were split into two groups (obese and non-obese) based on the sample BMI status as defined by the WHO classification (see Table~\ref{tab:whobmiclass}).
+The difference in the gene expression between these two sample groups were analysed by fitting a linear model to the data, using the \textit{limma} package in R.
+The top 799 differentially expressed gene probes that had p-value \textless{} 0.01 were identified.
+
+The process above was repeated on the same data, with same group (obese and non-obese) after controlling for all other clinical variables that were not related to the sample BMI (age, sex, race, tumour grade, lymph node (LN) status, progesterone receptor (PR) status, and estrogen receptor (ER) status).
+The top 799 differentially expressed gene probes that had p-value \textless{} 0.01 between the two groups were identified.
+
+To completely remove the effect of race on the analysis, the analysis was repeated using only the Caucasian samples.
+This reduced the sample size from 103 samples to (70 something) samples (check sample size).
+The data was controlled for all other clinical variables except the variables that were related to BMI (age, sex, tumour grade, LN status, PR status, and ER status).
+Again, the top 799 differentially expressed genes between the two groups with p-value \textless{} 0.01 were identified.
+
+\subsection{Comparison of \citet{Creighton2012} gene probes and identified gene probes}
+\label{subsec:cfgeneprobes}
+
+The obesity associated gene probes identified by \citet{Creighton2012} were compared with each of the three sets of gene probes identified from the differential gene expression analysis (see \ref{subsec:deg} \nameref{subsec:deg}).
+The presence of any of the \citet{Creighton2012} genes were checked for in each of the gene sets identified at a p-value \textless{} 0.05, as well as the top 799 gene probes at p-Value \textless{} 0.01.
 
 \subsection{Singular Value Decomposition}
+\label{subsec:svd}
+
+(need svd derivation and why I'm using it)
+
+\subsection{Correlation between obesity associated metagene and sample BMI}
+\label{subsec:metavsbmi}
+
+In order to check whether the metagenes were actually specific to the BMI status of the samples, the metagenes were plotted against both the BMI statuses and the BMI values of the samples.
+The metagenes were ranked for each sample, and a box plot and a scatter plot were plotted for sample BMI statuses and BMI values, respectively.
 
 \section{Applying metagene to other cancer types}
+\label{sec:metagene}
+
+(How you could transfer the metagene to other cancer data)
 
 \subsection{Transformation matrix}
+\label{subsec:transmat}
+
+(why transformation matrix is used)
+(refer back to eqiations in the  svd section)
 
 \section{Pathway Enrichment Analysis}
+\label{sec:pathenrich}

### 2016-04-22 summary_report/summary.tex
started writing up a summary in prep to committee meeting; some work done on the thesis

\documentclass[a4paper, 11pt]{article}

\usepackage[utf8]{inputenc}
\begin{document}

\section*{Summary so far}

\subsection*{Creighton data}

Creighton et al. found obesity-associated genetic signatures that are specific to obese samples, but not in the lean or overweight samples, in a study using breast tumour samples from patients.

\end{document}

### 2016-04-22 thesis/methods/methods.tex
started writing up a summary in prep to committee meeting; some work done on the thesis

@@
-\chapter{Methods}
+\chapter{Methods (draft)}
 \label{ch:methods}
@@
 Any cancer type with no weight and/or height data of the samples were excluded from the project, as no BMI information can be obtained without these data.
-This gave a total of eight cancer types (**insert all the names of cancer types here) to be used for further analyses.
+This gave a total of eight cancer types (bladder urothelial carcinoma (BLCA), cervical squamous cell carcinoma and endocervical adenocarcinoma (CESC), colon adenocarcinoma (COAD), kidney renal clear cell carcinoma (KIRP), liver hepatocellular carcinoma (LIHC), rectum adenocarcinoma (READ), skin cutaneous melanoma (SKCM), and uterine corpus endometrial carcinoma (UCEC)) to be used for further analyses.
 
@@
 The raw microarray data for all 103 samples in the study were normalised using Robust Multi Array (RMA) method from the \textit{affy} package in R.
+The data was then standardised so that each of the genes had a mean of 0 and a standard deviation of 1.
+(mention scaling anything over 3 were made to be 3 for heatmap scaling)
 
@@
 
-The process above was repeated on the same data, with same group (obese and non-obese) after controlling for all other clinical variables that were not related to the sample BMI (age, sex, race, tumour grade, lymph node (LN) status, progesterone receptor (PR) status, and estrogen receptor (ER) status).
+The process above was repeated on the same data, with the same group (obese and non-obese) after controlling for all other clinical variables that were not related to the sample BMI (age, sex, race, tumour grade, lymph node (LN) status, progesterone receptor (PR) status, and estrogen receptor (ER) status).
 The top 799 differentially expressed gene probes that had p-value \textless{} 0.01 between the two groups were identified.

### 2016-04-22 thesis/results/results.tex
started writing up a summary in prep to committee meeting; some work done on the thesis

@@
-\chapter{Results from \citet{Creighton2012} metagene}
+\chapter{Results from \citet{Creighton2012} metagene (draft)}
 
-\section{SVD}
+\section{Validation of previously identified obesity-associated genetic signatures}
+
+(draft)\\
+To validate the ability for the obesity-associated biomarkers to categorise the obese samples from non-obese samples, singular value decomposition with the previously identified 799 gene probes was used on the raw data.
+The metagene produced from this was used to check whether the obesity-associated biomarkers were able to identify the difference in obesity groups within the samples.
+(insert first figure - gene expression sorted by metagene)
+The resulting heatmap shown in figure~(fig:crobsheat) shows how the overall gene expression of each sample is represented well by the metagene created from the 799 gene probes identified by \citet{Creighton2012}.
+
+To assess whether the metagene was actually correlating with the sample BMI values and statuses, a scatter plot and a box plot were plotted against the metagene, respectively.
+These plots proved that the obesity-associated signatures identified by \citet{Creighton2012} were able to classify the samples based on their BMI status (specifically the obese samples from the non-obese samples), confirming the results fromthe study.
+
+\section{Reproducing the obesity-associated genetic signatures}
+
+(draft - are we able to find the same/similar genes as creighton, using the same data?)\\
+To see whether we were able to re-identify the same, or similar, gene probes as \citet{Creighton2012}, differential gene expression analyses were carried out between the obese and non-obese samples, with different statistical conditions.
+
+\section{Application of the obesity-associated genetic signatures to other cancer types}
+
+(draft - found nothing)
+
+\section{\citet{Fuentes-Mattei2014} gene signatures on \citet{Creighton2012} data and vice versa}
+
+(draft)
+
+\section{\citet{Fuentes-Mattei2014} gene signatures on TCGA data and vice versa}
+
+(draft)

### 2016-04-25 summary_report/summary.tex
tmp commit

@@
 Creighton et al. found obesity-associated genetic signatures that are specific to obese samples, but not in the lean or overweight samples, in a study using breast tumour samples from patients.
+Total of 103 samples were used in the study, majority (about 70 odd samples) were caucasian, and rest of them black or asian.
+Expression of genes from the samples were quantified using a microarray chip.
+The samples were split into three groups: obese, overweight, and lean samples.
+
+They initially looked at the difference in the gene expression between lean vs. non-lean, but they switched it to obese vs. non-obese group (read paper again).
+(I think they did an ANOVA to see the expression difference between all three groups)
+From this, they found 799 genetic signatures (600 odd unique genes) that were specific to the BMI status of the samples (i.e.
+the genetic signatures were able to identify obese  samples from the non-obese samples).
+These were identified as obesity specific genes, as they did not correlate (?) with any other clinical variable (judged by the heatmap??).
+
+They identified the genes to be involved in AKT/mTOR pathway, and further validated the results using drugs specific to these on mice with breast cancer (?, maybe it was the Fuentes-Mattei paper).
+With these results, we decided to see if these obesity-associated genetic signatures were able to predict the BMI status of the samples/patients in other tumour types from TCGA database.
+
+\subsection{TCGA}
 

### 2016-04-26 summary_report/summary.tex
tmp commit

--- 2016-04-22 summary_report/summary.tex
@@
 
-\subsection*{Creighton data}
+\section*{Creighton study}
 
-Creighton et al. found obesity-associated genetic signatures that are specific to obese samples, but not in the lean or overweight samples, in a study using breast tumour samples from patients.
+\subsection*{Intro and methods}
+
+Creighton et al. found obesity-associated genetic signature that is specific to obese samples, but not in the lean or overweight samples.
+The study used pretreatment breast tumour biopsies from 103 patient (77 Caucasian, 16 Black, and 10 Asian patients).
+They noticed that the Black patients were more likely to be obese compared to the Caucasian patients (13/16 vs. 22/77), but larger sample size would have made it more convincing.
+Race specific BMI classification may have given a different result (less Black patients may have been classified as obese).
+
+The quality of the tumour (tumour cell enrichment per biopsy - i.e.
+ratio of tumour cells and normal cells) was controlled by trimming the sample in such a way that the tumour cells in each sample were $\geq$ 70\% enriched.
+Expression of genes from these samples were quantified using an affymetrix HG-U133A microarray chip.
+
+The samples were split into three groups using the patient BMI status.
+``Tumours from obese patients were not particularly enriched for a specific molecular subtype, as defined by ER and HER2 status''.
+
+\subsection*{Results}
+
+First, they looked for any genes that were differentially expressed in any of the three BMI groups.
+Unsupervised hierarchical clustering of the genes separated the tumours based on the ER status rather than BMI status.
+They used ANOVA to look for any genes differentially expressed in any of the three BMI groups.
+From this, they got 658 probes at p \textless 0.01, which greatly exceeded the number you would get by chance (approx. 22000 genes $\times$ 0.01 = 220 genes).
+In a heatmap of these 658 gene probes, they noticed that almost all of the genes were differentially expressed in the obese group compared with both overweight and lean group (overweight and lean group appeared homogeneous).
+
+Since the overweight and lean groups looked similar to each other, they grouped these two into a single group and looked for differentially expressed genes between the obese and non-obese groups.
+They found 799 gene probes that were p \textless 0.01 and fold change \textgreater 1.2.
+They only found a median of 82 gene probes when they chose random samples to do gene expression analysis (1000 permutations done), showing that the 799 gene probes were statistically significant.
+
+The heatmap with these genes (fig1 of their paper) showed that the expression was uniform in the obese group, and not specific to tumour grade, ER/PR/HER2 status.
+This was validated in the supplementary figure 4 (haven't looked at it, but I assume it's a figure of box plots showing no significant difference between obese and non-obese groups).\\
+
+They then decided to to use other publicly available breast tumour data to see the relevance of their signature in other data.
+They ranked each sample according to the obesity-associated gene signature they have found (metagene and transformation matrix on other data).
+The results showed that high obesity gene signature scores correlated with a shorter time to metastasis and may represent genes that promote cancer progression only in van de Vijver dataset.
+``... this association with worse outcome may be dependent on the specific patient population under consideration.''
+
+There was no overlap in the genes activated by insulin growth factor (IGF-1) and other oncogenic pathways (MAPK and PI3K) and the genes highly expressed in obesity-associated tumours.
+There was some overlap (\textless 12\%) in genes repressed by the same pathways and the genes low in obese samples.
+They concluded that the obesity-associated genetic signature was distinct from oncogenic pathway signature (different oncogenic mechanism?).
+
+They did a metagene analysis with their data and other data with the IGF-1 signature.
+With their dataset, high IGF-1 score was observed in obese tumour group, compared to the rest of the tumours.
+In other datasets, they found that high IGF-1 scores correlated with high obesity-associated scores (table 2 in their paper).
+They didn't mention the correlation of obesity-associated scores and IGF-1 scores in their original data.
+Shows that both IGF1 and obesity-associated signatures are able to split the samples based on their BMI status, and/or contribute to cancer progression, but in an independent way?
+Maybe IGF1 uses mechanism x, whereas obesity-associated genes use mechanism y, but both contribute to cancer progression.
+
+\section*{Fuentes-Mattei study}
+
+coming soon
+
+\section*{What I have been doing (chronologically)}
+
+\subsection*{Data access and processing}
+
+First, I downloaded all the clinical data for all cancer types available online from the ICGC database and checked which cancer types had patient height and weight data.
+There were 14 cancer types in total that had both height and weight data, but only 8 cancer types from this list had RNA-seq data available.
+This gave me 8 cancer types to work with:
+\begin{description}
+	\item[BLCA] Bladder urothelial carcinoma
+	\item[CESC] Cervical squamous cell carcinoma and endocervical adenocarcinoma
+	\item[COAD] Colon adenocarcinoma
+	\item[KIRP] Kidney renal papillary cell carcinoma
+	\item[LIHC] Liver hepatocellular carcinoma
+	\item[READ] Rectal adenocarcinoma
+	\item[SKCM] Skin cutaneous melanoma
+	\item[UCEC] Uterine corpus endometrial carcinoma
+\end{description}
+
+The RNA-seq data for these cancer types were downloaded from TCGA database.
+
+\subsubsection*{Normalisation and standardisation}
+
+Creighton and Fuentes-Mattei data were imported into R and normalised using rma function.
+These data were standardised to have a mean of 0 and standard deviation of 1.
+
+The cancer data were imported into R and checked for the samples that had both height and weight data.
+Any samples without either height and/or weight data were removed from the analysis (removed from the clinical data and RNA-seq data).
+After this, all of the cancer data were normalised using Voom function  from limma (after it has been logged to the base 10).
+These were then standardised just like the Creighton and Fuentes-Mattei data (mean = 0, sd = 1).
+
+\subsection*{Metagene analysis and transformation matrix}
+
+\subsubsection*{Creighton's obesity-associated genes}
+
+799 gene probes identified by the study were used to see if the obesity-associated genetic signature was able to separate the samples based on their BMI status in different cancer types.
+First, I checked that these gene probes were able to separate the samples based on their BMI status like they have said in their paper.
+After normalising and standardising the data, a metagene  was made using SVD.
+Metagene represented each of the sample's obesity-associated gene expression level (i.e. obesity-associated gene scores that was described in the Creighton paper).
+
+The gene expression level of each sample correlated with their metagene score (i.e. high/low expression of the genes when metagene was high/low).
+To confirm that the metagene score was actually separating the samples based on the BMI status, a boxplot was made with the BMI status against the metagene score, as well as a scatter plot of the metagene against raw BMI value of each sample.
+These plots showed that the metagene scores were able to separate the samples into obese or non-obese groups, showing that the obesity-associated gene signature was real in their data.
+
+\subsubsection*{Transformation of Creighton obesity genes to TCGA cancer types}
+
+To transform the cancer data using a list of genes from another data, the transformation matrix from the SVD must have the same number of genes in both datasets (due to matrix multiplications).
+So, to make sure that both the Creighton and TCGA data had the same genes, the probe IDs of the obesity-associated genes were converted into gene symbols first (as TCGA data is RNA-seq data, not microarray).
+
+The gene symbols were matched against the genes present in one of the cancer types downloaded (all cancer types had the same genes, so the resulting gene list was applicable to other types).
+Any duplicates of gene symbols or gene probe IDs  were removed using the collapseRows function from WGCNA package.
+
+Transformation matrix was formed from the resulting list of genes that were common in both the Creighton and TCGA cancer data.
+The cancer data were transformed using this matrix, and the first column of the resulting matrix was used as the metagene for that cancer type.
+
+Heatmaps were made for each cancer type, using the metagene from the transformation.
+The overall gene expression profile for each cancer type correlated with the metagene score (i.e. high/low metagene score corresponded to high/low expression level of the selected genes).
+To confirm whether this association was based on the BMI status of the samples, a boxplot and a scatter plot was plotted with the metagene score against the BMI status and raw BMI value of the samples.
+
+Surprisingly, the apparent correlation of the overall gene expression level with metagene scores were not related to the samples BMI status -- the metagene scores were not able to separate the samples based on their BMI status like in the Creighton dataset.
+
+Since clinical data was not available  for Fuentes-Mattei  dataset, it was not possible to validate the Creighton obesity-associated genes on this dataset.
+However, assuming that the genetic signatures identified by Fuentes-Mattei were obesity-associated, it was possible to see whether their signatures were able to separate the samples based on BMI status.
+
+\subsubsection*{Transformation of Fuentes-Mattei obesity genes to Creighton and TCGA cancer types}
+
+Like with Creighton data, the gene signature from Fuentes-Mattei study was used to make a metagene and a transformation matrix.
+Again, any genes not present in the TCGA data were removed from the signature to make sure that both data sets had the same genes for transformation.
+
+Metagene was created in the Fuentes-Mattei data first and heatmap was produced.
+The heatmap showed that the overall gene expression for each sample correlated with the metagene score, which was expected.
+Since there was no clinical data available for the data set, the ability for the signature to separate the samples into obese or non-obese was not possible.
+
+To see if the Fuentes-Mattei signature was able to be used in other data, Creighton and TCGA data were transformed using the Fuentes-Mattei signature.
+Data was transformed and heatmaps, boxplot, and scatter plots were made for each data set to see if the signature correlated with overall gene expression and/or BMI status.
+
+The heatmaps for all the data showed similar results in that the overall gene expression of the samples correlated with the signature, but was not able to separate the obese samples from non-obese samples.
+This result showed that the signature from Fuentes-Mattei study captured the overall gene expression of the samples, but the signature was not specific to the BMI status of the samples.
+
+These results suggested that the signatures from previous studies are not able to identify the obese samples from the non-obese samples in different cancer types, and possibly not even in the same tumour type.
+
+\subsubsection*{What I think about the results}
+
+The lack of identification in different tumour types can be explained by different tumour biology for different types of tumour, together with tumour heterogeneity and other factors that are specific to the local environment of the tumour.
+
+However, the lack of evidence in the same tumour types were unexpected, as the tumour biology should be similar to one another, if not the same.
+Obvious difference between the two data sets was that Fuentes-Mattei study used tumour samples from ER positive patients, whereas the Creighton study used all the tumour types available at the time.
+It could be that the Fuentes-Mattei signature was specific to ER+ samples (repeat the same analysis but use only ER+ samples from Creighton data).
+
+Another reason could be that, since the tumours are known to be heterogeneous, the samples used in either studies were significantly different from each other.
+This may result in signatures that they thought was ``tumour specific'', but in fact it was quite different to what they have expected.
+I guess if those studies used more samples, you might be able to capture the effect of tumour heterogeneity more and dampen the effect that is complicating the signature's ability to identify obese samples.\\
+
+\newline
+
+Since the signatures identified by either studies could not separate the obese samples from non-obese samples, we decided to see whether there were any differentially expressed genes common in all cancer types (that are specific to obesity).
+If there was any, then use these genes to see if it can identify the obese samples from non-obese samples in Creighton data set.
+
+\subsection*{Gene expression analysis}
+
+\subsubsection*{Creighton data}
+
+I first started the analysis on the Creighton data to check their results.
+I have splitted the samples into two groups like in the study, but in two different ways: lean vs. overweight/obese, and lean/overweight vs. obese.
+
+\begin{center}
+	\begin{tabular}{c|c|c}
+                                  & P \textless{} 0.05 & P \textless{} 0.01 \\
+		\hline
+		Lean vs. overweight/obese & 759              & 104              \\
+		Lean/overweight vs obese  & 5278             & 1781             \\
+	\end{tabular}
+\end{center}
+
+The adjusted p-values were not used, as no genes showed up when adjusted.
+These results were different from the original Creighton data, which was mildly concerning...
+(Didn't check for the overlaps of the obesity-associated genes with the genes I found, but I did that later on).
+
+\subsubsection*{Common genes in TCGA cancer types}
 

### 2016-04-27 summary_report/summary.tex
thinking about future directions

@@
 \usepackage[utf8]{inputenc}
+\usepackage{booktabs}
+
 \begin{document}
@@
 This may result in signatures that they thought was ``tumour specific'', but in fact it was quite different to what they have expected.
-I guess if those studies used more samples, you might be able to capture the effect of tumour heterogeneity more and dampen the effect that is complicating the signature's ability to identify obese samples.\\
-
+I guess if those studies used more samples, you might be able to capture the effect of tumour heterogeneity more and dampen the effect that is complicating the signature's ability to identify obese samples.
+(Creighton's group did mention that their signature ``may be dependent on the specific patient population under consideration''.)\\
 \newline
@@
 \begin{center}
-	\begin{tabular}{c|c|c}
+	\begin{tabular}{ccc}
                                   & P \textless{} 0.05 & P \textless{} 0.01 \\
@@
 
+To see if there were any genes common across all cancer types, I did a gene expression analysis on each cancer type.
+The samples were split into two groups as in the Creighton study: obese vs. non-obese.
+Each cancer type was logged to the base 10 and Voom normalised before a linear model was fitted to the data.
+The names of the significantly expressed genes with p-value \textless{} 1 for each cancer type was recorded and compared with the other cancer types (forgot to check the fold changes).
+
+\begin{center}
+	\begin{tabular}{cccc}
+             & Lean & Overweight & Obese \\
+		BLCA & 103    & 97         & 61 \\
+		CESC & 85     & 62         & 77 \\
+		COAD & 75     & 80         & 71 \\
+		KIRP & 26     & 53         & 45 \\
+		LIHC & 140    & 75         & 49 \\
+		READ & 22     & 36         & 15 \\
+		SKCM & 72     & 73         & 73 \\
+		UCEC & 90     & 106        & 286 \\
+	\end{tabular}
+\end{center}
+
+The number of genes that were present in any two cancer types were noted down, and was repeated for the genes present in any three cancer types, and so on.
+This was repeated for groups with randomly chosen samples and were repeated 100 times (the proportion of the groups were kept the same - refer to the table above).
+
+\begin{center}
+	\begin{tabular}{ccccccccc}
+		& \multicolumn{8}{c}{Number of cancer types the gene overlapped} \\
+                                     &    1 &    2 &   3 &  4 & 5 & 6 & 7 & 8 \\
+		\midrule
+		         Obese vs. non-obese & 6865 & 1907 & 289 & 26 & 1 & 0 & 0 & 0 \\
+		Randomly split sample groups & 5012 &  870 &  89 &  6 & 0 & 0 & 0 & 0 \\
+	\end{tabular}
+\end{center}
+
+This table showed how the number of differentially expressed genes across all cancer types were consistently greater  than the amount of genes found between randomly split groups of samples.
+I think I had a number for the 95$^{th}$ percentile, but I can't find it right now -- might have to re-fo it... (I think the result shown was still over the 95$^{th}$ percentile mark).
+(I think I did one with 1000 iterations -- still looking for the results).
+
+This result shows that the number of differentially expressed genes observed between the obese and non-obese groups were greater than the number of genes one would expect by chance.
+However, as you can see from the table, there were no differentially expressed genes that were common across all cancer types, suggesting that there may be no general cancer markers that are specific to obesity.\\
+
+After this, we decided to see whether there were any significant pathways expressed in any of the cancer types, and if there is any, check the relevance of the pathway in obesity and/or cancer and use the genes that are involved in those pathways as a starting point for further analysis.
+
+\subsection*{Pathway enrichment analysis}
+
+First, I made a matrix of genes by pathway for each of the pathway databases (reactome, KEGG, and GO).
+Since the GO database had the most pathways and genes recorded, I decided to use this for the analysis.
+For each cancer type, I listed all of the significantly expressed genes between obese vs. non-obese groups.
+These genes were used to pull out the relevant genes from the matrix that I have made, which was used to do Fisher's exact test on.
+The results were corrected for multiple testing, using FDR control.
+
+The results (not shown) showed that there were  no significantly enriched pathway in any of the cancer types investigated.
+To see whether this was due to the genes being ``almost'' significant (and thus not included in the Fisher's test), I used the rank-based approach.
+To do this, I used the camera function from the limma package (as geneSetTest function was ``out-dated'' according to the manual).
+However, this still gave the same result as the Fisher's test, where none of the pathways were enriched in any of the cancer types.
+
+At this point, we went back to the Creighton data to see whether their obesity-associated genes were actually related to obesity and no other clinical variables.
+
+\subsection*{Creighton data revisited}
+
+Firstly, I re-did the gene expression analysis on the Creighton data and compared the genes that I found significant with their obesity-associated genes.
+Although all of Creighton's obesity-associated genes showed up in the list of genes that were p \textless{} 0.01, not all gene were in the top 799 genes in that list.
+
+The 799 genes that I have found were able to separate the obese group from the non-obese group like in the Creighton paper, even though the genes that were present in the list were quite different to the obesity-associated genes from Creighton's paper.
+To support this, the metagene scores from my 799 genes were compared with the metagene scores from the Creighton's obesity-associated genes on a scatter plot.
+This showed that both metagene scores were similar to one another, showing how both sets of genes are ``markers'' for obese patients' cancer, even if the genes they are derive from are different.
+
+When the genes that were common in both sets of genes (my genes and Creighton's genes) were used, it showed slightly better seapration of the obese samples from the non-obese samples (239 genes used).\\
+
+To make sure that the genes were not affected by any other clinical variables that are not related to obesity, a linear model was constructed with those variables included in the model.
+The variables included age, race, menopause status, tumour grade, LN status, ER status, PR status, and HER2 status.
+The residual data (remaining data after controlling for these varaibles) were used to get the differentially expressed genes between the obese and non-obese groups.
+
+In Creighton's study, they stated that they found 799 gene probes that were p \textless{} 0.01 and fold change \textgreater{} 1.2 between the obese and non-obese groups.
+However, there were not many gene probes with fold change greater than 1.2 (132 genes) even after logging/exponentiating, and there were significantly greater number of genes with p \textless{} 0.01 (more than 799 gene probes).
+Again, all of the genes that Creighton's team had identified were in the list of significant genes, but not all of them showed up in the top 799 genes.
+
+The heatmaps of the metagene scores from the top 799 genes (from residual data), Creighton genes that were in the top 799 genes (216 genes) showed similar results as the genes identified before controlling for other clinical variables.
+There were 140 genes present in the 799 residual genes that were also in the 239 genes found to be common in both Creighton and 799 genes (from raw data).
+The metagene scores generated from these genes showed better separation of the groups compared to either one on its own, similar to the results with 239 genes.\\
+
+To completely control for the effect of race in the analysis, differentially expressed genes were also obtained using only the Caucasian samples (n = 77).
+The top 799 genes were obtained, as well as the common genes from the Creighton genes that were in those 799 genes (x genes).\\
+
+Now, all of the genes mentioned above were used to get construct a transformation matrix in the raw data (unadjusted for clinical variables).
+These matrices were used to transform the cancer data and metagene scores were obtained as described in previous section.
+Unfortunately, none of the gene sets used were able to separate the samples based on their BMI status.
+
+\section*{Summary (of a summary)}
+
+It seems like that none of the genes that are differentially expressed in obese samples can not be used as a marker for obesity and cancer in other cancer types, nor in the same cancer type but with a different sample population.
+This was mentioned (very briefly) in the Creighton paper, when they were talking about other publicly available human breast tumour data.
+
+As mentioned earlier, this could be due to different cancer types having different tumour biology to one another, together with the tumour heterogeneity within and between the samples.
+Sample quality of the tumours used in the studies may have been problematic as well, especially in the TCGA cancer data as we don't know how those samples were taken and sequenced.
+Greater sample size and/or quality of the samples would help us focus on the true obesity-associated gene signature.\\
+
+I guess the results from the pathway enrichment analysis supports the idea of sample variability and quality.
+If all the cancer samples were developing in a similar fashion to one another, then there would be at least a single pathway coming up as significant.
+Seeing that there are no pathways coming up as significant, it suggests that the sample size and/or quality was too poor to find anything in common.
+
+An alternative explanation (or together with poor sample quality and size) could be that there were no significantly enriched pathways because the samples were so different from one another.
+It could be that each tumour sample have different pathways activated compared to other samples, but still getting the same net result (cancer progression).
+If this was true, this would suggest that there are a huge number of independent pathways that a tumour could take to progress through cancer, with no overlap or ``bottleneck'' pathway.
+However, I doubt that this would be the case as there are very limited pathways for cell cycle/proliferation/transcription which the cancer will eventually have to take.\\
+
+Although all of Creighton's obesity-associated genes came up in the list of significantly differentially expressed genes, it was concerning to find that not all genes found by Creighton were not in the top 799 genes of the genes that I have found.
+Also, there were not many genes that were expressed at a fold change greater than 1.2 like they have claimed in their paper.
+The difference in these results were probably due to the difference in the methods used, rather than other clinical variables affecting tha analysis.
+
+The fact that none of the obesity-associated gene sets found in the analyses (with or without controlling for other clinical variables) were able to separate the obese samples from the non-obese samples strongly suggests that there may not be any obesity-associated genetic signature for cancer in general (at least in the samples currently available to us).\\
+
+\section*{Future directions}
+
+(Re-)do everything that I have missed so far.
+
 \end{document}

### 2016-04-28 summary_report/summary.tex
minor proof reading

@@
 The study used pretreatment breast tumour biopsies from 103 patient (77 Caucasian, 16 Black, and 10 Asian patients).
-They noticed that the Black patients were more likely to be obese compared to the Caucasian patients (13/16 vs. 22/77), but larger sample size would have made it more convincing.
-Race specific BMI classification may have given a different result (less Black patients may have been classified as obese).
+They noticed that the Black patients were more likely to be obese compared to the Caucasian patients (13/16 vs. 22/77).
 
@@
 There was no overlap in the genes activated by insulin growth factor (IGF-1) and other oncogenic pathways (MAPK and PI3K) and the genes highly expressed in obesity-associated tumours.
-There was some overlap (\textless 12\%) in genes repressed by the same pathways and the genes low in obese samples.
-They concluded that the obesity-associated genetic signature was distinct from oncogenic pathway signature (different oncogenic mechanism?).
+There was some overlap (\textless 12\%) in genes repressed by IGF1/MAPK/PI3K pathways and the genes low in obese samples.
+They concluded that the obesity-associated genetic signature was distinct from oncogenic pathway signature (different oncogenic mechanism in obese samples?).
 
@@
 With their dataset, high IGF-1 score was observed in obese tumour group, compared to the rest of the tumours.
-In other datasets, they found that high IGF-1 scores correlated with high obesity-associated scores (table 2 in their paper).
+In other publicly available datasets, they found that high IGF-1 scores correlated with high obesity-associated scores (table 2 in their paper).
 They didn't mention the correlation of obesity-associated scores and IGF-1 scores in their original data.
@@
 The cancer data were imported into R and checked for the samples that had both height and weight data.
-Any samples without either height and/or weight data were removed from the analysis (removed from the clinical data and RNA-seq data).
+Any samples without height and/or weight data were removed from the analysis (removed from both the clinical data and RNA-seq data).
 After this, all of the cancer data were normalised using Voom function  from limma (after it has been logged to the base 10).
@@
 
-The gene expression level of each sample correlated with their metagene score (i.e. high/low expression of the genes when metagene was high/low).
+The overall gene expression level of each sample correlated with their metagene score (i.e. high/low expression of the genes when metagene was high/low).
 To confirm that the metagene score was actually separating the samples based on the BMI status, a boxplot was made with the BMI status against the metagene score, as well as a scatter plot of the metagene against raw BMI value of each sample.
@@
 The overall gene expression profile for each cancer type correlated with the metagene score (i.e. high/low metagene score corresponded to high/low expression level of the selected genes).
-To confirm whether this association was based on the BMI status of the samples, a boxplot and a scatter plot was plotted with the metagene score against the BMI status and raw BMI value of the samples.
+To confirm whether this association was based on the BMI status of the samples, a boxplot and a scatter plot was plotted with the metagene scores against the BMI status and raw BMI value of the samples.
 
-Surprisingly, the apparent correlation of the overall gene expression level with metagene scores were not related to the samples BMI status -- the metagene scores were not able to separate the samples based on their BMI status like in the Creighton dataset.
+Surprisingly, the apparent correlation of the overall gene expression level with metagene scores were not related to the sample BMI status -- the metagene scores were not able to separate the samples based on their BMI status like in the Creighton dataset.
 
 Since clinical data was not available  for Fuentes-Mattei  dataset, it was not possible to validate the Creighton obesity-associated genes on this dataset.
-However, assuming that the genetic signatures identified by Fuentes-Mattei were obesity-associated, it was possible to see whether their signatures were able to separate the samples based on BMI status.
+However, assuming that the genetic signatures identified by Fuentes-Mattei were obesity-associated, it was possible to see whether their signature was able to separate the samples based on BMI status.
 
@@
 The heatmap showed that the overall gene expression for each sample correlated with the metagene score, which was expected.
-Since there was no clinical data available for the data set, the ability for the signature to separate the samples into obese or non-obese was not possible.
+Since there was no clinical data available for the data set, the ability for the signature to separate the samples into obese or non-obese was not checked.
 
@@
 
-The heatmaps for all the data showed similar results in that the overall gene expression of the samples correlated with the signature, but was not able to separate the obese samples from non-obese samples.
+The heatmaps for all the data showed similar results as before, in that the overall gene expression of the samples correlated with the signature, but was not able to separate the obese samples from non-obese samples.
 This result showed that the signature from Fuentes-Mattei study captured the overall gene expression of the samples, but the signature was not specific to the BMI status of the samples.
@@
 
-The lack of identification in different tumour types can be explained by different tumour biology for different types of tumour, together with tumour heterogeneity and other factors that are specific to the local environment of the tumour.
+The lack of identification in different tumour types can be explained by different tumour biology in different tumours, together with tumour heterogeneity and other factors that are specific to the local environment of the tumour.
 
-However, the lack of evidence in the same tumour types were unexpected, as the tumour biology should be similar to one another, if not the same.
+However, the lack of evidence in the same tumour types was unexpected, as the tumour biology should be similar to one another, if not the same.
 Obvious difference between the two data sets was that Fuentes-Mattei study used tumour samples from ER positive patients, whereas the Creighton study used all the tumour types available at the time.
-It could be that the Fuentes-Mattei signature was specific to ER+ samples (repeat the same analysis but use only ER+ samples from Creighton data).
+It could be that the Fuentes-Mattei signature was specific to ER+ samples (Note: repeat the same analysis but use only ER+ samples from Creighton data).
 
@@
 Each cancer type was logged to the base 10 and Voom normalised before a linear model was fitted to the data.
-The names of the significantly expressed genes with p-value \textless{} 1 for each cancer type was recorded and compared with the other cancer types (forgot to check the fold changes).
+The names of the significantly expressed genes with p-value \textless{} 0.01 for each cancer type was recorded and compared with the other cancer types (forgot to check the fold changes).
 
@@
 This table showed how the number of differentially expressed genes across all cancer types were consistently greater  than the amount of genes found between randomly split groups of samples.
-I think I had a number for the 95$^{th}$ percentile, but I can't find it right now -- might have to re-fo it... (I think the result shown was still over the 95$^{th}$ percentile mark).
-(I think I did one with 1000 iterations -- still looking for the results).
+I think I had a number for the 95$^{th}$ percentile, but I can't find it right now -- might have to re-do it... (I think the result shown was still over the 95$^{th}$ percentile mark).
+(I think I have a result from 1000 iterations somewhere -- still looking for the results).
 
@@
 To see whether this was due to the genes being ``almost'' significant (and thus not included in the Fisher's test), I used the rank-based approach.
-To do this, I used the camera function from the limma package (as geneSetTest function was ``out-dated'' according to the manual).
-However, this still gave the same result as the Fisher's test, where none of the pathways were enriched in any of the cancer types.
+I used the camera function from the limma package (as geneSetTest function was ``out-dated'' according to the manual).
+However, this still gave the same result as the Fisher's test, where none of the pathways were significantly enriched in any of the cancer types.
 
-At this point, we went back to the Creighton data to see whether their obesity-associated genes were actually related to obesity and no other clinical variables.
+At this point, we went back to the Creighton data to see whether their obesity-associated genes were actually related to obesity and no other clinical variables were not affecting the signature.
 
@@
 The 799 genes that I have found were able to separate the obese group from the non-obese group like in the Creighton paper, even though the genes that were present in the list were quite different to the obesity-associated genes from Creighton's paper.
-To support this, the metagene scores from my 799 genes were compared with the metagene scores from the Creighton's obesity-associated genes on a scatter plot.
-This showed that both metagene scores were similar to one another, showing how both sets of genes are ``markers'' for obese patients' cancer, even if the genes they are derive from are different.
+To support this, the metagene scores from my 799 genes were compared with the metagene scores from the Creighton's obesity-associated genes by plotting it on a scatter plot.
+This showed that both metagene scores were similar to one another, showing how both sets of genes are ``markers'' for obese patients' cancer, even if the genes they are derived from are different.
 
@@
 The heatmaps of the metagene scores from the top 799 genes (from residual data), Creighton genes that were in the top 799 genes (216 genes) showed similar results as the genes identified before controlling for other clinical variables.
-There were 140 genes present in the 799 residual genes that were also in the 239 genes found to be common in both Creighton and 799 genes (from raw data).
+There were 140 genes present in the 799 residual genes that were also in the 239 genes found to be common in both Creighton and my 799 genes (from raw data).
 The metagene scores generated from these genes showed better separation of the groups compared to either one on its own, similar to the results with 239 genes.\\
@@
 
-Now, all of the genes mentioned above were used to get construct a transformation matrix in the raw data (unadjusted for clinical variables).
+Now, all of the genes mentioned above were used to construct a transformation matrix in the raw data (unadjusted for clinical variables).
 These matrices were used to transform the cancer data and metagene scores were obtained as described in previous section.
@@
 
-It seems like that none of the genes that are differentially expressed in obese samples can not be used as a marker for obesity and cancer in other cancer types, nor in the same cancer type but with a different sample population.
+It seems like that none of the genes that are differentially expressed in obese samples can not be used as a marker for obesity in other cancer types, nor in the same cancer type but with a different sample population.
 This was mentioned (very briefly) in the Creighton paper, when they were talking about other publicly available human breast tumour data.
@@
 As mentioned earlier, this could be due to different cancer types having different tumour biology to one another, together with the tumour heterogeneity within and between the samples.
-Sample quality of the tumours used in the studies may have been problematic as well, especially in the TCGA cancer data as we don't know how those samples were taken and sequenced.
-Greater sample size and/or quality of the samples would help us focus on the true obesity-associated gene signature.\\
+Sample quality of the tumours used in the studies may have been problematic as well, especially in the TCGA cancer data, as we don't know how those samples were taken and sequenced.
+Greater sample size and/or quality of the samples would help us focus on the true obesity-associated gene signature. (maybe).\\
 
-I guess the results from the pathway enrichment analysis supports the idea of sample variability and quality.
+I guess the results from the pathway enrichment analysis support the idea of sample variability and quality.
 If all the cancer samples were developing in a similar fashion to one another, then there would be at least a single pathway coming up as significant.
@@
 An alternative explanation (or together with poor sample quality and size) could be that there were no significantly enriched pathways because the samples were so different from one another.
-It could be that each tumour sample have different pathways activated compared to other samples, but still getting the same net result (cancer progression).
+It could be that each tumour sample has different pathways activated compared to other samples, but still getting the same net result (cancer progression).
 If this was true, this would suggest that there are a huge number of independent pathways that a tumour could take to progress through cancer, with no overlap or ``bottleneck'' pathway.
-However, I doubt that this would be the case as there are very limited pathways for cell cycle/proliferation/transcription which the cancer will eventually have to take.\\
+However, I doubt that this would be the case as there are very limited number of major oncogenic pathways for cell cycle/proliferation/transcription which the cancer will eventually have to take.\\
 
@@
 Also, there were not many genes that were expressed at a fold change greater than 1.2 like they have claimed in their paper.
-The difference in these results were probably due to the difference in the methods used, rather than other clinical variables affecting tha analysis.
+The difference in these results was probably due to the difference in the methods used, rather than other clinical variables affecting the analysis.
 
-The fact that none of the obesity-associated gene sets found in the analyses (with or without controlling for other clinical variables) were able to separate the obese samples from the non-obese samples strongly suggests that there may not be any obesity-associated genetic signature for cancer in general (at least in the samples currently available to us).\\
+The fact that none of the obesity-associated gene sets found in the analyses (with or without controlling for other clinical variables) were able to separate the obese samples from the non-obese samples strongly suggests that there may not be any obesity-associated genetic signature for cancer that can be applied in general (at least in the samples currently available to us).\\
 
@@
 
-(Re-)do everything that I have missed so far.
+\begin{itemize}
+	\item (Re-)do everything that I have missed so far -- i.e. side notes throughout this summary.
+		\item Still working on what to do next.
+\end{itemize}
 

### 2016-05-09 bioc192/intro.tex
editing bioc notes

@@
-This book is intended to provide a basic overview of BIOC192 that will have more than enough details to tutor BIOC192 students.
+This book is intended to provide a ``basic'' overview of BIOC192 that will have enough details to tutor BIOC192 students.
 It is highly recommended to listen to the podcasts and add your own notes to this book to get the maximal use out of this book.
+
+This book is organised in such a way that the contents/topics of each lecture are more in a logical order (in my opinion), rather than in the order the lecturer presents the material.
+I feel like some of the lecturers skip some details, go back to it later on, and go back and forth every now and then.
+This  is quite understandable as they only have an hour to go through a certain aspect of the overall concept, and hence after a block of lectures, you usually understand most of what they were saying in the previous lectures.
+Since this book is not time limited, I have organised it into  a ``logical order'' as best as I can.
 
@@
 Riku Takei \\
-4$^{th}$ year Biochemistry MSc student\\
+5$^{th}$ year Biochemistry MSc student\\

### 2016-05-09 bioc192/lect01.tex
editing bioc notes

@@
 Proteins are the workhorses of the cell -- they are required for every process a cell has to do (e.g. DNA replication, apoptosis, etc.)
-Proteins are very small (50$\sim$100\AA{}, or 5$\sim$10nm big; a ``nanostructure''), yet they all have unique structure and functions that makes them important for cell survival and growth, and therefore often referred to as ``molecular machines''.
+Proteins are very small (50$\sim$100\AA{}, or 5$\sim$10nm big; a ``nanostructure''), yet they all have a unique structure and function that makes them important for cell survival and growth, and therefore often referred to as ``molecular machines''.
 
 Biochemistry is the study of the molecular details of the biology of the cells/living organisms, including proteins, DNA and RNA.
-To understand what components are involved and how they function, you need to know the structure of the thing you're trying to describe -- which, in our case, is the protein.
+\textbf{To understand what components are involved and how they function, you need to know the structure of the thing you're trying to describe}, which, in our case, is the protein.
 
@@
 
-Amino acids are the building blocks of a protein.
+\textbf{Amino acids are the building blocks of a protein.}
 Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).
 
-There are 20 amino acids in our cells that make up the proteins.
+There are 20 (major) types of amino acids in our cells that make up the proteins.
 All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$).
-Due to the variable side chains, all amino acids except glycine are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
+Due to the variable side chains, all amino acids (except glycine) are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
 The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.
 
-\includegraphics[width=\textwidth]{aastructure}
+% \includegraphics[width=\textwidth]{aastructure}
 
@@
 
-As mentioned earlier, all 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
-This means that each of the 20 amino acids have different chemical properties, depending on the side chain group it has.
-Cells take advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster (i.e.
-it gives the protein its functionality, as well as its unique structure).
+All 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
+\textbf{This means that each of the 20 amino acids have different chemical properties, depending on the side chain group it has.}
+The cell takes advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster.
+\textbf{In other words, it gives the protein its functionality, as well as its unique structure.}
+
+\section{Charge of Amino Acid Side Chain}
+
+Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
+The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
+\textbf{When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.}
+Therefore, depending on the pH of the solution, there will be more or less protonated (or deprotonated) molecules in a single amino acid.
+
+The amount of de/protonated amino acids affect the net charge of the protein, and therefore the isoelectric point (pI) of the protein.
+\textbf{The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).}
+
+When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to be in a \textbf{protonated} state.
+Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to be in a \textbf{deprotonated} state.
+
+So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
+
+\begin{center}
+    \textbf
+        {pH \textless{} pK$_a$, then the side chain is protonated \\
+        pH \textgreater{} pK$_a$, then the side chain is deprotonated}
+\end{center}
+
+Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
+However, some amino acid side chains may be charged under physiological condition and this contributes to the net charge of the amino acid, and therefore the charge of the  protein.
+This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.
+
+\begin{center}
+% \includegraphics[width=0.8\textwidth]{aapka}
+\end{center}
 
@@
 
-\includegraphics[width=\textwidth]{20aa}
-
-\section{Charge of Amino Acids}
-
-Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
-The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
-When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.
-Depending on the pH of the solution, there will be more or less protonated molecules.
-
-The amount of de/protonated amino acids affect the net charge of the protein, and therefore the isoelectric point (pI) of the protein.
-The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).
-
-When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to become protonated.
-Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to become deprotonated.
-So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).
-
-\begin{center}
-    \textbf
-        {pH \textless{} pK$_a$, then the side chain is protonated \\
-        pH \textgreater{} pK$_a$, then the side chain is deprotonated}
-\end{center}
-
-Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
-However, some amino acid side chains may be charged under physiological condition and contribute to the net charge of the amino acid, and therefore the protein.
-This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.
-
-\begin{center}
-\includegraphics[width=0.8\textwidth]{aapka}
-\end{center}
+% \includegraphics[width=\textwidth]{20aa}
 
@@
 
-Some amino acids are modified after translation -- this is called the post-translational modification (PTM).
+Some amino acids are modified after the protein has been translated -- this is called the post-translational modification (PTM) of the protein.
 Types of PTM includes:
@@
 \begin{center}
-\includegraphics[width=\textwidth]{amidebond}
+% \includegraphics[width=\textwidth]{amidebond}
 \end{center}
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Proteins are made up from amino acids
+	\item There are 20 types of amino acids
+	\item Each amino acid has a carboxyl group, an amino group, a hydrogen atom, and a variable R group attached to the alpha--carbon (C$_{\alpha}$)
+	\item The charge of amino acid side chain can be altered with the pH of the environment
+		\begin{itemize}
+			\item When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated (i.e. ``neutral charge'')
+			\item When pH \textless{} pK$_a$, then the side chain is protonated
+			\item When pH \textgreater{} pK$_a$, then the side chain is deprotonated
+		\end{itemize}
+	\item The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).
+	\item There are four subgroups of amino acids: Non-polar (hydrophobic), polar uncharged, positively charged, and negatively charged amino acids
+	\item Post translational modification is when the amino acids are modified after they have been translated into a protein
+\end{itemize}

### 2016-05-09 bioc192/lect02.tex
added graphics
editing bioc notes

@@
 
-\section{Bond Rotation}
+\section{Secondary Structure of a Protein}
+
+There are three main secondary structures in a protein: $\alpha$-helix, $\beta$-sheets, and turns.
+These secondary structures of a protein has to be made in accordance with the allowed bond angles of the amide plane of the peptide backbone.
+
+\subsection{Bond Rotation}
 
@@
 
-\subsection{Phi ($\phi$) Angle}
+\subsubsection{Phi ($\phi$) Angle}
 
@@
 
-\subsection{Psi ($\Psi$) Angle}
+\subsubsection{Psi ($\Psi$) Angle}
 
@@
 
-\subsection{Omega ($\omega$) Angle}
+\subsubsection{Omega ($\omega$) Angle}
 
-mega angle is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
+Omega angle is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
 This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).
 
-\subsection{Chi ($\chi$) Angle}
+\subsubsection{Chi ($\chi$) Angle}
 
 Chi angle is the bond angle rotation of the side chain bond (C$_{\alpha}$--R bond).
+
+\subsection{$\alpha$-Helices}
+
+Specifically speaking, $\alpha$-helix is just one of the many types of helices that may be present in a protein -- there are other types of helices (e.g.
+polyproline helix), but $\alpha$-helix is the most common type.
+
+In $\alpha$-helix, the carbonyl group of the n$^{th}$ residue forms a H-bond to the amino group of the n+4$^{th}$ residue (e.g.
+the carbonyl group of the 1$^{st}$ residue H-bonds to the amino group of the 5$^{th}$ residue)
+There is about 3.6 residues per turn, which is about 5.4\AA{} per turn.
+The distance (or height) between 2 residues is about 1.5\AA.
+The side chains point \textbf{outwards} from the helix (this provides certain chemical properties on the sides of the helix, e.g. hydrophobic side).
+$\alpha$-helices are \textbf{dipole} in nature, due to the amide bond polarity, all of which points in a single direction.
+
+There are some residues that disrupt the helical structure.
+These amino acids are glycine and proline.
+Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
+This flexibility can cause disruption in the helical structure.
+Proline has a unique side chain that connects back to its amino group (forming an ``imino group''), meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.
+
+\begin{center}
+% \includegraphics[width=0.7\textwidth]{alphahelix}
+\end{center}
+
+\subsection{$\beta$-Sheets}
+
+$\beta$-sheets, or $\beta$-pleated sheets, are formed from H-bond formation between adjacent polypeptide chain (or strands), therefore $\beta$-sheets are formed from 2$\sim$10 strands (but it can have more strands).
+Each strand has about 15 amino acids, and the $\beta$-sheets can be either parallel (both strands go from N to C terminal) or anti-parallel (strands go in an opposite direction).
+$\beta$-sheets are extended, but pleated (i.e. it's not completely flat), and the side chains point above and below.
+
+\begin{center}
+% \includegraphics[width=0.5\textwidth]{betasheet}
+% \includegraphics[width=0.4\textwidth]{turns}
+\end{center}
+
+\subsection{Turns}
+
+Turns are sharp hairpin turns in a protein, and it has a high glycine and proline content (due to its flexibility and unique structure).
+About 30\% of the residues in a protein is involved in turns.
+It is common for the residues to form H-bonds across the turn.
+There are many types of turns (\textgreater16), but type I and II are the most common.
 
@@
 
-\section{Secondary Structure of a Protein}
-
-There are three main secondary structures in a protein: $\alpha$-helix, $\beta$-sheets, and turns.
-
-\subsection{$\alpha$-Helices}
-
-Specifically speaking, $\alpha$-helix is a type of helix that is present in a protein -- there are other types of helices (e.g. polyproline helix), but $\alpha$-helix is the most common type.
-
-In $\alpha$-helix, the carbonyl group of the n$^{th}$ residue forms a H-bond to the amino group of the n+4$^{th}$ residue (e.g.
-the carbonyl group of the 1$^{st}$ residue H-bonds to the amino group of the 5$^{th}$ residue)
-There is about 3.6 residues per turn, which is about 5.4\AA per turn.
-The distance (or height) between 2 residues is about 1.5\AA.
-The side chains point outwards from the helix (this provides certain chemical properties on the sides of the helix, e.g. hydrophobic side). 
-$\alpha$-helices are dipole in nature, due to the amide bond polarity, all of which points at a single direction.
-
-There are some residues that disrupt the helical structure, and these are glycine and proline.
-Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
-This flexibility can cause disruption in the helical structure.
-Proline has a unique side chain that connects back to its amino group, forming an imino group, meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.
-
-\subsection{$\beta$-Sheets}
-
-$\beta$-sheets, or $\beta$-pleated sheets, are formed from H-bond formation between adjacent polypeptide chain (or strands), therefore $\beta$-sheets are formed from 2$\sim$10 strands (but itcan have more strands).
-Each strand has about 15 amino acids, and the $\beta$-sheets can be either parallel (both strands go from N to C terminal) or anti-parallel (strands go in an opposite direction).
-$\beta$-sheets are extended, but pleated (i.e. it's not completely flat), and the side chains point above and below.
-
-\subsection{Turns}
-
-Turns are sharp hairpin turns in a protein, and it has a high glycine and proline content.
-About 30\% of the residues in a protein is involved in turns.
-It is common for the residues to form H-bonds across the turn.
-There are many types of turns (\textgreater16), but type I and II are common.
-
 \subsection{Relationship between Ramachandron Plots and Secondary Structures}
@@
 Secondary structures have certain phi/psi angle pairs, and when you overlay these to the plot, it falls into the allowed regions of the plot at a distinct region.
-This is not surprising, because if the secondary structures fall in the forbidden region, how did the structure form in the first place?
-It can't -- the structure cannot form due to steric hinderance.
+Secondary structures cannot contain any forbidden angle paris, because if they did contain such angle pair(s) there would be a steric hindrance.
 
-This essentially shows that secondary structures have unique phi/psi angle pairs.
+This graph essentially shows that secondary structures have unique phi/psi angle pairs.
+
+% \includegraphics[width=\textwidth]{ramachandron}
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Primary structure is the amino acid sequence of a protein
+		\begin{itemize}
+			\item Primary structure/sequence of the protein determines the 3D structure of the protein
+		\end{itemize}
+	\item Secondary structure is the regular folds/structure that are stabilised by H-bonds between the peptide backbone
+	\item Tertiary structure is the 3D structure of a protein and can have biological function(s)
+	\item Quaternary structure is a group of tertiary structure combined together as a single protein
+	\item Amide bonds of the peptide backbone has a partial double bond characteristic, and therefore cannot be rotated around this bond (amide plane)
+	\item Rotation is allowed between the amide planes, but not all angles are allowed due to steric hindrance
+	\item Relative to the C$_{\alpha}$ atom, phi $\phi$ angle is the rotation of the amide plane on the amino group side, whereas the psi $\psi$ angle is the rotation of the amide plane on the other side (i.e.
+the carboxyl side)
+	\item Omega $\omega$ angle is the angle of the peptide bond and can only be \textit{cis} or \textit{trans}
+	\item Chi $\chi$ angle is the angle of rotation of the side chain group
+	\item $\alpha$-helix properties:
+		\begin{itemize}
+			\item H-bond between the carbonyl group of the n$^{th}$ residue with the amino group of the n+4$^{th}$ residue
+			\item 3.6 residues per turn, 5.4\AA{} per turn
+			\item 1.5\AA{} between any two amino acid in the helix
+			\item Side chains point outwards
+			\item Dipole due to the dipole nature of the amide bonds
+			\item Glycine and proline are the common helix breakers
+		\end{itemize}
+	\item $\beta$-helix properties:
+		\begin{itemize}
+			\item H-bond between adjacent polypeptide chains/strands
+			\item Formed from 2$\sim$10 strands
+			\item Each strand has about 15 amino acids
+			\item It can be parallel or anti-parallel
+			\item Side chains points up or down
+		\end{itemize}
+	\item Properties of a turn:
+		\begin{itemize}
+			\item About 30\% of residues are involved in turns
+			\item proline and glycine rich
+			\item There are many types of turns, but types I and II are most common
+		\end{itemize}
+	\item Ramachandron Plot shows the allowed Phi/Psi angle pairs in a protein
+	\item Secondary structures have their own unique phi/psi angle pairs, and therefore have a distinct area on the Ramachandron plot.
+\end{itemize}

### 2016-05-09 bioc192/lect03.tex
added graphics
editing bioc notes

@@
 Secondary structures such as $\alpha$-helix, $\beta$-sheets, and turns combine together to form a specific supersecondary structures, or `motifs'.
+
+\begin{center}
+% \includegraphics[width=0.7\textwidth]{supersecondary}
+\end{center}
 
@@
 
-Quaternary structure refers to the non-covalent interactions between different domains, which forms a larger protein structure.
-The domains are called `subunits', and these assemblies are called dimer (2 subunits), trimers (3 subunits), etc.
+Quaternary structure refers to the larger protein structure formed from non-covalent interactions between different domains.
+The domains are called `subunits', and these assemblies are called dimer (2 subunits), trimer (3 subunits), etc.
 
@@
 
+% \includegraphics[width=\textwidth]{stabilisinginteraction}
+
 \section{Chaperones and Chaperonin}
@@
 If the proteins are not folded properly, it can lead to serious diseases like Alzheimer's and prion.
+
+\begin{center}
+% \includegraphics[width=0.4\textwidth]{chaperone}
+\end{center}
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Protein domains have hydrophobic core and hydrophilic surface, and is stabilised (mainly) by non-covalent interaction
+	\item Protein folding is directed predominantly by the hydrophobic forces from the non-polar residues
+	\item Protein folding is not a random process
+	\item Some proteins require chaperones and/or chaperonnins to fold correctly
+	\item Protein misfolding can lead to serious diseases such as Alzheimer's and prion
+\end{itemize}

### 2016-05-09 bioc192/lect04.tex
finished lecture 6, onto lecture 7
added graphics
editing bioc notes

@@
+This lecture part was to give you an idea of how the tertiary and quaternary structures provide the functionality of  a protein, using myoglobin and haemoglobin as an example.
+
 \section{Myoglobin}
@@
 Myoglobin is able to reversibly bind O$_2$, and have high affinity for O$_2$ as well.
+
+\begin{center}
+% \includegraphics[width=0.5\textwidth]{myoglobin}
+\end{center}
 
@@
 
+\begin{center}
+% \includegraphics[width=0.4\textwidth]{haem}
+\end{center}
+
 \section{Haemoglobin}
 
-Haemoglobin is the tetrameric (4 subunit) form of myoglobin, and it is a blood O$_2$ transporter.
+Haemoglobin is the tetrameric (4 subunit) form of myoglobin that consists of two $\alpha$ and two $\beta$ subunits, and it is a blood O$_2$ transporter.
 Since haemoglobin is a tetramer, it has 4 haem group in total (one on each subunit), and therefore able to carry 4 O$_2$ molecules.
 Slight conformational change occurs when the O$_2$ binds to the haemoglobin molecule, which will become important in explaining the cooperative effect of the subunits.
+
+\begin{center}
+% \includegraphics[width=0.4\textwidth]{haemoglobin}
+\end{center}
 
@@
 
-In contrast, haemoglobin's O$_2$ binding affinity is not that high compared with myoglobin.
+In contrast, haemoglobin's O$_2$ binding affinity is not that high compared to myoglobin.
 As you can see from the O$_2$ saturation curve, it requires quite a high pO$_2$ for haemoglobin to be saturated with O$_2$.
@@
 Of course, there are other factors that contribute to the allosteric binding/release of O$_2$ molecules.
+
+\begin{center}
+% \includegraphics[width=0.4\textwidth]{saturationcurve}
+\end{center}
 
@@
 
-\textbf{In sequential model, binding of one substrate changes the conformation of other subunits, which makes it easier for the other subunits to bind the substrate}
+\begin{center}
+% \includegraphics[width=\textwidth]{sequential}
+\end{center}
+
+\textbf{In sequential model, binding of one substrate changes the conformation of the other subunits, which makes it easier for the other subunits to bind the substrate}
 
@@
 The binding of substrate causes the equilibrium to shift to the R state and the substrates are more likely to bind to the protein.}
+
+\begin{center}
+% \includegraphics[width=0.3\textwidth]{concerted}
+\end{center}
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Myoglobin:
+		\begin{itemize}
+			\item Resides in muscles/tissues for O$_2$ storage
+			\item Consists of 8 $\alpha$-helices
+			\item Haem group located in the middle
+			\item Reversible binfing of O$_2$
+			\item Hyperbolic O$_2$ binding curve
+		\end{itemize}
+	\item Haem:
+		\begin{itemize}
+			\item Four pyrrole rings linked together with Fe$^{2+}$ at the center of the molecule
+			\item Fe$^{2+}$ is slightly below the plane of the porphyrin ring when O$_2$ is not bound
+			\item Attached to the myoglobin via the His F8
+			\item His E7 affects the O$_2$ binding affinity of the haem
+		\end{itemize}
+	\item Cyanide binds to the Fe$^{3+}$ form, not the Fe$^{2+}$ form
+	\item Cyanide's main inhibitory effect is at the cytochrome Fe$^{3+}$, which affects the electron transport system
+	\item Haemoglobin:
+		\begin{itemize}
+			\item Tetrameric form of myglobin
+			\item Consists of 2 $\alpha$ and 2 $\beta$ subunits
+			\item Blood O$_2$ transporter and carries 4 O$_2$ molecules at a time
+			\item Slight conformational change occurs when O$_2$ binds to a subunit
+			\item Sigmoidal O$_2$ binding curve
+		\end{itemize}
+	\item Myoglobin has a higher O$_2$ binding affinity than haemoglobin
+	\item Allosterism -- Change in binding affinity caused by a conformational change of the protein at a site that is different from where the substrate binds to
+	\item In sequential model, binding of one substrate changes the conformation of the other subunits, which makes it easier for the other subunits to bind the substrate
+	\item In concerted model, the T and R states are at equilibrium and substrates can bind at any time to any conformational state.
+The binding of substrate causes the equilibrium to shift to the R state and the substrates are more likely to bind to the protein
+\end{itemize}

### 2016-05-09 bioc192/lect05.tex
added graphics
editing bioc notes

@@
 However, there are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
-Since these molecules have effects on haemoglobin, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.
+Since these molecules have effects on haemoglobin's binding affinity, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.
 
-\vspace{0.5cm}
+\vspace{0.3cm}
 
@@
 
+\begin{center}
+% \includegraphics[width=0.3\textwidth]{bpg}
+\end{center}
+
 \section{H$^+$ and CO$_2$}
@@
 Similarly, as CO$_2$ concentration increases, CO$_2$ is more likely to bind to the haemoglobin and decrease the oxygen affinity of haemoglobin.
+
+\begin{center}
+% \includegraphics[width=0.3\textwidth]{satcurve2}
+% \includegraphics[width=0.6\textwidth]{satcurveph}
+\end{center}
 
@@
 Also note that BPG is released at the lungs when O$_2$ binds to the haemoglobin, and binds at the tissue/muscle when O$_2$ is released and CO$_2$ and H$^+$ binds to it, contributing to the binding and release of O$_2$.(insert the saturation curve of +/- allosteric effectors)
+
+% \includegraphics[width=\textwidth]{bicarbonate}
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Allosteric effectors are molecules that bind to the allosteric protein at a place other than/different to the normal substrate binds to.
+	\item
+	\item
+	\item
+	\item
+	\item
+	\item
+	\item
+	\item
+\end{itemize}

### 2016-05-09 bioc192/lect04.tex
tmp commit

@@
 
-In concerted model, all of the subunits are either in the T state, or the R state, and the states are at equilibrium (i.e. no `in between' state).
+In concerted model, all of the subunits are either in the T state, or the R state, and the states are at equilibrium (i.e.\  no `in between' state).
 The O$_2$ molecules can bind to either state at any time.
@@
 
-\textbf{ In concerted model, the T and R states are at equilibrium and substrates can bind at any time to any conformational state.
+\textbf{In concerted model, the T and R states are at equilibrium and substrates can bind at any time to any conformational state.
 The binding of substrate causes the equilibrium to shift to the R state and the substrates are more likely to bind to the protein.}

### 2016-05-09 bioc192/lect05.tex
tmp commit

@@
 2,3-Bisphosphoglycerate (BPG) is a small, negatively charged (-5 charge) molecule that binds to the deoxy-haemoglobin at the center of the tetramer.
-Binding of BPG causes the haemoglobin to be in T state, i.e.
-in the dexygenated state, and decreases the haemoglobin's binding affinity for O$_2$, meaning that the haemoglobin with BPG bound to it will require higher pO$_2$ to oxygenate it.
+Binding of BPG causes the haemoglobin to be in T state, i.e.\  in the dexygenated state, and decreases the haemoglobin's binding affinity for O$_2$, meaning that the haemoglobin with BPG bound to it will require higher pO$_2$ to oxygenate it.
 Binding of BPG at the muscle/tissue site ensures that the O$_2$ molecules are released at the muscles/tissues.
@@
 Both H$^+$ and CO$_2$ binds to haemoglobin at a different site as the substrate (O$_2$) and affects the oxygen affinity of haemoglobin.
-Both molecules favour the T state of haemoglobin, and therefore favours the deoxygenated state (i.e. decreased O$_2$ affinity).
+Both molecules favour the T state of haemoglobin, and therefore favours the deoxygenated state (i.e.\  decreased O$_2$ affinity).
 Note that both of these molecules are related to the pH of the environment, and the blood buffer system is also involved in controlling the concentration of these molecules.
@@
 
-Technically speaking, the Bohr effect is the effect of pH (i.e. [H$^+$]) and [CO$_2$] on the oxygen affinity of haemoglobin.
+Technically speaking, the Bohr effect is the effect of pH (i.e.\  [H$^+$]) and [CO$_2$] on the oxygen affinity of haemoglobin.
 As pH decreases (therefore [H$^+$] increases), the net charge of haemoglobin changes due to different amino acids getting protonated, which leads to slight conformational changes in the haemoglobin structure, and therefore decrease in haemoglobin's oxygen affinity.
@@
 
-Also note that BPG is released at the lungs when O$_2$ binds to the haemoglobin, and binds at the tissue/muscle when O$_2$ is released and CO$_2$ and H$^+$ binds to it, contributing to the binding and release of O$_2$.(insert the saturation curve of +/- allosteric effectors)
+Also note that BPG is released at the lungs when O$_2$ binds to the haemoglobin, and binds at the tissue/muscle when O$_2$ is released and CO$_2$ and H$^+$ binds to it, contributing to the binding and release of O$_2$ (insert the saturation curve of +/- allosteric effectors).
 
@@
 \begin{itemize}
-	\item Allosteric effectors are molecules that bind to the allosteric protein at a place other than/different to the normal substrate binds to.
-	\item
-	\item
-	\item
-	\item
-	\item
-	\item
-	\item
-	\item
+	\item Allosteric effectors are molecules that bind to the allosteric protein at a place other than/different to where the normal substrate binds to
+	\item BPG is a negatively charged small molecule that binds at the center of the haemoglobin tetramer and favours the T state
+	\item Both H$^+$ and CO$_2$ are allosteric effectors that favours the T state of haemoglobin
+	\item H$^+$ and CO$_2$ are closely related to the pH of the environment, so these molecules will have different effect in different parts of the body
 \end{itemize}

### 2016-05-09 bioc192/lect06.tex
added graphics
editing bioc notes
tmp commit

@@
 
+\begin{center}
+% \includegraphics[width=0.5\textwidth]{hbs}
+\end{center}
+
 \subsection{Met-Haemoglobin (HbM)}
@@
 Haemoglobin C (HbC) is common in West Africa, and is caused by a point mutation that causes a substitution mutation.
-The substitution is at the same place as HbS: glutamic acid to lysine at position 6 (E6K).
+The substitution is at the same place as HbS\@ glutamic acid to lysine at position 6 (E6K).
 As a consequence of this substitution, the charge is swapped from negatively charged glutamic acid to positively charged lysine.
@@
 However, individuals with homo or heterozygous HbC have less adherence, and thus have some resistance to malaria infection.
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Remember the haemoglobin variants mentioned in this note in some details
+\end{itemize}

### 2016-05-09 bioc192/lect07.tex
added graphics
editing bioc notes
tmp commit

@@
 
-Most proteins, if not all proteins, in eukaryotic cells are post-translationally modified.
+Most, if not all,  proteins in eukaryotic cells are post-translationally modified.
 Post-translational modification increases the number of distinct forms of proteins without increasing the number of genes in the genome.
@@
 
-The diversity of the proteins can be expanded by mRNA splicing and PTM.
+The diversity of the proteins can be expanded by mRNA splicing and PTM\@ (mRNA splicing will not be covered).
 PTM can be categorised into two categories: the covalent addition of chemical group to amino acid side chain, or covalent hydrolysis of the peptide bond in a protein.
@@
 
-\section{Phoshphorylation}
+\subsection{Phoshphorylation}
 
@@
 
-Quite often, phosphorylation state is used to `switch' the conformational state of the protein to enable a biological process in a cell (e.g. facilitating interaction with other proteins).
+Quite often, phosphorylation state is used to `switch' the conformational state of the protein to enable a biological process in a cell (e.g.\ facilitating interaction with other proteins).
 
-\subsection{Insulin Receptor}
+\subsubsection{Insulin Receptor}
 
@@
 The exposed tyrosine is then targeted by kinases, and get phosphorylated by it.
-Phosphorylated tyrosine can then cause responses (e.g. signal transduction, GLUT4 transporter expression).
+Phosphorylated tyrosine can then cause responses (e.g.\ signal transduction, GLUT4 transporter expression).
 
-\subsection{Na$^+$/K$^+$ Pump}
+\begin{center}
+% \includegraphics[width=0.7\textwidth]{insulinreceptor}
+\end{center}
+
+\subsubsection{Na$^+$/K$^+$ Pump}
 
@@
 
-\section{Gamma-carboxy Glutamic Acid (Gla)}
+\begin{center}
+% \includegraphics[width=0.7\textwidth]{nakpump}
+\end{center}
+
+\subsection{Gamma-carboxy Glutamic Acid (Gla)}
 
@@
 
-\section{Hydroxylation}
+\subsection{Hydroxylation}
 
@@
 This disease is known as scurvy.
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Phosphorylation occurs on the hydroxyl group of the amino acid side chains
+	\item Kinases phosphorylates amino acids, and phosphatases dephosphorylates amino acids
+	\item Phosphorylation results in an addition of a large negatively charged molecule to the protein
+	\item Gamma-carboxy glutamic acid has an extra corboxyl group added onto glutamic acid
+	\item Gamma-carboxy Glutamic Acid is done in some blood co-agulation proteins
+	\item Hydroxylation is an addition of the hydroxyl group to proline or lysine
+	\item Hydroxylation provides greater possibilities of hydrogen bonding with other atoms, and therefore provides strength/order in structures like collagen
+\end{itemize}

### 2016-05-09 bioc192/lect09.tex
added graphics
editing bioc notes
tmp commit

@@
 
-Therapeutic proteins were extracted from relevant sources (e.g. tissues, serum) and was used to treat the patients.
+Therapeutic proteins were extracted from relevant sources (e.g.\ tissues, serum) and was used to treat the patients.
 The problems with this approach was the risk of identified and unidentified pathogen infections from the source (e.g. HIV).
@@
 
-Recombinant antibodies are designed to control the effects of some proteins in your body (e.g. some cell receptors).
+Recombinant antibodies are designed to control the effects of some proteins in your body (e.g.\ some cell receptors).
 Mouse AB can be made, but it is recognised as foreign material in humans and get rejected.
 This lead to the formation of chimeric antibodies, so there will be no immune response.
+
+\begin{center}
+% \includegraphics[width=0.8\textwidth]{chimericab}
+\end{center}
 
@@
 The use of live animals to express/make recombinant proteins, for example, in milk.
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Read through this lecture section notes
+\end{itemize}

### 2016-05-09 bioc192/lect10.tex
added topic2 graphics
editing bioc notes
tmp commit

@@
 
+\begin{center}
+% \includegraphics[width=0.7\textwidth]{geo_stereo}
+\end{center}
 Geometric specificity is where the substrate must be made of same component/structure for it to be recognised by the enzyme.
@@
 \begin{center}
-\ch{ E + S <=> ES <=> [{T}] EP <=> E + P }
+\ch{E + S <=> ES <=> [{T}] EP <=> E + P}
 \end{center}
@@
 There are two models that describes how the substrate binds to the enzyme: the lock-and-key model, or the induced-fit model.
+
+\begin{center}
+% \includegraphics[width=0.6\textwidth]{enzymemodels}
+\end{center}
 
@@
 \section{Enzymes, Activation Energy, and \\Catalytic Mechanisms}
+
+\begin{center}
+% \includegraphics[width=0.5\textwidth]{enzcatalreac}
+\end{center}
 
@@
 \end{enumerate}
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Geometric specificity refers to the enzyme being specific  to the component/structure of the substrate
+	\item Stereospecificity refers to the enzyme being specific to the orientation of the substrate
+	\item Enzyme is not altered after an enzymatic reaction
+	\item Lock-and-key model is where the substrate perfectly matches the enzyme's active site
+	\item Induced-fit model is where the binding of the substrate to the enzyme causes a conformational change in both the substrate and the enzyme
+	\item Know the six catalytic mechanisms of an enzymatic reaction
+	\item Co-factors are molecules that are used in the enzymatic reaction, but is changed after the reaction completes
+	\item Know the ``numbers'' associated with each of the enzyme classes
+\end{itemize}

### 2016-05-09 bioc192/lect11.tex
added topic2 graphics
editing bioc notes
tmp commit

@@
 \section{Measuring Enzyme Catalysis}
+
+\begin{center}
+% \includegraphics[width=0.5\textwidth]{progresscurve}
+\end{center}
 
@@
 
-Consider the situation where there is an excess substrate (i.e. unlimited source of substrate).
+\begin{center}
+% \includegraphics[width=0.5\textwidth]{vvse}
+\end{center}
+
+Consider the situation where there is an excess substrate (i.e.\ unlimited source of substrate).
 As the concentration of the enzyme increases, the rate of reaction increases.
@@
 
-Consider the situation where the concentration of enzyme is fixed.
+\begin{center}
+% \includegraphics[width=0.4\textwidth]{vvss}
+\end{center}
+
+Consider the situation where the concentration of the enzyme is fixed.
 As we increase the concentration of substrate, the rate of reaction will initially increase linearly, but as more and more substrates are introduced, the rate of reaction reaches a `maximum'.
 This is because all of the active site of the enzymes become occupied as the substrate concentration increases -- it reaches the `maximum capacity' of the enzyme, and the enzymes cannot increase the rate of reaction no matter how hard it tries.
+
+\begin{center}
+% \includegraphics[width=0.4\textwidth]{zerofirstkinetics}
+\end{center}
 
@@
 Note: do NOT get this mixed up with ``K$_M$ = $\frac{V_{max}}{2}$'' -- remember that K$_M$ is the \textbf{substrate concentration} when V = $\frac{V_{max}}{2}$.
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Progress curve shows how much product is made over time
+	\item Only the initial part of the curve is used, so other factors don't limit the reaction as it proceeds
+	\item Reaction velocity is proportional to the enzyme concentration when there is excess substrate
+	\item First-order kinetics is where the rate of reaction is proportional to the concentration of the substrate
+	\item zero-order kinetics is where the rate of reaction is constant with changing substrate concentration
+\end{itemize}

### 2016-05-09 bioc192/lect12.tex
finished lecture 13, onto lecture 14
added topic2 graphics
editing bioc notes
tmp commit

@@
 \section{Lineweaver-Burk Plot}
+
+\begin{center}
+% \includegraphics[width=0.8\textwidth]{lineweaverburk}
+\end{center}
 
@@
 
-\section{V$_{max}$ k$_{cat}$}
+\section{V$_{max}$ and k$_{cat}$}
 
@@
 The catalytic efficiency will therefore have the highest value when k$_{cat}$ is high (high turnover), and low K$_M$ (high binding affinity).
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item K$_M$ is defined as the substrate concentration at which the velocity of the reaction is half the maximum velocity
+	\item K$_M$ is also defined as the ratio of the breakdown and the formation of the enzyme-substrate complex
+	\item K$_M$ characterises a single enzyme-substrate pair, meaning that the value of K$_M$ is different for every substrate and/or enzyme
+	\item V$_{max}$ is defined as the maximum velocity of the reaction when an enzyme is saturated with the substrate
+	\item Know the Michaelis-Menten equation, and its assumptions
+	\item Lineweaver-Burk plot is the V vs [S] curve
+	\item k$_{cat}$ is the number of moles of substrate converted into product per mole of enzyme per second
+	\item Catalytic efficiency of an enzyme is measured as the ratio of k$_{cat}$ and K$_M$ (i.e.\ how quickly a product is formed relative to how well the substrate binds to the enzyme)
+\end{itemize}

### 2016-05-09 bioc192/lect13.tex
added topic2 graphics
editing bioc notes
tmp commit

@@
 
+\begin{center}
+% \includegraphics[width=0.6\textwidth]{compinh}
+\end{center}
+
 These inhibitors bind to the enzyme at the active site, and competes with the substrate molecule.
@@
 \subsection{Non-competitive Reversible Inhibitor}
+
+\begin{center}
+% \includegraphics[width=0.6\textwidth]{noncompinh}
+\end{center}
 
@@
 \subsection{Allosteric Enzyme Regulation}
+
+\begin{center}
+% \includegraphics[width=0.6\textwidth]{allosterism}
+\end{center}
 
@@
 Isoenzyme is a form of transcriptional regulation of cellular activity, as some cells in your body has to catalyse the reaction differently to another part of your body.
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Irreversible inhibitors covalently binds to the enzyme and permanently inactivates it
+	\item Competitive reversible inhibitors increases the K$_M$ but doesn't affect the V$_{max}$ of the enzyme
+	\item Non-competitive reversible inhibitors decreases the V$_{max}$ but doesn't affect the K$_M$ of the enzyme
+\end{itemize}

### 2016-05-09 bioc192/lect14.tex
added images for topic 3
editing bioc notes
tmp commit

@@
 
+\begin{center}
+% \includegraphics[width=0.4\textwidth]{physrec1}
+% \includegraphics[width=0.4\textwidth]{physrec2}
+\end{center}
+
 Physiological receptors are normal receptors that are present on the cell surface.
@@
 
+\begin{center}
+% \includegraphics[width=1.0\textwidth]{fourreceptors}
+\end{center}
+
 \section{Ligand-Gated Ion Channels}
+
+\begin{center}
+% \includegraphics[width=0.7\textwidth]{ligandgated}
+\end{center}
 
@@
 
+\begin{center}
+% \includegraphics[width=1.0\textwidth]{voltagegated}
+\end{center}
+
 These receptors are made fromm four subunits, and opened in response to changes in voltage across the membrane (e.g. voltage change caused by ligand-gated ion channels).
@@
 The drugs that can block these receptors literally blocks the ion channel and prevents the ions to move through the pores (e.g. tetrodotoxin).
+
+\subsection*{What you should get out of this lecture}
+
+\begin{itemize}
+	\item Drug is a chemical that changes the behaviour of the cell
+	\item Agonists are the drugs or molecules that causes the full biological response
+	\item Antagonists are the drugs or molecules that prevents the full biological response (opposite of an agonist)
+	\item Drug receptor is anything that a drug binds to (DNA/RNA/protein/peptide)
+	\item Physiological receptor refers specifically to the  receptors located on the cell surface
+	\item Know how the gated channels work
+\end{itemize}

### 2016-09-12 presentation/beamer_theme/ctcrSep2016.tex
draft ctcr presentation

% \documentclass[notes]{beamer}
% \documentclass[notes=hide]{beamer}
% \documentclass[notes=only]{beamer}
\documentclass[handout]{beamer}

% \usepackage{pgfpages}
% \pgfpagesuselayout{2 on 1}[a4paper,border shrink=3mm]

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{xcolor,colortbl}

\definecolor{LightCyan}{rgb}{0.88,1,1}

\usetheme{riku}

\def\logo{
	{\includegraphics[width=20mm]{logo}}
}

\title[BMI and Cancer]{Body Mass Index and Pathway Dysregulation in Cancer}
\subtitle{CTCR Talk}
\author[Riku Takei]{Riku Takei (MSc)\\Supervisor: Assoc. Prof. Mik Black}
\institute{Department of Biochemistry, University of Otago}

\begin{document}

{
	\setbeamertemplate{footline}{}
	\begin{frame}[noframenumbering]
		\titlepage
	\end{frame}
}

\begin{frame}
	\frametitle{Introduction}
	\begin{center}
		\includegraphics[scale=0.35]{./images/obesityworldmap}
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Introduction}
	\begin{center}
		\includegraphics[width=0.48\paperwidth]{./images/ob1999}\\
		1999\\
		\includegraphics[width=0.48\paperwidth]{./images/ob2014}\\
		2014
	\end{center}
\end{frame}

\begin{frame}
	\frametitle{Body Mass Index and Cancer}
	\begin{columns}
		\column{0.6\textwidth}
		{\footnotesize
			Cancers are caused by dysregulation in various pathways/mechanisms that allow them to survive and proliferate
			\begin{itemize}
				\item Commonly know as the Hallmarks of Cancer
			\end{itemize}
			Many reports on obesity being a major risk for cancer and poor prognosis
		}
		\column{0.5\textwidth}
		\includegraphics[width=0.45\paperwidth]{./images/cancerhallmark}
	\end{columns}
\end{frame}

\begin{frame}
	\frametitle{Aims}
	\begin{enumerate}
		\item Investigate the association between patient BMI and tumour molecular characteristics
		\item Determine whether pathway commonalities exist across tumours that arise from other tissues
	\end{enumerate}
\end{frame}

\begin{frame}
	\frametitle{Obesity-associated Genetic Signatures}
	Creighton \textit{et al.} (2012)
	\begin{itemize}
		\item Identified obesity-associated genetic signatures from the tumours of breast cancer patients
		\item Showed that some compounds were able to specifically regulate the expression of these genetic signatures (potential for targeted therapy)
	\end{itemize}
\end{frame}

\begin{frame}{{\normalsize Obesity-associated genes from Creighton \textit{et al.} (2012) }}
	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
	{\footnotesize
		\begin{itemize}
			\item Majority of these genes were downregulated in obese patients
		\end{itemize}
	}
	\begin{center}
		\includegraphics[width=0.6\paperwidth]{./images/creighton}
	\end{center}
\end{frame}

\begin{frame}{Obesity-associated genes from Fuentes-Mattei \textit{et al.}}
	\begin{columns}
		\column{0.5\textwidth}
		Fuentes-Mattei \textit{et al.} (2014)
		\begin{itemize}
			\item 137 ER+ breast cancer patients in the study, 130 obesity-associated genes identified
			\item Direct evidence of how obesity/adiposity contributes to tumour growth (using mouse model)
		\end{itemize}
		\column{0.5\textwidth}
		\includegraphics[width=0.45\paperwidth]{./images/fmcircos}
	\end{columns}
\end{frame}

\begin{frame}
	\frametitle{Methods}
	Downloaded RNA-seq data from 8 different cancer types from ICGC
	\vskip2ex
	Microarray data from Creighton's paper
	\vskip2ex
	Microarray data from Fuentes-Mattei's paper
	\vskip2ex
	Microarray data from Cris Print
	\vskip2ex
	\includegraphics[width=0.4\textwidth]{./images/tcga}
	\hfill
	\includegraphics[width=0.4\textwidth]{./images/icgc}
\end{frame}

\begin{frame}{Validation of Methodology -- Creighton \textit{et al.} (2012)}
	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
	\vskip2ex
	Metagene was obtained from SVD using the genes that were common in both Creighton \textit{et al.} and other cancer data sets\\
	\begin{center}
		\includegraphics[width=0.6\paperwidth]{./images/creighton}
	\end{center}
\end{frame}

\begin{frame}{Singular Value Decomposition (SVD) recap}
	For each sample in the data set, you need to assign a ``score'' that represents the overall gene expression of the gene set of interest
	\begin{itemize}
		\item i.e. a score that represent the 799 obesity-associated gene probes identified by Creighton \textit{et al.}
	\end{itemize}
	\vskip2ex
	You can do this by applying SVD to the data set of interest and get a metagene out from this
\end{frame}

\begin{frame}{\normalsize Creighton \textit{et al.} metagene correlates with Gene Expression}
	\begin{center}
		\includegraphics[width=0.45\paperwidth]{./images/crge1}
	\end{center}
\end{frame}

\begin{frame}{Creighton \textit{et al.} metagene correlates with BMI}
	\hspace{-4.0ex}
	\includegraphics[width=0.45\paperwidth]{./images/crmetaclin0}
	\includegraphics[width=0.45\paperwidth]{./images/crmetaclin1}
\end{frame}

\begin{frame}{Transformation Matrix}
	Singular Value Decomposition of a dataset X gives:
	\begin{equation*}
		X = UDV'
	\end{equation*}
	Rearranging this equation gives:
	\begin{equation*}
		V' = D^{-1}U'X
	\end{equation*}
	Where $D^{-1}U'$ is the transformation matrix and $V'$ is the metagene for the data set X
	{\footnotesize
		\begin{itemize}
			\item This means that the transformation matrix can be applied to other data sets to obtain the metagene for those data sets
		\end{itemize}
	}
\end{frame}

\begin{frame}{Transformation Matrix}
	\begin{center}
		\includegraphics[height=0.7\paperheight,width=0.40\paperwidth]{./images/tmp/transmatexp1}
	\end{center}
\end{frame}

\begin{frame}{Transformation Matrix}
	\begin{center}
		\includegraphics[width=0.70\paperwidth]{./images/tmp/transmatexp2}
	\end{center}
\end{frame}

\begin{frame}{Metagene in ICGC data}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga2}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga7}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga12}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga17}\\
	\includegraphics[width=0.23\paperwidth]{./images/crtcga22}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga27}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga32}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga37}
\end{frame}

\begin{frame}{Does the metagene correlate with BMI?}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga3}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga8}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga13}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga18}\\
	\includegraphics[width=0.23\paperwidth]{./images/crtcga23}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga28}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga33}
	\includegraphics[width=0.23\paperwidth]{./images/crtcga38}
\end{frame}

\begin{frame}{{\normalsize Obesity-associated genes from Fuentes-Mattei \textit{et al.}}}
	\begin{columns}
		\column{0.5\textwidth}
		137 ER+ breast cancer samples in the study, 130 unique genes identified\\
		\vskip2ex
		Data were normalised  with RMA normalisation method\\
		\vskip2ex
		\column{0.5\textwidth}
		\includegraphics[width=0.45\paperwidth]{./images/fmcircos}
	\end{columns}
\end{frame}

\begin{frame}{FM metagene in Creighton \textit{et al.} data}
	\begin{center}
		\includegraphics[width=0.45\paperwidth]{./images/fmmetacr3}
	\end{center}
\end{frame}

\begin{frame}{FM metagene in Creighton \textit{et al.} data}
	\includegraphics[width=0.45\paperwidth]{./images/fmmetacr4}
	\includegraphics[width=0.45\paperwidth]{./images/fmmetacr5}
\end{frame}

\begin{frame}{FM metagene in other cancer types}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat2}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat7}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat12}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat17}\\
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat22}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat27}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat32}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat37}
\end{frame}

\begin{frame}{FM metagene and BMI}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi3}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi8}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi13}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi18}\\
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi23}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi28}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi33}
	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi38}
\end{frame}

\begin{frame}{Gene Expression Analysis}
	Maybe the obesity-associated genes identified by either group were not related to obesity, but related to a different clinical variable
	\vskip2ex
	Are the differentially expressed genes between the obese and non-obese groups (after correcting for other clinical variables) similar to the obesity-associated genes?
	{\footnotesize
		\begin{itemize}
			\item How do these DEGs compare with the Creighton's or Fuentes-Mattei's obesity-associated genes?
		\end{itemize}
	}
\end{frame}

\begin{frame}{Creighton \textit{et al.} data revisited}
	Gene expression analysis was carried out on the Creighton \textit{et al.} data:
	\begin{itemize}
		\item Original Creighton data
		\item Caucasian-only Creighton data
	\end{itemize}
	\vskip2ex
	Gene expression analysis was repeated with the residual version of the above data
	\vskip2ex
	Each ``obesity-associated genes'' were checked for any overlaps with the original Creighton metagene
\end{frame}

\begin{frame}{Completely remove the effect of ethnicity}
	African American patients are more likely to have higher BMI than Cauacasian patients
	\vskip2ex
	Removed the effect of ethnicity by removing the African American patients' data (total of 77 patient data to work with)
	{\footnotesize
		\begin{itemize}
			\item All of the other clinical variables were adjusted as before
		\end{itemize}
	}
\end{frame}

\begin{frame}{DEGs in residual data}
	\begin{center}
		\includegraphics[width=0.42\paperwidth]{./images/venn0}
		\includegraphics[width=0.45\paperwidth]{./images/venn1}
	\end{center}
\end{frame}

\begin{frame}{Metagene Analysis}
	All of the gene probe sets I have found (8 in total) were used to create metagenes as before
	\begin{itemize}
		\item Checked in Creighton \textit{et al.} data
		\item Transferred into other cancer data sets
	\end{itemize}
\end{frame}

\begin{frame}{Initial results -- Checking in Creighton \textit{et al.} data}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr2}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr7}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr12}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr17}\\
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr22}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr27}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr32}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr37}
\end{frame}

\begin{frame}{Initial results -- Checking in Creighton \textit{et al.} data}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr3}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr8}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr13}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr18}\\
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr23}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr28}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr33}
	\includegraphics[width=0.23\paperwidth]{./images/degmetacr38}
\end{frame}

\begin{frame}{Do the metagenes correlate with GE in ICGC data?}
	\includegraphics[width=0.23\paperwidth]{./images/rawobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/resobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/caobsmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/caresobsmeta2}\\
	\includegraphics[width=0.23\paperwidth]{./images/crolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/rescrolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/cacrolmeta2}
	\includegraphics[width=0.23\paperwidth]{./images/carescrolmeta2}
\end{frame}

\begin{frame}{Do the metagenes correlate with BMI?}
	\includegraphics[width=0.23\paperwidth]{./images/rawobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/resobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/caobsclin3}
	\includegraphics[width=0.23\paperwidth]{./images/caresobsclin3}\\
	\includegraphics[width=0.23\paperwidth]{./images/crolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/rescrolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/cacrolclin3}
	\includegraphics[width=0.23\paperwidth]{./images/carescrolclin3}
\end{frame}

% \begin{frame}
% 	Repeated this with Cris' data, but same results as in Creighton and ICGC data
% \end{frame}

% \begin{frame}{Do the metagenes correlate with each other?}
% 	\begin{center}
% 		\includegraphics[width=0.45\paperwidth]{./images/cor1}
% 	\end{center}
% \end{frame}

% \begin{frame}{Common genes across multiple cancer types}
% 	DEGs between the obese and non-obese groups were identified for each cancer types
% 	\vskip2ex
% 	For each cancer types, samples were randomly split into two groups (same proportion as obese vs. non-obese samples)
% 	\vskip2ex
% 	{\footnotesize
% 		\begin{tabular}{ccccccccc}
% 			& \multicolumn{8}{c}{No. of cancer types the gene overlapped}\\
% 			\rule{0pt}{2.25ex}                    & 1    & 2    & 3   & 4  & 5 & 6 & 7 & 8 \\
% 			\hline
% 			\rule{0pt}{2.25ex}Obese vs. Non-obese & 6865 & 1907 & 289  & 26   & 1 & 0 & 0 & 0 \\
% 			Randomly split groups                 & 6202 & 1488 & 198  & 18   & 1 & 0 & 0 & 0 \\
% 			No. of genes found                    & 5012 & 870  & 89   & 6    & 0 & 0 & 0 & 0 \\
% 			on average                            &      &      &      &      &   &   &   & \\
% 			Empirical FDR                         & 0.73 & 0.46 & 0.31 & 0.23 & 0 & 0 & 0 & 0 \\
% 		\end{tabular}
% 	}
% \end{frame}

% \begin{frame}{Pathway Enrichment Analysis}
% 	Maybe each cancer type have a different set of genes activated
% 	\vskip2ex
% 	Are there any specific pathways enriched in any specific cancer type?
% 	\vskip2ex
% 	If so, is it common across other cancer types?
% 	\vskip2ex
% 	How relevant is the pathway to obesity and/or cancer?
% 	\vskip2ex
% 	Can we use it to narrow down our search for obesity-associated signature?
% \end{frame}

% \begin{frame}{Pathway Enrichment Analysis}
% 	Gene Ontology (GO) database was used to check for the function(s) being enriched by the DEGs in each cancer type (between obese and non-obese groups)
% 	\vskip2ex
% 	Used a rank-based approach and controlled the error rate using FDR
% 	\vskip2ex
% 	\begin{center}
% 		\textbf {No pathways showed up as significantly enriched}
% 	\end{center}
% \end{frame}

% \begin{frame}{Common genes and pathway enrichment analysis}
% 	There were no obesity-associated genes that were common across all the ICGC cancer types
% 	\vskip2ex
% 	There were no pathways enriched in any of the obesity-associated genes found so far
% \end{frame}

\begin{frame}{Relevance of BMI in Cris' data}
	\begin{center}
		\includegraphics[width=0.60\paperwidth]{./images/tmp/mikres}
	\end{center}
\end{frame}

\begin{frame}{So \ldots BMI is not related to cancer?}
	All the results so far suggests that there is no association between BMI and cancer
	\vskip2ex
	However, the metagenes are picking up gene signatures  that are common across all cancer types, but not related to BMI
	\vskip2ex
	So what signature/mechanism is the metagene picking up?
\end{frame}

\begin{frame}{Gatza \textit{et al.} (2010)}
	Used 18 pathways to define sub-class of breast tumours, based on the expression of the pathway signatures
	\begin{itemize}
		\item AKT, ER, PR, HER2, PI3K, p53, MYC, RAS, etc...
	\end{itemize}
	\vskip2ex
	Each pathway signature was obtained by overexpressing a single gene in an appropriate cell line, and then differentially expressed genes were noted down
	\begin{itemize}
			\item Use these pathways to see what the obesity-associated genes are most similar to
	\end{itemize}
\end{frame}

\begin{frame}{Tricky thing about metagene direction}
	Need to consider how the original data was manipulated:
	\begin{enumerate}
		\item Normalisation methods (RMA vs MAS5)
		\item Scaling the data or not
		\item Metagene scaling (probit vs rank-based)
	\end{enumerate}
	\vskip2ex
	Need to check the direction of each pathway metagenes with its gene (e.g. EGFR gene with EGFR pathway metagene)
	\vskip2ex
	Need to consider the grouping/clustering of the pathway metagenes
\end{frame}

\begin{frame}{What I mean about directions...}
	\includegraphics[width=0.45\paperwidth]{./images/output-0}
	\includegraphics[width=0.45\paperwidth]{./images/output-1}
\end{frame}

\begin{frame}{Some weird genes}
	\includegraphics[width=0.45\paperwidth]{./images/output-2}
	\includegraphics[width=0.45\paperwidth]{./images/output-3}
\end{frame}

\begin{frame}{Grouping/clustering of different pathways}
	\begin{center}
		\includegraphics[width=0.45\paperwidth]{./images/gatzares1}
	\end{center}
\end{frame}

\begin{frame}{Re-making results from Gatza \textit{et al.} paper}
	\includegraphics[width=0.45\paperwidth]{./images/out2}
	\includegraphics[width=0.40\paperwidth]{./images/gatzares1}
\end{frame}

\begin{frame}{Predicting obesity metagene with pathway metagene}
	Make a linear model (in Cris' data) with sample BMI and/or pathway metagenes and use this to predict Creighton's obesity metagene
	\vskip2ex
	\begin{itemize}
		\item If any of the pathway metagenes are significant in the model, the pathway is a good predictor of the obesity metagene
	\end{itemize}
	\vskip2ex
	Then use this model to predict the obesity metagene in Creighton data
	\begin{itemize}
		\item Compare the predicted obesity metagene with the SVD-generated obesity metagene
	\end{itemize}
\end{frame}

\begin{frame}{Results}
	Sample BMI was not significant  in the model
	\vskip2ex
	Only pathway that was a good predictor of obesity metagene was PR pathway metagene
\end{frame}

\begin{frame}{\normalsize Comparison of the predicted obesity metagene with true value}
	\begin{center}
		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction1}
		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction2}
	\end{center}
\end{frame}

\begin{frame}{\normalsize Comparison of the predicted obesity metagene with true value}
	\begin{center}
		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction3}
		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction4}
	\end{center}
\end{frame}

\begin{frame}{Overall Summary}
	\textbf{Creighton/FM metagene analysis}
	\begin{itemize}
		\item Both Creighton and FM metagenes correlated with the overall gene expression in any cancer data set
		\item The metagene score does not correlate with the BMI status of the samples
	\end{itemize}
	\vskip2ex
	\textbf{Gene expression and metagene analysis}
	\begin{itemize}
		\item Again, all of the gene sets found correlated with the overall gene expression in any data set, but none of them correlated with the BMI status
	\end{itemize}
\end{frame}

\begin{frame}{Overall Summary}
	\textbf{Gatza pathway metagene analysis}
	\begin{itemize}
		\item Sample BMI was not a good predictor of the obesity metagene, but PR pathway metagene was
		\item However, the model was not great at predicting the obesity metagene
			\begin{itemize}
				\item Other mechanism contributing to obesity metagene?
			\end{itemize}
	\end{itemize}
\end{frame}

% \begin{frame}{Discussion}
% 	Tumour heterogeneity within and between samples
% 	{\footnotesize
% 		\begin{itemize}
% 				% TODO rewrite the following sentence
% 			\item Not finding any significant genes or pathways because the samples are so different from one another
% 		\end{itemize}
% 	}
% 	\vskip2ex
% 	Tumour sample quality and size
% 	{\footnotesize
% 		\begin{itemize}
% 			\item Picking up too much noise and/or don't have enough samples to find anything significant (i.e. shit data)
% 		\end{itemize}
% 	}
% \end{frame}

% \begin{frame}{Discussion -- Inter-tumour heterogeneity 1}
% 	Each patient has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
% 	\vskip2ex
% 	Each data set will have a different collection of patients, which in turn will have a different collection of dysregulated pathways
% 	{\footnotesize
% 		\begin{itemize}
% 			\item Hence DEGs show up, but not the pathways
% 		\end{itemize}
% 	}
% \end{frame}

% \begin{frame}{Discussion -- Inter-tumour heterogeneity 2}
% 	Some of the high BMI patients in the data set will have tumours that are BMI driven, but BMI may not be the driver for some of these patients
% 	\begin{itemize}
% 		\item Only subset of the patients with high BMI are actually driven by BMI
% 	\end{itemize}
% 	\vskip2ex
% 	Other data sets may have less of these truly BMI driven tumours
% 	\begin{itemize}
% 		\item Therefore no association with the signatures
% 	\end{itemize}
% \end{frame}

\begin{frame}{Acknowledgement}
	\begin{columns}
		\column{0.5\textwidth}
		My supervisor: Mik\\
		Members of the Black lab\\
		Cancer Genetics Lab\\
		SYSKA/Mozilla Study Group\\
		ICGC and TCGA\\
		\column{0.5\textwidth}
		\includegraphics[width=0.3\paperwidth]{./images/tcga}\\
		\includegraphics[width=0.3\paperwidth]{./images/icgc}
	\end{columns}
\end{frame}

% \begin{frame}
% 	\frametitle{References}
% 	\bibliographystyle{BiocRefStyle}
% 	\bibliography{/home/riku/Documents/References/BibTeX/MSc}
% \end{frame}

% \begin{frame}{test}
% 	\citet{Fuentes-Mattei2014}\\
% 	\citet{Creighton2012}\\
% 	\citet{Alexandrov2013}
% 	\cite{Gatza2010a}
% \end{frame}

\end{document}

### 2016-09-12 presentation/beamer_theme/example.tex
draft ctcr presentation

--- 2016-09-12 presentation/beamer_theme/ctcrSep2016.tex
@@
 
-% \usepackage{pgfpages}
-% \pgfpagesuselayout{2 on 1}[a4paper,border shrink=3mm]
+\usepackage{pgfpages}
+\pgfpagesuselayout{2 on 1}[a4paper,border shrink=3mm]
 
@@
 
-\title[BMI and Cancer]{Body Mass Index and Pathway Dysregulation in Cancer}
-\subtitle{CTCR Talk}
+\title[BMI and Cancer Pathway]{Body Mass Index and Pathway Dysregulation in Cancer}
+\subtitle{Committee Meeting 2016}
 \author[Riku Takei]{Riku Takei (MSc)\\Supervisor: Assoc. Prof. Mik Black}
@@
 }
+
+\begin{frame}{Overview}
+	\begin{enumerate}
+		\item Introduction to obesity and cancer
+		\item Results from last committee meeting
+		\item Results since the last committee meeting
+		\item Future directions
+	\end{enumerate}
+\end{frame}
 
@@
 			\begin{itemize}
-				\item Commonly know as the Hallmarks of Cancer
+				\item Commonly know as the Hallmarks of Cancer \citep{Hanahan2011}
 			\end{itemize}
@@
 \begin{frame}
+	\frametitle{Obesity-associated Genetic Signatures}
+	Creighton \textit{et al.} (2012)
+	\begin{itemize}
+		\item Identified obesity-associated genetic signatures from the tumours of breast cancer patients
+		\item Showed that some compounds were able to specifically regulate the expression of these genetic signatures (potential for targeted therapy)
+	\end{itemize}
+	Fuentes-Mattei \textit{et al.} (2014)
+	\begin{itemize}
+		\item Identified obesity-associated genetic signatures from the tumours of ER+ breast cancer patients
+		\item Direct evidence of how obesity/adiposity contributes to tumour growth (using mouse model)
+	\end{itemize}
+\end{frame}
+
+% \begin{frame}
+% 	\frametitle{Genetic Signatures Across Multiple Cancer Types}
+% 	\begin{center}
+% 		\includegraphics[width=0.85\paperwidth]{./images/multicancersignature}
+% 	\end{center}
+% \end{frame}
+
+\begin{frame}
 	\frametitle{Aims}
@@
 
-\begin{frame}
-	\frametitle{Obesity-associated Genetic Signatures}
-	Creighton \textit{et al.} (2012)
+\begin{frame}{Progress Overview}
+	Where I was at the last committee meeting:
+	{\footnotesize
 		\begin{itemize}
-		\item Identified obesity-associated genetic signatures from the tumours of breast cancer patients
-		\item Showed that some compounds were able to specifically regulate the expression of these genetic signatures (potential for targeted therapy)
+			\item Applied Creighton metagene to ICGC data
+			\item Applied Fuentes-Mattei metagene to Creighton \textit{et al.} and ICGC data
+			\item Found no association with BMI status
 		\end{itemize}
+	}
+	\vskip2ex
+	What I have done since the last committee meeting:
+	{\footnotesize
+		\begin{itemize}
+			\item Gene expression analysis in Creighton \textit{et al.} data
+			\item Gene expression analysis in ICGC data
+			\item Pathway enrichment analysis in ICGC data
+			\item Comparison of the obesity metagenes with other pathway metagenes from Gatza \textit{et al.} (preliminary results)
+		\end{itemize}
+	}
 \end{frame}
 
-\begin{frame}{{\normalsize Obesity-associated genes from Creighton \textit{et al.} (2012) }}
+\begin{frame}
+	\frametitle{Methods}
+	Download sequence (RNA-seq) data and clinical data from publicly available databases
+	{\footnotesize
+		\begin{itemize}
+			\item The Cancer Genome Atlas (TCGA)
+			\item International Cancer Genome Consortium (ICGC)
+		\end{itemize}
+	}
+	\vskip2ex
+	\includegraphics[width=0.4\textwidth]{./images/tcga}
+	\hfill
+	\includegraphics[width=0.4\textwidth]{./images/icgc}
+\end{frame}
+
+\begin{frame}{Clinical Data}
+	Download all the clinical data available from TCGA databases
+	\begin{itemize}
+		\item {\footnotesize Clinical data available for 33 cancer types}\\
+	\end{itemize}
+	\vskip2ex
+	Only interested in cancer types that have both height and weight data\\
+	\begin{equation*}
+		BMI = \frac{Weight (kg)}{Height^2 (m^2)}
+	\end{equation*}
+\end{frame}
+
+\begin{frame}{Clinical Data with Patient BMI}
+	Out of 33 cancer types, 14 types had both height and weight data\\
+	{\footnotesize
+		\begin{itemize}
+			\item 8 cancer types had RNA-seq data\\
+		\end{itemize}
+	}
+	\vskip2ex
+	\begin{center}
+		\begin{tabular}{lcccl}
+			& Normal & Overweight & Obese & Total\\
+			\rowcolor{LightCyan}
+			BLCA & 103  & 97         & 61    & 261\\
+			CESC & 85   & 62         & 77    & 224   \\
+			\rowcolor{LightCyan}
+			COAD & 75   & 80         & 71    & 226   \\
+			KIRP & 26   & 53         & 45    & 124\\
+			\rowcolor{LightCyan}
+			LIHC & 140  & 75         & 49    & 264\\
+			READ & 22   & 36         & 15    & 73\\
+			\rowcolor{LightCyan}
+			SKCM & 72   & 73         & 73    & 218\\
+			UCEC & 90   & 106        & 286   & 482\\
+			\hline
+			\rule{0pt}{2.25ex}Total & 613 & 582 & 677 & 1872\\
+		\end{tabular}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Validation of Methodology -- Creighton \textit{et al.} (2012)}
 	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
@@
 
-\begin{frame}{Obesity-associated genes from Fuentes-Mattei \textit{et al.}}
-	\begin{columns}
-		\column{0.5\textwidth}
-		Fuentes-Mattei \textit{et al.} (2014)
-		\begin{itemize}
-			\item 137 ER+ breast cancer patients in the study, 130 obesity-associated genes identified
-			\item Direct evidence of how obesity/adiposity contributes to tumour growth (using mouse model)
-		\end{itemize}
-		\column{0.5\textwidth}
-		\includegraphics[width=0.45\paperwidth]{./images/fmcircos}
-	\end{columns}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Methods}
-	Downloaded RNA-seq data from 8 different cancer types from ICGC
-	\vskip2ex
-	Microarray data from Creighton's paper
-	\vskip2ex
-	Microarray data from Fuentes-Mattei's paper
-	\vskip2ex
-	Microarray data from Cris Print
-	\vskip2ex
-	\includegraphics[width=0.4\textwidth]{./images/tcga}
-	\hfill
-	\includegraphics[width=0.4\textwidth]{./images/icgc}
-\end{frame}
-
 \begin{frame}{Validation of Methodology -- Creighton \textit{et al.} (2012)}
-	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
+	Normalised the data using RMA normalisation method (\texttt{affy} package)\\
 	\vskip2ex
@@
 
-\begin{frame}{Transformation Matrix}
-	\begin{center}
-		\includegraphics[height=0.7\paperheight,width=0.40\paperwidth]{./images/tmp/transmatexp1}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Transformation Matrix}
-	\begin{center}
-		\includegraphics[width=0.70\paperwidth]{./images/tmp/transmatexp2}
-	\end{center}
-\end{frame}
-
 \begin{frame}{Metagene in ICGC data}
@@
 
-\begin{frame}{{\normalsize Obesity-associated genes from Fuentes-Mattei \textit{et al.}}}
+\begin{frame}{Obesity-associated genes from Fuentes-Mattei \textit{et al.}}
 	\begin{columns}
@@
 		\vskip2ex
+		Metagene was formed from the common genes between the Fuentes-Mattei \textit{et al.} and other cancer data sets (116 genes)
 		\column{0.5\textwidth}
@@
 \end{frame}
+
+% \begin{frame}{Can the FM metagene predict BMI status?}
+% 	If both Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} found obesity-associated genetic signature, then these signatures should be similar
+% 	{\footnotesize
+% 		\begin{itemize}
+% 			\item Therefore should be able to separate the samples based on BMI
+% 		\end{itemize}
+% 	}
+% 	\vskip2ex
+% 	No BMI information is available for Fuentes-Mattei \textit{et al.} data set, so you need to transform the Creighton \textit{et al.} with the metagene\\
+% 	{\footnotesize
+% 		\begin{itemize}
+% 			\item Check whether the metagene correlates with BMI status
+% 		\end{itemize}
+% 	}
+% \end{frame}
 
@@
 
+\begin{frame}{Summary 1}
+	Metagene from Creighton \textit{et al.} correlated with gene expression and patient BMI in Creighton data set
+	\vskip2ex
+	Metagene derived from Creighton's obesity-associated genes  correlated with the gene expression but not the patient BMI in ICGC data
+	\vskip2ex
+	Metagene derived from Fuentes-Mattei obesity-associated genes correlated with the gene expression but not the patient BMI in both Creighton \textit{et al.} and ICGC data
+\end{frame}
+
 \begin{frame}{Gene Expression Analysis}
@@
 \begin{frame}{Creighton \textit{et al.} data revisited}
-	Gene expression analysis was carried out on the Creighton \textit{et al.} data:
+	Gene expression analysis was carried out on the Creighton \textit{et al.} data
+	{\footnotesize
 		\begin{itemize}
-		\item Original Creighton data
-		\item Caucasian-only Creighton data
+			\item in their paper, they found 799 gene probes with p \textless{} 0.01 and fold change \textgreater{} 1.2
 		\end{itemize}
+	}
 	\vskip2ex
-	Gene expression analysis was repeated with the residual version of the above data
+	1781 differentially expressed genes (p \textless{} 0.01) were found between obese vs. non-obese
+	{\footnotesize
+		\begin{itemize}
+			\item very few genes had fold change \textgreater{} 1.2 (132 genes)
+			\item no genes were found when p-value was adjusted
+			\item top 799 genes were chosen from this list
+		\end{itemize}
+	}
 	\vskip2ex
-	Each ``obesity-associated genes'' were checked for any overlaps with the original Creighton metagene
+	239 gene probes were present in the top 799 gene probes that I have found
+\end{frame}
+
+\begin{frame}{Residual data}
+	All of the non-BMI clinical variables were controlled for in the linear model
+	{\footnotesize
+		\begin{itemize}
+			\item Age, ethnicity, menopause status, grade, LN status, ER/PR/HER2 statuses
+		\end{itemize}
+	}
+	\vskip2ex
+	Residual data (or remaining data after controlling for the variables) was used to find the DEGs
+	\vskip2ex
+	Again, more than 799 gene probes showed up (1104 genes)
+	{\footnotesize
+		\begin{itemize}
+			\item 168 genes from Creighton \textit{et al.} were found in the top 799 genes
+		\end{itemize}
+	}
+\end{frame}
+
+\begin{frame}{DEGs in residual data}
+	\begin{center}
+		\includegraphics[width=0.50\paperwidth]{./images/venn0}
+	\end{center}
 \end{frame}
@@
 	}
+	\vskip2ex
+	Top 799 genes were identified in this data set (both raw and residual)
+	{\footnotesize
+		\begin{itemize}
+			\item  148 and 92 genes were in the top 799 genes in the raw and residual data, respectively
+		\end{itemize}
+	}
 \end{frame}
 
-\begin{frame}{DEGs in residual data}
+\begin{frame}{DEGs in Cauacasian-only data}
 	\begin{center}
-		\includegraphics[width=0.42\paperwidth]{./images/venn0}
-		\includegraphics[width=0.45\paperwidth]{./images/venn1}
+		\includegraphics[width=0.50\paperwidth]{./images/venn1}
 	\end{center}
@@
 \begin{frame}{Metagene Analysis}
-	All of the gene probe sets I have found (8 in total) were used to create metagenes as before
+	All of the gene probe sets I have found (8 in total) were used to create metagenes and to transform other cancer types
 	\begin{itemize}
@@
 
-% \begin{frame}
-% 	Repeated this with Cris' data, but same results as in Creighton and ICGC data
-% \end{frame}
-
-% \begin{frame}{Do the metagenes correlate with each other?}
-% 	\begin{center}
-% 		\includegraphics[width=0.45\paperwidth]{./images/cor1}
-% 	\end{center}
-% \end{frame}
-
-% \begin{frame}{Common genes across multiple cancer types}
-% 	DEGs between the obese and non-obese groups were identified for each cancer types
-% 	\vskip2ex
-% 	For each cancer types, samples were randomly split into two groups (same proportion as obese vs. non-obese samples)
-% 	\vskip2ex
-% 	{\footnotesize
-% 		\begin{tabular}{ccccccccc}
-% 			& \multicolumn{8}{c}{No. of cancer types the gene overlapped}\\
-% 			\rule{0pt}{2.25ex}                    & 1    & 2    & 3   & 4  & 5 & 6 & 7 & 8 \\
-% 			\hline
-% 			\rule{0pt}{2.25ex}Obese vs. Non-obese & 6865 & 1907 & 289  & 26   & 1 & 0 & 0 & 0 \\
-% 			Randomly split groups                 & 6202 & 1488 & 198  & 18   & 1 & 0 & 0 & 0 \\
-% 			No. of genes found                    & 5012 & 870  & 89   & 6    & 0 & 0 & 0 & 0 \\
-% 			on average                            &      &      &      &      &   &   &   & \\
-% 			Empirical FDR                         & 0.73 & 0.46 & 0.31 & 0.23 & 0 & 0 & 0 & 0 \\
-% 		\end{tabular}
-% 	}
-% \end{frame}
-
-% \begin{frame}{Pathway Enrichment Analysis}
-% 	Maybe each cancer type have a different set of genes activated
-% 	\vskip2ex
-% 	Are there any specific pathways enriched in any specific cancer type?
-% 	\vskip2ex
-% 	If so, is it common across other cancer types?
-% 	\vskip2ex
-% 	How relevant is the pathway to obesity and/or cancer?
-% 	\vskip2ex
-% 	Can we use it to narrow down our search for obesity-associated signature?
-% \end{frame}
-
-% \begin{frame}{Pathway Enrichment Analysis}
-% 	Gene Ontology (GO) database was used to check for the function(s) being enriched by the DEGs in each cancer type (between obese and non-obese groups)
-% 	\vskip2ex
-% 	Used a rank-based approach and controlled the error rate using FDR
-% 	\vskip2ex
-% 	\begin{center}
-% 		\textbf {No pathways showed up as significantly enriched}
-% 	\end{center}
-% \end{frame}
-
-% \begin{frame}{Common genes and pathway enrichment analysis}
-% 	There were no obesity-associated genes that were common across all the ICGC cancer types
-% 	\vskip2ex
-% 	There were no pathways enriched in any of the obesity-associated genes found so far
-% \end{frame}
-
-\begin{frame}{Relevance of BMI in Cris' data}
+\begin{frame}{Do the metagenes correlate with each other?}
 	\begin{center}
-		\includegraphics[width=0.60\paperwidth]{./images/tmp/mikres}
+		\includegraphics[width=0.45\paperwidth]{./images/cor0}
+		\includegraphics[width=0.45\paperwidth]{./images/cor1}
 	\end{center}
@@
 
-\begin{frame}{So \ldots BMI is not related to cancer?}
-	All the results so far suggests that there is no association between BMI and cancer
+\begin{frame}{Metagenes are correlating with GE but not BMI}
+	All of the metagenes found so far have correlated with the overall gene expression of the samples, but not with the sample BMI
 	\vskip2ex
-	However, the metagenes are picking up gene signatures  that are common across all cancer types, but not related to BMI
+	The metagenes are picking up gene signatures  that are common across all cancer types, but not related to BMI
 	\vskip2ex
-	So what signature/mechanism is the metagene picking up?
-\end{frame}
-
-\begin{frame}{Gatza \textit{et al.} (2010)}
-	Used 18 pathways to define sub-class of breast tumours, based on the expression of the pathway signatures
-	\begin{itemize}
-		\item AKT, ER, PR, HER2, PI3K, p53, MYC, RAS, etc...
-	\end{itemize}
-	\vskip2ex
-	Each pathway signature was obtained by overexpressing a single gene in an appropriate cell line, and then differentially expressed genes were noted down
-	\begin{itemize}
-			\item Use these pathways to see what the obesity-associated genes are most similar to
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Tricky thing about metagene direction}
-	Need to consider how the original data was manipulated:
-	\begin{enumerate}
-		\item Normalisation methods (RMA vs MAS5)
-		\item Scaling the data or not
-		\item Metagene scaling (probit vs rank-based)
-	\end{enumerate}
-	\vskip2ex
-	Need to check the direction of each pathway metagenes with its gene (e.g. EGFR gene with EGFR pathway metagene)
-	\vskip2ex
-	Need to consider the grouping/clustering of the pathway metagenes
-\end{frame}
-
-\begin{frame}{What I mean about directions...}
-	\includegraphics[width=0.45\paperwidth]{./images/output-0}
-	\includegraphics[width=0.45\paperwidth]{./images/output-1}
-\end{frame}
-
-\begin{frame}{Some weird genes}
-	\includegraphics[width=0.45\paperwidth]{./images/output-2}
-	\includegraphics[width=0.45\paperwidth]{./images/output-3}
-\end{frame}
-
-\begin{frame}{Grouping/clustering of different pathways}
 	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/gatzares1}
+		\textbf{So what signature/mechanism is the metagene picking up?}
 	\end{center}
@@
 
-\begin{frame}{Re-making results from Gatza \textit{et al.} paper}
-	\includegraphics[width=0.45\paperwidth]{./images/out2}
-	\includegraphics[width=0.40\paperwidth]{./images/gatzares1}
-\end{frame}
-
-\begin{frame}{Predicting obesity metagene with pathway metagene}
-	Make a linear model (in Cris' data) with sample BMI and/or pathway metagenes and use this to predict Creighton's obesity metagene
+\begin{frame}{Pathway signatures from Gatza \textit{et al.} (2010)}
+	Made metagenes from the pathway signatures from the Gatza \textit{et al.} paper in ICGC data
 	\vskip2ex
+	Compared these pathway metagenes with the BMI metagenes
+	{\footnotesize
 		\begin{itemize}
-		\item If any of the pathway metagenes are significant in the model, the pathway is a good predictor of the obesity metagene
+			\item Is the BMI metagene similar to any of the pathway signatures?
 		\end{itemize}
+	}
 	\vskip2ex
-	Then use this model to predict the obesity metagene in Creighton data
-	\begin{itemize}
-		\item Compare the predicted obesity metagene with the SVD-generated obesity metagene
-	\end{itemize}
+	All of the samples from the ICGC data were used (1872 samples in total)
 \end{frame}
@@
 \begin{frame}{Results}
-	Sample BMI was not significant  in the model
-	\vskip2ex
-	Only pathway that was a good predictor of obesity metagene was PR pathway metagene
-\end{frame}
-
-\begin{frame}{\normalsize Comparison of the predicted obesity metagene with true value}
 	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction1}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction2}
+		\includegraphics[width=0.90\paperwidth]{./images/allmeta1}
 	\end{center}
@@
 
-\begin{frame}{\normalsize Comparison of the predicted obesity metagene with true value}
+\begin{frame}{Results}
 	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction3}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction4}
+		\includegraphics[width=0.90\paperwidth]{./images/allmeta2}
 	\end{center}
@@
 
-\begin{frame}{Overall Summary}
+\begin{frame}{Results}
+	\begin{center}
+		\includegraphics[width=0.45\paperwidth]{./images/allmetacor1}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Summary 2}
+	Metagenes found in Creighton \textit{et al.} data correlates with GE and BMI status in Creighton data, but not in  the ICGC data
+	\vskip2ex
+	The metagenes do not correlate with any other pathway metagenes
+	\vskip2ex
+	\begin{center}
+		\textbf{Are there any common genes and/or pathways being dysregulated across all of the cancer types?}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Common genes across multiple cancer types}
+	DEGs between the obese and non-obese groups were identified for each cancer types
+	\vskip2ex
+	For each cancer types, samples were randomly split into two groups (same proportion as obese vs. non-obese samples)
+	\vskip2ex
+	{\footnotesize
+		\begin{tabular}{ccccccccc}
+			& \multicolumn{8}{c}{No. of cancer types the gene overlapped}\\
+			\rule{0pt}{2.25ex}                    & 1    & 2    & 3   & 4  & 5 & 6 & 7 & 8 \\
+			\hline
+			\rule{0pt}{2.25ex}Obese vs. Non-obese & 6865 & 1907 & 289  & 26   & 1 & 0 & 0 & 0 \\
+			Randomly split groups                 & 6202 & 1488 & 198  & 18   & 1 & 0 & 0 & 0 \\
+			No. of genes found                    & 5012 & 870  & 89   & 6    & 0 & 0 & 0 & 0 \\
+			on average                            &      &      &      &      &   &   &   & \\
+			Empirical FDR                         & 0.73 & 0.46 & 0.31 & 0.23 & 0 & 0 & 0 & 0 \\
+		\end{tabular}
+	}
+\end{frame}
+
+\begin{frame}{Pathway Enrichment Analysis}
+	Maybe each cancer type have a different set of genes activated
+	\vskip2ex
+	Are there any specific pathways enriched in any specific cancer type?
+	\vskip2ex
+	If so, is it common across other cancer types?
+	\vskip2ex
+	How relevant is the pathway to obesity and/or cancer?
+	\vskip2ex
+	Can we use it to narrow down our search for obesity-associated signature?
+\end{frame}
+
+\begin{frame}{Pathway Enrichment Analysis}
+	Gene Ontology (GO) database was used to check for the function(s) being enriched by the DEGs in each cancer type (between obese and non-obese groups)
+	\vskip2ex
+	Used a rank-based approach and controlled the error rate using FDR
+	\vskip2ex
+	\begin{center}
+		\textbf {No pathways showed up as significantly enriched}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Summary}
 	\textbf{Creighton/FM metagene analysis}
@@
 	\end{itemize}
-	\vskip2ex
 	\textbf{Gene expression and metagene analysis}
@@
 	\end{itemize}
-\end{frame}
-
-\begin{frame}{Overall Summary}
-	\textbf{Gatza pathway metagene analysis}
+	\textbf{Pathway enrichment analysis}
 	\begin{itemize}
-		\item Sample BMI was not a good predictor of the obesity metagene, but PR pathway metagene was
-		\item However, the model was not great at predicting the obesity metagene
-			\begin{itemize}
-				\item Other mechanism contributing to obesity metagene?
-			\end{itemize}
+		\item There were no pathway significantly enriched in any cancer type
 	\end{itemize}
@@
 
-% \begin{frame}{Discussion}
-% 	Tumour heterogeneity within and between samples
-% 	{\footnotesize
-% 		\begin{itemize}
-% 				% TODO rewrite the following sentence
-% 			\item Not finding any significant genes or pathways because the samples are so different from one another
-% 		\end{itemize}
-% 	}
-% 	\vskip2ex
-% 	Tumour sample quality and size
-% 	{\footnotesize
-% 		\begin{itemize}
-% 			\item Picking up too much noise and/or don't have enough samples to find anything significant (i.e. shit data)
-% 		\end{itemize}
-% 	}
-% \end{frame}
+\begin{frame}{Discussion}
+	Different methods used to find DEGs
+	{\footnotesize
+		\begin{itemize}
+			\item Found more genes than Creighton \textit{et al.} because they were using different programs and normalisation method in their analyses
+		\end{itemize}
+	}
+	\vskip2ex
+	Tumour heterogeneity within and between samples
+	{\footnotesize
+		\begin{itemize}
+			\item Not finding any significant genes or pathways because the samples are so different from one another
+		\end{itemize}
+	}
+	\vskip2ex
+	Tumour sample quality and size
+	{\footnotesize
+		\begin{itemize}
+			\item Picking up too much noise and/or don't have enough samples to find anything significant
+		\end{itemize}
+	}
+\end{frame}
 
-% \begin{frame}{Discussion -- Inter-tumour heterogeneity 1}
-% 	Each patient has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
-% 	\vskip2ex
-% 	Each data set will have a different collection of patients, which in turn will have a different collection of dysregulated pathways
-% 	{\footnotesize
-% 		\begin{itemize}
-% 			\item Hence DEGs show up, but not the pathways
-% 		\end{itemize}
-% 	}
-% \end{frame}
+\begin{frame}{Discussion -- Inter-tumour heterogeneity 1}
+	Each patient has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
+	\vskip2ex
+	Each data set will have a different collection of patients, which in turn will have a different collection of dysregulated pathways
+	{\footnotesize
+		\begin{itemize}
+			\item Hence DEGs show up, but not the pathways
+		\end{itemize}
+	}
+\end{frame}
 
-% \begin{frame}{Discussion -- Inter-tumour heterogeneity 2}
-% 	Some of the high BMI patients in the data set will have tumours that are BMI driven, but BMI may not be the driver for some of these patients
-% 	\begin{itemize}
-% 		\item Only subset of the patients with high BMI are actually driven by BMI
-% 	\end{itemize}
-% 	\vskip2ex
-% 	Other data sets may have less of these truly BMI driven tumours
-% 	\begin{itemize}
-% 		\item Therefore no association with the signatures
-% 	\end{itemize}
-% \end{frame}
+\begin{frame}{Discussion -- Inter-tumour heterogeneity 2}
+	Some of the high BMI patients in the data set will have tumours that are BMI driven, but BMI may not be the driver for some of these patients
+	\begin{itemize}
+		\item Only subset of the patients with high BMI are actually driven by BMI
+	\end{itemize}
+	\vskip2ex
+	Other data sets may have less of these truly BMI driven tumours
+	\begin{itemize}
+		\item Therefore no association with the signatures
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Conclusion}
+	Obesity-associated genetic signatures cannot be used in other cancer types, nor in the same cancer type with different sample population
+	\vskip2ex
+	Gene expression and pathway enrichment analyses showed that there may not be any obesity-associated signature across multiple cancer types
+\end{frame}
+
+\begin{frame}{Future Directions}
+	Explore the biology of the ``obesity-associated'' metagene
+	{\footnotesize
+		\begin{itemize}
+			\item Use other pre-defined pathway signatures and see if any of them correlates with the metagene
+		\end{itemize}
+	}
+	\vskip2ex
+	Effect of BMI on methylation status in different cancer types
+\end{frame}
 
@@
 
-% \begin{frame}
-% 	\frametitle{References}
-% 	\bibliographystyle{BiocRefStyle}
-% 	\bibliography{/home/riku/Documents/References/BibTeX/MSc}
-% \end{frame}
+\begin{frame}
+	\frametitle{References}
+	\bibliographystyle{BiocRefStyle}
+	\bibliography{/home/riku/Documents/References/BibTeX/MSc}
+\end{frame}
 
-% \begin{frame}{test}
-% 	\citet{Fuentes-Mattei2014}\\
-% 	\citet{Creighton2012}\\
-% 	\citet{Alexandrov2013}
-% 	\cite{Gatza2010a}
-% \end{frame}
+\begin{frame}{test}
+	\citet{Fuentes-Mattei2014}\\
+	\citet{Creighton2012}\\
+	\citet{Alexandrov2013}
+	\cite{Gatza2010a}
+\end{frame}
 

### 2016-09-12 presentation/beamer_theme/images/tmp/pdf/colloquia/beamer_theme/example.tex
draft ctcr presentation

--- 2016-09-12 presentation/beamer_theme/example.tex
@@
 
-\begin{frame}{Overview}
-	\begin{enumerate}
-		\item Introduction to obesity and cancer
-		\item Results from last committee meeting
-		\item Results since the last committee meeting
-		\item Future directions
-	\end{enumerate}
-\end{frame}
-
 \begin{frame}
@@
 	\includegraphics[width=0.48\paperwidth]{./images/ob1999}\\
-		1999\\
+	1993\\
 	\includegraphics[width=0.48\paperwidth]{./images/ob2014}\\
@@
 	\begin{itemize}
-		\item Identified obesity-associated genetic signatures from the tumours of breast cancer patients
+		\item Identified obesity-associated genetic signatures in breast cancer patients
 		\item Showed that some compounds were able to specifically regulate the expression of these genetic signatures (potential for targeted therapy)
@@
 	\begin{itemize}
-		\item Identified obesity-associated genetic signatures from the tumours of ER+ breast cancer patients
+		\item Identified obesity-associated genetic signatures in ER+ breast cancer patients
 		\item Direct evidence of how obesity/adiposity contributes to tumour growth (using mouse model)
@@
 		\begin{itemize}
-			\item Gene expression analysis in Creighton \textit{et al.} data
-			\item Gene expression analysis in ICGC data
-			\item Pathway enrichment analysis in ICGC data
+			\item Gene expression analysis using Creighton \textit{et al.} data
+			\item Gene expression analysis using ICGC data
+			\item Pathway enrichment analysis using ICGC data
 			\item Comparison of the obesity metagenes with other pathway metagenes from Gatza \textit{et al.} (preliminary results)
@@
 		\begin{tabular}{lcccl}
-			& Normal & Overweight & Obese & Total\\
+			& Lean & Overweight & Obese & total \\
 			\rowcolor{LightCyan}
@@
 		\begin{itemize}
-			\item Majority of these genes were downregulated in obese patients
+			\item These gene probes were used to assign an obesity-associated gene signature score to each sample
 		\end{itemize}
@@
 	\vskip2ex
-	Metagene derived from Creighton's obesity-associated genes  correlated with the gene expression but not the patient BMI in ICGC data
+	Metagene from Creighton \textit{et al.} correlated with gene expression but not the patient BMI in ICGC data
 	\vskip2ex
-	Metagene derived from Fuentes-Mattei obesity-associated genes correlated with the gene expression but not the patient BMI in both Creighton \textit{et al.} and ICGC data
+	Metagene from Fuentes-Mattei \textit{et al.} correlated with gene expression but not the patient BMI in both Creighton \textit{et al.} and ICGC data
 \end{frame}
@@
 	\vskip2ex
-	Are the differentially expressed genes between the obese and non-obese groups (after correcting for other clinical variables) similar to the obesity-associated genes?
+	Does the differentially expressed genes between the obese and non-obese groups (after correcting for other clinical variables) similar to the obesity-associated genes?
 	{\footnotesize
@@
 		\begin{itemize}
-			\item Age, ethnicity, menopause status, grade, LN status, ER/PR/HER2 statuses
+			\item Age, ethnicity, menopause status, grade, LN/ER/PR/HER2 status
 		\end{itemize}
@@
 \begin{frame}{Completely remove the effect of ethnicity}
-	African American patients are more likely to have higher BMI than Cauacasian patients
+	African American patients are more likely to be obese than Cauacasian patients
 	\vskip2ex
@@
 \begin{frame}{Metagene Analysis}
-	All of the gene probe sets I have found (8 in total) were used to create metagenes and to transform other cancer types
+	All of the new/novel gene probe sets (8 in total) were used to create metagenes and used to transform other cancer types
 	\begin{itemize}
@@
 
-\begin{frame}{Do the metagenes correlate with GE in ICGC data?}
+\begin{frame}{Do the metagenes correlate with gene expression in ICGC data?}
 	\includegraphics[width=0.23\paperwidth]{./images/rawobsmeta2}
@@
 	\includegraphics[width=0.23\paperwidth]{./images/caresobsmeta2}\\
-	\includegraphics[width=0.23\paperwidth]{./images/crolmeta2}
+	\includegraphics[width=0.23\paperwidth]{./images/rawcrolmeta2}
 	\includegraphics[width=0.23\paperwidth]{./images/rescrolmeta2}
@@
 	\includegraphics[width=0.23\paperwidth]{./images/caresobsclin3}\\
-	\includegraphics[width=0.23\paperwidth]{./images/crolclin3}
+	\includegraphics[width=0.23\paperwidth]{./images/rawcrolclin3}
 	\includegraphics[width=0.23\paperwidth]{./images/rescrolclin3}
@@
 	\vskip2ex
-	The metagenes are picking up gene signatures  that are common across all cancer types, but not related to BMI
+	The metagenes are picking up gene signatures  that are common across all cancer types, and not related to BMI
 	\vskip2ex
-	\begin{center}
 	\textbf{So what signature/mechanism is the metagene picking up?}
-	\end{center}
 \end{frame}
@@
 	\begin{center}
-		\includegraphics[width=0.90\paperwidth]{./images/allmeta1}
+		\includegraphics[width=0.90\paperwidth, height=0.60\paperheight]{./images/allmeta1}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=0.90\paperwidth]{./images/allmeta2}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Results}
-	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/allmetacor1}
+		\includegraphics[width=0.50\paperwidth]{./images/allmeta2}
 	\end{center}
@@
 	\vskip2ex
-	\begin{center}
 	\textbf{Are there any common genes and/or pathways being dysregulated across all of the cancer types?}
-	\end{center}
 \end{frame}
@@
 	\vskip2ex
-	Are there any specific pathways enriched in any specific cancer type?
+	Are there any pathways enriched by these genes?
 	\vskip2ex
@@
 			\item Found more genes than Creighton \textit{et al.} because they were using different programs and normalisation method in their analyses
+				\begin{itemize}
+					\item Logged the data instead of RMA normalisation
+				\end{itemize}
 		\end{itemize}
@@
 
-\begin{frame}{Discussion -- Inter-tumour heterogeneity 1}
-	Each patient has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
+\begin{frame}{Discussion -- Tumour heterogeneity}
+	Each sample has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
+	{\footnotesize
+		\begin{itemize}
+			\item Each data set will have more/less of these samples
+			\item Therefore picking up DEGs that are significant in that one particular data set but not the other data sets
+		\end{itemize}
+	}
 	\vskip2ex
-	Each data set will have a different collection of patients, which in turn will have a different collection of dysregulated pathways
+	Each data set will have a different collection of samples, which in turn have a different collection of dysregulated pathways
 	{\footnotesize
@@
 	}
-\end{frame}
-
-\begin{frame}{Discussion -- Inter-tumour heterogeneity 2}
-	Some of the high BMI patients in the data set will have tumours that are BMI driven, but BMI may not be the driver for some of these patients
-	\begin{itemize}
-		\item Only subset of the patients with high BMI are actually driven by BMI
-	\end{itemize}
-	\vskip2ex
-	Other data sets may have less of these truly BMI driven tumours
-	\begin{itemize}
-		\item Therefore no association with the signatures
-	\end{itemize}
 \end{frame}
@@
 	\citet{Alexandrov2013}
-	\cite{Gatza2010a}
+	\citet{Gatza2010}
 \end{frame}

### 2016-09-12 presentation/ctcr/beamer_theme/example.tex
draft ctcr presentation

--- 2016-09-12 presentation/beamer_theme/ctcrSep2016.tex
@@
-% \documentclass[notes]{beamer}
-% \documentclass[notes=hide]{beamer}
-% \documentclass[notes=only]{beamer}
-\documentclass[handout]{beamer}
-
-% \usepackage{pgfpages}
-% \pgfpagesuselayout{2 on 1}[a4paper,border shrink=3mm]
+\documentclass{beamer}
 
@@
 \usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{natbib}
-\usepackage{xcolor,colortbl}
-
-\definecolor{LightCyan}{rgb}{0.88,1,1}
 
@@
 	\frametitle{Introduction}
-	\begin{center}
-		\includegraphics[scale=0.35]{./images/obesityworldmap}
-	\end{center}
+	blahabaldkfja
+	\includegraphics{obesityworldmap.png}
 \end{frame}
 
-\begin{frame}
-	\frametitle{Introduction}
-	\begin{center}
-		\includegraphics[width=0.48\paperwidth]{./images/ob1999}\\
-		1999\\
-		\includegraphics[width=0.48\paperwidth]{./images/ob2014}\\
-		2014
-	\end{center}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Body Mass Index and Cancer}
-	\begin{columns}
-		\column{0.6\textwidth}
-		{\footnotesize
-			Cancers are caused by dysregulation in various pathways/mechanisms that allow them to survive and proliferate
-			\begin{itemize}
-				\item Commonly know as the Hallmarks of Cancer
-			\end{itemize}
-			Many reports on obesity being a major risk for cancer and poor prognosis
-		}
-		\column{0.5\textwidth}
-		\includegraphics[width=0.45\paperwidth]{./images/cancerhallmark}
-	\end{columns}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Aims}
-	\begin{enumerate}
-		\item Investigate the association between patient BMI and tumour molecular characteristics
-		\item Determine whether pathway commonalities exist across tumours that arise from other tissues
-	\end{enumerate}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Obesity-associated Genetic Signatures}
-	Creighton \textit{et al.} (2012)
-	\begin{itemize}
-		\item Identified obesity-associated genetic signatures from the tumours of breast cancer patients
-		\item Showed that some compounds were able to specifically regulate the expression of these genetic signatures (potential for targeted therapy)
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{{\normalsize Obesity-associated genes from Creighton \textit{et al.} (2012) }}
-	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
-	{\footnotesize
-		\begin{itemize}
-			\item Majority of these genes were downregulated in obese patients
-		\end{itemize}
-	}
-	\begin{center}
-		\includegraphics[width=0.6\paperwidth]{./images/creighton}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Obesity-associated genes from Fuentes-Mattei \textit{et al.}}
-	\begin{columns}
-		\column{0.5\textwidth}
-		Fuentes-Mattei \textit{et al.} (2014)
-		\begin{itemize}
-			\item 137 ER+ breast cancer patients in the study, 130 obesity-associated genes identified
-			\item Direct evidence of how obesity/adiposity contributes to tumour growth (using mouse model)
-		\end{itemize}
-		\column{0.5\textwidth}
-		\includegraphics[width=0.45\paperwidth]{./images/fmcircos}
-	\end{columns}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Methods}
-	Downloaded RNA-seq data from 8 different cancer types from ICGC
-	\vskip2ex
-	Microarray data from Creighton's paper
-	\vskip2ex
-	Microarray data from Fuentes-Mattei's paper
-	\vskip2ex
-	Microarray data from Cris Print
-	\vskip2ex
-	\includegraphics[width=0.4\textwidth]{./images/tcga}
-	\hfill
-	\includegraphics[width=0.4\textwidth]{./images/icgc}
-\end{frame}
-
-\begin{frame}{Validation of Methodology -- Creighton \textit{et al.} (2012)}
-	103 samples used in the study, 799 obesity-associated gene probes (695 unique genes) identified\\
-	\vskip2ex
-	Metagene was obtained from SVD using the genes that were common in both Creighton \textit{et al.} and other cancer data sets\\
-	\begin{center}
-		\includegraphics[width=0.6\paperwidth]{./images/creighton}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Singular Value Decomposition (SVD) recap}
-	For each sample in the data set, you need to assign a ``score'' that represents the overall gene expression of the gene set of interest
-	\begin{itemize}
-		\item i.e. a score that represent the 799 obesity-associated gene probes identified by Creighton \textit{et al.}
-	\end{itemize}
-	\vskip2ex
-	You can do this by applying SVD to the data set of interest and get a metagene out from this
-\end{frame}
-
-\begin{frame}{\normalsize Creighton \textit{et al.} metagene correlates with Gene Expression}
-	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/crge1}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Creighton \textit{et al.} metagene correlates with BMI}
-	\hspace{-4.0ex}
-	\includegraphics[width=0.45\paperwidth]{./images/crmetaclin0}
-	\includegraphics[width=0.45\paperwidth]{./images/crmetaclin1}
-\end{frame}
-
-\begin{frame}{Transformation Matrix}
-	Singular Value Decomposition of a dataset X gives:
-	\begin{equation*}
-		X = UDV'
-	\end{equation*}
-	Rearranging this equation gives:
-	\begin{equation*}
-		V' = D^{-1}U'X
-	\end{equation*}
-	Where $D^{-1}U'$ is the transformation matrix and $V'$ is the metagene for the data set X
-	{\footnotesize
-		\begin{itemize}
-			\item This means that the transformation matrix can be applied to other data sets to obtain the metagene for those data sets
-		\end{itemize}
-	}
-\end{frame}
-
-\begin{frame}{Transformation Matrix}
-	\begin{center}
-		\includegraphics[height=0.7\paperheight,width=0.40\paperwidth]{./images/tmp/transmatexp1}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Transformation Matrix}
-	\begin{center}
-		\includegraphics[width=0.70\paperwidth]{./images/tmp/transmatexp2}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Metagene in ICGC data}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga2}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga7}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga12}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga17}\\
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga22}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga27}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga32}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga37}
-\end{frame}
-
-\begin{frame}{Does the metagene correlate with BMI?}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga3}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga8}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga13}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga18}\\
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga23}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga28}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga33}
-	\includegraphics[width=0.23\paperwidth]{./images/crtcga38}
-\end{frame}
-
-\begin{frame}{{\normalsize Obesity-associated genes from Fuentes-Mattei \textit{et al.}}}
-	\begin{columns}
-		\column{0.5\textwidth}
-		137 ER+ breast cancer samples in the study, 130 unique genes identified\\
-		\vskip2ex
-		Data were normalised  with RMA normalisation method\\
-		\vskip2ex
-		\column{0.5\textwidth}
-		\includegraphics[width=0.45\paperwidth]{./images/fmcircos}
-	\end{columns}
-\end{frame}
-
-\begin{frame}{FM metagene in Creighton \textit{et al.} data}
-	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/fmmetacr3}
-	\end{center}
-\end{frame}
-
-\begin{frame}{FM metagene in Creighton \textit{et al.} data}
-	\includegraphics[width=0.45\paperwidth]{./images/fmmetacr4}
-	\includegraphics[width=0.45\paperwidth]{./images/fmmetacr5}
-\end{frame}
-
-\begin{frame}{FM metagene in other cancer types}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat2}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat7}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat12}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat17}\\
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat22}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat27}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat32}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetaheat37}
-\end{frame}
-
-\begin{frame}{FM metagene and BMI}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi3}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi8}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi13}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi18}\\
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi23}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi28}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi33}
-	\includegraphics[width=0.23\paperwidth]{./images/fmmetabmi38}
-\end{frame}
-
-\begin{frame}{Gene Expression Analysis}
-	Maybe the obesity-associated genes identified by either group were not related to obesity, but related to a different clinical variable
-	\vskip2ex
-	Are the differentially expressed genes between the obese and non-obese groups (after correcting for other clinical variables) similar to the obesity-associated genes?
-	{\footnotesize
-		\begin{itemize}
-			\item How do these DEGs compare with the Creighton's or Fuentes-Mattei's obesity-associated genes?
-		\end{itemize}
-	}
-\end{frame}
-
-\begin{frame}{Creighton \textit{et al.} data revisited}
-	Gene expression analysis was carried out on the Creighton \textit{et al.} data:
-	\begin{itemize}
-		\item Original Creighton data
-		\item Caucasian-only Creighton data
-	\end{itemize}
-	\vskip2ex
-	Gene expression analysis was repeated with the residual version of the above data
-	\vskip2ex
-	Each ``obesity-associated genes'' were checked for any overlaps with the original Creighton metagene
-\end{frame}
-
-\begin{frame}{Completely remove the effect of ethnicity}
-	African American patients are more likely to have higher BMI than Cauacasian patients
-	\vskip2ex
-	Removed the effect of ethnicity by removing the African American patients' data (total of 77 patient data to work with)
-	{\footnotesize
-		\begin{itemize}
-			\item All of the other clinical variables were adjusted as before
-		\end{itemize}
-	}
-\end{frame}
-
-\begin{frame}{DEGs in residual data}
-	\begin{center}
-		\includegraphics[width=0.42\paperwidth]{./images/venn0}
-		\includegraphics[width=0.45\paperwidth]{./images/venn1}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Metagene Analysis}
-	All of the gene probe sets I have found (8 in total) were used to create metagenes as before
-	\begin{itemize}
-		\item Checked in Creighton \textit{et al.} data
-		\item Transferred into other cancer data sets
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Initial results -- Checking in Creighton \textit{et al.} data}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr2}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr7}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr12}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr17}\\
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr22}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr27}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr32}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr37}
-\end{frame}
-
-\begin{frame}{Initial results -- Checking in Creighton \textit{et al.} data}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr3}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr8}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr13}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr18}\\
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr23}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr28}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr33}
-	\includegraphics[width=0.23\paperwidth]{./images/degmetacr38}
-\end{frame}
-
-\begin{frame}{Do the metagenes correlate with GE in ICGC data?}
-	\includegraphics[width=0.23\paperwidth]{./images/rawobsmeta2}
-	\includegraphics[width=0.23\paperwidth]{./images/resobsmeta2}
-	\includegraphics[width=0.23\paperwidth]{./images/caobsmeta2}
-	\includegraphics[width=0.23\paperwidth]{./images/caresobsmeta2}\\
-	\includegraphics[width=0.23\paperwidth]{./images/crolmeta2}
-	\includegraphics[width=0.23\paperwidth]{./images/rescrolmeta2}
-	\includegraphics[width=0.23\paperwidth]{./images/cacrolmeta2}
-	\includegraphics[width=0.23\paperwidth]{./images/carescrolmeta2}
-\end{frame}
-
-\begin{frame}{Do the metagenes correlate with BMI?}
-	\includegraphics[width=0.23\paperwidth]{./images/rawobsclin3}
-	\includegraphics[width=0.23\paperwidth]{./images/resobsclin3}
-	\includegraphics[width=0.23\paperwidth]{./images/caobsclin3}
-	\includegraphics[width=0.23\paperwidth]{./images/caresobsclin3}\\
-	\includegraphics[width=0.23\paperwidth]{./images/crolclin3}
-	\includegraphics[width=0.23\paperwidth]{./images/rescrolclin3}
-	\includegraphics[width=0.23\paperwidth]{./images/cacrolclin3}
-	\includegraphics[width=0.23\paperwidth]{./images/carescrolclin3}
-\end{frame}
-
-% \begin{frame}
-% 	Repeated this with Cris' data, but same results as in Creighton and ICGC data
-% \end{frame}
-
-% \begin{frame}{Do the metagenes correlate with each other?}
-% 	\begin{center}
-% 		\includegraphics[width=0.45\paperwidth]{./images/cor1}
-% 	\end{center}
-% \end{frame}
-
-% \begin{frame}{Common genes across multiple cancer types}
-% 	DEGs between the obese and non-obese groups were identified for each cancer types
-% 	\vskip2ex
-% 	For each cancer types, samples were randomly split into two groups (same proportion as obese vs. non-obese samples)
-% 	\vskip2ex
-% 	{\footnotesize
-% 		\begin{tabular}{ccccccccc}
-% 			& \multicolumn{8}{c}{No. of cancer types the gene overlapped}\\
-% 			\rule{0pt}{2.25ex}                    & 1    & 2    & 3   & 4  & 5 & 6 & 7 & 8 \\
-% 			\hline
-% 			\rule{0pt}{2.25ex}Obese vs. Non-obese & 6865 & 1907 & 289  & 26   & 1 & 0 & 0 & 0 \\
-% 			Randomly split groups                 & 6202 & 1488 & 198  & 18   & 1 & 0 & 0 & 0 \\
-% 			No. of genes found                    & 5012 & 870  & 89   & 6    & 0 & 0 & 0 & 0 \\
-% 			on average                            &      &      &      &      &   &   &   & \\
-% 			Empirical FDR                         & 0.73 & 0.46 & 0.31 & 0.23 & 0 & 0 & 0 & 0 \\
-% 		\end{tabular}
-% 	}
-% \end{frame}
-
-% \begin{frame}{Pathway Enrichment Analysis}
-% 	Maybe each cancer type have a different set of genes activated
-% 	\vskip2ex
-% 	Are there any specific pathways enriched in any specific cancer type?
-% 	\vskip2ex
-% 	If so, is it common across other cancer types?
-% 	\vskip2ex
-% 	How relevant is the pathway to obesity and/or cancer?
-% 	\vskip2ex
-% 	Can we use it to narrow down our search for obesity-associated signature?
-% \end{frame}
-
-% \begin{frame}{Pathway Enrichment Analysis}
-% 	Gene Ontology (GO) database was used to check for the function(s) being enriched by the DEGs in each cancer type (between obese and non-obese groups)
-% 	\vskip2ex
-% 	Used a rank-based approach and controlled the error rate using FDR
-% 	\vskip2ex
-% 	\begin{center}
-% 		\textbf {No pathways showed up as significantly enriched}
-% 	\end{center}
-% \end{frame}
-
-% \begin{frame}{Common genes and pathway enrichment analysis}
-% 	There were no obesity-associated genes that were common across all the ICGC cancer types
-% 	\vskip2ex
-% 	There were no pathways enriched in any of the obesity-associated genes found so far
-% \end{frame}
-
-\begin{frame}{Relevance of BMI in Cris' data}
-	\begin{center}
-		\includegraphics[width=0.60\paperwidth]{./images/tmp/mikres}
-	\end{center}
-\end{frame}
-
-\begin{frame}{So \ldots BMI is not related to cancer?}
-	All the results so far suggests that there is no association between BMI and cancer
-	\vskip2ex
-	However, the metagenes are picking up gene signatures  that are common across all cancer types, but not related to BMI
-	\vskip2ex
-	So what signature/mechanism is the metagene picking up?
-\end{frame}
-
-\begin{frame}{Gatza \textit{et al.} (2010)}
-	Used 18 pathways to define sub-class of breast tumours, based on the expression of the pathway signatures
-	\begin{itemize}
-		\item AKT, ER, PR, HER2, PI3K, p53, MYC, RAS, etc...
-	\end{itemize}
-	\vskip2ex
-	Each pathway signature was obtained by overexpressing a single gene in an appropriate cell line, and then differentially expressed genes were noted down
-	\begin{itemize}
-			\item Use these pathways to see what the obesity-associated genes are most similar to
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Tricky thing about metagene direction}
-	Need to consider how the original data was manipulated:
-	\begin{enumerate}
-		\item Normalisation methods (RMA vs MAS5)
-		\item Scaling the data or not
-		\item Metagene scaling (probit vs rank-based)
-	\end{enumerate}
-	\vskip2ex
-	Need to check the direction of each pathway metagenes with its gene (e.g. EGFR gene with EGFR pathway metagene)
-	\vskip2ex
-	Need to consider the grouping/clustering of the pathway metagenes
-\end{frame}
-
-\begin{frame}{What I mean about directions...}
-	\includegraphics[width=0.45\paperwidth]{./images/output-0}
-	\includegraphics[width=0.45\paperwidth]{./images/output-1}
-\end{frame}
-
-\begin{frame}{Some weird genes}
-	\includegraphics[width=0.45\paperwidth]{./images/output-2}
-	\includegraphics[width=0.45\paperwidth]{./images/output-3}
-\end{frame}
-
-\begin{frame}{Grouping/clustering of different pathways}
-	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/gatzares1}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Re-making results from Gatza \textit{et al.} paper}
-	\includegraphics[width=0.45\paperwidth]{./images/out2}
-	\includegraphics[width=0.40\paperwidth]{./images/gatzares1}
-\end{frame}
-
-\begin{frame}{Predicting obesity metagene with pathway metagene}
-	Make a linear model (in Cris' data) with sample BMI and/or pathway metagenes and use this to predict Creighton's obesity metagene
-	\vskip2ex
-	\begin{itemize}
-		\item If any of the pathway metagenes are significant in the model, the pathway is a good predictor of the obesity metagene
-	\end{itemize}
-	\vskip2ex
-	Then use this model to predict the obesity metagene in Creighton data
-	\begin{itemize}
-		\item Compare the predicted obesity metagene with the SVD-generated obesity metagene
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Results}
-	Sample BMI was not significant  in the model
-	\vskip2ex
-	Only pathway that was a good predictor of obesity metagene was PR pathway metagene
-\end{frame}
-
-\begin{frame}{\normalsize Comparison of the predicted obesity metagene with true value}
-	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction1}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction2}
-	\end{center}
-\end{frame}
-
-\begin{frame}{\normalsize Comparison of the predicted obesity metagene with true value}
-	\begin{center}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction3}
-		\includegraphics[width=0.45\paperwidth]{./images/tmp/prediction4}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Overall Summary}
-	\textbf{Creighton/FM metagene analysis}
-	\begin{itemize}
-		\item Both Creighton and FM metagenes correlated with the overall gene expression in any cancer data set
-		\item The metagene score does not correlate with the BMI status of the samples
-	\end{itemize}
-	\vskip2ex
-	\textbf{Gene expression and metagene analysis}
-	\begin{itemize}
-		\item Again, all of the gene sets found correlated with the overall gene expression in any data set, but none of them correlated with the BMI status
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Overall Summary}
-	\textbf{Gatza pathway metagene analysis}
-	\begin{itemize}
-		\item Sample BMI was not a good predictor of the obesity metagene, but PR pathway metagene was
-		\item However, the model was not great at predicting the obesity metagene
-			\begin{itemize}
-				\item Other mechanism contributing to obesity metagene?
-			\end{itemize}
-	\end{itemize}
-\end{frame}
-
-% \begin{frame}{Discussion}
-% 	Tumour heterogeneity within and between samples
-% 	{\footnotesize
-% 		\begin{itemize}
-% 				% TODO rewrite the following sentence
-% 			\item Not finding any significant genes or pathways because the samples are so different from one another
-% 		\end{itemize}
-% 	}
-% 	\vskip2ex
-% 	Tumour sample quality and size
-% 	{\footnotesize
-% 		\begin{itemize}
-% 			\item Picking up too much noise and/or don't have enough samples to find anything significant (i.e. shit data)
-% 		\end{itemize}
-% 	}
-% \end{frame}
-
-% \begin{frame}{Discussion -- Inter-tumour heterogeneity 1}
-% 	Each patient has an obesity driven pathway dysregulated, as well as different ``core'' pathways that lead to cancer progression
-% 	\vskip2ex
-% 	Each data set will have a different collection of patients, which in turn will have a different collection of dysregulated pathways
-% 	{\footnotesize
-% 		\begin{itemize}
-% 			\item Hence DEGs show up, but not the pathways
-% 		\end{itemize}
-% 	}
-% \end{frame}
-
-% \begin{frame}{Discussion -- Inter-tumour heterogeneity 2}
-% 	Some of the high BMI patients in the data set will have tumours that are BMI driven, but BMI may not be the driver for some of these patients
-% 	\begin{itemize}
-% 		\item Only subset of the patients with high BMI are actually driven by BMI
-% 	\end{itemize}
-% 	\vskip2ex
-% 	Other data sets may have less of these truly BMI driven tumours
-% 	\begin{itemize}
-% 		\item Therefore no association with the signatures
-% 	\end{itemize}
-% \end{frame}
-
-\begin{frame}{Acknowledgement}
-	\begin{columns}
-		\column{0.5\textwidth}
-		My supervisor: Mik\\
-		Members of the Black lab\\
-		Cancer Genetics Lab\\
-		SYSKA/Mozilla Study Group\\
-		ICGC and TCGA\\
-		\column{0.5\textwidth}
-		\includegraphics[width=0.3\paperwidth]{./images/tcga}\\
-		\includegraphics[width=0.3\paperwidth]{./images/icgc}
-	\end{columns}
-\end{frame}
-
-% \begin{frame}
-% 	\frametitle{References}
-% 	\bibliographystyle{BiocRefStyle}
-% 	\bibliography{/home/riku/Documents/References/BibTeX/MSc}
-% \end{frame}
-
-% \begin{frame}{test}
-% 	\citet{Fuentes-Mattei2014}\\
-% 	\citet{Creighton2012}\\
-% 	\citet{Alexandrov2013}
-% 	\cite{Gatza2010a}
-% \end{frame}
-
 \end{document}

### 2016-10-22 presentation/beamer_theme/ctcrSep2016.tex
tmp commit

@@
 \title[BMI and Cancer]{Body Mass Index and Pathway Dysregulation in Cancer}
-\subtitle{CTCR Talk}
+\subtitle{CGL Talk}
 \author[Riku Takei]{Riku Takei (MSc)\\Supervisor: Assoc. Prof. Mik Black}
@@
 	\vskip2ex
-	Microarray data from Creighton's paper
-	\vskip2ex
-	Microarray data from Fuentes-Mattei's paper
-	\vskip2ex
-	Microarray data from Cris Print
+	Microarray data from:
+	\begin{itemize}
+		\item Creighton's paper
+		\item Fuentes-Mattei's paper
+		\item Cris Print
+	\end{itemize}
 	\vskip2ex
@@
 \begin{frame}{Creighton \textit{et al.} data revisited}
-	Gene expression analysis was carried out on the Creighton \textit{et al.} data:
+	Gene expression analysis was carried out on:
 	\begin{itemize}
 		\item Original Creighton data
-		\item Caucasian-only Creighton data
+		\item Creighton data (Caucasian samples only)
 	\end{itemize}
@@
 		\item Checked in Creighton \textit{et al.} data
-		\item Transferred into other cancer data sets
+		\item Transferred into other cancer data sets using transformation matrix
 	\end{itemize}
@@
 
-\begin{frame}{Relevance of BMI in Cris' data}
+\begin{frame}{So \ldots BMI is not related to cancer?}
 	\begin{center}
-		\includegraphics[width=0.60\paperwidth]{./images/tmp/mikres}
+		\includegraphics[width=0.55\paperwidth]{./images/tmp/mikres}
 	\end{center}
@@
 \begin{frame}{Gatza \textit{et al.} (2010)}
-	Used 18 pathways to define sub-class of breast tumours, based on the expression of the pathway signatures
+	Used 18 pathways to define sub-classes of breast tumours, based on the expression of the pathway signatures
 	\begin{itemize}
@@
 	\vskip2ex
-	Each pathway signature was obtained by overexpressing a single gene in an appropriate cell line, and then differentially expressed genes were noted down
+	Each pathway signature was obtained by overexpressing a single gene in a model cell line, and then differentially expressed genes were noted down
 	\begin{itemize}
-			\item Use these pathways to see what the obesity-associated genes are most similar to
+			\item Use these `pathway genes' to see what the obesity-associated genes are most similar to
 	\end{itemize}
@@
 
-\begin{frame}{Predicting obesity metagene with pathway metagene}
+\begin{frame}{\normalsize Predicting obesity metagene with pathway metagene}
 	Make a linear model (in Cris' data) with sample BMI and/or pathway metagenes and use this to predict Creighton's obesity metagene
-	\vskip2ex
 	\begin{itemize}
@@
 \begin{frame}{Results}
+	Made 4 models:
+	\begin{itemize}
+		\item BMI-only
+		\item BMI and pathway metagenes
+		\item Pathways-only
+		\item PR pathway only
+	\end{itemize}
+	\vskip2ex
 	Sample BMI was not significant  in the model
@@
 	\begin{itemize}
-		\item Again, all of the gene sets found correlated with the overall gene expression in any data set, but none of them correlated with the BMI status
+		\item Again, all of the gene sets correlated with the overall gene expression in any data set, but none of them correlated with the BMI status
 	\end{itemize}
@@
 		\item Sample BMI was not a good predictor of the obesity metagene, but PR pathway metagene was
-		\item However, the model was not great at predicting the obesity metagene
+		\item None of the models were great at predicting the obesity metagene
 			\begin{itemize}

### 2016-10-22 thesis_plan/plan.tex
added thesis plan

\documentclass[a4paper, 10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[margin=1.5cm]{geometry}

\begin{document}
	\section*{Introduction}
	\begin{itemize}
		\item BMI Intro.
		\item Creighton and Funtes-Mattei intro
		\item Gene signature intro + microarray/NGS intro
		\item BMI and cancer
		\item Mechanism of cancer progression in obese patient
		\item TCGA/ICGC/GEO
		\item Gatza pathway studies (old, 18 pathways and new papers)
		\item SVD/transformation matrix intro (maybe in methods?)
		\item Aim of research
	\end{itemize}
	\section*{Methods}
	\begin{itemize}
		\item WHO BMI classification
		\item Data download
		\item Normalisation methods (RMA/MAS5)
		\item Batch effect
		\item Standardising/scaling
		\item SVD/transformation matrix
		\item Gene expression analysis
		\item Pathway enrichment analysis
		\item Metagene direction
		\item Obesity metagene prediction with Gatza pathway (using linear model)
	\end{itemize}
	\section*{Results part 1 -- Creighton/FM metagene and Cris/ICGC data}
	\begin{itemize}
		\item Original Creighton obesity metagene in Creighton data
		\item Original Creighton obesity metagene in ICGC/Cris data
		\item Original FM obesity metagene in FM and Creighton data
		\item Original FM obesity metagene in ICGC/Cris data
		\item DEGs in Creighton data
		\item DEG obesity metagenes from Creighton data in ICGC/Cris data
		\item Common DEGs in ICGC data (per-cancer and combined)
		\item Path enrichment analysis in ICGC data (per-cancer and combined)
	\end{itemize}
	\section*{Results part 2 -- Gatza pathway metagenes and obesity metagenes}
	\begin{itemize}
		\item Gatza pathway metagene direction in Gatza data
		\item All metagenes in all data set, and correlation between SVD and transformation matrix
			\begin{itemize}
				\item Determining which data set the transformation matrices is being made in
			\end{itemize}
		\item Predicting obesity metagene (in Cris' data) using linear model that was made from Gatza pathways
	\end{itemize}
\end{document}

### 2017-04-02 notes/glossary/glossary.tex
Separated tips and notes directory

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Glossary terms
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newglossaryentry{rnaseq}{
	name=RNA-seq,
	description={Sequence data using the transcriptome}
}

\newglossaryentry{r}{
	name=R,
	description={Statistical programming language}
}

\newglossaryentry{immunoediting}{
	name=Immunoediting,
	text=immunoediting,
	description={Process where the immune system successfully eliminates the highly immunogenic cancer cells during surveillance, but fails to eliminate the weakly immunogenic cells}
}

\newglossaryentry{metagene}{
	name=Metagene,
	text=metagene,
	description={Summary score of a sample that represents the overall gene expression of the genes that was used to obtain the scores}
}

\newglossaryentry{ldp}{
	name=Limited duration prevalence,
	text=limited duration prevalence,
	description={Number of patients diagnosed with a disease within a fixed period of time in the past \citep{Bray2013}}
}

\newglossaryentry{apoptosis}{
	name=Apoptosis,
	text=apoptosis,
	description={The death of cells which occurs as a normal and controlled part of an organism's growth or development}
}

\newglossaryentry{chromothripsis}{
	name=Chromothripsis,
	text=chromothripsis,
	description={An all-at-once mutational pattern in which a large number of rearrangements are confined to local regions of one or more chromosomes \citep{Stephens2011}}
}

\newglossaryentry{kataegis}{
	name=Kataegis,
	text=kataegis,
	description={Showers of single nucleotide changes confined to local regions, often near sites of chromosome rearrangements \citep{Stephens2011}}
}

\newglossaryentry{adipocyte}{
	name=Adipocyte,
	text=adipo\-cyte,
	plural=adipo\-cytes,
	description={Adipose tissue cell}
}

\newglossaryentry{type1}{
	name=Type I error,
	text=Type I error,
	plural=Type I errors,
	description={Incorrect rejection of a true null hypothesis; false positive}
}

\newglossaryentry{type2}{
	name=Type II error,
	text=Type II error,
	plural=Type II errors,
	description={Incorrect retaining of a false null hypothesis; false negative}
}

\newglossaryentry{power}{
	name=Power,
	text=power,
	description={Probability that it will reject a false null hypothesis; it is inversely related to the probability of making a Type II error}
}

\newglossaryentry{ontology}{
	name=Ontology,
	text=ontology,
	description={A set of concepts and categories in a subject area or domain that shows their properties and the relations between them}
}

\newglossaryentry{carcinogen}{
	name=Carcinogen,
	text=carcinogen,
	plural=carcinogens,
	description={A substance capable of causing cancer}
}

\newglossaryentry{passenger}{
	name=Passanger mutation,
	text=passanger,
	description={Mutations that have neither a selective advantage nor contribute to tumour progression}
}

\newglossaryentry{driver}{
	name=Driver mutation,
	text=driver,
	description={Mutations that helps tumorigenesis and its subsequent progression}
}

\newglossaryentry{somatic}{
	name=Somatic mutation,
	text=somatic mutation,
	plural=somatic mutations,
	description={Mutations that are acquried by the cell over the course of the lifetime of an individual}
}

\newglossaryentry{germline}{
	name=Germline mutation,
	text=germline mutation,
	plural=germline mutations,
	description={Mutations that are passed on from the parents}
}

\newglossaryentry{stability}{
	name=Stability gene,
	text=stability gene,
	plural=stability genes,
	description={Genes that are involved in \acrshort{dna} repair mechanisms}
}

\newglossaryentry{allele}{
	name=Allele,
	text=allele,
	plural=alleles,
	description={Variant form of a given gene}
}

\newglossaryentry{suppressor}{
	name=Tumour suppressor gene,
	text=tumour suppressor,
	description={Genes that prevent or ``decelerate'' the process of tumorigenesis and/or tumour progression}
}

\newglossaryentry{oncogene}{
	name=Oncogene,
	text=oncogene,
	plural=oncogenes,
	description={Genes that promote or ``accelerate'' the process of tumorigenesis and/or tumour progression}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acronym terms
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newacronym[description={Disability-Adjusted Life Year},longplural={Disability-Adjusted Life Years}]{daly}{DALY}{Disability-Adjusted Life Year}
\newacronym[description={Genome wide association study},longplural={genome wide association studies},\glsshortpluralkey={GWAS}]{gwas}{GWAS}{genome wide association study}
\newacronym[description={Adrenocorticotropic hormone}]{acth}{ACTH}{adrenocorticotropic hormone}
\newacronym[description={Agouti-related protein}]{agrp}{AGRP}{agouti-related protein}
\newacronym[description={Activation-induced cytidine deaminase}]{aid}{AID}{activation-induced cytidine deaminase}
\newacronym[description={Analysis of variance}]{anova}{ANOVA}{analysis of variance}
\newacronym[description={Apolipoprotein B mRNA editing enzyme, catalytic polypeptide-like}]{apobec}{APOBEC}{apolipoprotein B mRNA editing enzyme, catalytic polypeptide-like}
\newacronym[description={Adenosine triphosphate}]{atp}{ATP}{adenosine triphosphate}
\newacronym[description={$\beta$-catenin}]{bcat}{BCAT}{$\beta$-catenin}
\newacronym[description={B-cell lymphoma 2}]{bcl2}{Bcl-2}{B-cell lymphoma 2}
\newacronym[description={Base-excision repair}]{ber}{BER}{base-excision repair}
\newacronym[description={Bcl-2 homology 3}]{bh3}{BH3}{Bcl-2 homology 3}
\newacronym[description={Bayesian Information Criterion}]{bic}{BIC}{Bayesian Information Criterion}
\newacronym[description={Bladder urothelial carcinoma}]{blca}{BLCA}{bladder urothelial carcinoma}
\newacronym[description={Body mass index}]{bmi}{BMI}{body mass index}
\newacronym[description={Correlation adjusted mean rank gene set test}]{camera}{CAMERA}{correlation adjusted mean rank gene set test}
\newacronym[description={Cocaine and amphetamine related transcript}]{cart}{CART}{cocaine and amphetamine related transcript}
\newacronym[description={Complimentary DNA}]{cdna}{cDNA}{complimentary DNA}
\newacronym[description={Cervical squamous cell carcinoma and endocervical adenocarcinoma}]{cesc}{CESC}{cervical squamous cell carcinoma and endocervical adenocarcinoma}
\newacronym[description={Colon adenocarcinoma}]{coad}{COAD}{colon adenocarcinoma}
\newacronym[description={Counts per million}]{cpm}{cpm}{counts per million}
\newacronym[description={Cytochrome P450}]{cyp}{CYP}{cytochrome P450}
\newacronym[description={Differentially expressed gene}]{deg}{DEG}{differentially expressed gene}
\newacronym[description={Deoxyribonucleic acid}]{dna}{DNA}{deoxyribonucleic acid}
\newacronym[description={Epidermal growth factor receptor}]{egfr}{EGFR}{epidermal growth factor receptor}
\newacronym[description={Endothelial-mesenchymal transition}]{emt}{EMT}{endothelial-mesenchymal transition}
\newacronym[description={Extracellular signal regulated kinase}]{erk}{ERK}{extracellular signal regulated kinase}
\newacronym[description={Estrogen receptor}]{er}{ER}{estrogen receptor}
\newacronym[description={Functional class scoring}]{fcs}{FCS}{functional class scoring}
\newacronym[description={Fold change}]{fc}{FC}{fold change}
\newacronym[description={False discovery rate}]{fdr}{FDR}{false discovery rate}
\newacronym[description={Fat mass and obesity associated}]{fto}{FTO}{fat mass and obesity associated}
\newacronym[description={Family-wise error rate}]{fwer}{FWER}{family-wise error rate}
\newacronym[description={Gap 1}]{g1}{G1}{Gap 1}
\newacronym[description={$\gamma$-aminobutyric acid}]{gaba}{GABA}{$\gamma$-aminobutyric acid}
\newacronym[description={Gene Expression Omnibus}]{geo}{GEO}{Gene Expression Omnibus}
\newacronym[description={Growth hormone secretagogue receptor}]{ghsr}{GHSR}{growth hormone secretagogue receptor}
\newacronym[description={Gene Ontology}]{go}{GO}{Gene Ontology}
\newacronym[description={G-protein coupled receptor}]{gpcr}{GPCR}{G-protein coupled receptor}
\newacronym[description={Gene Set Enrichment Analysis}]{gsea}{GSEA}{Gene Set Enrichment Analysis}
\newacronym[description={Human epidermal growth factor receptor 2}]{her2}{HER2}{human epidermal growth factor receptor 2}
\newacronym[description={International Cancer Genome Consortium}]{icgc}{ICGC}{International Cancer Genome Consortium}
\newacronym[description={Interferon-$\alpha$}]{ifna}{IFN$\alpha$}{interferon-$\alpha$}
\newacronym[description={Interferon-$\gamma$}]{ifny}{IFN$\gamma$}{interferon-$\gamma$}
\newacronym[description={Insulin-like growth factor binding protein}]{igfbp}{IGFBP}{insulin-like growth factor binding protein}
\newacronym[description={Insulin-like growth factor}]{igf}{IGF}{insulin-like growth factor}
\newacronym[description={Interleukin-10}]{il10}{IL-10}{interleukin-10}
\newacronym[description={Interleukin-6}]{il6}{IL-6}{interleukin-6}
\newacronym[description={Insulin receptor substrate}]{irs}{IRS}{insulin receptor substrate}
\newacronym[description={Janus kinase}]{jak}{JAK}{janus kinase}
\newacronym[description={Kyoto Encyclopedia of Genes and Genomes}]{kegg}{KEGG}{Kyoto Encyclopedia of Genes and Genomes}
\newacronym[description={Kidney renal clear cell carcinoma}]{kirp}{KIRP}{kidney renal clear cell carcinoma}
\newacronym[description={Liver hepatocellular carcinoma}]{lihc}{LIHC}{liver hepatocellular carcinoma}
\newacronym[description={Lymph node}]{ln}{LN}{lymph node}
\newacronym[description={Microarray analysis suite 5.0}]{mas}{MAS5}{microarray analysis suite 5.0}
\newacronym[description={Melanocortin 1 receptor}]{mc1r}{MC1R}{melanocortin 1 receptor}
\newacronym[description={Melanocortin 4 receptor}]{mc4r}{MC4R}{melanocortin 4 receptor}
\newacronym[description={Mesenchymal-endothelial transition}]{met}{MET}{mesenchymal-endothelial transition}
\newacronym[description={Mismatch repair}]{mmr}{MMR}{mismatch repair}
\newacronym[description={Mismatch}]{mm}{MM}{mismatch}
\newacronym[description={Messenger RNA}]{mrna}{mRNA}{messenger RNA}
\newacronym[description={Melanocyte-stimulating hormone}]{msh}{MSH}{melanocyte-stimulating hormone}
\newacronym[description={Mammalian target of rapamycin}]{mtor}{mTOR}{mammalian target of rapamycin}
\newacronym[description={Mean}]{m}{$\mu$}{mean}
\newacronym[description={Nucleotide-excision repair}]{ner}{NER}{nucleotide-excision repair}
\newacronym[description={Nuclear factor $\kappa$B}]{nfkb}{NF-$\kappa$B}{nuclear factor $\kappa$B}
\newacronym[description={Next generation sequencing}]{ngs}{NGS}{next generation sequencing}
\newacronym[description={4-(methylnitrosamino)-1-(3-pyridyl)-1-butanone}]{nnk}{NNK}{4-(methylnitrosamino)-1-(3-pyridyl)-1-butanone}
\newacronym[description={Neuropeptide Y}]{npy}{NPY}{neuropeptide Y}
\newacronym[description={Neuropeptide Y1 receptor}]{nyr1}{Y1R}{neuropeptide Y1 receptor}
\newacronym[description={New Zealand Breast Cancer}]{nzbc}{NZBC}{New Zealand Breast Cancer}
\newacronym[description={New Zealand}]{nz}{NZ}{New Zealand}
\newacronym[description={Obesity policy action}]{opa}{OPA}{obesity policy action}
\newacronym[description={Over representation analysis}]{ora}{ORA}{over representation analysis}
\newacronym[description={Polycyclic aromatic hydrocarbon}]{pah}{PAH}{polycyclic aromatic hydrocarbon}
\newacronym[description={Prohormone convertase 1}]{pc1}{PC1}{prohormone convertase 1}
\newacronym[description={Phosphoenol pyruvate}]{pep}{PEP}{phosphoenol pyruvate}
\newacronym[description={Phosphatidylinositol 3-kinase}]{pi3k}{PI3K}{phosphatidylinositol 3-kinase}
\newacronym[description={Perfect match}]{pm}{PM}{perfect match}
\newacronym[description={Pro-opiomelanocortin}]{pomc}{POMC}{pro-opiomelanocortin}
\newacronym[description={Pattern recognition receptor}]{prr}{PRR}{pattern recognition receptor}
\newacronym[description={Progesterone receptor}]{pr}{PR}{progesterone receptor}
\newacronym[description={3-36}]{pyy}{PYY$_{3-36}$}{peptide YY$_{3-36}$}
\newacronym[description={Retinoblastoma-associated}]{rb}{RB}{retinoblastoma-associated}
\newacronym[description={Rectum adenocarcinoma}]{read}{READ}{rectum adenocarcinoma}
\newacronym[description={Robust multi array}]{rma}{RMA}{robust multi array}
\newacronym[description={Ribonucleic acid}]{rna}{RNA}{ribonucleic acid}
\newacronym[description={Reactive oxygen species}]{ros}{ROS}{reactive oxygen species}
\newacronym[description={Standard deviation}]{sd}{$\sigma$}{standard deviation}
\newacronym[description={Skin cutaneous melanoma}]{skcm}{SKCM}{skin cutaneous melanoma}
\newacronym[description={Single nucleotide polymorphism}]{snp}{SNP}{single nucleotide polymorphism}
\newacronym[description={Single-stranded DNA}]{ssdna}{ssDNA}{single-stranded DNA}
\newacronym[description={Signal transducer and activator of transcription 3}]{stat3}{STAT3}{signal transducer and activator of transcription 3}
\newacronym[description={Signal transducer and activator of transcription}]{stat}{STAT}{signal transducer and activator of transcription}
\newacronym[description={Singular value decomposition}]{svd}{SVD}{singular value decomposition}
\newacronym[description={The Cancer Genome Atlas}]{tcga}{TCGA}{The Cancer Genome Atlas}
\newacronym[description={Transforming growth factor-$\beta$}]{tgfb}{TGF$\beta$}{transforming growth factor-$\beta$}
\newacronym[description={Toll-like receptor}]{tlr}{TLR}{Toll-like receptor}
\newacronym[description={Tumour necrosis factor-$\alpha$}]{tnfa}{TNF$\alpha$}{tumour necrosis factor-$\alpha$}
\newacronym[description={Thrombospondin-1}]{tsp1}{TSP-1}{thrombospondin-1}
\newacronym[description={Uterine corpus endometrial carcinoma}]{ucec}{UCEC}{uterine corpus endometrial carcinoma}
\newacronym[description={United States of America}]{usa}{USA}{United States of America}
\newacronym[description={Ultraviolet}]{uv}{UV}{ultraviolet}
\newacronym[description={Vascular endothelial growth factor-A}]{vegfa}{VEGF-A}{vascular endothelial growth factor-A}
\newacronym[description={Vascular endothelial growth factor}]{vegf}{VEGF}{vascular endothelial growth factor}
\newacronym[description={Variance inflation factor}]{vif}{VIF}{variance inflation factor}
\newacronym[description={World Health Organization}]{who}{WHO}{World Health Organization}
\newacronym[description={Wilcoxon-Mann-Whitney}]{wmw}{WMW}{Wilcoxon-Mann-Whitney}

### 2017-04-02 notes/notebook.tex
Separated tips and notes directory

% My master tips book
\documentclass[a4paper, 12pt] {report}

% All of the packages imported for this document:
\usepackage[utf8]{inputenc}
\usepackage{glossaries}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{url}
\usepackage{multicol}

\makeglossaries

\begin{document}

\include{./title/title}
\include{./glossary/glossary}

\tableofcontents

\printglossaries
\clearpage

\include{./subnotes/intro}

\include{./subnotes/gwas}

\end{document}

### 2017-04-02 notes/subnotes/intro.tex
Separated tips and notes directory

%--------------------------------------------------------------------------------
%     File Name           :     subnotes/intro.tex
%     Created By          :     rikutakei
%     Creation Date       :     [2017-04-03 10:27]
%     Last Modified       :     [2017-04-03 10:37]
%     Description         :     Intro for my notebook
%--------------------------------------------------------------------------------

\chapter{Introduction}
\label{cha:introduction}

This notebook is for noting down all of the concepts and background information I need to know for my current and future projects.
I will try and maintain all of the references and keep this note as updated as possible.
\\

\noindent
This notebook is free of use for everyone, as long as it complies with the licence file associated with this Git repository.

\vfill

Riku Takei

### 2017-04-02 tips/vim/chapter1/intro.tex
Separated tips and notes directory

--- 2016-03-22 tips_and_notes/vim/intro.tex
@@
 
-As I read more and more about vim, the urge to learn vim properly became larger.
+As I read more and more about vim, I had an urge to learn vim properly.
 I had plenty of other options to learn before vim (for example, another programming language), but I chose vim to be my first target - if I could master how to edit text faster, then whatever that follows will also (probably) become faster.
@@
 This book is just a bunch of tips I pulled out from his book.
+
+\section{Before you start}
+
+Firstly, make sure you can touch type.
+It doesn't have to be super fast or anything like that - just know where to find the keys that you are looking for (including special symbols like \%).
+A good website I found was \url{www.keybr.com}.
+It's free to sign up, and keeps all the record of what you have typed and what keys you need to work on.
+
+Secondly, do the Vim tutor on command line (type \verb|vimtutor| in terminal).
+This tutorial will get you through the basics of vim and how to get around files, save files, close files, etc.
+Vim tutor will be a good starting point before using vim.
+
+Lastly, have a google around for the spf13-vim plugin.
+This plugin downloads other plugins that make vim awesome (e.g. tab completion).
+Starting off with this plugin will make your life a lot easier, but remember to make your own .vimrc file once you get the hang of all the commands and (re-)mappings.

### 2017-04-02 tips/vim/chapter3/normal.tex
Separated tips and notes directory

--- 2016-03-30 tips_and_notes/vim/chapter3/normal.tex
@@
     h/j/k/l & default cursor movement\\
-    w/W     & word or WORD\\
+    w/W     & go to next word or WORD\\
     b/B     & go back a word or WORD\\

### 2017-04-02 tips/vim/chapter5/simple.tex
Separated tips and notes directory

\chapter{Navigation within files}

\section{Basic navigation}

The idea here is to keep your fingers on the home row, no matter what.
Moving your hands from the keyboard to the mouse is going to waste your time.

\subsection{Moving letter-wise}

To move around in the document one cursor (i.e. one letter/character) at a time, press the \verb|h/j/k/l| keys, which corresponds to the left/down/up/right arrow keys.
It is easy to reach out onto your arrow keys instead of using these keys, but, like with reaching for the mouse, it will waste your time.

To force yourself to use the \verb|h/j/k/l| keys, you should map the arrow keys to do nothing: \verb|noremap <Up> <Nop>| (and do it for left/right/down keys).

\subsection{Moving word-wise}

You can move around bits of text character-wise, but moving word-wise would be faster.
The \verb|w/b/e/ge| commands will allow you to move word-wise through the text, each offering slightly different movement.
\verb|w/b| moves your cursor forward/backward to the start of the next word, and \verb|e/ge| moves your cursor forward/backward to the end of the next word.

\subsection{word vs WORD}

WORDS are sequence of any non-blank characters, whereas words are sequence of letters, digits, and underscores, or as a sequence of other non-blank characters.
For example, \verb|www.google.com| has 5 words, but only 1 WORD.
To move WORD-wise instead of word-wise, you will have to press the uppercase version of the word-wise commands (i.e. \verb|W/B/E/gE|).

\subsection{Moving to a specific position on a line}

Press \verb|0/\$| to move to the beginning/end of the line.
To go to the first non-blank character of the line, press \verb|^|.

### 2017-04-02 tips/vim/chapter9/substitution.tex
Separated tips and notes directory

\chapter{Substitution}

Substitution is very much like search, but allows you to replace the search pattern with another string.

\section{Syntax of substitute command}

\begin{center}
    \LARGE{\verb|:[range]s/{pattern}/{string}/[flags]|}
\end{center}

\section{Tweaking substitute command using flags}

There are different flags you could use in the substitute command to suit the requirement(s).
Some of the flags are listed below (see \verb|:h :s_flags| for a full reference).
\newline

\begin{tabular}{c|l}
\verb|g| & use substitution command globally on the current line\\
\verb|c| & allow you to confirm or reject the changes each time\\
\verb|n| & suppresses the normal substitution and report the number of occurrences\\
\verb|&| & reuse the previous flags\\
\verb|e| & suppress error messages\\
\end{tabular}

\section{Special characters for the replacement string}

There are some special characters you can use in the replacement string.
Some of them are listed below (see \verb|:h sub-replace-special| for full reference).
\newline

\begin{tabular}{c|l}
\verb|\r|             & \verb|<CR>| \\
\verb|\t|             &  tab\\
\verb|\\|             & single backslash \\
\verb|\{digit}|       & submatches \\
\verb|\0|             & entire matched pattern\\
\verb|&|  & entire matched pattern\\
\verb|~|              &  use the \verb|{string}| from the previous substitution\\
\verb|\={vim script}| & evaluate vim script and replace \verb|{string}| with the result of script\\
\end{tabular}

\section{Find and replace every match in a file}

Note in the previous section that the \verb|g| flag only works on the current line.
To act on the whole file, you need to use \verb|%| (all the lines in a file) as the range.

\section{Check each substitution}

Use the \verb|c| flag to check/confirm if you want to substitute it or not (e.g. \verb|:%s/{pattern}/{string}/gc|)
\newline

\begin{tabular}{c|l}
\verb|y| & ``yes''/substitute\\
\verb|b| & ``no''/don't substitute\\
\verb|q| & quit substitution\\
\verb|l| & ``last'' substitution (substitute this match and quit)\\
\verb|a| & ``all'' (substitute this and any remaining matches)\\
\verb|<ctrl-e>| & scroll screen up\\
\verb|<ctrl-y>| & scroll screen down\\
\end{tabular}

\section{Re-using the last search pattern}

Leave the \verb|{pattern}| field empty to use the last search pattern.
If you want to, you can paste it into the substitute command with \verb|<ctrl-r>/| (register where the search pattern is stored).
This will keep the full record of what you have substituted on the command history, rather than a blank line.

\section{Replace pattern with the contents of a register}

You can do this by typing \verb|<ctrl-r>{register}|.
This may cause problems of the contents of that register contains special characters or have multiple lines (in which case you will have to escape each character).
Instead of pasting the contents directly into the substitute command, you can reference the register bu evaluating it as a vim script (e.g. \verb|:%s//\=@0/g|).

The \verb|\=| tells vim to evaluate a vim script expression, therefore the command tells vim to substitute the pattern using the contents of \verb|@0| (yank register).
By using the  register reference, you run the risk of substituting different patterns with different words when we originally ran the \verb|:%s//\=@0/g| command.
But we can take advantage of this behaviour by searching for a different pattern, then yanking a different replacement.
When you run the exact same command, it will replace your search pattern with the yanked text.

\section{Repeating the previous substitute command}

If you missed the \verb|%| symbol to make it a file-wise substitution, you can just press \verb|g&|, which is equivalent to \verb|:%s///~/&|.
Some useful commands are listed below (see pg 227 for good mappings):
\newline

\begin{tabular}{c|l}
\verb|:&| & repeat the last substitution command on this line (flag ignored)\\
\verb|:&&| & same as \verb|:&| but use the flags in previous substitution command.\\
& Note that this can be used with visual selection\\
\verb|:%&&| & same as \verb|:&&| but act file-wise (equivalent to \verb|g&|)\\
\end{tabular}

### 2017-04-03 notes/glossary/glossary.tex
tmp commit

@@
 	description={Statistical programming language}
+}
+
+\newglossaryentry{ld}{
+	name=Linkage disquilibrium,
+	text=linkage disquilibrium,
+	description={Correlation between nearby variants such that the alleles at neighbouring markers (on the same chromosome) are associated within a population more often than if they were unlinked}
+}
+
+\newglossaryentry{heritability}{
+	name=Heritability,
+	text=heritability,
+	description={The proportion of the variation in a given characteristic or state that can be attributed to (additive) genetic factors}
 }

### 2017-04-03 notes/notebook.tex
tmp commit

@@
 % All of the packages imported for this document:
-\usepackage[utf8]{inputenc}
-\usepackage{glossaries}
-\usepackage{hyperref}
-\usepackage{setspace}
-\usepackage{url}
-\usepackage{multicol}
+% \usepackage[utf8]{inputenc}
+% \usepackage{glossaries}
+% \usepackage{hyperref}
+% \usepackage{setspace}
+% \usepackage{url}
+% \usepackage{multicol}
+
+% Preamble:
+\usepackage[T1]{fontenc}                                 % package for specifying output font encoding
+\usepackage[acronym,style=super,nogroupskip,section=chapter]{glossaries} % package for making glossaries
+\usepackage[flushleft]{threeparttablex}                  % package to add notes at the bottom of the table
+\usepackage[font=footnotesize,format=hang]{caption}      % package for formatting figure caption
+\usepackage[hidelinks, colorlinks]{hyperref}             % package for hyperlinking references - hidelinks option removes the ugly boxes around the huperlinks
+\usepackage[md]{titlesec}                                % package for tweaking the chapter/section/subsection titles format
+\usepackage[noabbrev,capitalise]{cleveref}               % package for referencing multiple referencing in a single \ref command
+\usepackage[toc]{appendix}                               % package for making appendices
+\usepackage[utf8]{inputenc}                              % package for specifying input font encoding
+\usepackage{chngcntr}                                    % package for making continuous figure numbering, instead of section-based numbering
+\usepackage{graphicx}                                    % package for graphics handling (i.e. figure images)
+\usepackage{mathptmx}                                    % package for Times New Roman font
+\usepackage{natbib}                                      % package for reference management
+\usepackage{setspace}                                    % package for setting line space (used in title page)
+\usepackage{xcolor,colortbl}                             % package for making colours
+\usepackage{booktabs}                                    % package for better looking tables
+\usepackage{longtable}                                   % package for tables that span multiple pages
+\usepackage{siunitx}                                     % package for scientific notations and units
+\usepackage{etoolbox}                                    % package to 'robustify' \bfseries
+\usepackage{enumitem}                                    % package for tweaking lists
+
+% Set up the hyperlinks - change the colour of the links to blue:
+% \hypersetup{allcolors=blue}
+\hypersetup{allcolors=black} % uncomment this line to make the hyperlinks black
+
+\def\labelitemi{--}
 
@@
 
+\bibliographystyle{BiocRefStyle}
+
+\addcontentsline{toc}{chapter}{\bibname}
+
+% Print out the bibliography:
+\bibliography{./GWAS}
+
 \end{document}

### 2017-04-03 notes/subnotes/gwas.tex
tmp commit

\chapter{\Gls{gwas}}
\label{cha:gwas}

This chapter will be written using the information from the \citet{Hirschhorn2005} and \citet{McCarthy2008} reviews as a starting point/guide.

\section{Mendelian diseases}
\label{sec:mendelian_diseases}

Mendelian diseases, or monogenic diseases, are diseases caused by (often) rare gene variants that directly affects the phenotype of the patient.
These variants are usually penetrant, meaning that the carrier of the disease-causing allele will always show the disease phenotype.
Therefore, by tracking a ``marker'' at or near the disease-causing allele you are able to determine the disease status of an individual for a Mendelian disease (since these markers will co-segregate with the causal allele).

Another thing to note about Mendelian diseases is that the causal variants are usually rare.
This is because negative selection would have acted on the resulting phenotype (i.e.\ disease) which causes early-onset mortality and morbidity.
As a result, the frequencies of these variants observed in the population are dramatically reduced.
This also means that the variants that cause Mendelian diseases emerged reltively recently, in terms of evolutionary time scale, otherwise the variants would not have survived the negative selection over many generations.

Linkage analysis has been used to detect and identify the causal genes of Mendelian diseases.
In linkage studies, many markers are spread out along the genome and the segregation of these markers are tracked through generations.
Markers that are close to the disease-causing gene will segregate with the disease in the family since Mendelian diseases are highly penetrant (meaning that the individuals who have the disease allele will have the disease).
Therefore, markers that are close (within 10$\sim$20 cM) to  the disease-causing alleles will co-segregate with disease status, and able to be tracked and studied.

\section{Common diseases and quantitative traits}
\label{sec:common_diseases_and_quantitative_traits}

Common diseases and quantitative traits are not as simple as Mendelian diseases.
For common diseases and quantitative traits, it is hypothesised that the traits will be determined by variants that occur frequently in the population, but each only having a small impact on the resulting phenotype.
This is known as the ``common disease, common  variant'' hypothesis.
Since each individual variants contribute small amount to the phenotype (whether it is related to the disease or a trait), these variants would have gone through less strict negative selection process, unlike the rare Mendelian variants.
Therefore, most variants are likely to be ``ancient'' and have been present in the population for a long time.

Linkage analysis has been very successful in identifying genes that cause rare Mendelian diseases, but it was not so successful in identifying genes for common disease or quantitative trait.
There have been a few genes, gene regions and gene variants that have been identified that related to some diseases, but these variants usually explain only a small fraction of the \gls{heritability} of the disease.
For example, the identified variant may explain an excess risk of two-fold, where in reality there is a total excess risk of 30-fold, which shows that there are many other causal genes that are yet to be identified.

There are various reasons for this:
\begin{itemize}[noitemsep]
	\item low \gls{heritability} of most complex traits
	\item the inability of the standard set of microsatellite markers to extract complete information about inheritance
	\item the imprecise definition of phenotypes
	\item inadequately powered study designs
\end{itemize}
Linkage analysis has a poor power of detecting common alleles that have low penetrance and have modest effect on disease.
Therefore, an alternative strategy is required to study common diseases and traits.

\section{Candidate gene resequencing studies}
\label{sec:candidate_gene_resequencing_studies}

Candidate gene resequencing studies are where you select genes or gene regions of interest that may be important to the disease.
These genes are chosen based on their location in a region of linkage, or some other evidence that may affect the disease risk.
Once these genes or gene regions are resequenced, the variants or set of variants that are different between the cases and controls are searched.

This approach is expensive and labour-intensive, and therefore it is largely limited to the coding regions of one or a few candidate genes.

\section{Association studies}
\label{sec:association_studies}

In brief, association studies compare the frequency of common alleles or genotypes of  a particular variant between disease cases and controls.

\subsection{Candidate gene association studies}
\label{sub:candidate_gene_association_studies}

In this approach, a set of genes will be chosen based on the biological hypotheses, or the location of the genes that are within previously determined region of linkage.
These studies are made more efficient by the use of \gls{ld}, as less markers will be required to assess multiple variants or genomic regions.
Candidate gene association studies have identified many of the genes that are known to contribute to susceptability to common disease.
With that said, these studies can only identify  a fraction of genetic risk factors, even for diseases with well-known pathophysiology.
Furthermore, it is clear that if the physiological mechanism of a disease is unknown, this approach would not be able to fully explain the genetic basis of the disease.

\subsection{\Acrfull{gwas}}
\label{sub:gwas}

In contrast to the candidate gene association studies, \gls{gwas} looks at most of the genome for the causal variants.
Unlike candidate gene association studies, \gls{gwas} does not make any assumptions about the genomic location of the causal variants.
This means that you are able to carry out the association studies without having to know the causal genes of the disease.
``The genome-wide association approach therefore represents an unbiased yet fairly comprehensive option that can be attempted even in the absence of convincing evidence regarding the function or location of the causal genes'' \citep{Hirschhorn2005}.

### 2017-04-11 notes/subnotes/gwas.tex
tmp commit

@@
 This chapter will be written using the information from the \citet{Hirschhorn2005} and \citet{McCarthy2008} reviews as a starting point/guide.
+See \citet{Astle2009} for detailed discussion about relatedness.
 
@@
 These studies are made more efficient by the use of \gls{ld}, as less markers will be required to assess multiple variants or genomic regions.
-Candidate gene association studies have identified many of the genes that are known to contribute to susceptability to common disease.
+Candidate gene association studies have identified many of the genes that are known to contribute to susceptibility to common disease.
 With that said, these studies can only identify  a fraction of genetic risk factors, even for diseases with well-known pathophysiology.

### 2017-04-11 notes/subnotes/tmp.tex
tmp commit

\section{Types of markers used in \gls{gwas}}
\label{sec:types_of_markers_used_in_gwas}

\subsection{LD-based markers}
\label{sub:ld_based_markers}

As mentioned earlier, LD can be useful to make a \gls{gwas} more efficient.
From the HapMap studies, it was shown that most of the roughly 11 million SNPs in the genome have neighbouring groups that are nearly perfectly correlated with each other.
This means that the genotype of one SNP perfectly predicts the other SNPs that correlate with it (i.e.\ in high LD).
Since SNPs that are in high LD segregate together, genotyping a single marker SNP from the group of SNPs that are in high LD will be sufficient to track all of the SNPs for that group.
Using the information about these known LD patterns, you can choose a few ``tag SNPs'' that represent or capture the most common variation within the regions of high LD.
It is theorised that a well-chosen set of SNPs can be used to provide information about the genome without having to genotype the whole genome.
With that said, in a region of high LD, it is hard to fine-map a causal variant since there will be highly correlated markers each showing a similar strength of association with the phenotype.

There are some populations that have larger variation and less LD than other populations, such as the African populations.
For these populations, a larger number of SNPs is required to gain useful information about their genome.
This means that the LD-based markers may be appropriate for one population, but it  may fail to detect variation in a different population due to the difference in the LD pattern for that population.
Therefore, it is important to consider the ancestry of the population when choosing markers based on LD.

\subsection{Missense mutation markers}
\label{sub:missense_mutation_markers}

Missense mutations are mutations in the coding region of the gene that result in amino acid change.
Since there is a high proportion of missense mutations that cause Mendelian diseases, it is proposed that focussing on these missense mutations may be better and more efficient than genotyping the whole genome.
A typical gene contains one or two missense mutations, which means that only about 30000 to 60000 SNPs are required to be genotyped.
Because missense mutations are more likely to have functional consequences, this approach may be more efficient at identifying the cause of the disease.

However, Mendelian diseases are highly penetrant and cause severe phenotypes.
Furthermore, these diseases are caused by not only missense mutations, but frameshift, insertion and deletion, and other mutations.
In contrast, the alleles that contribute to complex phenotypes have more subtle effects, and are more likely to be caused by mutations in the non-coding regulatory regions rather than in the coding regions.
Since these causal alleles contribute modestly to the risk factors associated with the common diseases, these alleles are less likely to be subject to negative selection, unlike those alleles that cause Mendelian disorders.

SNPs in the coding regions are usually accepted as the ``answer'' when an association of the missense mutation is detected.
This may not be the case, as some nearby non-coding mutation may be more strongly associated with the disease than the SNPs in the  coding regions.
Therefore, regardless of where the SNPs are located (coding or non-coding), SNPs that are associated with the disease should be validated.

\subsection{Convenience-based markers}
\label{sub:convenience_based_markers}

In this approach, the markers are picked based on the ease and cost of genotyping.
This approach will be less efficient per variant for surveying the genome for disease alleles than a set based on LD of functional considerations.
One way to do this is to type a few SNPs in or near the coding region of each gene.
This will allow the detection of those chosen SNPs as well as those that are in high LD with these SNPs.
Limitation to this approach is that the SNPs that are chosen based on the physical proximity does not guarantee the detection of the SNPs that are nearby, and will almost certainly miss the SNPs that are further away from the SNPs.

\section{Study designs to make \gls{gwas} more efficient}
\label{sec:study_designs_to_make_gwas_more_efficient}

\subsection{Sample size}
\label{sub:sample_size}

\Gls{gwas} are very expensive and requires a lot of effort.
Due to this, \gls{gwas} try to keep the sample size low, but this results in the reduction in power.
Since variants that contribute to common traits are likely to have modest effect on the phenotype, large sample size is  required to have enough power  to detect anything interesting.
In addition to this, the need of large sample size is further emphasised by the large number of hypothesis testing done in \gls{gwas}.
Larger sample size will be required to keep the findings significant after correcting for multiple testing.

With that said, you are able to reduce the sample size if you lower the significance threshold (i.e.\ make it less stringent).
However, by doing this you run the risk of introducing more false positives in the results, and therefore require the identification of the true causal genes from the sea of false positives.
Depending on what the outcome you want from the study, the sample size should be chosen accordingly.

\subsection{Multi-stage approach}
\label{sub:multi_stage_approach}

One way to overcome the limitation of sample size and the accompanied false positives is to use a two- or three-stage screening process.
In the first stage, a subset of samples are genotyped for the whole genome with significance threshold set to a value such that it is enough to detect loci that is associated with the phenotype.
This will also allow reasonable number of false positives to be detected.

In the second stage, all of the markers that passed the threshold in the  first stage is then tested on an independent population sample with a more stringent threshold value (similar or larger sample size than the first stage).
Since the second stage only uses the markers that were significant in the first stage, the process  will be more efficient in the second stage.
Furthermore, second stage will be able to identify the true positives that are associated with the phenotype from the large number of false positives identified in the first stage.
This can be repeated for the third stage as well.

There are a number of ways in which the  thresholds for these sort of analyses can be defined.
The first is to use the Bonferroni correction, but this will generally give a ridiculously low p-value threshold due to its highly conservative nature.
Permutation approach can be used to determine the threshold by comparing  the probability of results observed from the analysis and the results observed by chance (from the permutation).
There are other approaches like frequentist and Bayesian approaches.

\subsection{Founder population}
\label{sub:founder_population}

Founder populations are those that have been recently derived (\textless{}100 generations ago) from a limited pool of individuals.
Founder population will therefore represent the genetic make up of the population that originated from these founders.
Furthermore, these founders will have had less recombination compared with their descendants, so the LD patterns in these founding populations will be larger (as recombination would have disrupted these LD sites in the descendants).
By using the founder populations, you are able to use less markers to genotype the samples, as the LD patterns will be larger in the founders (and therefore require less markers to obtain information about the genome).

This approach is better for localising genes that underlie Mendelian disorders (i.e.\ rare and recent variants), and may provide less use for common diseases.
This is because common alleles would have already entered the population multiple times, and so the LD pattern surrounding these alleles will already be smaller around these common variants.
Therefore, the number of markers required for these common alleles may not be reduced substantially.

\subsection{Pooled samples}
\label{sub:pooled_samples}

Pooled samples is when you pool equal amounts  of \acrshort{dna} from multiple individuals into a single ``sample'', and genotype this mixed sample.
Allele frequencies are estimated from the pooled genotypes and \gls{gwas} is carried out using this estimate.
Pooled samples are able to reduce the cost as less number of genotyping is required.
However, for some diseases, individual genotype data is required to carry out complex analyses of gene-gene or gene-environment interactions (you are taking one genotype that represents the phenotypes of those samples used for genotyping; in other words a single genotype is not representing a single phenotype).

\section{Limitations}
\label{sec:limitations}

\subsection{Rare alleles and \gls{gwas}}
\label{ssub:rare_alleles_and_gwas}

The ability or power to detect alleles increases with increasing frequency and increasing penetrance of the allele.
In other words,  the power to detect an allele depends on the proportion of the phenotypic variance in the population that is explained by a particular variant.
Therefore, highly penetrant and rare alleles are able to be detected, and so are common alleles that have low penetrance.
However, for those rare alleles that have modest effect on the phenotype are very difficult to detect with \gls{gwas}.
Furthermore, since tag SNPs are currently designed to tag common SNPs (\textgreater{}5\% frequencies), rare alleles are less well represented in the SNP databases.

One thing to note about rare alleles is that rare alleles are likely to have arisen relatively recently, and therefore there would have been less recombination and mutation that disrupt the haplotype on which they arose.
This also means that rare variants are expected to be on single, long haplotypes.

\subsection{Population stratification}
\label{sub:population_stratification}

In a large population group, the population will have a mix of ethnicities or sub-populations.
These different sub groups may have different disease prevalence or phenotypes compared with other groups, and the disease cases can be over-represented in one or more sub groups.
If this is the case, then the frequencies of the markers used will be different depending on the subgroups, and therefore lead to a bias and false positives.

One way to avoid this is by using a well-matched cases and  controls in the study.
This will prevent a large-scale population stratification in the study cohort, as the frequencies of the markers will be similar between the cases and controls when they are well-matched.
However, this method cannot completely remove population stratification, and mild stratification will remain.
Another way of preventing population stratification is by using family-based samples.
Since the offspring will have inherited the genes from the parents, there will be no problem of stratification between the cases and controls (all of the genetic variations observed in the offspring is represented in the genetic make up of the parents).
However, family-based approach will reduce the power, due to smaller sample size compared to normal \gls{gwas}.

\section{Extending \gls{gwas}}
\label{sec:extending_gwas}

\subsection{Meta-analysis}
\label{sub:meta_analysis}

Individual \gls{gwas} are underpowered to detect only those variants that have the biggest effect, and greater power is required to detect other susceptibility loci.
Meta-analysis can be carried out on the data from a comparable \gls{gwas}.
This is a low-cost approach to enhance the power for the main  as well as the joint (gene-gene and gene-environment) effects.
Furthermore, this allows you to replicate the results, get better idea of the SNPs for subsequent replication studies, and investigate potential sources of heterogeneity.

Summary-level data will be enough to carry out many meta-analysis, but access to individual-level data allows for more sophisticated analyses.
These include: haplotype and conditional analyses; imputation; examine joint effects of genes and environment; and explore phenotypic heterogeneity.

\section{\Acrfull{hwe}}
\label{sec:hwe}

\Gls{hwe} is a theory stating that, in a large random-mating population with no selection, mutation, or migration, the allele frequencies and the genotype frequencies are constant from generation to generation.
Furthermore, there is a simple relationship between the allele frequencies and the genotype frequencies.

Consider two alleles $A$ and $a$, where the frequency of allele $A$ in a population is given by $f(A) = p$ and frequency of allele $a$ is given by $f(a) = q$.
\Gls{hwe} states that $f(AA) = p^2$, $f(Aa) = 2pq$, and $f(aa) = q^2$ in a given population.
These proportions are known as the Hardy-Weinberg proportions.
If the proportions deviate from the \gls{hwe} (in other words, the observed genotype frequencies don't follow HW proportions), then this suggests that the assumption(s) underlying the \gls{hwe} is violated.
These assumptions are:
\begin{enumerate}[(a),noitemsep]
	\item Organisms are diploid
	\item Only sexual reproduction occur
	\item No overlapping generations
	\item Mating is random
	\item Population size is infinitely large
	\item Allele frequencies are equal in the sexes
	\item There is no migration, mutation or selection
\end{enumerate}

\section{Relatedness}
\label{sec:relatedness}

In linkage studies, you look for genetic loci that correlate between the phenotype of interest and the pattern of transmission over generations.
The signals that you identify in the so-called ``family-based association studies'' are specific to the particular family, and it cannot be generalised to the larger population.
In contrast, \gls{gwas} searches for loci that have significant correlation between the phenotypes and genotypes of unrelated individuals.
Since \gls{gwas} is carried out at a population-level, the findings can be generalised to that population.
However, if for some reason the samples happen to be related with one another in the sample population used in the \gls{gwas}, the findings can no longer be generalised, as the signal may have come from those related samples (i.e.\ only generalisable in that related population).
Therefore, relatedness between the samples are essential for a linkage studies, but it is nuisance in \gls{gwas}, as it violates the assumption that all of the samples are unrelated to one another.

There are two forms of relatedness that can confound an association study: population structure and cryptic relatedness.
Population structure is when there are large-scale systematic differences in ancestry.
For exmple, groups of individuals with more recent shared ancestors than one would expect in a random-mating population.
Shared ancestry corresponds to relatedness (or kinship) and therefore population structure can be defined in terms of patterns of kinship among groups of individuals (i.e.\ subpopulations, or ``islands'').
Cryptic relatedness refers to the presence of close relatives in a sample of seemingly unrelated individuals.
Where population structure refers to the relatedness of large groups of individuals, cryptic relatedness refers to recent common ancestry among smaller groups of individuals (often pairs).

\subsection{\Acrfull{ibd}}
\label{sub:ibd}

\Gls{ibd} is the probability of four alleles from two diploid individuals at a given genetic locus is from the same ancestral allele without an intermediate mutation.
Consider two individuals $A$ and $B$.
At a given genetic locus, $A$ may have alleles 
If one or more of these four alleles ($a_1$,  $a_2$, $b_1$ and  $b_2$) was from a single ancestor, then that allele will be \gls{ibd}.
I know the naming of the alleles are confusing, but say for example $x_1$ was an allele present in one the ancestors.
This $x_1$ allele may have been passed down to individual $A$ at the given locus (for example, at $a_1$) through many generations.
Similarly, $x_1$ may have also been passed down to individual $B$ at the locus as $b_2$.
Since both $a_1$ and $b_2$ alleles originated from the same ancestral allele ($x_1$), these two alleles are said to be \gls{ibd}.

\subsection{\Acrfull{ibs}}
\label{sub:ibs}

\Gls{ibs} is the probability that a sequence of \acrshort{dna}  are identical between two individuals.
Consider a region of \acrshort{dna} with $M$ markers, containing $1\dots k\dots M$ SNPs.
The \gls{ibs} between two individuals $i$ and $j$ is given by:
\begin{equation*}
	IBS_{ij} = 1 - \frac{1}{2M}\sum_k \lvert G_{ik} - G_{jk} \rvert
\end{equation*}
where $G_{ik}$ is the number of minor allele (0,1 or 2) carried by the $i^{th}$ individual at SNP $k$.
If individuals $i$ and $j$ have an identical SNP at position $k$, then $\lvert G_{ik} - G_{jk} \rvert$ will equal to 0.
Therefore, \gls{ibs} of 100\% (or close to 100\%) means that the individuals are identical.

If the \acrshort{dna} segment is \gls{ibd}, then that segment will also be \gls{ibs} (since the segment would have come from the same ancestor).
However, there is  a chance that a segment can be \gls{ibs} but not \gls{ibd}.
This can occur if the individuals obtained the sequence from two different ancestors that had the same mutation, and therefore the same sequence.

\subsection{Ascertainment bias}
\label{sub:ascertainment_bias}

\subsection{Controlling for relatedness}
\label{sub:controlling_for_relatedness}

### 2017-05-18 notes/glossary/glossary.tex
tmp commit

@@
 \newacronym[description={Genome wide association study},longplural={genome wide association studies},\glsshortpluralkey={GWAS}]{gwas}{GWAS}{genome wide association study}
+\newacronym[description={Trans-ancestry meta-analysis}]{tama}{TAMA}{trans-ancestry meta analysis}
+\newacronym[description={Quantile-Quantile plot}]{qq}{QQ}{Quantile-Quantile}
+\newacronym[description={Principal Component Analysis}]{pca}{PCA}{principal component analysis}
+\newacronym[description={Minor allele frequency}]{maf}{MAF}{minor allele frequency}
 \newacronym[description={Adrenocorticotropic hormone}]{acth}{ACTH}{adrenocorticotropic hormone}
+\newacronym[description={Identity by state}]{ibs}{IBS}{identity by state}
+\newacronym[description={Identical by descent}]{ibd}{IBD}{identical by descent}
+\newacronym[description={Hardy-Weinberg equilibrium}]{hwe}{HWE}{Hardy-Weinberg equilibrium}
 \newacronym[description={Agouti-related protein}]{agrp}{AGRP}{agouti-related protein}

### 2017-05-18 notes/subnotes/gwas.tex
tmp commit

@@
 As a result, the frequencies of these variants observed in the population are dramatically reduced.
-This also means that the variants that cause Mendelian diseases emerged reltively recently, in terms of evolutionary time scale, otherwise the variants would not have survived the negative selection over many generations.
+This also means that the variants that cause Mendelian diseases emerged relatively recently in terms of evolutionary time scale, otherwise the variants would not have survived the negative selection over many generations.
 
@@
 Linkage analysis has been very successful in identifying genes that cause rare Mendelian diseases, but it was not so successful in identifying genes for common disease or quantitative trait.
-There have been a few genes, gene regions and gene variants that have been identified that related to some diseases, but these variants usually explain only a small fraction of the \gls{heritability} of the disease.
+There have been a few genes, gene regions and gene variants that have been identified that related to some diseases, but these variants usually explain only a small fraction of the total \gls{heritability} of the disease.
 For example, the identified variant may explain an excess risk of two-fold, where in reality there is a total excess risk of 30-fold, which shows that there are many other causal genes that are yet to be identified.
@@
 
-In contrast to the candidate gene association studies, \gls{gwas} looks at most of the genome for the causal variants.
+In contrast to the candidate gene association studies, \gls{gwas} look at most of the genome for the causal variants.
 Unlike candidate gene association studies, \gls{gwas} does not make any assumptions about the genomic location of the causal variants.

### 2017-05-18 notes/subnotes/tmp.tex
tmp commit

@@
 
-\section{Extending \gls{gwas}}
-\label{sec:extending_gwas}
-
-\subsection{Meta-analysis}
-\label{sub:meta_analysis}
-
-Individual \gls{gwas} are underpowered to detect only those variants that have the biggest effect, and greater power is required to detect other susceptibility loci.
-Meta-analysis can be carried out on the data from a comparable \gls{gwas}.
-This is a low-cost approach to enhance the power for the main  as well as the joint (gene-gene and gene-environment) effects.
-Furthermore, this allows you to replicate the results, get better idea of the SNPs for subsequent replication studies, and investigate potential sources of heterogeneity.
-
-Summary-level data will be enough to carry out many meta-analysis, but access to individual-level data allows for more sophisticated analyses.
-These include: haplotype and conditional analyses; imputation; examine joint effects of genes and environment; and explore phenotypic heterogeneity.
-
 \section{\Acrfull{hwe}}
@@
 Consider two alleles $A$ and $a$, where the frequency of allele $A$ in a population is given by $f(A) = p$ and frequency of allele $a$ is given by $f(a) = q$.
-\Gls{hwe} states that $f(AA) = p^2$, $f(Aa) = 2pq$, and $f(aa) = q^2$ in a given population.
+\Gls{hwe} states that the allele frequencies are $f(AA) = p^2$, $f(Aa) = 2pq$, and $f(aa) = q^2$ in a given population (that is infinitely large and randomly mating).
 These proportions are known as the Hardy-Weinberg proportions.
@@
 \end{enumerate}
+
+In terms of quality control in \gls{gwas}, significant deviation of a sample from the \gls{hwe} is most likely indicative of genotyping error (i.e.\ technical error), especially if the sample is believed to come from the same genetic population.
+Other causes of deviation from the \gls{hwe} could be due to the samples coming from, for example,  distinct genetic populations (unlikely to be unnoticed) or possibly due to inbreeding.
+In either case, the Hardy-Weinberg proportion of allele frequencies will be altered.
+
+\section{Effective number of tests in \gls{gwas}}
+\label{sec:effective_number_of_tests_in_gwas}
+
+In \gls{gwas}, you test the association of all of the \glspl{snp} with the phenotype of interest, meaning that there will be as many hypothesis tests done as the number of \glspl{snp} in the analysis.
+Consider a \gls{gwas} with 1,000,000 \glspl{snp} and a significance threshold of p < 0.05.
+In this scenario, you would expect 50,000 \glspl{snp} to show up as being associated with the disease by chance (i.e.\ false positives), which is not ideal.
+
+One way to reduce the number of false positives is to adjust the significance threshold with \gls{fwer} control.
+For the situation above, you would adjust the p-value with the number of tests carried (Bonferroni adjustment):
+\begin{equation*}
+	p = \frac{0.05}{1000000} = 5\times10^{-8}
+\end{equation*}
+This p-value is used in many \glspl{gwas}.
+However, this approach may be too conservative and may lead to loss in power (increased false negatives).
+One way to work around this problem is to reduce the number of tests carried out in the \gls{gwas}.
+\citet{Duggal2008} suggested the use of a single \gls{snp} per \gls{ld}, and use the p-value calculated from the reduced number of tests.
+
+Another, perhaps more reliable way, is to carry out a permutation test.
+With permutation test, the samples are ``shuffled'' in such a way that the link between the genotype and the phenotype of the samples are separated.
+This is usually done by assigning cases and controls randomly, using the same samples as the original \gls{gwas} data.
+\gls{gwas} is then carried out on this shuffled data $n$ times, shuffling the samples every time.
+The lowest p-value from each analysis is recorded, as it represents the most significant p-value you would obtain by chance.
+%TODO: finish explaining how to get the permuted p-value
+These p-values are compared with the lowest p-value from the original \gls{gwas}.
 
@@
 Consider two individuals $A$ and $B$.
-At a given genetic locus, $A$ may have alleles 
-If one or more of these four alleles ($a_1$,  $a_2$, $b_1$ and  $b_2$) was from a single ancestor, then that allele will be \gls{ibd}.
-I know the naming of the alleles are confusing, but say for example $x_1$ was an allele present in one the ancestors.
+At a given genetic locus, $A$ may have alleles $a_1$ and $a_2$, and individual $B$ may have alleles $b_1$ and  $b_2$.
+If one or more of these four alleles ($a_1$,  $a_2$, $b_1$ or $b_2$) was from a single ancestor, then that allele will be \gls{ibd}.
+I know the naming of the alleles are confusing, but say for example $x_1$ was an allele present in one of the ancestors.
 This $x_1$ allele may have been passed down to individual $A$ at the given locus (for example, at $a_1$) through many generations.
@@
 
-\subsection{Controlling for relatedness}
-\label{sub:controlling_for_relatedness}
+\section{Controlling for relatedness}
+\label{sec:controlling_for_relatedness}
+
+There are several ways to control for relatedness (population substructure and cryptic relatedness) among the samples.
+One way is to remove the samples that are related to one another, based on their \gls{ibs} and/or \gls{ibd}.
+However, this will reduce the sample size and therefore reduce the power to detect association.
+Alternatively, you could match the samples in the case and control groups so the population groups are matched between the cases and controls.
+This will control for the major effect that population substructure may have on the data, but it will not account for the effects of small, fine scale population effect.
+For example, you get an apparent association between \glspl{snp} and Type II diabetes in Pima Indians (WTCCC training course material).
+
+Another way to control for the relatedness is to include families into the study, where affected offspring will be in the case group and parents in the control group.
+This approach will control for the population structure, and also allows you to look at which parents the causal allele came from (which is often difficult in other association studies).
+Again, this approach reduces the power and increase the cost, as it is difficult to gather enough families and two parents are required to make one control.
+
+There are statistical methods that control for population substructure and relatedness between the samples in the study, which will be discussed below.
+These methods control for sample relatedness without removing the samples, and therefore retains the power to detect association.
+
+\subsection{\gls{qq} plot}
+\label{sub:qq_plot}
+
+\Gls{qq} plot is a plot that compares two probability distributions, and it can be used to see whether the data set has any evidence of population structure.
+The expected negative log p-values for each \gls{snp}  (which is the rank of the \gls{snp} divided by the number of markers) is along the x-axis and the observed p-values for the \glspl{snp} are on the y-axis.
+In this plot, the ``tail'' end of the plot represent the most significant p-value, and therefore a potentially important \gls{snp}.
+
+If there is no population structure in the data set, the dots will be along the null distribution (i.e.\ y = x).
+However, if there is a population structure present in the data set, then the p-values are ``inflated'' by a constant factor $\lambda$, and the values move away from the null distribution.
+This inflation factor is used in the genomic control method to control for the population structure.
+
+\subsection{Genomic control}
+\label{sub:genomic_control}
+
+Genomic control is a method developed by \citet{Devlin1999} that is easy to apply on the data set to control for population effect.
+\citet{Devlin1999} argued that the Armitage test statistics inflated linearly by a constant factor $\lambda$ when there was evidence of population substructure.
+Therefore, by adjusting all of the test statistics by a  constant factor $\lambda$, genomic control is able to adjust for the confounding effect of the population structure.
+A disadvantage of genomic control is that it adjusts each marker by uniform value ($\lambda$), which may not be suitable for markers that differ significantly in their allele frequencies across ancestral populations and/or under selection \citep{Price2006}.
+In other words, uniform adjustment across all the markers is not suitable when a marker is significantly different across ancestries.
+
+\subsection{Adjustments using \gls{pca}}
+\label{sub:adjustments_using_pca}
+
+Instead of adjusting the markers with a constant factor, \gls{pca}-based methods aim to adjust the genotype and phenotype of the samples based on the ancestry of the samples.
+First, \gls{pca} is applied to the genotype data (matrix of \glspl{snp} by samples) to get  a summary of the genotype data to obtain a continuous axes of genetic variation (i.e.\ principal components).
+These axes explain the genetic variation (or the ancestry) of the samples.
+Next, the genotypes and  the phenotypes of the samples are continuously adjusted by the amount attributable to ancestry, by taking the residuals from the linear models \citep{Price2006}.
+Lastly, the adjusted data is used for the \gls{gwas}.
+
+Although \gls{pca}-based methods (such as EIGENSTRAT) are able to account for the majority of the population structure, it does not account for all of it.
+This is because these methods only utilise a few principal components to capture the sample structure, which will explain majority (but not all) of the genetic variation of the population.
+
+\subsection{Mixed linear models}
+\label{sub:mixed_linear_models}
+
+As mentioned earlier, \gls{pca}-based adjustments may not be sufficient to correct for all of the population substructure.
+If the complete genealogy of the samples were known, this would be possible by applying variance component to correct for the relatedness.
+However, this is not usually the case.
+With that said, high-density genotype information of the samples can be  used to estimate the level of relatedness between individuals and make a ``kinship matrix'' or ``empirical relatedness matrix''.
+This kinship matrix can be included in the mixed linear model and be accounted for as ``random effect'' in the model, thus able to focus on the effect of the locus on the disease without the confounding effect from the population structure.
+The following description of the methodology for mixed model is from \citet{Kang2010}.
+
+Let $Z_{ij}$ be the contribution of factor $j$ to the phenotype of individual $i$.
+Then the phenotype of individual $i$, $y_i$ is:
+\begin{equation*}
+	y_i = \sum_{j=1} Z_{ij} + \varepsilon_i
+\end{equation*}
+where $\varepsilon_i$ represent the random variable (i.e.\ environment factor) that affect the phenotype and $E(\varepsilon_i) = 0$.
+
+Now let $Y = \{y_1, y_2,\ldots,y_n\}$ (vector of phenotype of individuals), and assume the environmental components are uncorrelated and the variance covariance structure of $Y$ depends on the number of genes shared among individuals.
+Then in the absence of dominance effects:
+\begin{equation*}
+	Var(Y) = 2\sigma^2_a\Phi + \sigma^2_eI
+\end{equation*}
+where $\Phi$ = kinship coefficients, $I$ = identity matrix, $\sigma_a^2$ for additive genetic variance and $\sigma_e^2$ for random environmental variance.
+
+In association studies, you test each locus $k$ for an association by testing whether $H_0:\beta_k = 0$ using the model:
+\begin{equation*}
+	y_i = \beta_0 + \beta_kX_{ik} + \eta_{ik}
+\end{equation*}
+where the error term $\eta_{ik} = \sum \beta_sX_{is} + \varepsilon_i$ and $X$ is the \gls{maf} counts at a given locus.
+If the relatedness is known, then the variance covariance of $\eta_{ik}$ can be approximated in the form of $Var(Y) = 2\sigma^2_a\Phi + \sigma^2_eI$ (see below), and therefore able to focus on the effect of locus by accounting for the relatedness as a ``random effect'' in the mixed model.
+
+Genotype data can be used to construct an $n\times n$ genetic relatedness matrix (e.g.\ \gls{ibs} or Balding-Nichols matrix), $\hat{S}$, and normalise it to have sample variance of 1, using a Gower's centered matrix:
+\begin{equation*}
+	\hat{S}_N = \frac{(n-1)\hat{S}}{Tr(P\hat{S}P)}
+\end{equation*}
+\begin{equation*}
+	P = I-ll'/n
+\end{equation*}
+where $l$ is a vector of 1's.
+Use variance component model $Var(Y) = 2\sigma^2_a\hat{S}_N + \sigma^2_eI$ to estiimate the restricted maximum likelihood parameters of $\sigma^2_a$ and $\sigma^2_e$, and test for $H_0:\sigma^2_a = 0$.
+If this is rejected, use generalised least square F-test (or any score test) to estimate the effects of $\beta_k$, and test whether $\beta_k = 0$ using the model:
+\begin{equation*}
+	y_i = \beta_0 + \beta_kX_{ik} + \eta_{i}
+\end{equation*}
+\begin{equation*}
+	Var(\eta) = V \propto2\hat{\sigma}^2_a\hat{S}_N + \hat{\sigma}^2_eI
+\end{equation*}
+
+\section{Meta-analysis}
+\label{sec:meta_analysis}
+
+Individual \gls{gwas} can be underpowered to detect only those variants that have the biggest effect, and greater power is required to detect other susceptibility loci.
+To increase the power to detect causal variants, meta-analysis can be carried out on the data from comparable \gls{gwas}.
+This is a low-cost approach to enhance the power for the main effect as well as the joint (gene-gene and gene-environment) effects.
+Furthermore, this allows you to replicate the results, get better idea of the SNPs for subsequent replication studies, and investigate potential sources of heterogeneity.
+
+Summary-level data will be enough to carry out many meta-analysis, but access to individual-level data allows for more sophisticated analyses.
+These include: haplotype and conditional analyses; imputation; examine joint effects of genes and environment; and explore phenotypic heterogeneity.
+Broadly speaking, there are two types of models that underly the methods for meta-analyses: fixed-effects model and random-effects model.
+
+\subsection{Imputation}
+\label{sub:imputation}
+
+\subsection{Fixed-effects model}
+\label{sub:fixed_effects_model}
+
+Under the fixed-effects model, all of the populations included in the meta analysis are assumed to have the same effect size.
+This means that in the fixed-effects model, all of the populations are assumed to be homogeneous and therefore require similar studies for the meta analysis.
+Due to this assumption, studies are carefully selected to make sure the effect sizes are similar across different studies.
+However, this does not necessarily maximise the sample size which is the main goal of meta analysis in the first place.
+
+Fixed-effects model often uses inverse-variance weighting to adjust the effect sizes from different studies.
+Let $X_1,\ldots, X_n$ be the effect size estimate of study 1 to $n$ (e.g.\ log odds ratio), variance of the effect size of study $i$ be $V_i = SE(X_i)^2$, and the weight of the study $i$ be $W_i = V_i^{-1}$.
+Then the inverse-variance weighted effect size estimator is:
+\begin{equation*}
+	\bar{X} = \frac{\sum W_iX_i}{\sum W_i}
+\end{equation*}
+$Z$-scores can then be generated from this, taking into account the number of cases and controls in each study:
+\begin{equation*}
+	Z_{FE} = \frac{\bar{X}}{SE(\bar{X})} = \frac{\sum W_iX_i}{\sqrt{\sum W_i}}
+\end{equation*}
+Or, for weighted $Z$-score:
+\begin{equation*}
+	Z_{WS} = \frac{\sum \sqrt{N_ip_i(1-p_i)}Z_i}{\sum \sqrt{N_ip_i(1-p_i)}}
+\end{equation*}
+Where $p_i$ is the \gls{maf}, and $N_i$ is the effective sample size of study $i$.
+For more details, see \citet{Han2011}.
+
+\subsection{Random-effects model}
+\label{sub:random_effects_model}
+
+In contrast, the random-effects model assumes that all of the populations included in the meta analysis have different effect sizes.
+This will allow for the heterogeneity between the effect size estimates between the studies.
+However, as \citet{Han2011} pointed out, random-effects model assumes this heterogeneity even under the null hypothesis where the effect sizes of the studies are all 0, and therefore should have no heterogeneity under the null hypothesis.
+This results in overly conservative p-values and consequently results in lower p-values compared to the fixed-effects model.
+\citet{Han2011} developed a new random-effects model that assumed no heterogeneity under the null hypothesis (RE-HE).
+
+\section{Trans-ancestry meta-analysis}
+\label{sec:trans_ancestry_meta_analysis}
+
+\gls{gwas} and meta-analyses have been carried out extensively in the European population, and it has provided significant amount of knowledge about variants that cause diseases in European population.
+However, these variants may not necessarily be present or have significance in other populations.
+Furthermore, the variants identified in European population alone does not explain the overall disease risk (check citation).
+Therefore, it is important to carry out \gls{tama} to identify novel and population-specific variants.
+
+\subsection{Advantages of \gls{tama}}
+\label{sub:advantages_of_tama}
+
+Most of the advantages of \gls{tama} are the same as in normal meta-analysis: increased sample size, increased power to detect variants, and replication of results.
+However, there are a couple of additional advantages in \gls{tama}.
+First, by including multiple different populations with different ancestries, \gls{tama} allows you to identify local and global variants.
+Local variants will be specific to a population, whereas global variants are those that are present in all of the populations included in the studies (check citation).
+By identifying these variants, it will be possible to customise clinical treatments for specific groups of people in the future.
+
+Second, a mix of different ancestral populations allow one to fine map certain regions of the genome for a more precise identification of causal \glspl{snp}.
+This is due to the fact that different ancestry will have different \gls{ld} patterns, and by overlapping these regions of \gls{ld}, you can fine map certain regions of genome.
+Fine mapping of genetic regions allow you to narrow down the region where the causal variant may be, which will be particularly useful for regions with sparse \gls{ld} patterns (?; check if correct).
+
+\subsection{Challenges of trans-ancestry meta-analysis}
+\label{sub:challenges_of_trans_ancestry_meta_analysis}
+
+\section{Methods to carry out trans-ancestry meta-analysis}
+\label{sec:methods_to_carry_out_trans_ancestry_meta_analysis}
+
+Similar methods to normal meta-analysis can be used, but population structure, cryptic relatedness and difference in ancestries will have to be accounted for, as well as between study differences.
+
+\subsection{Bayesian partition model (MANTRA)}
+\label{sub:bayesian_partition_model}
+
+Both fixed-effects and random-effects models do not fit well with the meta analysis design.
+Fixed-effects model assumes that all studies
+
+\section{Fine-mapping}
+\label{sec:fine_mapping}
+
+One of the limitations with a single population \gls{gwas} or meta-analysis is that there may be many \glspl{snp} that show association with the disease in a given region of the genome.
+This is usually due to the large \gls{ld} patterns that are specific to the population, making it difficult to identify which \gls{snp} within the \gls{ld} region is actually causal.
+The purpose of fine-mapping is to narrow down this region of genome to reduce the number of candidate \glspl{snp} that could be relevant to the disease.
+
+\subsection{Credible sets}
+\label{sub:credible_sets}
+
+Credible set was first defined in the study done by \citet{wtccc2012}, using Bayes factors and posterior probability to describe how likely a \gls{snp} is causal.
+A credible set is the minimal number of \glspl{snp} in a genomic region that are most likely to be causal of the disease \citep{wtccc2012}.
+\glspl{snp} in the credible sets are selected based on the likelihood of a \gls{snp} accounting for the association signal within that gene region.
+For example, a 99\% credible set will have \glspl{snp} that are 99\% likely to account for the association signal within that genomic region.
+As the authors mention in their study, credible sets are analogous to confidence interval, where in a particular gene region, you can be confident that the true causal \gls{snp} will be included in the relevant credible set \citep{wtccc2012}.
+Therefore, the smaller the credible set, the more confident the true causal variant is included in the credible set of that genomic region (kind of?).
+
+Under the assumptions that there is only one causal \gls{snp} in a gene region, and that the causal \gls{snp} was genotyped in the study, the authors were able to calculate the posterior probability of a given \gls{snp} being causal of the association signal \citep{wtccc2012}.
+First, the strength of evidence for association for each of the \glspl{snp} were determined based on the Bayes factor.
+Let $X_i$ be the genotype data at locus $i$ of a gene region with $k$ loci.
+Consider two models, $M_0$ and $M_i$, where $M_0$ is the null model of no association with locus $i$, and $M_i$ is the alternative model where the \gls{snp} at locus $i$ is the only causal allele within that region.
+Now, the Bayes factor of the locus $i$, $BF_i$, is given by:
+\begin{equation*}
+	BF_i = \frac{Pr(X_i|M_i)}{Pr(X_i|M_0)}
+\end{equation*}
+where $Pr(X_i|M_i)$ is the probability that the \gls{snp} at locus $i$ is causal in the region and $Pr(X_i|M_0)$ is the probability that there is no association.
+
+The Bayes factor for a region ($BF_{reg}$) is:
+\begin{equation*}
+	BF_{reg} = \sum^{k}_{i = 1}BF_iPr(M_i|M)
+\end{equation*}
+where $M$ represents all of the $k$ models for $k$ loci in that gene region.
+Assuming there is a uniform prior that any particular \gls{snp} in a region can be causal, the mean Bayes factor for the region is:
+\begin{align*}
+	Pr(M_i|M) = \frac{1}{k}\\
+	\therefore BF_{reg} = \frac{1}{k}\sum_{i = 1}^{k}BF_i
+\end{align*}
+Now, the posterior probability that a given \gls{snp} is the causal \gls{snp} in that region is:
+\begin{equation*}
+	Pr(M_i|X,M) = \frac{1}{k}\frac{BF_i}{BF_{reg}}
+\end{equation*}
+where $X$ is the genotype of the $k$ loci in the region, and $M$ the corresponding models.
+Note that the posterior probability of the \gls{snp} being causal is proportional to the Bayes factor of that \gls{snp}.
+In other words, the stronger the evidence of association, the higher the probability of that \gls{snp} is included in the credible set.

### 2018-10-09 notes/glossary/glossary.tex
tmp commit

--- 2017-04-02 notes/glossary/glossary.tex
@@
 	description={Statistical programming language}
+}
+
+\newglossaryentry{ldg}{
+	name=Linkage disquilibrium,
+	text=linkage disquilibrium,
+	description={Correlation between nearby variants such that the alleles at neighbouring markers (on the same chromosome) are associated within a population more often than if they were unlinked}
+}
+
+\newglossaryentry{heritability}{
+	name=Heritability,
+	text=heritability,
+	description={The proportion of the variation in a given characteristic or state that can be attributed to (additive) genetic factors}
 }
@@
 \newacronym[description={Wilcoxon-Mann-Whitney}]{wmw}{WMW}{Wilcoxon-Mann-Whitney}
+\newglossaryentry{ld}{type=\acronymtype,name={LD},first={Linkage Disequilibrium (LD)\glsadd{ldg}},description={Linkage disquilibrium},see=[Glossary:]{ldg}}

### 2018-10-09 notes/notebook.tex
tmp commit

@@
-% My master tips book
-\documentclass[a4paper, 12pt] {report}
-
-% All of the packages imported for this document:
-% \usepackage[utf8]{inputenc}
-% \usepackage{glossaries}
-% \usepackage{hyperref}
-% \usepackage{setspace}
-% \usepackage{url}
-% \usepackage{multicol}
+\documentclass[a4paper, 12pt, oneside]{report} % for final draft compilation
+% \documentclass[a4paper, 12pt, oneside, draft]{report} % for draft compilation
 
@@
 \usepackage[T1]{fontenc}                                 % package for specifying output font encoding
-\usepackage[acronym,style=super,nogroupskip,section=chapter]{glossaries} % package for making glossaries
+\usepackage[acronym,style=super,nogroupskip]{glossaries} % package for making glossaries
 \usepackage[flushleft]{threeparttablex}                  % package to add notes at the bottom of the table
@@
 
+% Define colour(s) for table rows:
+\definecolor{LightCyan}{rgb}{0.88,1,1}
+
+% Change the itemize bullet points to dash:
+\def\labelitemi{--}
+
 % Set up the hyperlinks - change the colour of the links to blue:
@@
 
-\def\labelitemi{--}
+% Change the default cref name for equations:
+\creflabelformat{equation}{#2\textup{#1}#3}
 
+% Change the default cref name for appendix:
+\crefname{appcha}{Appendix}{Appendices}
+\crefname{appsec}{Appendix}{Appendices}
+
+% Make siunitx detect bold/italics/etc:
+\robustify\bfseries
+\sisetup{detect-all}
+
+% Set the width of captions for threeparttable to \linewidth:
+\captionsetup[threeparttable]{width=\linewidth}
+
+% Set up continuous figures:
+\DeclareCaptionFormat{cont}{#1 (cont.)\par}
+
+% Change the default url font:
+\urlstyle{rm}
+
+% change figure numbering from section-based to continuous:
+\counterwithout{figure}{chapter}
+\counterwithout{equation}{chapter}
+\counterwithout{table}{chapter}
+
+% Change the title of bibliography to references:
+\renewcommand{\bibname}{References}
+
+% Change the title of Table of Contents:
+\renewcommand{\contentsname}{Table of Contents}
+
+% Change the title of Table of Contents:
+% \renewcommand{\thesubfigure}{\Alph{subfigure}}
+
+% Italicise the subsubsection title:
+\titleformat{\subsubsection}{\normalfont\itshape\bfseries}{\thesubsubsection}{1em}{}
+\titleformat{\paragraph}{\normalfont\itshape}{\theparagraph}{\noindent}{}
+
+% Remove unnecessary spaces between the figures from different chapters in LoF and LoT:
+\newcommand*{\noaddvspace}{\renewcommand*{\addvspace}[1]{}}
+\addtocontents{lof}{\protect\noaddvspace}
+\addtocontents{lot}{\protect\noaddvspace}
+
+% For generating glossaries:
+\include{./glossary/glossary}
 \makeglossaries
+
+% Specify graphics path:
+\graphicspath{{./images/}}
+\DeclareGraphicsExtensions{.pdf,.png,.jpg}
+
+% Define some alias for very long citations:
+% \defcitealias{TCGA}{NCI and NHGRI, 2016}
+% \defcitealias{ICGC}{ICGC, 2016}
+% \defcitealias{GEO}{NCBI, 2016}
 
@@
 \include{./title/title}
-\include{./glossary/glossary}
+
+\pagenumbering{roman} % set page numbering to roman numeral
+
+\include{./subnotes/preface}
+
+\addcontentsline{toc}{section}{Preface}
+
+% \include{./misc/acknowledgement}
+
+\addcontentsline{toc}{section}{\contentsname}
 
@@
 
-\printglossaries
 \clearpage
 
-\include{./subnotes/intro}
+% \addcontentsline{toc}{section}{\listfigurename}
+
+% \listoffigures
+
+% \clearpage
+
+% \addcontentsline{toc}{section}{\listtablename}
+
+% \listoftables
+
+% \clearpage
+
+\addcontentsline{toc}{section}{Glossary}
+
+\printglossary[style=super]
+
+% Need to set line spacing to single space, otherwise it inserts a blank page for an unknown reason:
+% \singlespacing
+
+\clearpage
+
+\addcontentsline{toc}{section}{Acronyms}
+
+\printglossary[type=\acronymtype,style=super]
+
+\clearpage
+
+\pagenumbering{arabic} % revert the page numbering to normal numbering
+\glsresetall
 
@@
 
+% \include{./methods/methods}
+
+% \include{./results/results}
+
+% \include{./results/results2}
+
+% \include{./discussion/discussion}
+
+% \include{./misc/appendix}
+
+% \include{./misc/appendix2}
+
+% Use the reference style specified in BiocRefStyle.bst:
 \bibliographystyle{BiocRefStyle}

### 2018-10-09 notes/subnotes/gwas.tex
tmp commit

--- 2017-04-03 notes/subnotes/gwas.tex
@@
 ``The genome-wide association approach therefore represents an unbiased yet fairly comprehensive option that can be attempted even in the absence of convincing evidence regarding the function or location of the causal genes'' \citep{Hirschhorn2005}.
+
+\section{\glsdesc{ld}}
+\label{sec:ld}
+
+Consider two alleles or gene loci, A and B, that are on the same chromosome.
+If these alleles are ``dependent'', then A and B will always segregate together.
+Mathematically, the probability of both A and B segregating together is $P(AB)$.
+Now, if A and B are ``independent'', then A being passed on to the offspring has no influence on B being passed on, and vice versa.
+Mathematically, the probability of both A and B segregating in this case is given by $P(A)P(B)$.
+
+\Gls{ld}, by definition, is the correlation between nearby variants such that the alleles at neighbouring markers on the same chromosome are associated within a population more often than they were unlinked.
+In other words, {\bfseries \gls{ld} is a measure of how likely two alleles or loci are passed on to the offspring together}.
+Mathematically, this is represented as:
+\begin{equation}
+	\label{eq:ld}
+D = P(AB) - P(A)P(B)
+\end{equation}
+where $D$ is the coefficient of \gls{ld}.
+When $D = 0$, A and B are said to be in equilibrium and will not segregate together.
+On the other hand, when $D = 1$, A and B are said to be in perfect \gls{ld}, where A and B will always segregate together.
+
+By measuring the \gls{ld} of two alleles, or in our case two markers, you are able to determine whether those alleles will be passed on together.
+This means that only one marker is required to detect or track both alleles, if they are in strong \gls{ld}.
+Therefore, you can choose several markers that represent several alleles that are in high \gls{ld} with the marker.
+Furthermore, if that marker is associated with a disease, it is likely that at least one of the markers that are in high \gls{ld} with that marker is also associated with the disease.
+This allows us to use less markers without losing any power to detect variants that are associated with the disease.

### 2018-10-09 coding/algorithms/algorithm.tex
started writing algorithm a week

--- 2017-04-02 notes/notebook.tex
@@
-% My master tips book
-\documentclass[a4paper, 12pt] {report}
+\documentclass[a4paper, 12pt, oneside]{report} % for final draft compilation
+% \documentclass[a4paper, 12pt, oneside, draft]{report} % for draft compilation
 
-% All of the packages imported for this document:
-\usepackage[utf8]{inputenc}
-\usepackage{glossaries}
-\usepackage{hyperref}
-\usepackage{setspace}
-\usepackage{url}
-\usepackage{multicol}
+% Preamble:
+\usepackage[T1]{fontenc}                                 % package for specifying output font encoding
+\usepackage[acronym,style=super,nogroupskip]{glossaries} % package for making glossaries
+\usepackage[hidelinks, colorlinks]{hyperref}             % package for hyperlinking references - hidelinks option removes the ugly boxes around the huperlinks
+\usepackage[md]{titlesec}                                % package for tweaking the chapter/section/subsection titles format
+\usepackage[utf8]{inputenc}                              % package for specifying input font encoding
+\usepackage{mathptmx}                                    % package for Times New Roman font
 
-\makeglossaries
+% Change the itemize bullet points to dash:
+\def\labelitemi{--}
+
+% Set up the hyperlinks - change the colour of the links to blue:
+% \hypersetup{allcolors=blue}
+\hypersetup{allcolors=black} % uncomment this line to make the hyperlinks black
+
+% Change the default url font:
+\urlstyle{rm}
 
@@
 
-\include{./title/title}
-\include{./glossary/glossary}
+\include{./titlepage}
 
-\tableofcontents
-
-\printglossaries
-\clearpage
-
-\include{./subnotes/intro}
-
-\include{./subnotes/gwas}
+\include{./preface}
 
